### 2015-10-30 2016-pldi/2016-pldi-key_to_data_parallel_compiler.tex
Add in initial transfer of 2015 ICFP paper to PLDI 2016

\documentclass[pldi]{sigplanconf-pldi15}
% \usepackage[utf8]{inputenc}
\usepackage{fontspec}
\setmainfont[Ligatures=TeX]{STIX}
\setmonofont{APL385 Unicode}
\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
\usepackage{url}                  % format URLs
\usepackage{listings}          % format code
\usepackage{enumitem}      % adjust spacing in enums
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI

\begin{document}

\title{The Key to a Data Parallel Compiler}

\maketitle
\begin{abstract}
By choosing an appropriate representation for the AST, we can implement a compiler for a modern functional 
language using only data-parallel array operations under simple functional composition without the use of 
branching, recursion, or other means of explicit control flow. In this pearl we elucidate the core data structure 
and an operator, called Key, that enable a surprisingly direct implementation of non-trivial transformations. 
The result permits arbitrary computation over sub-trees chosen by parent-child and other inter-node 
relationships in a pure data-parallel fashion without complex control flow. We give direction on the use of this 
method through examples of function lifting and expression flattening.
\end{abstract}

%[of]:Introduction
\section{Introduction}

When compiling a high-level language into a low-level language, we often implement a variety of 
flattening transformations to normalize more complex code exhibiting nesting into flat programs 
that are semantically equivalent. Function lifting, for example, takes a program with nested functions 
into an equivalent program in which all functions are at the root of the program. The following function
f always returns 14.

\begin{verbatim}
f←λ.(λ ⍵.(⍵+⍵))7
\end{verbatim}

A function lifting pass would transform this into the following:

\begin{verbatim}
fn1←λ⍵.⍵+⍵
f←λ.fn1 7
\end{verbatim}

Expression flattening works in a similar fasion. Assuming all expressions evaluate right to left, this expression:

\begin{verbatim}
v←(a+b)÷c×d
\end{verbatim}

Becomes this:

\begin{verbatim}
v←c×d
t←a+b
v←t÷v
\end{verbatim}

Characteristic of many compiler passes, most of the work is manipulating the tree, rather than, say, number 
crunching. Almost universally, compilers use a combination of recursion and branching to implement these 
transformations, such as using the visitor pattern, Nanopass framework [19], or direct structural recursion. 
The “work” happens in the control flow of the program text. The ubiquity of this approach warrants note. 
Let us savour such rare solidarity for a moment. Now, let us challenge it.

Let us arbitrarily restrict ourselves to a data-flow, data-parallel model. Specifically, consider a language 
whose core primitives are pointwise functions over n-dimensional arrays. We have a set of operators that 
take functions and apply them over different sub-parts of arrays. We can compose these functions together. 
That is all. Can we implement a compiler in such a language? The obvious answer is, yes, given arbitrary 
primitives. What if we restrict ourselves to standard and general-purpose primitives? The answer is still 
yes. Furthermore, the method is surprisingly direct and concise. The language we use is a subset of Dyalog 
APL 14.0 and the reader may follow the examples in the following sections using TryAPL.org.

The trick to such a compiler is solving the issue illustrated by the previous examples, expressing tree 
manipulation without embedding the core logic in the control flow. We accomplish this by choosing an 
appropriate representation of the AST and a property of each node we call its node coordinate. These 
coordinates allow us to locally answer the inter-node questions we need without complex control flow. When 
we combine this with a general-purpose operator called “Key,” we can perform arbitrary computation over 
sub-trees that we partition based on these inter-node relationships, such as whether one node is an ancestor 
of another. When combined with more mundane array programming, we can implement the complete core of a 
compiler, including lexical scope, function lifting, expression flattening/normalization, loop fusion, 
frame/register allocation, and so on. The result is a completely data-parallel compiler modulo parsing and 
some aspects of code generation.

The notation used throughout this paper is introduced piecemeal as necessary in each relevant section in 
which the notation is first used. All functions are applied infix and take either one or two arguments. All 
expressions evaluate from right to left. Tables 1 and 2 provide a reference for the notation and primitives 
as a reference aid.

%[of]:Contributions
\paragraph{Contributions}

\begin{itemize}[noitemsep]
\item We introduce node coordinates as a means to locally compute inter-node properties without using complex 
control flow.
\item We demonstrate methods to perform arbitrary computation over sub-trees partitioned by these 
inter-node properties in a pure data-flow, data-parallel style.
\item The presented methods are direct and concise; the data structures, simple.
\item These methods require no specialized operations, but instead, use only established, general-purpose 
array programming.
\end{itemize}
%[cf]
%[cf]
%[of]:A Definition of Node Coordinates
\section{A Definition of Node Coordinates}

Every node in an AST has a node coordinate, named because a coordinate is a precise location in a space. 
We can imagine all the nodes arranged inside some multi-dimensional space, leading to a specific set of 
node coordinates. Many such arrangements exist, of varying usefulness. In our case, any arrangement 
should allow us to answer the following questions about any two arbitrary nodes in a tree:

\begin{enumerate}[noitemsep]
\item Are the nodes the same?
\item Are they siblings?
\item Does one appear “earlier” in the tree?
\item Are they at the same depth?
\item Is one an ancestor of another?
\end{enumerate}

In short, we care about the relative position of each node in a tree relative to any other. We define a node 
coordinate as follows.

A node coordinate is a vector whose length is the depth of the tree. Its elements are natural numbers. 
The count of non-zero elements in the vector is equal to the depth of that node in the tree. All zero elements
appear after the non-zero elements. That is, a coordinate is zero-padded on the right. When ordered 
lexicographically, the nodes for each coordinate appear in order according to a depth-first pre-order 
traversal of the tree. Each coordinate uniquely identifies a single node. Every ancestor’s coordinate is a 
prefix of any child’s coordinate, ignoring zeros. From the above it follows that every node is lexicographically 
greater than its left sibling and differs from it by exactly one non-zero element, and that this element is the 
last non-zero element in the coordinate.

To understand how we construct these coordinates we must consider how we represent the AST.
%[cf]
%[of]:Encoding the AST
\section{Encoding the AST}

We represent the AST as a matrix of 3 columns and n rows, one for each node in the tree. The first column 
contains the inter-node relationships in the form of a depth vector. The second column is a vector of the 
node types, while the third contains the “value” of the node, such as the name in a variable.

Parsing the examples above gives the following ASTs:

\begin{verbatim}
      F(f)              E(v)
       │         ┌───────┼──────┐
       E         E      P(÷)    E
     ┌─┴─┐  ┌────┼────┐    ┌────┼────┐
     F   A V(a) P(+) V(b) V(c) P(×) V(d)
     │   │
     E  N(7) 
 ┌───┼────┐
V(⍵) P(+) V(⍵)
\end{verbatim}

The depth vector for these trees we name Fd and Ed, respectively:

\begin{verbatim}
Fd←0 1 2 3 4 4 4 2 2 3
Ed←0 1 2 2 2 1 1 2 2 2
\end{verbatim}

The node types we call Ft and Et:

\begin{verbatim}
Ft←’FEFEVPVAN’
Et←’EEVPVPEVPV’
\end{verbatim}

And finally, we call the values vectors Fv and Ev:

\begin{verbatim}
Fv←’f’ 0 0 0 ‘⍵’ ‘+’ 0 7
Ev←’v’ 0 ‘a’ ‘+’ ‘b’ ‘÷’ 0 ‘c’ ‘×’ ‘d’
\end{verbatim}

We combine these to form the respective AST matrices. We write A,B to catenate arrays A and B along their 
last axes and ⍪A to turn a vector into a 1-column matrix. Thus, the two tree matrices are given by the 
following expressions:

\begin{verbatim}
      Fd,Ft,⍪Fv
0 F f
1 E 0
2 F 0
3 E 0
4 V ⍵
4 P +
4 V ⍵
2 A 0
3 N 7
      Ed,Et,Ev
0 E v
1 E 0
2 V a
2 P +
2 V b
1 P ÷
1 E 0
2 V c
2 P ×
2 V d
\end{verbatim}

Note especially that all inter-node information lives in the depth vector, but that this information requires 
non-local access as described in the previous section to use it. Constructing node coordinates fixes this issue.
%[cf]
%[of]:Constructing Node Coordinates
\section{Constructing Node Coordinates}

In building the node coordinates, our goal is to build a matrix where each row is a node 
coordinate satisfying our previously described requirements. We end up with an n d 
shaped matrix where n is the number of rows and d the depth of the tree. As noted above, 
the depth vector contains all the necessary information, but in the wrong form.

We write f⌿A to reduce the first axis of A using function f. Thus, +⌿V is the sum of the 
elements in vector V. The function x⌈y gives the maximum of its two arguments. So, the 
depth of each tree is given by the following:

\begin{verbatim}
      1+⌈⌿Fd
5
      1+⌈⌿Ed
3
\end{verbatim}

We can obtain the ordered sequence  by writing \verb;⍳n;:

\begin{verbatim}
      ⍳3
0 1 2
\end{verbatim}

So the depths of all nodes that appear in the depth vectors is thus:

\begin{verbatim}
      ⍳1+⌈⌿Fd
0 1 2 3 4 5
      ⍳1+⌈⌿Ed
0 1 2
\end{verbatim}

The function table or outer product of f over vectors U and V is written U ∘.f V giving a U V 
shaped matrix as a result. Thus, (⍳3)∘.×⍳3 gives a small multiplication table:

\begin{verbatim}
      (⍳3)∘.×⍳3
0 0 0
0 1 2
0 2 4
\end{verbatim}

If we use = instead, we have a Boolean identity matrix:

\begin{verbatim}
      (⍳3)∘.=⍳3
1 0 0
0 1 0
0 0 1
\end{verbatim}

If we use ∘.= on the depth vector and its set of depths instead, we see an expanded Boolean 
representation of the depth vector:

\begin{verbatim}
      Fd∘.=⍳1+⌈⌿Fd
1 0 0 0 0
0 1 0 0 0
0 0 1 0 0
0 0 0 1 0
0 0 0 0 1
0 0 0 0 1
0 0 0 0 1
0 0 1 0 0
0 0 0 1 0
      Ed∘.=⍳1+⌈⌿Ed
1 0 0
0 1 0
0 0 1
0 0 1
0 0 1
0 1 0
0 1 0
0 0 1
0 0 1
0 0 1
\end{verbatim}

These matrices let us see the nesting features of each tree more visually, but also 
suggest another step. We can compute a sum scan with +⍀, also called a prefix sum, 
along the first axis. Applying this function on the above matrices leads to an interesting 
result:

\begin{verbatim}
      +⍀Fd∘.=⍳1+⌈⌿Fd
1 0 0 0 0
1 1 0 0 0
1 1 1 0 0
1 1 1 1 0
1 1 1 1 1
1 1 1 1 2
1 1 1 1 3
1 1 2 1 3
1 1 2 2 3
      +⍀Ed∘.=⍳1+⌈⌿Ed
1 0 0
1 1 0
1 1 1
1 1 2
1 1 3
1 2 3
1 3 3
1 3 4
1 3 5
1 3 6
\end{verbatim}

These matrices are lexicographically ordered, and each ancestor shares a common prefix 
with its descendants. They are also unique coordinates. Only the spurious digits at the end 
of each coordinate prevent these matrices from meeting all our requirements for valid node 
coordinates.

The expression V f⍤¯1⊢M applies f to corresponding elements of V and rows of M:

\begin{verbatim}
      3 3⍴⍳9
0 1 2
3 4 5
6 7 8
      (⍳3)+⍤¯1⊢3 3⍴⍳9
0 1  2
4 5  6
8 9 10
\end{verbatim}

If n↑V takes the first n elements of V, then we can obtain coordinate matrices from the 
prefix sums by noting that the spurious digits all come after column d+1 where d is the 
depth corresponding to that coordinate. The following gives a complete expression for 
computing a node coordinate matrix from a depth vector, shown using Fd and Fe:

\begin{verbatim}
      ⊢Fc←(1+Fd)↑⍤¯1⊢+⍀Fd∘.=⍳1+⌈⌿Fd
1 0 0 0 0
1 1 0 0 0
1 1 1 0 0
1 1 1 1 0
1 1 1 1 1
1 1 1 1 2
1 1 1 1 3
1 1 2 0 0
1 1 2 2 0
      ⊢Ec←(1+Ed)↑⍤¯1⊢+⍀Ed∘.=⍳1+⌈⌿Ed
1 0 0
1 1 0
1 1 1
1 1 2
1 1 3
1 2 0
1 3 0
1 3 4
1 3 5
1 3 6
\end{verbatim}

A careful study of the definition of a node coordinate and the above construction should 
reveal why this works. Intuitively, we are creating a multi-dimensional space or a number 
system in which each digit place or dimension contains or circumscribes a smaller space in 
which are contained all the descendant nodes that appear lower in the tree. Each coordinate 
is a sort of special path through the tree encoded to have desirable properties relative to 
other paths.
%[cf]
%[of]:Operations on Node Coordinates
\section{Operations on Node Coordinates}

The simplest operation over a node coordinate is to extract the depth of the node. The 
expression C⍳0 finds the first occurrence of 0 in C and returns the index of that 
occurrence. (The application of ⍳ here is its dyadic form, meaning that it receives two 
arguments. Previous uses of ⍳ have used its monadic form, which generates all indices, 
instead of finding a specific index. Most APL functions have a dyadic and monadic form.) 
Thus, the depth of a node is given by the following:

\begin{verbatim}
      C←1 1 2 0 0
      ¯1+C⍳0
2
\end{verbatim}

The primary calculation we care about in order to do function lifting, expression flattening, 
and other sorts of flattening transformations is to determine which nodes are ancestors of 
another node. The core of this computation determines whether one node is a child of another 
based on their node coordinates. This amounts to computing whether one node is a prefix of 
another, ignoring zeros.

We write (f g h) to represent the composition of functions f, g, and h as a function train. This i
s used with another shorthand of composition according to the following equivalences:

\begin{verbatim}
A(f g h)B ←→ (A f B) g (A h B)
A(0 f h)B ←→ 0 f (A h B) ⍝ Any constant for 0
A(f g)B   ←→ f (A g B)
\end{verbatim}

With this, we can write a function to compute whether a given coordinate is a prefix of another.

\begin{verbatim}
      P←1 1 0 0 0
      C(=∨0=⊢)P
1 1 1 1 1
      ∧⌿C(=∨0=⊢)P
1
\end{verbatim}

Here P is the coordinate we wish to check against C, to determine whether C is a descendant. The
 functions are all standard logical functions extended to apply point-wise over arrays. The 
 function ⊢ always returns its right argument. The function (=∨0=⊢) reads intuitively as, “equal or 
 right element is zero.” In this case to make this a predicate we combine this with the “for all” 
 reduction using ∧⌿. This particular pattern is a special case of inner product, which we can compute 
 using f.g. Thus, +.× is a function that computes traditional matrix-matrix multiplication when applied 
 to two matrices. However, we can also use it to compute the above reduction and application:

\begin{verbatim}
      C∧.(=∨0=⊢)P
1
\end{verbatim}

This extends the reduction to allow us to use matrix values for C and P instead of vectors, and thus 
determine whether rows in C are descendants of multiple candidates given by P. This becomes important 
to using the prefix function. In our examples, we want to determine the nearest ancestor of a certain 
node type to which a node belongs. In the case of function lifting, we want to determine the nearest 
ancestor function, in the case of expression flattening, we want to determine the nearest ancestor 
expression node. We can extract the rows we care about for each example using ⌷, where i⌷M returns 
the i-th row of M. In the following expressions, we use 0 2 rather than ¯1 in ⍤ to indicate that ⌷ will 
receive the entire matrix on the right for each scalar index on the left.

\begin{verbatim}
      ⊢Fp←0 2⌷⍤0 2⊢Fc
1 0 0 0 0
1 1 1 0 0
      ⊢Ep←0 1 6⌷⍤0 2⊢Ec
1 0 0
1 1 0
1 3 0
\end{verbatim}

The matrices Fp and Ep correspond to the function and expression node coordinates in our two 
examples, respectively. We could now use Fp and Ep to compare against Fc and Ec and determine 
which nodes contained which nodes in the tree. However, our prefix function will return 1 when we 
ask whether a node is a prefix of itself. In this case we don’t want that. The solution to this is to 
drop the last non-zero element from the coordinate. This will not prevent it from matching against 
any of its ancestors, but will prevent matching against itself. We can use the depth vector with the 
↑ function to take all but the last non-zero element. We need to remember to extend the returned 
array with an extra zero column, since the shape will be too small otherwise.

\begin{verbatim}
      ⊢Fcp←(Fd↑⍤¯1⊢Fc),0
0 0 0 0 0
1 0 0 0 0
1 1 0 0 0
1 1 1 0 0
1 1 1 1 0
1 1 1 1 0
1 1 1 1 0
1 1 0 0 0
1 1 2 0 0
      ⊢Ecp←(Ed↑⍤¯1⊢Ec),0
0 0 0
1 0 0
1 1 0
1 1 0
1 1 0
1 0 0
1 0 0
1 3 0
1 3 0
1 3 0
\end{verbatim}

We can now use Fcp and Ecp to determine which Function and Expression nodes match for each node:

\begin{verbatim}
     Fcp∧.(=∨0=⊢)⍉Fp
0 0
1 0
1 0
1 1
1 1
1 1
1 1
1 0
1 0
      Ecp∧.(=∨0=⊢)⍉Ep
0 0 0
1 0 0
1 1 0
1 1 0
1 1 0
1 0 0
1 0 0
1 0 1
1 0 1
1 0 1
\end{verbatim}

We use the ⍉ function, which computes the transpose of its right argument, to ensure that the row 
and column sizes match up. With the above, we can determine the “greatest” match, which is the closest. 
We can replace each 1 in the above matrices by their column numbers (the number of the column in which 
that 1 occurs), and then we can take the maximum column number in each row to determine an index in 
either Fp or Ep that is the appropriate “parent” coordinate for that node. This works because we have 
made sure to order the Ep and Fp matrices lexicographically. The following shows this computation for 
the expression example; we elide the function example as redundant at this point. We also use ⌈/ instead 
of ⌈⌿ to reduce along the last, rather than the first, axis.

\begin{verbatim}
      ⊢Ei←⌈/(⍳3)×⍤1⊢Ecp∧.(=∨0=⊢)⍉Ep
0 0 1 1 1 0 0 2 2 2
      ⊢Ek←Ei⌷⍤0 2⊢Ep 
1 0 0
1 0 0
1 1 0
1 1 0
1 1 0
1 0 0
1 0 0
1 3 0
1 3 0
1 3 0
\end{verbatim}

At this point we have two values, Ek and Fk, which indicate the closest containing node that we care about 
for each node in the tree, using its node coordinate.

\begin{verbatim}
      Fk
1 0 0 0 0
1 0 0 0 0
1 0 0 0 0
1 1 1 0 0
1 1 1 0 0
1 1 1 0 0
1 1 1 0 0
1 0 0 0 0
1 0 0 0 0
\end{verbatim}

We will use these keys to perform computation over the AST and accomplish our tasks of function lifting and 
expression flattening in the next section, but we make a final note here, that we can imagine many other 
operations which might be used throughout the compiler when we care about how nodes relate to each 
other. The use of operators like reduction, scan, inner and outer products allow us to obtain the necessary 
information from the node coordinate matrix. Sometimes these results might not be obvious, but the 
example of the “belongs to” relationship given above demonstrates a pattern that arises repeatedly 
throughout our data parallel compiler, and is therefore particularly useful to understand. 
%[cf]
%[of]:Computing with Node Coordinates
\section{Computing with Node Coordinates}

Given the above keys and grouping information, we now have all the information we need to do 
function lifting and expression flattening, but we lack a suitable operator to complete the task. 
Indeed, it isn’t immediately apparent that a single, general purpose operator over arrays exists 
which can accomplish such a task. However, the Key operator, while masquerading as a useful business 
analytics operator, represents just such a useful tree transformation tool.

%[of]:The Key Operator
\subsection{The Key Operator}

The Key operator (written \verb;⌸;), takes a single function that expects two arguments and returns 
a function which takes two arguments. The left argument is a set of keys, and the right argument 
is a set of corresponding elements associated with those keys. In our case, we provide matrices 
for both of these arguments, so each pair of rows corresponds to a key-value pair. A simple example 
of the key operation at work is to compute a histogram (≢ here can be thought of as “tally” or count):

\begin{verbatim}
       ⊢X←?10⍴5
1 1 4 0 1 3 1 1 2 1
      X(⊣,(≢⊢))⌸X
1 6
4 1
0 1
3 1
2 1
\end{verbatim}

To understand a bit better how the Key operator applies its function, consider the function {⍺ ⍵} 
which returns the pair of its right and left arguments. If we apply it to the same value as above, we 
get the following:

\begin{verbatim}
      X{⍺ ⍵}⌸X
1  1 1 1 1 1 1 
4  4           
0  0           
3  3           
2  2
\end{verbatim}

In our case, we use either Fk or Ek as our keys applied to the corresponding AST. We also will drop off 
the first row in each AST using 1↓ since this node “contains” everything. In a complete AST this is 
usually the Module boundary node which contains the entire set of functions and values in the module.
%[cf]
%[of]:Function Lifting
\subsection{Function Lifting}

If we use the key operator with Fk, we get the following (the monadic use of ↓ converts Fc from a 
matrix to a vector of vectors):

\begin{verbatim}
      (1↓Fk){⍺ ⍵}⌸1↓Fd,Ft,Fv,⍪↓Fc
┌─────────┬─────────────────┐
│1 0 0 0 0│┌─┬─┬─┬─────────┐│
│         ││1│E│0│1 1 0 0 0││
│         │├─┼─┼─┼─────────┤│
│         ││2│F│0│1 1 1 0 0││
│         │├─┼─┼─┼─────────┤│
│         ││2│A│0│1 1 2 0 0││
│         │├─┼─┼─┼─────────┤│
│         ││3│N│7│1 1 2 2 0││
│         │└─┴─┴─┴─────────┘│
├─────────┼─────────────────┤
│1 1 1 0 0│┌─┬─┬─┬─────────┐│
│         ││3│E│0│1 1 1 1 0││
│         │├─┼─┼─┼─────────┤│
│         ││4│V│⍵│1 1 1 1 1││
│         │├─┼─┼─┼─────────┤│
│         ││4│P│+│1 1 1 1 2││
│         │├─┼─┼─┼─────────┤│
│         ││4│V│⍵│1 1 1 1 3││
│         │└─┴─┴─┴─────────┘│
└─────────┴─────────────────┘
\end{verbatim}

Notice that we have now grouped all of the relevant parts of the tree according to which nodes would 
appear in their respective functions after lifting. Refer to the original lifting example in the 
introduction to verify this. Indeed. The second row in the above example shows the internal function 
complete and ready to name. Each element in the second column corresponds to the body of one of our 
lifted functions. In the case of the first function, the outer function, we have a spurious function node 
in the body. This is intentional. When we lift these functions, we will replace each spurious function 
node with a variable node referring to the function’s generated name.

Each of these function bodies has a specific coordinate associated with it. Because these coordinates 
are uniquely identifying, we can use these as input into a name generator to generate names that we 
know are unique for each function body. Furthermore, because we retain this information in the 
corresponding function nodes that appear in the body of each function to be lifted, we know exactly what 
name that function has been given, and we can replace the function node with a variable node referencing 
that name instead, without referring to any state outside of the immediate information given to the 
function lifter. Indeed, each row in the above matrix represents a function lifting task that can be 
completed without any additional information. That is, there are no dependencies between rows to perform 
lifting. This gives us a straightforward parallel execution of function lifting.

The final task to complete function lifting of each function body before lifting is to shift the depths in 
the depth vectors to correspond to those of a function lifted to the top-level, and to attach a function 
node to the top of each of the bodies. At that point, we simply recombine all of the newly created functions 
into a single top level. 
%[cf]
%[of]:Expression Flattening
\subsection{Expression Flattening}

If we use Ek as the key for the expression example, we get the following:

\begin{verbatim}
      (1↓Ek){⍺ ⍵}⌸1↓Ed,Et,Ev,⍪↓Ec
┌─────┬─────────────┐
│1 0 0│┌─┬─┬─┬─────┐│
│     ││1│E│0│1 1 0││
│     │├─┼─┼─┼─────┤│
│     ││1│P│÷│1 2 0││
│     │├─┼─┼─┼─────┤│
│     ││1│E│0│1 3 0││
│     │└─┴─┴─┴─────┘│
├─────┼─────────────┤
│1 1 0│┌─┬─┬─┬─────┐│
│     ││2│V│a│1 1 1││
│     │├─┼─┼─┼─────┤│
│     ││2│P│+│1 1 2││
│     │├─┼─┼─┼─────┤│
│     ││2│V│b│1 1 3││
│     │└─┴─┴─┴─────┘│
├─────┼─────────────┤
│1 3 0│┌─┬─┬─┬─────┐│
│     ││2│V│c│1 3 4││
│     │├─┼─┼─┼─────┤│
│     ││2│P│×│1 3 5││
│     │├─┼─┼─┼─────┤│
│     ││2│V│d│1 3 6││
│     │└─┴─┴─┴─────┘│
└─────┴─────────────┘
\end{verbatim}

Again, we can see immediately that we have grouped each set of nodes according to the expressions 
that are to be lifted. Just as in the case of function lifting, we can adjust the depths of each 
expression to the correct depth and we can replace each expression node with a variable reference
 based on that node’s coordinate. Each expression can be given a unique name based on its 
 coordinate. A later compiler pass can reduce these names down to the minimum actually required to 
 represent the expression if desired.

The only extra issue involved here is to ensure that the order of evaluation matches. In our case, we 
are assuming that the order of evaluation is right to left, which means that the above order is actually 
backwards of the desired order. During recombination, we simply reverse these orders and this fixes 
that problem. More work would be required to take into consideration a specific precedence hierarchy. 
%[cf]
%[cf]
%[of]:Scaling up, Considerations and Thoughts
\section{Scaling up, Considerations and Thoughts}

How well do these techniques work in practice? We have applied these techniques to an entire compiler, 
with the same restrictions as used here. The result was a success, and the core of the compiler, which 
is everything but parsing and code generation, is contained in a single file consisting of only 40 lines. 
In 12 point font, the compiler easily fits on a single sheet of paper. It produces C and CUDA code.

Not every pass in the compiler requires this sort of operation. Indeed, if flattening occurs early and 
in the correct order, it’s possible to use much simpler methods of grouping using only a prefix sum 
scan instead of operating over the node coordinates. It’s also interesting to note that we do not 
need to recompute the node coordinates throughout the compiler. Indeed, this would defeat their 
purpose somewhat, as the node coordinates allow us to understand the original position of a node, 
even if that node has now been lifted elsewhere. We use this information when resolving lexical scope, 
for instance.

When writing data-parallel code, duplicated work is common. The existing compiler contains little to 
no inherently duplicated work. Most of the duplicated work exists for brevity and clarity of the code, 
rather than for any required reason, and could easily be factored out.

Finally, it is interesting to note that the simplified control flow gives rise to some positive side-effects. 
Besides brevity and directness, the complexity of each operation and the composition of operations is 
well known. We have not yet tried to do this, but it is straightforward to derive a worst-case complexity 
bound on code written in this style, so simple, in fact, that it can be accomplished automatically. It is 
conceivable then, that the compiler could produce not only an executable program from code written 
in this style, but also a complexity bound on its execution. 
%[cf]
%[of]:Related Work
\section{Related Work}

The J programming language [18] was the first practical, general-purpose programming language 
to introduce the key operator as a primitive operator with the presumption of its general 
usefulness. The rank operator (⍤) used throughout this treatment also derives from the J 
traditional, receiving particular interest throughout the APL community [1, 17].

Fritz Henglein demonstrated a class of operations, called discriminators, of which the Key operator 
is a member. [16] Namely, a discriminator performs the same grouping computation as Key, but does 
not apply a function over these groups with their keys. Henglein provides a linear implementation 
of these operations.

The EigenCFA effort [22] demonstrated significant performance improvements of a 0-CFA flow 
analysis by utilizing similar techniques to those demonstrated here. In particular, encoding the 
AST and using accessor functions have a very similar feel to the node coordinates and AST 
encoding given here, though they have a different formulation and spend considerable effort 
understanding the trade-offs of performance associated with the different encodings, whereas 
the encodings here were chosen for their clarity and directness, rather than their performance.

Mendez-Lojo, et al. implemented a GPU version of Inclusion-based Points-to Analysis [20] that also 
focuses on adapting data structures and algorithms to efficiently execute on the GPU. In particular, 
they use similar techniques of prefix sums and sorts to achieve some of their adaptation to the GPU, 
Additionally, they have clever and efficient methods of representing graphs on the GPU which enable 
dynamic rewriting of the graph.

The APEX compiler [2] developed vectorized approaches to handling certain analyses to compile 
traditional APL, including a SIMD tokenizer [4]. It uses a SSA representation, and converts the
dynamic scope of traditional APL functions into a static form early on. It also uses a matrix 
format to represent the AST. Traditional APL did not have nested function definitions, however, 
and thus the APEX compiler does not have any specific approaches to dealing with function lifting.

Bernecky further identified methods of reducing or optimizing the computational complexity or cost 
of certain array operations, allowing improved performance of easy to understand array expressions [3].

Timothy Budd implemented a compiler [5, 6] for APL which targeted vector processors as well as one 
for C code. They used a method of lazy evaluation to avoid intermediate data copying. Budd provided 
thoughts and some ideas on how the compiler might be implemented in parallel as well.

Walter Schwarz implemented an APL to C compiler for the ACORN system targeting the CM-2 machine, 
demonstrating performance potential for APL as a massively parallel language [21].

W. Ching and D. Ju have spent significant work on the ELI language and other APL-class language
 implementations, especially on parallelized code and optimization. [8, 9, 10, 11, 13, 14, 15]

J. D. Bunda and J. A. Gerth presented a method for doing table driven parsing of APL which 
suggested a parallel optimization for parsing, but did not elucidate the algorithm [7].
%[cf]
%[of]:Conclusion
\section{Conclusion}

We have derived a method of performing computation over sub-trees selected on the basis of 
inter-node properties through the use of the Key (⌸) operator and node coordinates, which enable 
local computation of these inter-node properties. This method is both general and direct, and when 
combined with traditional and more mundane array programming, suffices to implement the 
complete core of a compiler, modulo parsing and code generation. The method requires no special 
operations or unique special casing primitives in the language. Moreover, it is strictly data-parallel 
and data-flow, without any complex control flow, which not only results in concise code, but also 
enables much easier analysis of properties such as runtime complexity and verification of correctness.

We have demonstrated the technique and the core insights behind the data structures involved. It 
presents a solution to a very old and traditional problem in a very uncommon light, by eschewing 
the common practices that underlie every other significant and general solution found in modern 
compilers today and replacing them with an entirely different paradigm centred on parallelism 
and aggregate operations.
%[cf]
%[of]:Acknowledgements
%[cf]

% Bibliography goes here

\end{document}

### 2015-10-30 2016-pldi/2016-pldi-key_to_data_parallel_compiler.tex
Add some notes and outline
Rewrite the Introduction

@@
 \begin{abstract}
-By choosing an appropriate representation for the AST, we can implement a compiler for a modern functional 
-language using only data-parallel array operations under simple functional composition without the use of 
-branching, recursion, or other means of explicit control flow. In this pearl we elucidate the core data structure 
-and an operator, called Key, that enable a surprisingly direct implementation of non-trivial transformations. 
-The result permits arbitrary computation over sub-trees chosen by parent-child and other inter-node 
-relationships in a pure data-parallel fashion without complex control flow. We give direction on the use of this 
-method through examples of function lifting and expression flattening.
+TBW
 \end{abstract}
+
+%[c]Proposed new outline:
+%[c]
+%[c]I. 	Generalization on manipulations over Sub-tree computation
+%[c]  A.	Key + Encoding → Data parallel, Array-based computation
+%[c]  B.	Easy to write/read
+%[c]  C.	Uniform across passes
+%[c]II.	Case Studies
+%[c]  A.	Function Lifting
+%[c]  B.	Expression Flattening
+%[c]  C.	Lexical Scope Resolution
+%[c]  D.	Illustrates the problem space
+%[c]III.	Other passes/Issues
+%[c]  A.	Discussion of other passes
+%[c]  B.	Beyond the Key Operator
+%[c]IV.	In practice/Co-dfns Compiler
+%[c]V.	Performance/Results
+%[c]
+%[c]Questions that must be addressed:
+%[c]
+%[c]1. Why the formalization in APL?
+%[c]2. Why parallelize a compiler?
+%[c]3. How well does it work/scale/generalize
 
@@
 \section{Introduction}
+
+Compilers are everywhere, but their design and structure does not scale well to modern 
+parallel architectures. Traditional compilers are very difficult to parallelize, and the benefits of 
+modern architectures is often seen only through the use of distributed build systems which invoke 
+the single threaded compilers on multiple ``documents'' at the same time. Future CPU architectures 
+promise to be even more data parallel, such as the Xeon Phi and the GPGPU architectures. On these 
+systems it is even harder to obtain good performance for compilers written in the traditional fashion.
+
+Fundamentally, a compiler is a translator that operates over some sort of tree and transforms it to 
+another tree of an usually radically different shape but equivalent ``semantics.'' These algorithms and 
+transformations show up everywhere, from document processing on the Web, XML-based publishing 
+workflows, to the programming language compilers with which most developers are intimately familar. 
+These transformations differ from other widely studied parallel graph problems in that they tend to greatly 
+alter the fundamental structure of the graph/tree, unlike, say, breadth-first search or single-source 
+shortest path algorithms. This rapidly shifting graph structure requires different approaches to 
+achieving good parallelism.
+
+There has been some interesting use of GPUs to accelerate certain compiler analyses and passes. 
+Unfortunately, attempts to parallelize compilers usually takes the form of a great deal of work spent to 
+highly optimize a single compiler pass. The results are often fragile and require a great deal of effort 
+to maintain and keep in line with the rest of the compiler. This integration problem makes it difficult to 
+scale the "ad hoc" parallelization effort of individual compiler passes to a complete system. This is 
+particularly true in situations where the compiler passes may change rapidly, such as with XSLT or 
+document publishing, where it is very common for the document preparer to write or tweak transformation 
+passes to suit the specific class of documents being handled. In these cases, the expertise and time 
+involved in parallelizing such transformations makes the ad hoc approach mostly intractable.
+
+Furthermore, compilers are increasingly becoming a part of a "live" and interactive feedback loop. 
+Just in time compilation and on the fly static compilation are both used to provide custom binaries tuned 
+to specific tasks on the fly. These techniques are used in web services and large scale databases, where 
+in-memory databases may construct functions on the fly to execute. There is great interest in moving 
+such systems on to the GPU or other highly data-parallel architectures, but it would significantly hamper 
+the performance of the system if the compilation phase of query execution were required to occur on the 
+CPU, especially at scale where there may be many more GPU accelerators than CPU resources.
+
+We propose a language-driven approach to compiler construction that provides a framework for the 
+development of arbitrary compiler transformations that are data-parallel by construction. These 
+techniques intentionally optimize for the data-parallel case, which permits a wide degree of platform 
+independent performance compared to the traditional approach to compiler design, which requires 
+significant effort and possibly signficiant transformation of the source code to achieve good parallel 
+performance on high-parallel architectures.
+
+We use a mathematically oriented, array based language that is accessible and popular with information 
+scientists. We restrict this language to permit only the composition of data-flow oriented array based 
+programs whose core primitives are all data-parallel by definition. The permitted compositions of these 
+primitives ensures that the resulting programs are both analyzable and parallel by construction, permitting 
+a wide array of whole program and transformation specific optimizations that have been long studied. 
+The notation is intentially close to mathematical notation, and as such, the resulting compiler passes 
+more closely resemble the compact formal descriptions of the transformations rather than complex 
+recursive algorithms for transforming trees.
+
+At the core of such techniques, we must have some method for transforming trees in a controlled and 
+predictable manner. We build on the Nanopass paradigm for compiler construction, wherein a compiler is 
+developed as a series of small, functional tree transformations (passes) that have an input AST language 
+and an output AST language. Through progressive refinement of the tree, the composition of these 
+passes results in the complete compiler. This is a highly functional approach, with very little mutation. 
+Fundamentally, each pass must identify regions of the tree upon which it works, do some sort of computation 
+on that sub-tree that has been matched (often through pattern matching on the nodes) and reassemble 
+the tree in some new order that matches the desired structure. In other words, a compiler nanopass 
+can be thought of as an arbitrary computation over a given selection of sub-trees in an AST, where 
+this computation could move, relocate, or change the sub-tree.
+
+The key to our language-drive approach is a set of techniques that enable a data-parallel expression of 
+these arbitrary transformations. By using a higher-order function called Key, together with basic array 
+programming techniques and an encoding for the tree connections called a Node Coordinate matrix, we can 
+identify sub-trees and select them for computation by their parent-child relationships in a data-parallel 
+fashion, and then compute on these sub-trees using traditional array programming techniques.
+
+We have constructed a complete compiler for a superset of our language called Co-dfns. This compiler is 
+part of a commercial effort to develop platform independent performance for array programs. The 
+compiler is developed completely in a data-parallel by construction as described in this paper, and uses 
+only traditional and well known array primitives together with function composition to form the entire 
+core compiler passes and generator. As a result, the compiler takes the form of a pure data-flow graph 
+without mutation and has an extremely uniform and concise definition. It uses no recursion, pattern 
+matching, branching, or other forms of explicit looping. The compiler, together with its runtime libraries 
+and parser, comes in under 1000 lines of code.
+
+In this paper we fully elucidate three compiler passes from the Co-dfns compiler to demonstrate the 
+techniques used and we provide further analysis of how these techniques are used in the other compiler 
+passes. We further provide a comparative analysis of these passes against their counterparts implemented 
+using the Nanopass compiler framework, comparing performance both in terms of programmer effort and 
+portability/maintainability, as well as runtime performance.
+
+\paragraph{Contributions}
+
+\begin{itemize}[noitemsep]
+\item We describe a method of computing over arbitrary sub-trees selected by their parent-child 
+relationships in a data-parallel manner using the Key operator and Node Coordinates.
+\item We situate this technique into a broader, language-driven compiler construction strategy 
+that enables the development of parallel by construction compilers that exhibit desirable programmability 
+features in terms of code size, complexity, portability, and maintainability. 
+\item We provide analysis of these techniques as used in a commercial compiler project called Co-dfns and 
+report on the results.
+\item We compare these techniques to the Nanopass approach to compiler design in terms of programmer 
+productivity and compiler runtime performance. 
+\item We enumerate a set of programming techniques, patterns, and strategies that complement the 
+Key operator and Node Coordinates in the construction of tree transformations in general and give some 
+intuitions on the use of these techniques on arbitrary transformations.
+\end{itemize}
+%[cf]
+%[of]:Introduction
+\section{Introduction (Old)}
 

### 2015-11-11 2016-pldi/2016-pldi-key_to_data_parallel_compiler.tex
Reorganize things in the PLDI paper to match the new outline.
Add in a table for compiler passes.

@@
 %[cf]
-%[of]:Introduction
-\section{Introduction (Old)}
+%[of]:Generalized Sub-tree Computation
+\section{Data Parallel Sub-tree Computation}
 
-When compiling a high-level language into a low-level language, we often implement a variety of 
-flattening transformations to normalize more complex code exhibiting nesting into flat programs 
-that are semantically equivalent. Function lifting, for example, takes a program with nested functions 
-into an equivalent program in which all functions are at the root of the program. The following function
-f always returns 14.
+%[of]:The Key Operator
+\subsection{The Key Operator}
+
+The Key operator (written \verb;⌸;), takes a single function that expects two arguments and returns 
+a function which takes two arguments. The left argument is a set of keys, and the right argument 
+is a set of corresponding elements associated with those keys. In our case, we provide matrices 
+for both of these arguments, so each pair of rows corresponds to a key-value pair. A simple example 
+of the key operation at work is to compute a histogram (≢ here can be thought of as “tally” or count):
 
 \begin{verbatim}
-f←λ.(λ ⍵.(⍵+⍵))7
+       ⊢X←?10⍴5
+1 1 4 0 1 3 1 1 2 1
+      X(⊣,(≢⊢))⌸X
+1 6
+4 1
+0 1
+3 1
+2 1
 \end{verbatim}
 
-A function lifting pass would transform this into the following:
+To understand a bit better how the Key operator applies its function, consider the function {⍺ ⍵} 
+which returns the pair of its right and left arguments. If we apply it to the same value as above, we 
+get the following:
 
 \begin{verbatim}
-fn1←λ⍵.⍵+⍵
-f←λ.fn1 7
+      X{⍺ ⍵}⌸X
+1  1 1 1 1 1 1 
+4  4           
+0  0           
+3  3           
+2  2
 \end{verbatim}
 
-Expression flattening works in a similar fasion. Assuming all expressions evaluate right to left, this expression:
-
-\begin{verbatim}
-v←(a+b)÷c×d
-\end{verbatim}
-
-Becomes this:
-
-\begin{verbatim}
-v←c×d
-t←a+b
-v←t÷v
-\end{verbatim}
-
-Characteristic of many compiler passes, most of the work is manipulating the tree, rather than, say, number 
-crunching. Almost universally, compilers use a combination of recursion and branching to implement these 
-transformations, such as using the visitor pattern, Nanopass framework [19], or direct structural recursion. 
-The “work” happens in the control flow of the program text. The ubiquity of this approach warrants note. 
-Let us savour such rare solidarity for a moment. Now, let us challenge it.
-
-Let us arbitrarily restrict ourselves to a data-flow, data-parallel model. Specifically, consider a language 
-whose core primitives are pointwise functions over n-dimensional arrays. We have a set of operators that 
-take functions and apply them over different sub-parts of arrays. We can compose these functions together. 
-That is all. Can we implement a compiler in such a language? The obvious answer is, yes, given arbitrary 
-primitives. What if we restrict ourselves to standard and general-purpose primitives? The answer is still 
-yes. Furthermore, the method is surprisingly direct and concise. The language we use is a subset of Dyalog 
-APL 14.0 and the reader may follow the examples in the following sections using TryAPL.org.
-
-The trick to such a compiler is solving the issue illustrated by the previous examples, expressing tree 
-manipulation without embedding the core logic in the control flow. We accomplish this by choosing an 
-appropriate representation of the AST and a property of each node we call its node coordinate. These 
-coordinates allow us to locally answer the inter-node questions we need without complex control flow. When 
-we combine this with a general-purpose operator called “Key,” we can perform arbitrary computation over 
-sub-trees that we partition based on these inter-node relationships, such as whether one node is an ancestor 
-of another. When combined with more mundane array programming, we can implement the complete core of a 
-compiler, including lexical scope, function lifting, expression flattening/normalization, loop fusion, 
-frame/register allocation, and so on. The result is a completely data-parallel compiler modulo parsing and 
-some aspects of code generation.
-
-The notation used throughout this paper is introduced piecemeal as necessary in each relevant section in 
-which the notation is first used. All functions are applied infix and take either one or two arguments. All 
-expressions evaluate from right to left. Tables 1 and 2 provide a reference for the notation and primitives 
-as a reference aid.
-
-%[of]:Contributions
-\paragraph{Contributions}
-
-\begin{itemize}[noitemsep]
-\item We introduce node coordinates as a means to locally compute inter-node properties without using complex 
-control flow.
-\item We demonstrate methods to perform arbitrary computation over sub-trees partitioned by these 
-inter-node properties in a pure data-flow, data-parallel style.
-\item The presented methods are direct and concise; the data structures, simple.
-\item These methods require no specialized operations, but instead, use only established, general-purpose 
-array programming.
-\end{itemize}
-%[cf]
-%[cf]
-%[of]:A Definition of Node Coordinates
-\section{A Definition of Node Coordinates}
-
-Every node in an AST has a node coordinate, named because a coordinate is a precise location in a space. 
-We can imagine all the nodes arranged inside some multi-dimensional space, leading to a specific set of 
-node coordinates. Many such arrangements exist, of varying usefulness. In our case, any arrangement 
-should allow us to answer the following questions about any two arbitrary nodes in a tree:
-
-\begin{enumerate}[noitemsep]
-\item Are the nodes the same?
-\item Are they siblings?
-\item Does one appear “earlier” in the tree?
-\item Are they at the same depth?
-\item Is one an ancestor of another?
-\end{enumerate}
-
-In short, we care about the relative position of each node in a tree relative to any other. We define a node 
-coordinate as follows.
-
-A node coordinate is a vector whose length is the depth of the tree. Its elements are natural numbers. 
-The count of non-zero elements in the vector is equal to the depth of that node in the tree. All zero elements
-appear after the non-zero elements. That is, a coordinate is zero-padded on the right. When ordered 
-lexicographically, the nodes for each coordinate appear in order according to a depth-first pre-order 
-traversal of the tree. Each coordinate uniquely identifies a single node. Every ancestor’s coordinate is a 
-prefix of any child’s coordinate, ignoring zeros. From the above it follows that every node is lexicographically 
-greater than its left sibling and differs from it by exactly one non-zero element, and that this element is the 
-last non-zero element in the coordinate.
-
-To understand how we construct these coordinates we must consider how we represent the AST.
+In our case, we use either Fk or Ek as our keys applied to the corresponding AST. We also will drop off 
+the first row in each AST using 1↓ since this node “contains” everything. In a complete AST this is 
+usually the Module boundary node which contains the entire set of functions and values in the module.
 %[cf]
 %[of]:Encoding the AST
-\section{Encoding the AST}
+\subsection{Encoding the AST}
 
@@
 %[cf]
+%[of]:Node Coordinates
+\subsection{Node Coordinates}
+
+%[of]:A Definition of Node Coordinates
+\subsubsection{A Definition of Node Coordinates}
+
+Every node in an AST has a node coordinate, named because a coordinate is a precise location in a space. 
+We can imagine all the nodes arranged inside some multi-dimensional space, leading to a specific set of 
+node coordinates. Many such arrangements exist, of varying usefulness. In our case, any arrangement 
+should allow us to answer the following questions about any two arbitrary nodes in a tree:
+
+\begin{enumerate}[noitemsep]
+\item Are the nodes the same?
+\item Are they siblings?
+\item Does one appear “earlier” in the tree?
+\item Are they at the same depth?
+\item Is one an ancestor of another?
+\end{enumerate}
+
+In short, we care about the relative position of each node in a tree relative to any other. We define a node 
+coordinate as follows.
+
+A node coordinate is a vector whose length is the depth of the tree. Its elements are natural numbers. 
+The count of non-zero elements in the vector is equal to the depth of that node in the tree. All zero elements
+appear after the non-zero elements. That is, a coordinate is zero-padded on the right. When ordered 
+lexicographically, the nodes for each coordinate appear in order according to a depth-first pre-order 
+traversal of the tree. Each coordinate uniquely identifies a single node. Every ancestor’s coordinate is a 
+prefix of any child’s coordinate, ignoring zeros. From the above it follows that every node is lexicographically 
+greater than its left sibling and differs from it by exactly one non-zero element, and that this element is the 
+last non-zero element in the coordinate.
+
+To understand how we construct these coordinates we must consider how we represent the AST.
+%[cf]
 %[of]:Constructing Node Coordinates
-\section{Constructing Node Coordinates}
+\subsubsection{Constructing Node Coordinates}
 
@@
 %[of]:Operations on Node Coordinates
-\section{Operations on Node Coordinates}
+\subsubsection{Operations on Node Coordinates}
 
@@
 %[cf]
-%[of]:Computing with Node Coordinates
-\section{Computing with Node Coordinates}
 
-Given the above keys and grouping information, we now have all the information we need to do 
-function lifting and expression flattening, but we lack a suitable operator to complete the task. 
-Indeed, it isn’t immediately apparent that a single, general purpose operator over arrays exists 
-which can accomplish such a task. However, the Key operator, while masquerading as a useful business 
-analytics operator, represents just such a useful tree transformation tool.
+%[cf]
+%[cf]
+%[of]:Case Studies
+\section{Case Studies}
 
-%[of]:The Key Operator
-\subsection{The Key Operator}
-
-The Key operator (written \verb;⌸;), takes a single function that expects two arguments and returns 
-a function which takes two arguments. The left argument is a set of keys, and the right argument 
-is a set of corresponding elements associated with those keys. In our case, we provide matrices 
-for both of these arguments, so each pair of rows corresponds to a key-value pair. A simple example 
-of the key operation at work is to compute a histogram (≢ here can be thought of as “tally” or count):
-
-\begin{verbatim}
-       ⊢X←?10⍴5
-1 1 4 0 1 3 1 1 2 1
-      X(⊣,(≢⊢))⌸X
-1 6
-4 1
-0 1
-3 1
-2 1
-\end{verbatim}
-
-To understand a bit better how the Key operator applies its function, consider the function {⍺ ⍵} 
-which returns the pair of its right and left arguments. If we apply it to the same value as above, we 
-get the following:
-
-\begin{verbatim}
-      X{⍺ ⍵}⌸X
-1  1 1 1 1 1 1 
-4  4           
-0  0           
-3  3           
-2  2
-\end{verbatim}
-
-In our case, we use either Fk or Ek as our keys applied to the corresponding AST. We also will drop off 
-the first row in each AST using 1↓ since this node “contains” everything. In a complete AST this is 
-usually the Module boundary node which contains the entire set of functions and values in the module.
+%[of]:Table of Compiler Passes
+\begin{table*}
+\centering
+\begin{tabular}{l p{2.4in} l l l}
+\textbf{Pass} & \textbf{Description} & \textbf{Core Patterns} 
+ & \textbf{Key?} & \textbf{Coordinate?} \\ \hline
+Record Node Coordinates & Adds a new field to each node containing that node's 
+ node coordinate. & Outer Product/Scan & No & Yes \\
+Record Function Depths & Adds a new field to each node recording how many 
+ functions surround the node. & Inner Product & No & Yes \\
+Drop Unnamed Functions & Eliminates some code that will not be evaluated at 
+ the top-level. & Filter & No & No \\
+Drop Unreachable Code & Eliminates some unreachable code. &
+ Filter/Inner Product & No & Yes \\
+Lift Functions & Moves all functions to the top-level. & Key & Yes & Yes \\
+Drop Redundant Nodes & Eliminates unnecessary nodes/nesting & Filter & No & No \\
+Flatten Expressions & Removes nesting from expressions. & Key & Yes & Yes \\
+Compress Atomic Nodes & Atomizes nested atom nodes & Amend & No & No \\
+Propagate Constants & Inlines all references to literal values. &
+ Amend/Rank & No & Yes \\
+Fold Constants & Converts constant expressions to literals & Amend & No & No \\
+Compress Expressions & Converts expression sub-trees into single nodes. & 
+ Amend & No & No \\
+Record Final Return Value & Records the value returned by each function. & 
+ Key & Yes & No \\
+Normalize Values Field & Normalizes the shape and size of the values field. &
+ Amend & No & No \\
+Lift Type-checking & Infers some type information at compile time &
+ Power Limit/Rank & No & No \\
+Allocate Value Slots & Does a form of frame allocation for variables &
+ Key & Yes & No \\
+Anchor Variables & Resolves lexically scoped variables &
+ Key & Yes & No \\
+Record Live Variables & Records the variables that are live at each point of 
+ execution. & Key & Yes & No \\
+Fuse Scalar Loops & Identifies Fusion opportunities and fuses expressions &
+ Key & Yes & No \\
+Type Specialization & Specializes each function for a series of potential inputs. 
+ & Key & Yes & No \\
+\end{tabular}
+\caption{A listing of some compiler passes in the Co-dfns compiler and their 
+ relationship with the Key operator and associated tree computation techniques}
+\label{table:passes}
+\end{table*}
 %[cf]
@@
 %[cf]
+%[of]:Lexical Scoping
+\subsection{Lexical Scoping}
 %[cf]
-%[of]:Scaling up, Considerations and Thoughts
-\section{Scaling up, Considerations and Thoughts}
-
-How well do these techniques work in practice? We have applied these techniques to an entire compiler, 
-with the same restrictions as used here. The result was a success, and the core of the compiler, which 
-is everything but parsing and code generation, is contained in a single file consisting of only 40 lines. 
-In 12 point font, the compiler easily fits on a single sheet of paper. It produces C and CUDA code.
-
-Not every pass in the compiler requires this sort of operation. Indeed, if flattening occurs early and 
-in the correct order, it’s possible to use much simpler methods of grouping using only a prefix sum 
-scan instead of operating over the node coordinates. It’s also interesting to note that we do not 
-need to recompute the node coordinates throughout the compiler. Indeed, this would defeat their 
-purpose somewhat, as the node coordinates allow us to understand the original position of a node, 
-even if that node has now been lifted elsewhere. We use this information when resolving lexical scope, 
-for instance.
-
-When writing data-parallel code, duplicated work is common. The existing compiler contains little to 
-no inherently duplicated work. Most of the duplicated work exists for brevity and clarity of the code, 
-rather than for any required reason, and could easily be factored out.
-
-Finally, it is interesting to note that the simplified control flow gives rise to some positive side-effects. 
-Besides brevity and directness, the complexity of each operation and the composition of operations is 
-well known. We have not yet tried to do this, but it is straightforward to derive a worst-case complexity 
-bound on code written in this style, so simple, in fact, that it can be accomplished automatically. It is 
-conceivable then, that the compiler could produce not only an executable program from code written 
-in this style, but also a complexity bound on its execution. 
+%[cf]
+%[of]:Other Passes
+\section{Other Passes}
+%[cf]
+%[of]:The Co-dfns Compiler
+\section{The Co-dfns Compiler}
+%[cf]
+%[of]:Performance
+\section{Performance}
+%[cf]
+%[of]:Future Work
+\section{Future Work}
 %[cf]
@@
 
+%[of]:Unintegrated Sections
+\section{Unintegrated Sections}
+
+%[of]:Scaling up, Considerations and Thoughts
+\subsection{Scaling up, Considerations and Thoughts}
+
+How well do these techniques work in practice? We have applied these techniques to an entire compiler, 
+with the same restrictions as used here. The result was a success, and the core of the compiler, which 
+is everything but parsing and code generation, is contained in a single file consisting of only 40 lines. 
+In 12 point font, the compiler easily fits on a single sheet of paper. It produces C and CUDA code.
+
+Not every pass in the compiler requires this sort of operation. Indeed, if flattening occurs early and 
+in the correct order, it’s possible to use much simpler methods of grouping using only a prefix sum 
+scan instead of operating over the node coordinates. It’s also interesting to note that we do not 
+need to recompute the node coordinates throughout the compiler. Indeed, this would defeat their 
+purpose somewhat, as the node coordinates allow us to understand the original position of a node, 
+even if that node has now been lifted elsewhere. We use this information when resolving lexical scope, 
+for instance.
+
+When writing data-parallel code, duplicated work is common. The existing compiler contains little to 
+no inherently duplicated work. Most of the duplicated work exists for brevity and clarity of the code, 
+rather than for any required reason, and could easily be factored out.
+
+Finally, it is interesting to note that the simplified control flow gives rise to some positive side-effects. 
+Besides brevity and directness, the complexity of each operation and the composition of operations is 
+well known. We have not yet tried to do this, but it is straightforward to derive a worst-case complexity 
+bound on code written in this style, so simple, in fact, that it can be accomplished automatically. It is 
+conceivable then, that the compiler could produce not only an executable program from code written 
+in this style, but also a complexity bound on its execution. 
+%[cf]
+%[of]:Computing with Node Coordinates
+\subsection{Computing with Node Coordinates}
+
+Given the above keys and grouping information, we now have all the information we need to do 
+function lifting and expression flattening, but we lack a suitable operator to complete the task. 
+Indeed, it isn’t immediately apparent that a single, general purpose operator over arrays exists 
+which can accomplish such a task. However, the Key operator, while masquerading as a useful business 
+analytics operator, represents just such a useful tree transformation tool. 
+%[cf]
+%[of]:Introduction (Old)
+\subsection{Introduction (Old)}
+
+When compiling a high-level language into a low-level language, we often implement a variety of 
+flattening transformations to normalize more complex code exhibiting nesting into flat programs 
+that are semantically equivalent. Function lifting, for example, takes a program with nested functions 
+into an equivalent program in which all functions are at the root of the program. The following function
+f always returns 14.
+
+\begin{verbatim}
+f←λ.(λ ⍵.(⍵+⍵))7
+\end{verbatim}
+
+A function lifting pass would transform this into the following:
+
+\begin{verbatim}
+fn1←λ⍵.⍵+⍵
+f←λ.fn1 7
+\end{verbatim}
+
+Expression flattening works in a similar fasion. Assuming all expressions evaluate right to left, this expression:
+
+\begin{verbatim}
+v←(a+b)÷c×d
+\end{verbatim}
+
+Becomes this:
+
+\begin{verbatim}
+v←c×d
+t←a+b
+v←t÷v
+\end{verbatim}
+
+Characteristic of many compiler passes, most of the work is manipulating the tree, rather than, say, number 
+crunching. Almost universally, compilers use a combination of recursion and branching to implement these 
+transformations, such as using the visitor pattern, Nanopass framework [19], or direct structural recursion. 
+The “work” happens in the control flow of the program text. The ubiquity of this approach warrants note. 
+Let us savour such rare solidarity for a moment. Now, let us challenge it.
+
+Let us arbitrarily restrict ourselves to a data-flow, data-parallel model. Specifically, consider a language 
+whose core primitives are pointwise functions over n-dimensional arrays. We have a set of operators that 
+take functions and apply them over different sub-parts of arrays. We can compose these functions together. 
+That is all. Can we implement a compiler in such a language? The obvious answer is, yes, given arbitrary 
+primitives. What if we restrict ourselves to standard and general-purpose primitives? The answer is still 
+yes. Furthermore, the method is surprisingly direct and concise. The language we use is a subset of Dyalog 
+APL 14.0 and the reader may follow the examples in the following sections using TryAPL.org.
+
+The trick to such a compiler is solving the issue illustrated by the previous examples, expressing tree 
+manipulation without embedding the core logic in the control flow. We accomplish this by choosing an 
+appropriate representation of the AST and a property of each node we call its node coordinate. These 
+coordinates allow us to locally answer the inter-node questions we need without complex control flow. When 
+we combine this with a general-purpose operator called “Key,” we can perform arbitrary computation over 
+sub-trees that we partition based on these inter-node relationships, such as whether one node is an ancestor 
+of another. When combined with more mundane array programming, we can implement the complete core of a 
+compiler, including lexical scope, function lifting, expression flattening/normalization, loop fusion, 
+frame/register allocation, and so on. The result is a completely data-parallel compiler modulo parsing and 
+some aspects of code generation.
+
+The notation used throughout this paper is introduced piecemeal as necessary in each relevant section in 
+which the notation is first used. All functions are applied infix and take either one or two arguments. All 
+expressions evaluate from right to left. Tables 1 and 2 provide a reference for the notation and primitives 
+as a reference aid.
+
+%[of]:Contributions
+\paragraph{Contributions}
+
+\begin{itemize}[noitemsep]
+\item We introduce node coordinates as a means to locally compute inter-node properties without using complex 
+control flow.
+\item We demonstrate methods to perform arbitrary computation over sub-trees partitioned by these 
+inter-node properties in a pure data-flow, data-parallel style.
+\item The presented methods are direct and concise; the data structures, simple.
+\item These methods require no specialized operations, but instead, use only established, general-purpose 
+array programming.
+\end{itemize}
+%[cf]
+%[cf]
+
+%[cf]
+
 % Bibliography goes here

### 2015-11-11 2016-pldi/2016-pldi-key_to_data_parallel_compiler.tex
Complete the other passes section.

--- 2015-10-30 2016-pldi/2016-pldi-key_to_data_parallel_compiler.tex
@@
 % \usepackage[utf8]{inputenc}
+\usepackage{array}
 \usepackage{fontspec}
@@
 %[cf]
-%[of]:Computing with Node Coordinates
-\section{Computing with Node Coordinates}
 
-Given the above keys and grouping information, we now have all the information we need to do 
-function lifting and expression flattening, but we lack a suitable operator to complete the task. 
-Indeed, it isn’t immediately apparent that a single, general purpose operator over arrays exists 
-which can accomplish such a task. However, the Key operator, while masquerading as a useful business 
-analytics operator, represents just such a useful tree transformation tool.
+%[cf]
+%[cf]
+%[of]:Case Studies
+\section{Case Studies}
 
-%[of]:The Key Operator
-\subsection{The Key Operator}
-
-The Key operator (written \verb;⌸;), takes a single function that expects two arguments and returns 
-a function which takes two arguments. The left argument is a set of keys, and the right argument 
-is a set of corresponding elements associated with those keys. In our case, we provide matrices 
-for both of these arguments, so each pair of rows corresponds to a key-value pair. A simple example 
-of the key operation at work is to compute a histogram (≢ here can be thought of as “tally” or count):
-
-\begin{verbatim}
-       ⊢X←?10⍴5
-1 1 4 0 1 3 1 1 2 1
-      X(⊣,(≢⊢))⌸X
-1 6
-4 1
-0 1
-3 1
-2 1
-\end{verbatim}
-
-To understand a bit better how the Key operator applies its function, consider the function {⍺ ⍵} 
-which returns the pair of its right and left arguments. If we apply it to the same value as above, we 
-get the following:
-
-\begin{verbatim}
-      X{⍺ ⍵}⌸X
-1  1 1 1 1 1 1 
-4  4           
-0  0           
-3  3           
-2  2
-\end{verbatim}
-
-In our case, we use either Fk or Ek as our keys applied to the corresponding AST. We also will drop off 
-the first row in each AST using 1↓ since this node “contains” everything. In a complete AST this is 
-usually the Module boundary node which contains the entire set of functions and values in the module.
-%[cf]
 %[of]:Function Lifting
@@
 %[cf]
+%[of]:Lexical Scoping
+\subsection{Lexical Scoping}
 %[cf]
-%[of]:Scaling up, Considerations and Thoughts
-\section{Scaling up, Considerations and Thoughts}
+%[cf]
+%[of]:Table of Compiler Passes
+\renewcommand{\arraystretch}{1.5}
+\begin{table*}
+\centering
+\begin{tabular}{l p{2.8in} l l l}
+\textbf{Pass} & \textbf{Description} & \textbf{Core Patterns} 
+ & \textbf{Coordinate?} \\ \hline
+Record Node Coordinates & Adds a new field to each node containing that node's 
+ node coordinate. & Outer Product/Scan & Yes \\
+Record Function Depths & Adds a new field to each node recording how many 
+ functions surround the node. & Inner Product & Yes \\
+Drop Unnamed Functions & Eliminates some code that will not be evaluated at 
+ the top-level. & Filter & No \\
+Drop Unreachable Code & Eliminates some unreachable code. &
+ Filter/Inner Product & Yes \\
+Lift Functions & Moves all functions to the top-level. & Key & Yes \\
+Drop Redundant Nodes & Eliminates unnecessary nodes/nesting & Filter & No \\
+Flatten Expressions & Removes nesting from expressions. & Key & Yes \\
+Compress Atomic Nodes & Atomizes nested atom nodes & Amend & No \\
+Propagate Constants & Inlines all references to literal values. &
+ Amend/Rank & Yes \\
+Fold Constants & Converts constant expressions to literals & Amend & No \\
+Compress Expressions & Converts expression sub-trees into single nodes. & 
+ Amend & No \\
+Record Final Return Value & Records the value returned by each function. & 
+ Key & No \\
+Normalize Values Field & Normalizes the shape and size of the values field. &
+ Amend & No \\
+Lift Type-checking & Infers some type information at compile time &
+ Power Limit/Rank & No \\
+Allocate Value Slots & Does a form of frame allocation for variables &
+ Key & No \\
+Anchor Variables & Resolves lexically scoped variables &
+ Key & Yes \\
+Record Live Variables & Records the variables that are live at each point of 
+ execution. & Key & No \\
+Fuse Scalar Loops & Identifies Fusion opportunities and fuses expressions &
+ Key & No \\
+Type Specialization & Specializes each function for a series of potential inputs. 
+ & Key & No \\
+\end{tabular}
+\caption{A listing of some compiler passes in the Co-dfns compiler and their 
+ relationship with the Key operator and associated tree computation techniques}
+\label{table:passes}
+\end{table*}
+%[cf]
+%[of]:Other Passes
+\section{Other Passes}
 
-How well do these techniques work in practice? We have applied these techniques to an entire compiler, 
-with the same restrictions as used here. The result was a success, and the core of the compiler, which 
-is everything but parsing and code generation, is contained in a single file consisting of only 40 lines. 
-In 12 point font, the compiler easily fits on a single sheet of paper. It produces C and CUDA code.
+Table \ref{table:passes} contains a selection of passes from the Co-dfns compiler 
+and breaks down their features based on how they traverse the tree and whether 
+or not they use the Key operator or Node Coordinates. They are oriented with 
+front-end passes higher in the table and passes further down the compiler 
+chain at the bottom of the table. These passes give a good idea of how the Key 
+operator plays a part in the complete compiler as well as how Node Coordinates 
+help to deal with complex sub-tree selections.
 
-Not every pass in the compiler requires this sort of operation. Indeed, if flattening occurs early and 
-in the correct order, it’s possible to use much simpler methods of grouping using only a prefix sum 
-scan instead of operating over the node coordinates. It’s also interesting to note that we do not 
-need to recompute the node coordinates throughout the compiler. Indeed, this would defeat their 
-purpose somewhat, as the node coordinates allow us to understand the original position of a node, 
-even if that node has now been lifted elsewhere. We use this information when resolving lexical scope, 
-for instance.
+We make a few notes on the table and the analysis. The Key operator can be used 
+with or without the node coordinates structure. In cases where the AST is 
+sufficiently flat and arranged in a sufficiently convenient order, the grouping 
+operations are very simple. In these cases, the Key operator amounts to a 
+map operation over some selected sub-trees selected at a specific depth. These 
+occur later in the compiler where the reader will note appearences of the Key 
+operator without the use of Node Coordinates.
 
-When writing data-parallel code, duplicated work is common. The existing compiler contains little to 
-no inherently duplicated work. Most of the duplicated work exists for brevity and clarity of the code, 
-rather than for any required reason, and could easily be factored out.
+Likewise, node coordinates find their place early on in the compiler where it 
+is necessary to deal with the more highly nested AST. They are useful for identifying 
+this information regardless of whether the Key operator is used to handle the 
+grouping or not. Thus passes like ``Drop Unreachable Code'' that remove whole 
+sub-trees from the AST may use the Node Coordinates to do the grouping once 
+the correct parent node has been identified.
 
-Finally, it is interesting to note that the simplified control flow gives rise to some positive side-effects. 
-Besides brevity and directness, the complexity of each operation and the composition of operations is 
-well known. We have not yet tried to do this, but it is straightforward to derive a worst-case complexity 
-bound on code written in this style, so simple, in fact, that it can be accomplished automatically. It is 
-conceivable then, that the compiler could produce not only an executable program from code written 
-in this style, but also a complexity bound on its execution. 
+In short, the goal of the structure and design of the compiler passes given 
+in table \ref{table:passes} is to utilize Node Coordinates to remove as quickly as possible 
+the nested structure of the AST into a flatter form that does not require complex 
+analysis to traverse. 
+%[cf]
+%[of]:The Co-dfns Compiler
+\section{The Co-dfns Compiler}
+%[cf]
+%[of]:Performance
+\section{Performance}
+%[cf]
+%[of]:Future Work
+\section{Future Work}
 %[cf]

### 2016-04-04 2016-spring-array/2016-array-key.tex
2016 Spring ARRAY submission

--- 2015-11-11 2016-pldi/2016-pldi-key_to_data_parallel_compiler.tex
@@
-\documentclass[pldi]{sigplanconf-pldi15}
+\documentclass[numbers,preprint]{sigplanconf}
 % \usepackage[utf8]{inputenc}
@@
 \newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
+\usepackage{booktabs}
 
@@
 \title{The Key to a Data Parallel Compiler}
+
+\conferenceinfo{ARRAY '16}{June, 2016, Santa Barbara, CA, USA}
+\copyrightyear{2016}
+
+\authorinfo{Aaron W. Hsu}{Indiana University}{awhsu@indiana.edu}
 
@@
 \end{itemize}
-%[cf]
-%[of]:Generalized Sub-tree Computation
+
+\begin{table}
+\centering
+\begin{tabular}{cll}
+\toprule
+Symbol                   & Monadic            & Dyadic \\
+\midrule
+\texttt{+}               & Identity           & Plus (Add) \\
+\texttt{-}               & Negative           & Minus (Subtract) \\
+\texttt{×}               & Direction (Signum) & Times (Multiply) \\
+\texttt{÷}               & Reciprocal         & Divide \\
+\texttt{|}               & Magnitude          & Residue (Modulo) \\
+\texttt{⌊}               & Floor              & Minimum \\
+\texttt{⌈}               & Ceiling            & Maximum \\
+\texttt{*}               & Exponential        & Power \\
+\texttt{⍟}               & Natural Logarithm  & Logarithm \\
+\texttt{○}               & Pi Times           & Circular (Trigonometric) \\
+\texttt{!}               & Factorial          & Binomial \\
+\texttt{\textasciitilde} & Not                & \\
+\texttt{?}               & Roll               & \\
+\texttt{\^}              &                    & And \\
+\texttt{∨}               &                    & Or \\
+\texttt{⍲}               &                    & Nand \\
+\texttt{⍱}               &                    & Nor \\
+\texttt{<}               &                    & Less \\
+\texttt{≤}               &                    & Less Or Equal \\
+\texttt{=}               &                    & Equal \\
+\texttt{≥}               &                    & Greater Or Equal \\
+\texttt{>}               &                    & Greater \\
+\texttt{≠}               &                    & Not Equal \\
+\end{tabular}
+\caption{Scalar Primitives in Proposed Language}
+\label{tab:scalarprims}
+\end{table}
+
+\begin{table}
+\centering
+\begin{tabular}{cll}
+\toprule
+Symbol     & Monadic       & Dyadic \\
+\midrule
+\texttt{⍴} & Reshape       & \\
+\texttt{,} & Ravel         & Catenate/Laminate \\
+\texttt{⍪} & Table         & Catenate First/Laminate \\
+\texttt{⌽} & Reverse       & Rotate \\
+\texttt{⊖} & Reverse First & Rotate First \\
+\texttt{⍉} & Transpose     & Transpose \\
+\texttt{↑} & Mix           & \\
+\texttt{↓} & Split         & \\
+\texttt{⊂} & Enclose       & Partitioned Enclose \\
+\texttt{∊} & Enlist        & \\
+\end{tabular}
+\caption{Structural Mixed Primitives in Proposed Language}
+\label{tab:structprims}
+\end{table}
+
+\begin{table}
+\centering
+\begin{tabular}{cll}
+\toprule
+Symbol                   & Monadic  & Dyadic \\
+\midrule
+\texttt{⊃}               & Disclose & Pick\\
+\texttt{↑}               &          & Take \\
+\texttt{↓}               &          & Drop \\
+\texttt{/}               &          & Replicate \\
+\texttt{⌿}               &          & Replicate First \\
+\texttt{\textbackslash}  &          & Expand \\
+\texttt{⍀}               &          & Expand First \\
+\texttt{\textasciitilde} &          & Without (Excluding) \\
+\texttt{∩}               &          & Intersection \\
+\texttt{∪}               & Unique   & Union \\
+\texttt{⊣}               & Same     & Left \\
+\texttt{⊢}               & Identity & Right\\
+\end{tabular}
+\caption{Selection Mixed Primitives in Proposed Language}
+\label{tab:selprims}
+\end{table}
+
+\begin{table}
+\centering
+\begin{tabular}{cll}
+\toprule
+Symbol     & Monadic         & Dyadic \\
+\midrule
+\texttt{⍳} & Index Generator & Index Of \\
+\texttt{∊} &                 & Membership \\
+\texttt{⍋} & Grade Up        & Grade Up \\
+\texttt{⍒} & Grade Down      & Grade Down \\
+\texttt{?} &                 & Deal \\
+\texttt{⍷} &                 & Find \\
+\end{tabular}
+\caption{Selector Mixed Primitives in Proposed Language}
+\label{tab:seltorprims}
+\end{table}
+
+\begin{table}
+\centering
+\begin{tabular}{cll}
+\toprule
+Symbol     & Monadic       & Dyadic \\
+\midrule
+\texttt{⍴} & Shape         & \\
+\texttt{≡} & Depth         & Match \\
+\texttt{≢} & Tally         & Not Match \\
+\texttt{⍎} & Execute       & Execute \\
+\texttt{⍕} & Format        & Format \\
+\texttt{⊥} &               & Decode (Base) \\
+\texttt{⊤} &               & Encode (Representation) \\
+\texttt{⌹} & Matrix Divide & Matrix Inverse \\
+\end{tabular}
+\caption{Miscellaneous Mixed Primitives in Proposed Language}
+\label{tab:miscprims}
+\end{table}
+
+\begin{table*}
+\centering
+\begin{tabular}{cll}
+\toprule
+Symbol     & Name & Description \\
+\midrule
+\texttt{⍨} & Commute & Swaps arguments or distributes
+ right argument to both sides \\
+\texttt{¨} & Each & Applies its operand point-wise over the left/right
+ arguments \\
+\texttt{/} & Reduce & Reduce along the last axis \\
+\texttt{⌿} & Reduce First & Reduce along the first axis \\
+\texttt{\textbackslash} & Scan & Scan along the last axis \\
+\texttt{⍀} & Scan First & Scan along the first axis \\
+\texttt{⌸} & Key & Apply operand once for each sub-array grouped by key \\
+\end{tabular}
+\caption{Primitive Adverbs in Proposed Language, each takes a single left
+operand and describes a verb operating over one or two arguments}
+\label{tab:adverbs}
+\end{table*}
+
+\begin{table*}
+\centering
+\begin{tabular}{cll}
+\toprule
+Symbol     & Name & Description \\
+\midrule
+\texttt{∘} & Compose & Composes two operands as in traditional mathematics \\
+\texttt{.} & Inner Product & Inner product operation, e.g.
+ \texttt{+.×} for matrix multiplication \\
+\texttt{∘.} & Outer Product & Cartesian product or ``function table'' \\
+\texttt{⍣} & Power & Iteration, Limited use only \\
+\texttt{⍤} & Rank & Apply a verb along cells of an array \\
+\end{tabular}
+\caption{Primitive Conjunctions in Proposed Language, each takes a left
+ and right operand and describes a verb operating over one or two arguments}
+\label{tab:conjunctions}
+\end{table*}
+
 \section{Data Parallel Sub-tree Computation}
@@
 
-%[of]:A Definition of Node Coordinates
-\subsubsection{A Definition of Node Coordinates}
-
 Every node in an AST has a node coordinate, named because a coordinate is a precise location in a space. 
@@
 To understand how we construct these coordinates we must consider how we represent the AST.
-%[cf]
+
 %[of]:Constructing Node Coordinates
@@
 \begin{tabular}{l p{2.8in} l l l}
-\textbf{Pass} & \textbf{Description} & \textbf{Core Patterns} 
- & \textbf{Coordinate?} \\ \hline
+\toprule
+Pass & Description & Core Patterns & Coordinate? \\
+\midrule
 Record Node Coordinates & Adds a new field to each node containing that node's 
@@
 
-%[of]:Unintegrated Sections
-\section{Unintegrated Sections}
-
-%[of]:Scaling up, Considerations and Thoughts
-\subsection{Scaling up, Considerations and Thoughts}
-
-How well do these techniques work in practice? We have applied these techniques to an entire compiler, 
-with the same restrictions as used here. The result was a success, and the core of the compiler, which 
-is everything but parsing and code generation, is contained in a single file consisting of only 40 lines. 
-In 12 point font, the compiler easily fits on a single sheet of paper. It produces C and CUDA code.
-
-Not every pass in the compiler requires this sort of operation. Indeed, if flattening occurs early and 
-in the correct order, it’s possible to use much simpler methods of grouping using only a prefix sum 
-scan instead of operating over the node coordinates. It’s also interesting to note that we do not 
-need to recompute the node coordinates throughout the compiler. Indeed, this would defeat their 
-purpose somewhat, as the node coordinates allow us to understand the original position of a node, 
-even if that node has now been lifted elsewhere. We use this information when resolving lexical scope, 
-for instance.
-
-When writing data-parallel code, duplicated work is common. The existing compiler contains little to 
-no inherently duplicated work. Most of the duplicated work exists for brevity and clarity of the code, 
-rather than for any required reason, and could easily be factored out.
-
-Finally, it is interesting to note that the simplified control flow gives rise to some positive side-effects. 
-Besides brevity and directness, the complexity of each operation and the composition of operations is 
-well known. We have not yet tried to do this, but it is straightforward to derive a worst-case complexity 
-bound on code written in this style, so simple, in fact, that it can be accomplished automatically. It is 
-conceivable then, that the compiler could produce not only an executable program from code written 
-in this style, but also a complexity bound on its execution. 
-%[cf]
-%[of]:Computing with Node Coordinates
-\subsection{Computing with Node Coordinates}
-
-Given the above keys and grouping information, we now have all the information we need to do 
-function lifting and expression flattening, but we lack a suitable operator to complete the task. 
-Indeed, it isn’t immediately apparent that a single, general purpose operator over arrays exists 
-which can accomplish such a task. However, the Key operator, while masquerading as a useful business 
-analytics operator, represents just such a useful tree transformation tool. 
-%[cf]
-%[of]:Introduction (Old)
-\subsection{Introduction (Old)}
-
-When compiling a high-level language into a low-level language, we often implement a variety of 
-flattening transformations to normalize more complex code exhibiting nesting into flat programs 
-that are semantically equivalent. Function lifting, for example, takes a program with nested functions 
-into an equivalent program in which all functions are at the root of the program. The following function
-f always returns 14.
-
-\begin{verbatim}
-f←λ.(λ ⍵.(⍵+⍵))7
-\end{verbatim}
-
-A function lifting pass would transform this into the following:
-
-\begin{verbatim}
-fn1←λ⍵.⍵+⍵
-f←λ.fn1 7
-\end{verbatim}
-
-Expression flattening works in a similar fasion. Assuming all expressions evaluate right to left, this expression:
-
-\begin{verbatim}
-v←(a+b)÷c×d
-\end{verbatim}
-
-Becomes this:
-
-\begin{verbatim}
-v←c×d
-t←a+b
-v←t÷v
-\end{verbatim}
-
-Characteristic of many compiler passes, most of the work is manipulating the tree, rather than, say, number 
-crunching. Almost universally, compilers use a combination of recursion and branching to implement these 
-transformations, such as using the visitor pattern, Nanopass framework [19], or direct structural recursion. 
-The “work” happens in the control flow of the program text. The ubiquity of this approach warrants note. 
-Let us savour such rare solidarity for a moment. Now, let us challenge it.
-
-Let us arbitrarily restrict ourselves to a data-flow, data-parallel model. Specifically, consider a language 
-whose core primitives are pointwise functions over n-dimensional arrays. We have a set of operators that 
-take functions and apply them over different sub-parts of arrays. We can compose these functions together. 
-That is all. Can we implement a compiler in such a language? The obvious answer is, yes, given arbitrary 
-primitives. What if we restrict ourselves to standard and general-purpose primitives? The answer is still 
-yes. Furthermore, the method is surprisingly direct and concise. The language we use is a subset of Dyalog 
-APL 14.0 and the reader may follow the examples in the following sections using TryAPL.org.
-
-The trick to such a compiler is solving the issue illustrated by the previous examples, expressing tree 
-manipulation without embedding the core logic in the control flow. We accomplish this by choosing an 
-appropriate representation of the AST and a property of each node we call its node coordinate. These 
-coordinates allow us to locally answer the inter-node questions we need without complex control flow. When 
-we combine this with a general-purpose operator called “Key,” we can perform arbitrary computation over 
-sub-trees that we partition based on these inter-node relationships, such as whether one node is an ancestor 
-of another. When combined with more mundane array programming, we can implement the complete core of a 
-compiler, including lexical scope, function lifting, expression flattening/normalization, loop fusion, 
-frame/register allocation, and so on. The result is a completely data-parallel compiler modulo parsing and 
-some aspects of code generation.
-
-The notation used throughout this paper is introduced piecemeal as necessary in each relevant section in 
-which the notation is first used. All functions are applied infix and take either one or two arguments. All 
-expressions evaluate from right to left. Tables 1 and 2 provide a reference for the notation and primitives 
-as a reference aid.
-
-%[of]:Contributions
-\paragraph{Contributions}
-
-\begin{itemize}[noitemsep]
-\item We introduce node coordinates as a means to locally compute inter-node properties without using complex 
-control flow.
-\item We demonstrate methods to perform arbitrary computation over sub-trees partitioned by these 
-inter-node properties in a pure data-flow, data-parallel style.
-\item The presented methods are direct and concise; the data structures, simple.
-\item These methods require no specialized operations, but instead, use only established, general-purpose 
-array programming.
-\end{itemize}
-%[cf]
-%[cf]
-
-%[cf]
-
 % Bibliography goes here

### 2016-04-05 2016-spring-array/2016-array-key.tex
Add in most of the other sections, tables, and other goodies.

@@
 \begin{abstract}
-TBW
+We present a language-driven strategy for the construction of 
+compilers that are inherently data-parallel in their design and 
+implementation. Using an encoding of the inter-node relationships 
+between nodes in an AST called a Node Coordinate Matrix, we demonstrate how 
+an operator called the Key operator, that applies a function over groupings 
+of array cells grouped and ordered by their keys, when used in conjunction 
+with a Node Coordinate Matrix, permits arbitrary computation over sub-trees 
+of an AST using purely data-parallel array programming techniques. We discuss 
+the use of these techniques when applied to an experimental commercial 
+compiler called Co-dfns, which is a fully data-parallel compiler developed 
+using the techniques discussed here.
 \end{abstract}
 
-%[c]Proposed new outline:
-%[c]
-%[c]I. 	Generalization on manipulations over Sub-tree computation
-%[c]  A.	Key + Encoding → Data parallel, Array-based computation
-%[c]  B.	Easy to write/read
-%[c]  C.	Uniform across passes
-%[c]II.	Case Studies
-%[c]  A.	Function Lifting
-%[c]  B.	Expression Flattening
-%[c]  C.	Lexical Scope Resolution
-%[c]  D.	Illustrates the problem space
-%[c]III.	Other passes/Issues
-%[c]  A.	Discussion of other passes
-%[c]  B.	Beyond the Key Operator
-%[c]IV.	In practice/Co-dfns Compiler
-%[c]V.	Performance/Results
-%[c]
-%[c]Questions that must be addressed:
-%[c]
-%[c]1. Why the formalization in APL?
-%[c]2. Why parallelize a compiler?
-%[c]3. How well does it work/scale/generalize
-
-%[of]:Introduction
+% Proposed new outline:
+% 
+% I. 	Generalization on manipulations over Sub-tree computation
+%   A.	Key + Encoding → Data parallel, Array-based computation
+%   B.	Easy to write/read
+%   C.	Uniform across passes
+% II.	Case Studies
+%   A.	Function Lifting
+%   B.	Expression Flattening
+%   C.	Lexical Scope Resolution
+%   D.	Illustrates the problem space
+% III.	Other passes/Issues
+%   A.	Discussion of other passes
+%   B.	Beyond the Key Operator
+% IV.	In practice/Co-dfns Compiler
+% V.	Performance/Results
+% 
+% Questions that must be addressed:
+% 
+% 1. Why the formalization in APL?
+% 2. Why parallelize a compiler?
+% 3. How well does it work/scale/generalize
+
 \section{Introduction}
 
-Compilers are everywhere, but their design and structure does not scale well to modern 
-parallel architectures. Traditional compilers are very difficult to parallelize, and the benefits of 
-modern architectures is often seen only through the use of distributed build systems which invoke 
-the single threaded compilers on multiple ``documents'' at the same time. Future CPU architectures 
-promise to be even more data parallel, such as the Xeon Phi and the GPGPU architectures. On these 
-systems it is even harder to obtain good performance for compilers written in the traditional fashion.
-
-Fundamentally, a compiler is a translator that operates over some sort of tree and transforms it to 
-another tree of an usually radically different shape but equivalent ``semantics.'' These algorithms and 
-transformations show up everywhere, from document processing on the Web, XML-based publishing 
-workflows, to the programming language compilers with which most developers are intimately familar. 
-These transformations differ from other widely studied parallel graph problems in that they tend to greatly 
-alter the fundamental structure of the graph/tree, unlike, say, breadth-first search or single-source 
-shortest path algorithms. This rapidly shifting graph structure requires different approaches to 
-achieving good parallelism.
-
-There has been some interesting use of GPUs to accelerate certain compiler analyses and passes. 
-Unfortunately, attempts to parallelize compilers usually takes the form of a great deal of work spent to 
-highly optimize a single compiler pass. The results are often fragile and require a great deal of effort 
-to maintain and keep in line with the rest of the compiler. This integration problem makes it difficult to 
-scale the "ad hoc" parallelization effort of individual compiler passes to a complete system. This is 
-particularly true in situations where the compiler passes may change rapidly, such as with XSLT or 
-document publishing, where it is very common for the document preparer to write or tweak transformation 
-passes to suit the specific class of documents being handled. In these cases, the expertise and time 
-involved in parallelizing such transformations makes the ad hoc approach mostly intractable.
-
-Furthermore, compilers are increasingly becoming a part of a "live" and interactive feedback loop. 
-Just in time compilation and on the fly static compilation are both used to provide custom binaries tuned 
-to specific tasks on the fly. These techniques are used in web services and large scale databases, where 
-in-memory databases may construct functions on the fly to execute. There is great interest in moving 
-such systems on to the GPU or other highly data-parallel architectures, but it would significantly hamper 
-the performance of the system if the compilation phase of query execution were required to occur on the 
-CPU, especially at scale where there may be many more GPU accelerators than CPU resources.
-
-We propose a language-driven approach to compiler construction that provides a framework for the 
-development of arbitrary compiler transformations that are data-parallel by construction. These 
-techniques intentionally optimize for the data-parallel case, which permits a wide degree of platform 
-independent performance compared to the traditional approach to compiler design, which requires 
-significant effort and possibly signficiant transformation of the source code to achieve good parallel 
-performance on high-parallel architectures.
-
-We use a mathematically oriented, array based language that is accessible and popular with information 
-scientists. We restrict this language to permit only the composition of data-flow oriented array based 
-programs whose core primitives are all data-parallel by definition. The permitted compositions of these 
-primitives ensures that the resulting programs are both analyzable and parallel by construction, permitting 
-a wide array of whole program and transformation specific optimizations that have been long studied. 
-The notation is intentially close to mathematical notation, and as such, the resulting compiler passes 
-more closely resemble the compact formal descriptions of the transformations rather than complex 
+Compilers are everywhere, but their design and structure does not scale
+well to modern parallel architectures. Traditional compilers are very
+difficult to parallelize, and the benefits of modern architectures is
+often seen only through the use of distributed build systems which
+invoke the single threaded compilers on multiple ``documents'' at
+the same time. Future CPU architectures promise to be even more data
+parallel, such as the Xeon Phi and the GPGPU architectures. On these
+systems it is even harder to obtain good performance for compilers
+written in the traditional fashion.
+
+Fundamentally, a compiler is a translator that operates over some sort
+of tree and transforms it to another tree of an usually radically
+different shape but equivalent ``semantics.'' These algorithms and
+transformations show up everywhere, from document processing on the
+Web, XML-based publishing workflows, to the programming language
+compilers with which most developers are intimately familar.
+These transformations differ from other widely studied parallel
+graph problems in that they tend to greatly alter the fundamental
+structure of the graph/tree, unlike, say, breadth-first search or
+single-source shortest path algorithms. This rapidly shifting graph
+structure requires different approaches to achieving good parallelism.
+
+There has been some interesting use of GPUs to accelerate certain
+compiler analyses and passes.  Unfortunately, attempts to parallelize
+compilers usually takes the form of a great deal of work spent
+to highly optimize a single compiler pass. The results are often
+fragile and require a great deal of effort to maintain and keep in
+line with the rest of the compiler. This integration problem makes it
+difficult to scale the "ad hoc" parallelization effort of individual
+compiler passes to a complete system. This is particularly true in
+situations where the compiler passes may change rapidly, such as with
+XSLT or document publishing, where it is very common for the document
+preparer to write or tweak transformation passes to suit the specific
+class of documents being handled. In these cases, the expertise and
+time involved in parallelizing such transformations makes the ad hoc
+approach mostly intractable.
+
+Furthermore, compilers are increasingly becoming a part of a "live"
+and interactive feedback loop.  Just in time compilation and on
+the fly static compilation are both used to provide custom binaries
+tuned to specific tasks on the fly. These techniques are used in web
+services and large scale databases, where in-memory databases may
+construct functions on the fly to execute. There is great interest
+in moving such systems on to the GPU or other highly data-parallel
+architectures, but it would significantly hamper the performance of
+the system if the compilation phase of query execution were required
+to occur on the CPU, especially at scale where there may be many more
+GPU accelerators than CPU resources.
+
+We propose a language-driven approach to compiler construction that
+provides a framework for the development of arbitrary compiler
+transformations that are data-parallel by construction. These
+techniques intentionally optimize for the data-parallel case, which
+permits a wide degree of platform independent performance compared to
+the traditional approach to compiler design, which requires significant
+effort and possibly signficiant transformation of the source code to
+achieve good parallel performance on high-parallel architectures.
+
+We use a mathematically oriented, array based language that is
+accessible and popular with information scientists. We restrict
+this language to permit only the composition of data-flow oriented
+array based programs whose core primitives are all data-parallel by
+definition. The permitted compositions of these primitives ensures
+that the resulting programs are both analyzable and parallel
+by construction, permitting a wide array of whole program and
+transformation specific optimizations that have been long studied.
+The notation is intentially close to mathematical notation, and
+as such, the resulting compiler passes more closely resemble the
+compact formal descriptions of the transformations rather than complex
 recursive algorithms for transforming trees.
 
-At the core of such techniques, we must have some method for transforming trees in a controlled and 
-predictable manner. We build on the Nanopass paradigm for compiler construction, wherein a compiler is 
-developed as a series of small, functional tree transformations (passes) that have an input AST language 
-and an output AST language. Through progressive refinement of the tree, the composition of these 
-passes results in the complete compiler. This is a highly functional approach, with very little mutation. 
-Fundamentally, each pass must identify regions of the tree upon which it works, do some sort of computation 
-on that sub-tree that has been matched (often through pattern matching on the nodes) and reassemble 
-the tree in some new order that matches the desired structure. In other words, a compiler nanopass 
-can be thought of as an arbitrary computation over a given selection of sub-trees in an AST, where 
-this computation could move, relocate, or change the sub-tree.
-
-The key to our language-drive approach is a set of techniques that enable a data-parallel expression of 
-these arbitrary transformations. By using a higher-order function called Key, together with basic array 
-programming techniques and an encoding for the tree connections called a Node Coordinate matrix, we can 
-identify sub-trees and select them for computation by their parent-child relationships in a data-parallel 
-fashion, and then compute on these sub-trees using traditional array programming techniques.
-
-We have constructed a complete compiler for a superset of our language called Co-dfns. This compiler is 
-part of a commercial effort to develop platform independent performance for array programs. The 
-compiler is developed completely in a data-parallel by construction as described in this paper, and uses 
-only traditional and well known array primitives together with function composition to form the entire 
-core compiler passes and generator. As a result, the compiler takes the form of a pure data-flow graph 
-without mutation and has an extremely uniform and concise definition. It uses no recursion, pattern 
-matching, branching, or other forms of explicit looping. The compiler, together with its runtime libraries 
-and parser, comes in under 1000 lines of code.
-
-In this paper we fully elucidate three compiler passes from the Co-dfns compiler to demonstrate the 
-techniques used and we provide further analysis of how these techniques are used in the other compiler 
-passes. We further provide a comparative analysis of these passes against their counterparts implemented 
-using the Nanopass compiler framework, comparing performance both in terms of programmer effort and 
-portability/maintainability, as well as runtime performance.
+At the core of such techniques, we must have some method for
+transforming trees in a controlled and predictable manner. We
+build on the Nanopass paradigm for compiler construction, wherein
+a compiler is developed as a series of small, functional tree
+transformations (passes) that have an input AST language and an
+output AST language. Through progressive refinement of the tree, the
+composition of these passes results in the complete compiler. This is a
+highly functional approach, with very little mutation.  Fundamentally,
+each pass must identify regions of the tree upon which it works,
+do some sort of computation on that sub-tree that has been matched
+(often through pattern matching on the nodes) and reassemble the tree
+in some new order that matches the desired structure. In other words,
+a compiler nanopass can be thought of as an arbitrary computation
+over a given selection of sub-trees in an AST, where this computation
+could move, relocate, or change the sub-tree.
+
+The key to our language-drive approach is a set of techniques
+that enable a data-parallel expression of these arbitrary
+transformations. By using a higher-order function called Key, together
+with basic array programming techniques and an encoding for the tree
+connections called a Node Coordinate matrix, we can identify sub-trees
+and select them for computation by their parent-child relationships
+in a data-parallel fashion, and then compute on these sub-trees using
+traditional array programming techniques.
+
+We have constructed a complete compiler for a superset of our language
+called Co-dfns. This compiler is part of a commercial effort to develop
+platform independent performance for array programs. The compiler is
+developed completely in a data-parallel by construction as described in
+this paper, and uses only traditional and well known array primitives
+together with function composition to form the entire core compiler
+passes and generator. As a result, the compiler takes the form of a
+pure data-flow graph without mutation and has an extremely uniform and
+concise definition. It uses no recursion, pattern matching, branching,
+or other forms of explicit looping. The compiler, together with its
+runtime libraries and parser, comes in under 1000 lines of code.
+
+In this paper we fully elucidate three compiler passes from the Co-dfns
+compiler to demonstrate the techniques used and we provide further
+analysis of how these techniques are used in the other compiler
+passes. We further provide a comparative analysis of these passes
+against their counterparts implemented using the Nanopass compiler
+framework, comparing performance both in terms of programmer effort
+and portability/maintainability, as well as runtime performance.
 
@@
 \begin{itemize}[noitemsep]
-\item We describe a method of computing over arbitrary sub-trees selected by their parent-child 
-relationships in a data-parallel manner using the Key operator and Node Coordinates.
-\item We situate this technique into a broader, language-driven compiler construction strategy 
-that enables the development of parallel by construction compilers that exhibit desirable programmability 
-features in terms of code size, complexity, portability, and maintainability. 
-\item We provide analysis of these techniques as used in a commercial compiler project called Co-dfns and 
-report on the results.
-\item We compare these techniques to the Nanopass approach to compiler design in terms of programmer 
-productivity and compiler runtime performance. 
-\item We enumerate a set of programming techniques, patterns, and strategies that complement the 
-Key operator and Node Coordinates in the construction of tree transformations in general and give some 
+
+\item We describe a method of computing over arbitrary sub-trees
+selected by their parent-child relationships in a data-parallel manner
+using the Key operator and Node Coordinates.
+
+\item We situate this technique into a broader, language-driven
+compiler construction strategy that enables the development of parallel
+by construction compilers that exhibit desirable programmability
+features in terms of code size, complexity, portability, and
+maintainability.
+
+\item We provide analysis of these techniques as used in a commercial
+compiler project called Co-dfns and report on the results.
+
+\item We compare these techniques to the Nanopass approach to
+compiler design in terms of programmer productivity and compiler
+runtime performance.
+
+\item We enumerate a set of programming techniques, patterns, and
+strategies that complement the Key operator and Node Coordinates
+in the construction of tree transformations in general and give some
 intuitions on the use of these techniques on arbitrary transformations.
 \end{itemize}
+
+\section{Notational Conventions}
+
+For space and convenience, we use a concise array notational convention to 
+describe our techniques and approach. The notation is executable and well 
+established in the array community (it is a limited subset of APL). This is 
+in fact the same notation used within the Co-dfns compiler itself, whose 
+source code is available online and provides a complete example of the use 
+of both these algorithms and this notation in the large.
+
+All code examples are given in the following form:
+
+\begin{verbatim}
+      5×3+4 ⍝ Right to Left Evaluation
+35
+\end{verbatim}
+
+That is, the sample code given indented by 6 spaces, followed by the value 
+of the expression without indent. All expressions are evaluated right to 
+left and all function application is unified and infix; that is, a function 
+application may apply a function (such as \verb;+;) to one or two arguments, 
+which must appear on the right and (optionally) on the left of the function 
+name, as in the above example. Tables \ref{tab:scalarprims} to 
+\ref{tab:conjunctions}
+provide a listing of all of the primitives used in the Co-dfns compiler, 
+a selection of which are used here.
+
+We separate out higher-order functions called operators, from functions that 
+operate over arrays that we call simply functions. Operators take a function 
+argument or two, and bind more strongly to the left than the right. An 
+operator may take one or two operands, either on the left, or on the 
+right and left, and will return a function. In the following 
+example, we apply a reduction operator to a function created using the 
+composition operator, as an example:
+
+\begin{verbatim}
+      +∘-/1 2 3 4 5
+3
+      +/1 2 3 4 5
+15
+\end{verbatim}
+
+Uses of the notation are introduced slowly throughout the treatment, so 
+the reader is encouraged to simply proceed through the treatment, referencing 
+the tables only when required.
 
@@
 \end{tabular}
-\caption{Scalar Primitives in Proposed Language}
+\caption{Scalar Primitives}
 \label{tab:scalarprims}
@@
 \end{tabular}
-\caption{Structural Mixed Primitives in Proposed Language}
+\caption{Structural Mixed Primitives}
 \label{tab:structprims}
@@
 \end{tabular}
-\caption{Selection Mixed Primitives in Proposed Language}
+\caption{Selection Mixed Primitives}
 \label{tab:selprims}
@@
 \end{tabular}
-\caption{Selector Mixed Primitives in Proposed Language}
+\caption{Selector Mixed Primitives}
 \label{tab:seltorprims}
@@
 \end{tabular}
-\caption{Miscellaneous Mixed Primitives in Proposed Language}
+\caption{Miscellaneous Mixed Primitives}
 \label{tab:miscprims}
@@
 \end{tabular}
-\caption{Primitive Adverbs in Proposed Language, each takes a single left
+\caption{Primitive Monadic/Unary Operators, each takes a single left
 operand and describes a verb operating over one or two arguments}
@@
 \end{tabular}
-\caption{Primitive Conjunctions in Proposed Language, each takes a left
+\caption{Primitive Dyadic/Binary Operators, each takes a left
  and right operand and describes a verb operating over one or two arguments}
@@
 
-%[of]:The Key Operator
+Computing over sub-trees normally involves recursing down the structure of 
+the AST, identifying the root nodes of each sub-tree, and then dispatching
+to a handler for that sub-tree, which will continue the recursion at that 
+point and return the modified tree, which will replace the previous 
+sub-tree. More complicated transformations may involved moving the root 
+nodes around in the tree or other large, distant structural modifications.
+
+We divide the work of sub-tree computation into three basic phases. Firstly, 
+we maintain a node coordinate with each node that uniquely identifies it 
+relative to others, we use these coordinates to select and group nodes for 
+work as a single sub-tree, and then we operate over these groupings using 
+the key operator.
+
 \subsection{The Key Operator}
 
-The Key operator (written \verb;⌸;), takes a single function that expects two arguments and returns 
-a function which takes two arguments. The left argument is a set of keys, and the right argument 
-is a set of corresponding elements associated with those keys. In our case, we provide matrices 
-for both of these arguments, so each pair of rows corresponds to a key-value pair. A simple example 
-of the key operation at work is to compute a histogram (≢ here can be thought of as “tally” or count):
+The Key operator (written \verb;⌸;), takes a single function
+that expects two arguments and returns a function which takes two
+arguments. The left argument is a set of keys, and the right argument
+is a set of corresponding elements associated with those keys. In our
+case, we provide matrices for both of these arguments, so each pair
+of rows corresponds to a key-value pair. A simple example of the key
+operation at work is to compute a histogram (≢ here can be thought
+of as “tally” or count):
 
@@
 
-To understand a bit better how the Key operator applies its function, consider the function {⍺ ⍵} 
-which returns the pair of its right and left arguments. If we apply it to the same value as above, we 
+To understand a bit better how the Key operator applies its function,
+consider the function {⍺ ⍵} which returns the pair of its right
+and left arguments. If we apply it to the same value as above, we
 get the following:
@@
 
-In our case, we use either Fk or Ek as our keys applied to the corresponding AST. We also will drop off 
-the first row in each AST using 1↓ since this node “contains” everything. In a complete AST this is 
-usually the Module boundary node which contains the entire set of functions and values in the module.
-%[cf]
-%[of]:Encoding the AST
+In our case, we use either Fk or Ek as our keys applied to the
+corresponding AST. We also will drop off the first row in each AST
+using 1↓ since this node “contains” everything. In a complete
+AST this is usually the Module boundary node which contains the entire
+set of functions and values in the module.
+
 \subsection{Encoding the AST}
 
-We represent the AST as a matrix of 3 columns and n rows, one for each node in the tree. The first column 
-contains the inter-node relationships in the form of a depth vector. The second column is a vector of the 
-node types, while the third contains the “value” of the node, such as the name in a variable.
+We represent the AST as a matrix of 3 columns and n rows, one for
+each node in the tree. The first column contains the inter-node
+relationships in the form of a depth vector. The second column is a
+vector of the node types, while the third contains the “value”
+of the node, such as the name in a variable.
 
@@
 
-We combine these to form the respective AST matrices. We write A,B to catenate arrays A and B along their 
-last axes and ⍪A to turn a vector into a 1-column matrix. Thus, the two tree matrices are given by the 
-following expressions:
+We combine these to form the respective AST matrices. We write A,B
+to catenate arrays A and B along their last axes and ⍪A to turn a
+vector into a 1-column matrix. Thus, the two tree matrices are given
+by the following expressions:
 
@@
 
-Note especially that all inter-node information lives in the depth vector, but that this information requires 
-non-local access as described in the previous section to use it. Constructing node coordinates fixes this issue.
-%[cf]
-%[of]:Node Coordinates
+Note especially that all inter-node information lives in the depth
+vector, but that this information requires non-local access as
+described in the previous section to use it. Constructing node
+coordinates fixes this issue.
+
 \subsection{Node Coordinates}
 
-Every node in an AST has a node coordinate, named because a coordinate is a precise location in a space. 
-We can imagine all the nodes arranged inside some multi-dimensional space, leading to a specific set of 
-node coordinates. Many such arrangements exist, of varying usefulness. In our case, any arrangement 
-should allow us to answer the following questions about any two arbitrary nodes in a tree:
+Every node in an AST has a node coordinate, named because a coordinate
+is a precise location in a space.  We can imagine all the nodes
+arranged inside some multi-dimensional space, leading to a specific
+set of node coordinates. Many such arrangements exist, of varying
+usefulness. In our case, any arrangement should allow us to answer
+the following questions about any two arbitrary nodes in a tree:
 
@@
 
-In short, we care about the relative position of each node in a tree relative to any other. We define a node 
-coordinate as follows.
-
-A node coordinate is a vector whose length is the depth of the tree. Its elements are natural numbers. 
-The count of non-zero elements in the vector is equal to the depth of that node in the tree. All zero elements
-appear after the non-zero elements. That is, a coordinate is zero-padded on the right. When ordered 
-lexicographically, the nodes for each coordinate appear in order according to a depth-first pre-order 
-traversal of the tree. Each coordinate uniquely identifies a single node. Every ancestor’s coordinate is a 
-prefix of any child’s coordinate, ignoring zeros. From the above it follows that every node is lexicographically 
-greater than its left sibling and differs from it by exactly one non-zero element, and that this element is the 
-last non-zero element in the coordinate.
-
-To understand how we construct these coordinates we must consider how we represent the AST.
-
-%[of]:Constructing Node Coordinates
+In short, we care about the relative position of each node in a tree
+relative to any other. We define a node coordinate as follows.
+
+A node coordinate is a vector whose length is the depth of the
+tree. Its elements are natural numbers.  The count of non-zero elements
+in the vector is equal to the depth of that node in the tree. All zero
+elements appear after the non-zero elements. That is, a coordinate
+is zero-padded on the right. When ordered lexicographically, the
+nodes for each coordinate appear in order according to a depth-first
+pre-order traversal of the tree. Each coordinate uniquely identifies a
+single node. Every ancestor’s coordinate is a prefix of any child’s
+coordinate, ignoring zeros. From the above it follows that every node
+is lexicographically greater than its left sibling and differs from
+it by exactly one non-zero element, and that this element is the last
+non-zero element in the coordinate.
+
+To understand how we construct these coordinates we must consider
+how we represent the AST.
+
 \subsubsection{Constructing Node Coordinates}
 
-In building the node coordinates, our goal is to build a matrix where each row is a node 
-coordinate satisfying our previously described requirements. We end up with an n d 
-shaped matrix where n is the number of rows and d the depth of the tree. As noted above, 
-the depth vector contains all the necessary information, but in the wrong form.
-
-We write f⌿A to reduce the first axis of A using function f. Thus, +⌿V is the sum of the 
-elements in vector V. The function x⌈y gives the maximum of its two arguments. So, the 
-depth of each tree is given by the following:
+In building the node coordinates, our goal is to build a matrix where
+each row is a node coordinate satisfying our previously described
+requirements. We end up with an n d shaped matrix where n is the number
+of rows and d the depth of the tree. As noted above, the depth vector
+contains all the necessary information, but in the wrong form.
+
+We write f⌿A to reduce the first axis of A using function f. Thus,
++⌿V is the sum of the elements in vector V. The function x⌈y
+gives the maximum of its two arguments. So, the depth of each tree
+is given by the following:
 
@@
 
-The function table or outer product of f over vectors U and V is written U ∘.f V giving a U V 
-shaped matrix as a result. Thus, (⍳3)∘.×⍳3 gives a small multiplication table:
+The function table or outer product of f over vectors U and V is
+written U ∘.f V giving a U V shaped matrix as a result. Thus,
+(⍳3)∘.×⍳3 gives a small multiplication table:
 
@@
 
-If we use ∘.= on the depth vector and its set of depths instead, we see an expanded Boolean 
-representation of the depth vector:
+If we use ∘.= on the depth vector and its set of depths instead,
+we see an expanded Boolean representation of the depth vector:
 
@@
 
-These matrices let us see the nesting features of each tree more visually, but also 
-suggest another step. We can compute a sum scan with +⍀, also called a prefix sum, 
-along the first axis. Applying this function on the above matrices leads to an interesting 
-result:
+These matrices let us see the nesting features of each tree more
+visually, but also suggest another step. We can compute a sum scan
+with +⍀, also called a prefix sum, along the first axis. Applying
+this function on the above matrices leads to an interesting result:
 
@@
 
-These matrices are lexicographically ordered, and each ancestor shares a common prefix 
-with its descendants. They are also unique coordinates. Only the spurious digits at the end 
-of each coordinate prevent these matrices from meeting all our requirements for valid node 
-coordinates.
-
-The expression V f⍤¯1⊢M applies f to corresponding elements of V and rows of M:
+These matrices are lexicographically ordered, and each ancestor
+shares a common prefix with its descendants. They are also unique
+coordinates. Only the spurious digits at the end of each coordinate
+prevent these matrices from meeting all our requirements for valid
+node coordinates.
+
+The expression V f⍤¯1⊢M applies f to corresponding elements of
+V and rows of M:
 
@@
 
-If n↑V takes the first n elements of V, then we can obtain coordinate matrices from the 
-prefix sums by noting that the spurious digits all come after column d+1 where d is the 
-depth corresponding to that coordinate. The following gives a complete expression for 
-computing a node coordinate matrix from a depth vector, shown using Fd and Fe:
+If n↑V takes the first n elements of V, then we can obtain coordinate
+matrices from the prefix sums by noting that the spurious digits
+all come after column d+1 where d is the depth corresponding to that
+coordinate. The following gives a complete expression for computing
+a node coordinate matrix from a depth vector, shown using Fd and Fe:
 
@@
 
-A careful study of the definition of a node coordinate and the above construction should 
-reveal why this works. Intuitively, we are creating a multi-dimensional space or a number 
-system in which each digit place or dimension contains or circumscribes a smaller space in 
-which are contained all the descendant nodes that appear lower in the tree. Each coordinate 
-is a sort of special path through the tree encoded to have desirable properties relative to 
-other paths.
-%[cf]
-%[of]:Operations on Node Coordinates
+A careful study of the definition of a node coordinate and the
+above construction should reveal why this works. Intuitively, we are
+creating a multi-dimensional space or a number system in which each
+digit place or dimension contains or circumscribes a smaller space
+in which are contained all the descendant nodes that appear lower in
+the tree. Each coordinate is a sort of special path through the tree
+encoded to have desirable properties relative to other paths.
+
 \subsubsection{Operations on Node Coordinates}
 
-The simplest operation over a node coordinate is to extract the depth of the node. The 
-expression C⍳0 finds the first occurrence of 0 in C and returns the index of that 
-occurrence. (The application of ⍳ here is its dyadic form, meaning that it receives two 
-arguments. Previous uses of ⍳ have used its monadic form, which generates all indices, 
-instead of finding a specific index. Most APL functions have a dyadic and monadic form.) 
-Thus, the depth of a node is given by the following:
+The simplest operation over a node coordinate is to extract the depth
+of the node. The expression C⍳0 finds the first occurrence of 0 in C
+and returns the index of that occurrence. (The application of ⍳ here
+is its dyadic form, meaning that it receives two arguments. Previous
+uses of ⍳ have used its monadic form, which generates all indices,
+instead of finding a specific index. Most APL functions have a
+dyadic and monadic form.)  Thus, the depth of a node is given by
+the following:
 
@@
 
-The primary calculation we care about in order to do function lifting, expression flattening, 
-and other sorts of flattening transformations is to determine which nodes are ancestors of 
-another node. The core of this computation determines whether one node is a child of another 
-based on their node coordinates. This amounts to computing whether one node is a prefix of 
-another, ignoring zeros.
-
-We write (f g h) to represent the composition of functions f, g, and h as a function train. This i
-s used with another shorthand of composition according to the following equivalences:
+The primary calculation we care about in order to do function lifting,
+expression flattening, and other sorts of flattening transformations
+is to determine which nodes are ancestors of another node. The core
+of this computation determines whether one node is a child of another
+based on their node coordinates. This amounts to computing whether
+one node is a prefix of another, ignoring zeros.
+
+We write (f g h) to represent the composition of functions f, g,
+and h as a function train. This i s used with another shorthand of
+composition according to the following equivalences:
 
@@
 A(f g h)B ←→ (A f B) g (A h B)
-A(0 f h)B ←→ 0 f (A h B) ⍝ Any constant for 0
+A(0 f h)B ←→ 0 f (A h B) ⍝ 0 is a constant
 A(f g)B   ←→ f (A g B)
@@
 
-With this, we can write a function to compute whether a given coordinate is a prefix of another.
+With this, we can write a function to compute whether a given
+coordinate is a prefix of another.
 
@@
 
-Here P is the coordinate we wish to check against C, to determine whether C is a descendant. The
- functions are all standard logical functions extended to apply point-wise over arrays. The 
- function ⊢ always returns its right argument. The function (=∨0=⊢) reads intuitively as, “equal or 
- right element is zero.” In this case to make this a predicate we combine this with the “for all” 
- reduction using ∧⌿. This particular pattern is a special case of inner product, which we can compute 
- using f.g. Thus, +.× is a function that computes traditional matrix-matrix multiplication when applied 
- to two matrices. However, we can also use it to compute the above reduction and application:
+Here P is the coordinate we wish to check against C, to determine
+whether C is a descendant. The functions are all standard logical
+functions extended to apply point-wise over arrays. The function
+⊢ always returns its right argument. The function (=∨0=⊢)
+reads intuitively as, “equal or right element is zero.” In this
+case to make this a predicate we combine this with the “for all”
+reduction using ∧⌿. This particular pattern is a special case
+of inner product, which we can compute using f.g. Thus, +.× is a
+function that computes traditional matrix-matrix multiplication when
+applied to two matrices. However, we can also use it to compute the
+above reduction and application:
 
@@
 
-This extends the reduction to allow us to use matrix values for C and P instead of vectors, and thus 
-determine whether rows in C are descendants of multiple candidates given by P. This becomes important 
-to using the prefix function. In our examples, we want to determine the nearest ancestor of a certain 
-node type to which a node belongs. In the case of function lifting, we want to determine the nearest 
-ancestor function, in the case of expression flattening, we want to determine the nearest ancestor 
-expression node. We can extract the rows we care about for each example using ⌷, where i⌷M returns 
-the i-th row of M. In the following expressions, we use 0 2 rather than ¯1 in ⍤ to indicate that ⌷ will 
-receive the entire matrix on the right for each scalar index on the left.
+This extends the reduction to allow us to use matrix values for
+C and P instead of vectors, and thus determine whether rows in C
+are descendants of multiple candidates given by P. This becomes
+important to using the prefix function. In our examples, we want
+to determine the nearest ancestor of a certain node type to which a
+node belongs. In the case of function lifting, we want to determine
+the nearest ancestor function, in the case of expression flattening,
+we want to determine the nearest ancestor expression node. We can
+extract the rows we care about for each example using ⌷, where i⌷M
+returns the i-th row of M. In the following expressions, we use 0 2
+rather than ¯1 in ⍤ to indicate that ⌷ will receive the entire
+matrix on the right for each scalar index on the left.
 
@@
 
-The matrices Fp and Ep correspond to the function and expression node coordinates in our two 
-examples, respectively. We could now use Fp and Ep to compare against Fc and Ec and determine 
-which nodes contained which nodes in the tree. However, our prefix function will return 1 when we 
-ask whether a node is a prefix of itself. In this case we don’t want that. The solution to this is to 
-drop the last non-zero element from the coordinate. This will not prevent it from matching against 
-any of its ancestors, but will prevent matching against itself. We can use the depth vector with the 
-↑ function to take all but the last non-zero element. We need to remember to extend the returned 
-array with an extra zero column, since the shape will be too small otherwise.
+The matrices Fp and Ep correspond to the function and expression
+node coordinates in our two examples, respectively. We could now
+use Fp and Ep to compare against Fc and Ec and determine which nodes
+contained which nodes in the tree. However, our prefix function will
+return 1 when we ask whether a node is a prefix of itself. In this
+case we don’t want that. The solution to this is to drop the last
+non-zero element from the coordinate. This will not prevent it from
+matching against any of its ancestors, but will prevent matching
+against itself. We can use the depth vector with the ↑ function to
+take all but the last non-zero element. We need to remember to extend
+the returned array with an extra zero column, since the shape will
+be too small otherwise.
 
@@
 
-We can now use Fcp and Ecp to determine which Function and Expression nodes match for each node:
+We can now use Fcp and Ecp to determine which Function and Expression
+nodes match for each node:
 
@@
 
-We use the ⍉ function, which computes the transpose of its right argument, to ensure that the row 
-and column sizes match up. With the above, we can determine the “greatest” match, which is the closest. 
-We can replace each 1 in the above matrices by their column numbers (the number of the column in which 
-that 1 occurs), and then we can take the maximum column number in each row to determine an index in 
-either Fp or Ep that is the appropriate “parent” coordinate for that node. This works because we have 
-made sure to order the Ep and Fp matrices lexicographically. The following shows this computation for 
-the expression example; we elide the function example as redundant at this point. We also use ⌈/ instead 
-of ⌈⌿ to reduce along the last, rather than the first, axis.
+We use the ⍉ function, which computes the transpose of its right
+argument, to ensure that the row and column sizes match up. With the
+above, we can determine the “greatest” match, which is the closest.
+We can replace each 1 in the above matrices by their column numbers
+(the number of the column in which that 1 occurs), and then we can
+take the maximum column number in each row to determine an index in
+either Fp or Ep that is the appropriate “parent” coordinate for
+that node. This works because we have made sure to order the Ep and Fp
+matrices lexicographically. The following shows this computation for
+the expression example; we elide the function example as redundant
+at this point. We also use ⌈/ instead of ⌈⌿ to reduce along
+the last, rather than the first, axis.
 
@@
 
-At this point we have two values, Ek and Fk, which indicate the closest containing node that we care about 
-for each node in the tree, using its node coordinate.
+At this point we have two values, Ek and Fk, which indicate the
+closest containing node that we care about for each node in the tree,
+using its node coordinate.
 
@@
 
-We will use these keys to perform computation over the AST and accomplish our tasks of function lifting and 
-expression flattening in the next section, but we make a final note here, that we can imagine many other 
-operations which might be used throughout the compiler when we care about how nodes relate to each 
-other. The use of operators like reduction, scan, inner and outer products allow us to obtain the necessary 
-information from the node coordinate matrix. Sometimes these results might not be obvious, but the 
-example of the “belongs to” relationship given above demonstrates a pattern that arises repeatedly 
-throughout our data parallel compiler, and is therefore particularly useful to understand. 
-%[cf]
-
-%[cf]
-%[cf]
-%[of]:Case Studies
-\section{Case Studies}
-
-%[of]:Function Lifting
-\subsection{Function Lifting}
-
-If we use the key operator with Fk, we get the following (the monadic use of ↓ converts Fc from a 
-matrix to a vector of vectors):
-
-\begin{verbatim}
-      (1↓Fk){⍺ ⍵}⌸1↓Fd,Ft,Fv,⍪↓Fc
-┌─────────┬─────────────────┐
-│1 0 0 0 0│┌─┬─┬─┬─────────┐│
-│         ││1│E│0│1 1 0 0 0││
-│         │├─┼─┼─┼─────────┤│
-│         ││2│F│0│1 1 1 0 0││
-│         │├─┼─┼─┼─────────┤│
-│         ││2│A│0│1 1 2 0 0││
-│         │├─┼─┼─┼─────────┤│
-│         ││3│N│7│1 1 2 2 0││
-│         │└─┴─┴─┴─────────┘│
-├─────────┼─────────────────┤
-│1 1 1 0 0│┌─┬─┬─┬─────────┐│
-│         ││3│E│0│1 1 1 1 0││
-│         │├─┼─┼─┼─────────┤│
-│         ││4│V│⍵│1 1 1 1 1││
-│         │├─┼─┼─┼─────────┤│
-│         ││4│P│+│1 1 1 1 2││
-│         │├─┼─┼─┼─────────┤│
-│         ││4│V│⍵│1 1 1 1 3││
-│         │└─┴─┴─┴─────────┘│
-└─────────┴─────────────────┘
-\end{verbatim}
-
-Notice that we have now grouped all of the relevant parts of the tree according to which nodes would 
-appear in their respective functions after lifting. Refer to the original lifting example in the 
-introduction to verify this. Indeed. The second row in the above example shows the internal function 
-complete and ready to name. Each element in the second column corresponds to the body of one of our 
-lifted functions. In the case of the first function, the outer function, we have a spurious function node 
-in the body. This is intentional. When we lift these functions, we will replace each spurious function 
-node with a variable node referring to the function’s generated name.
-
-Each of these function bodies has a specific coordinate associated with it. Because these coordinates 
-are uniquely identifying, we can use these as input into a name generator to generate names that we 
-know are unique for each function body. Furthermore, because we retain this information in the 
-corresponding function nodes that appear in the body of each function to be lifted, we know exactly what 
-name that function has been given, and we can replace the function node with a variable node referencing 
-that name instead, without referring to any state outside of the immediate information given to the 
-function lifter. Indeed, each row in the above matrix represents a function lifting task that can be 
-completed without any additional information. That is, there are no dependencies between rows to perform 
-lifting. This gives us a straightforward parallel execution of function lifting.
-
-The final task to complete function lifting of each function body before lifting is to shift the depths in 
-the depth vectors to correspond to those of a function lifted to the top-level, and to attach a function 
-node to the top of each of the bodies. At that point, we simply recombine all of the newly created functions 
-into a single top level. 
-%[cf]
-%[of]:Expression Flattening
-\subsection{Expression Flattening}
-
-If we use Ek as the key for the expression example, we get the following:
-
-\begin{verbatim}
-      (1↓Ek){⍺ ⍵}⌸1↓Ed,Et,Ev,⍪↓Ec
-┌─────┬─────────────┐
-│1 0 0│┌─┬─┬─┬─────┐│
-│     ││1│E│0│1 1 0││
-│     │├─┼─┼─┼─────┤│
-│     ││1│P│÷│1 2 0││
-│     │├─┼─┼─┼─────┤│
-│     ││1│E│0│1 3 0││
-│     │└─┴─┴─┴─────┘│
-├─────┼─────────────┤
-│1 1 0│┌─┬─┬─┬─────┐│
-│     ││2│V│a│1 1 1││
-│     │├─┼─┼─┼─────┤│
-│     ││2│P│+│1 1 2││
-│     │├─┼─┼─┼─────┤│
-│     ││2│V│b│1 1 3││
-│     │└─┴─┴─┴─────┘│
-├─────┼─────────────┤
-│1 3 0│┌─┬─┬─┬─────┐│
-│     ││2│V│c│1 3 4││
-│     │├─┼─┼─┼─────┤│
-│     ││2│P│×│1 3 5││
-│     │├─┼─┼─┼─────┤│
-│     ││2│V│d│1 3 6││
-│     │└─┴─┴─┴─────┘│
-└─────┴─────────────┘
-\end{verbatim}
-
-Again, we can see immediately that we have grouped each set of nodes according to the expressions 
-that are to be lifted. Just as in the case of function lifting, we can adjust the depths of each 
-expression to the correct depth and we can replace each expression node with a variable reference
- based on that node’s coordinate. Each expression can be given a unique name based on its 
- coordinate. A later compiler pass can reduce these names down to the minimum actually required to 
- represent the expression if desired.
-
-The only extra issue involved here is to ensure that the order of evaluation matches. In our case, we 
-are assuming that the order of evaluation is right to left, which means that the above order is actually 
-backwards of the desired order. During recombination, we simply reverse these orders and this fixes 
-that problem. More work would be required to take into consideration a specific precedence hierarchy. 
-%[cf]
-%[of]:Lexical Scoping
-\subsection{Lexical Scoping}
-%[cf]
-%[cf]
-%[of]:Table of Compiler Passes
-\renewcommand{\arraystretch}{1.5}
+We will use these keys to perform computation over the AST and
+accomplish our tasks of function lifting and expression flattening in
+the next section, but we make a final note here, that we can imagine
+many other operations which might be used throughout the compiler when
+we care about how nodes relate to each other. The use of operators
+like reduction, scan, inner and outer products allow us to obtain
+the necessary information from the node coordinate matrix. Sometimes
+these results might not be obvious, but the example of the “belongs
+to” relationship given above demonstrates a pattern that arises
+repeatedly throughout our data parallel compiler, and is therefore
+particularly useful to understand.
+
 \begin{table*}
@@
  relationship with the Key operator and associated tree computation techniques}
-\label{table:passes}
+\label{tab:passes}
 \end{table*}
-%[cf]
-%[of]:Other Passes
+
+\section{Case Studies}
+
+The following compiler passes are often found in any suitably complex 
+language being compiled, and represent a difficult problem for anyone attempting 
+to do data-parallel computation. Indeed, they were the primary motivating 
+factor for the development of the node coordinate matrix and its associated 
+techniques. They demonstrate how we can operate on highly nested structures, 
+usually with the intent of eliminating that structure to more closely map 
+the intended output language.
+
+These case studies come from the Co-dfns compiler, which is compiling a 
+lexically scoped language that contains both nested functions and 
+expressions.
+
+\subsection{Function Lifting}
+
+If we use the key operator with Fk, we get the following (the monadic
+use of ↓ converts Fc from a matrix to a vector of vectors):
+
+\begin{verbatim}
+      (1↓Fk){⍺ ⍵}⌸1↓Fd,Ft,Fv,⍪↓Fc
+┌─────────┬─────────────────┐
+│1 0 0 0 0│┌─┬─┬─┬─────────┐│
+│         ││1│E│0│1 1 0 0 0││
+│         │├─┼─┼─┼─────────┤│
+│         ││2│F│0│1 1 1 0 0││
+│         │├─┼─┼─┼─────────┤│
+│         ││2│A│0│1 1 2 0 0││
+│         │├─┼─┼─┼─────────┤│
+│         ││3│N│7│1 1 2 2 0││
+│         │└─┴─┴─┴─────────┘│
+├─────────┼─────────────────┤
+│1 1 1 0 0│┌─┬─┬─┬─────────┐│
+│         ││3│E│0│1 1 1 1 0││
+│         │├─┼─┼─┼─────────┤│
+│         ││4│V│⍵│1 1 1 1 1││
+│         │├─┼─┼─┼─────────┤│
+│         ││4│P│+│1 1 1 1 2││
+│         │├─┼─┼─┼─────────┤│
+│         ││4│V│⍵│1 1 1 1 3││
+│         │└─┴─┴─┴─────────┘│
+└─────────┴─────────────────┘
+\end{verbatim}
+
+Notice that we have now grouped all of the relevant parts of the
+tree according to which nodes would appear in their respective
+functions after lifting. Refer to the original lifting example in
+the introduction to verify this. Indeed. The second row in the
+above example shows the internal function complete and ready to
+name. Each element in the second column corresponds to the body of
+one of our lifted functions. In the case of the first function, the
+outer function, we have a spurious function node in the body. This
+is intentional. When we lift these functions, we will replace
+each spurious function node with a variable node referring to the
+function’s generated name.
+
+Each of these function bodies has a specific coordinate associated
+with it. Because these coordinates are uniquely identifying, we can
+use these as input into a name generator to generate names that we
+know are unique for each function body. Furthermore, because we
+retain this information in the corresponding function nodes that
+appear in the body of each function to be lifted, we know exactly
+what name that function has been given, and we can replace the
+function node with a variable node referencing that name instead,
+without referring to any state outside of the immediate information
+given to the function lifter. Indeed, each row in the above matrix
+represents a function lifting task that can be completed without any
+additional information. That is, there are no dependencies between
+rows to perform lifting. This gives us a straightforward parallel
+execution of function lifting.
+
+The final task to complete function lifting of each function
+body before lifting is to shift the depths in the depth vectors
+to correspond to those of a function lifted to the top-level, and
+to attach a function node to the top of each of the bodies. At that
+point, we simply recombine all of the newly created functions into
+a single top level.
+
+\subsection{Expression Flattening}
+
+If we use Ek as the key for the expression example, we get the
+following:
+
+\begin{verbatim}
+      (1↓Ek){⍺ ⍵}⌸1↓Ed,Et,Ev,⍪↓Ec
+┌─────┬─────────────┐
+│1 0 0│┌─┬─┬─┬─────┐│
+│     ││1│E│0│1 1 0││
+│     │├─┼─┼─┼─────┤│
+│     ││1│P│÷│1 2 0││
+│     │├─┼─┼─┼─────┤│
+│     ││1│E│0│1 3 0││
+│     │└─┴─┴─┴─────┘│
+├─────┼─────────────┤
+│1 1 0│┌─┬─┬─┬─────┐│
+│     ││2│V│a│1 1 1││
+│     │├─┼─┼─┼─────┤│
+│     ││2│P│+│1 1 2││
+│     │├─┼─┼─┼─────┤│
+│     ││2│V│b│1 1 3││
+│     │└─┴─┴─┴─────┘│
+├─────┼─────────────┤
+│1 3 0│┌─┬─┬─┬─────┐│
+│     ││2│V│c│1 3 4││
+│     │├─┼─┼─┼─────┤│
+│     ││2│P│×│1 3 5││
+│     │├─┼─┼─┼─────┤│
+│     ││2│V│d│1 3 6││
+│     │└─┴─┴─┴─────┘│
+└─────┴─────────────┘
+\end{verbatim}
+
+Again, we can see immediately that we have grouped each set of nodes
+according to the expressions that are to be lifted. Just as in the
+case of function lifting, we can adjust the depths of each expression
+to the correct depth and we can replace each expression node with a
+variable reference based on that node’s coordinate. Each expression
+can be given a unique name based on its coordinate. A later compiler
+pass can reduce these names down to the minimum actually required to
+represent the expression if desired.
+
+The only extra issue involved here is to ensure that the order of
+evaluation matches. In our case, we are assuming that the order of
+evaluation is right to left, which means that the above order is
+actually backwards of the desired order. During recombination, we
+simply reverse these orders and this fixes that problem. More work
+would be required to take into consideration a specific precedence
+hierarchy.
+
 \section{Other Passes}
 
-Table \ref{table:passes} contains a selection of passes from the Co-dfns compiler 
-and breaks down their features based on how they traverse the tree and whether 
-or not they use the Key operator or Node Coordinates. They are oriented with 
-front-end passes higher in the table and passes further down the compiler 
-chain at the bottom of the table. These passes give a good idea of how the Key 
-operator plays a part in the complete compiler as well as how Node Coordinates 
+Table \ref{tab:passes} contains a selection of passes from the
+Co-dfns compiler and breaks down their features based on how they
+traverse the tree and whether or not they use the Key operator or
+Node Coordinates. They are oriented with front-end passes higher in
+the table and passes further down the compiler chain at the bottom
+of the table. These passes give a good idea of how the Key operator
+plays a part in the complete compiler as well as how Node Coordinates
 help to deal with complex sub-tree selections.
 
-We make a few notes on the table and the analysis. The Key operator can be used 
-with or without the node coordinates structure. In cases where the AST is 
-sufficiently flat and arranged in a sufficiently convenient order, the grouping 
-operations are very simple. In these cases, the Key operator amounts to a 
-map operation over some selected sub-trees selected at a specific depth. These 
-occur later in the compiler where the reader will note appearences of the Key 
-operator without the use of Node Coordinates.
-
-Likewise, node coordinates find their place early on in the compiler where it 
-is necessary to deal with the more highly nested AST. They are useful for identifying 
-this information regardless of whether the Key operator is used to handle the 
-grouping or not. Thus passes like ``Drop Unreachable Code'' that remove whole 
-sub-trees from the AST may use the Node Coordinates to do the grouping once 
-the correct parent node has been identified.
-
-In short, the goal of the structure and design of the compiler passes given 
-in table \ref{table:passes} is to utilize Node Coordinates to remove as quickly as possible 
-the nested structure of the AST into a flatter form that does not require complex 
-analysis to traverse. 
-%[cf]
-%[of]:The Co-dfns Compiler
+We make a few notes on the table and the analysis. The Key operator
+can be used with or without the node coordinates structure. In cases
+where the AST is sufficiently flat and arranged in a sufficiently
+convenient order, the grouping operations are very simple. In these
+cases, the Key operator amounts to a map operation over some selected
+sub-trees selected at a specific depth. These occur later in the
+compiler where the reader will note appearences of the Key operator
+without the use of Node Coordinates.
+
+Likewise, node coordinates find their place early on in the compiler
+where it is necessary to deal with the more highly nested AST. They
+are useful for identifying this information regardless of whether the
+Key operator is used to handle the grouping or not. Thus passes like
+``Drop Unreachable Code'' that remove whole sub-trees from the AST
+may use the Node Coordinates to do the grouping once the correct
+parent node has been identified.
+
+In short, the goal of the structure and design of the compiler passes
+given in table \ref{tab:passes} is to utilize Node Coordinates to
+remove as quickly as possible the nested structure of the AST into
+a flatter form that does not require complex analysis to traverse.
+
 \section{The Co-dfns Compiler}
-%[cf]
-%[of]:Performance
-\section{Performance}
-%[cf]
-%[of]:Future Work
+
+The Co-dfns compiler is an experimental commercial compiler funded by 
+Dyalog, Ltd. It is intended to provide increased scaling and performance
+for APL programmers using the dfns syntax (lexically scoped, functionally 
+oriented APL). These techniques form one of the foundational elements to 
+the architecture of the compiler, which is built in the style of a Nanopass 
+compiler, with very small passes chained together through composition, with 
+the added requirements that each compiler pass must be fully data parallel.
+
+The compiler produces good code that performs closely with hand-written 
+C code for the same programs on the GPU and CPU. It produces platform 
+independent code, meaning that the compiler is able to target either the 
+GPU or CPU from the same program source and produce reasonable performance 
+on both.
+
+The compiler itself is composed of three main parts: the core compiler, 
+the parser, and the code generator. The code generator includes the runtime 
+library of the compiler. Parsing APL code in parallel is a well-studied 
+problem. The compiler currently uses parsing combinators to do its parsing, 
+though we plan to implement the Two-by-Two parser given in the literature. 
+The code generator itself is a simple parallel map over each node, as 
+each node may be generated independent from the other nodes, leading to a 
+fully parallel generator, though there are some branching elements in the 
+generator that still remain, currently.
+
+The core of the Co-dfns compiler is written entirely using function trains 
+and parallel operations in the style described above. We use these techniques 
+extensively throughout the compiler, and the code is publicly available 
+at our repository. The core of the compiler, including some of the large 
+tables comes in at around 250 lines of code. Without the tables, the core 
+logic comes in at around 150 lines of code.
+
+The entire compiler, together with its runtime libraries, parser, generator, 
+and core, comes in at around 1500 lines of commented code.
+
 \section{Future Work}
-%[cf]
-%[of]:Related Work
+
+Work remains on expanding the features and capabilities of the compiler, 
+to demonstrate the scalability of these techniques on more sophisticated 
+optimizations. The current compiler represents a non-trivial instance of 
+the successful application of these techniques, but it focuses on an untyped, 
+functionally oriented language. The type inference in the current version 
+of the compiler is somewhat naive. More work is required to expand the 
+type inferencer to include more sophisticated type inference. Likewise, 
+the compiler does not currently handle error handling such as signals or 
+exceptions.
+
+Further work remains to implement these techniques on an imperative and 
+object-oriented language instead of a functional one. It would also be 
+educational to implement a version for a lazy language.
+
+The performance of these techniques has not been a problem in practice, but 
+we do not have a large scale understanding of the performance characteristics 
+of these techniques in real life. Further implementation and analysis is 
+required to understand what sort of compiler optimizations and analyses are 
+necessary to compile the Co-dfns compiler efficiently for the GPU and CPU. 
+We current lack a good comparison against existing techniques for the same 
+language, since the current compiler is relatively unique in that no other 
+compilers to our knowledge compile the same language to the same targets 
+with the same sorts of optimizations. We intend to implement other paradigms 
+and other languages which do have existing compiler implementations to compare 
+relative performance on multiple platforms.
+
+While we have found this method of compiler construction to be very compact 
+and to permit working with a data-parallel compiler as easily or nearly as 
+easily as with a standard Nanopass compiler, we have not done extensive 
+studies to determine just how comparable in programmability, ease of 
+maintenance, or extensibility these techniques are with established 
+compiler construction methods, such as OOP Visitors, Nanopass, or the 
+type-directed functional style.
+
 \section{Related Work}
 
-The J programming language [18] was the first practical, general-purpose programming language 
-to introduce the key operator as a primitive operator with the presumption of its general 
-usefulness. The rank operator (⍤) used throughout this treatment also derives from the J 
-traditional, receiving particular interest throughout the APL community [1, 17].
-
-Fritz Henglein demonstrated a class of operations, called discriminators, of which the Key operator 
-is a member. [16] Namely, a discriminator performs the same grouping computation as Key, but does 
-not apply a function over these groups with their keys. Henglein provides a linear implementation 
-of these operations.
-
-The EigenCFA effort [22] demonstrated significant performance improvements of a 0-CFA flow 
-analysis by utilizing similar techniques to those demonstrated here. In particular, encoding the 
-AST and using accessor functions have a very similar feel to the node coordinates and AST 
-encoding given here, though they have a different formulation and spend considerable effort 
-understanding the trade-offs of performance associated with the different encodings, whereas 
-the encodings here were chosen for their clarity and directness, rather than their performance.
-
-Mendez-Lojo, et al. implemented a GPU version of Inclusion-based Points-to Analysis [20] that also 
-focuses on adapting data structures and algorithms to efficiently execute on the GPU. In particular, 
-they use similar techniques of prefix sums and sorts to achieve some of their adaptation to the GPU, 
-Additionally, they have clever and efficient methods of representing graphs on the GPU which enable 
+The J programming language [18] was the first practical,
+general-purpose programming language to introduce the key operator as a
+primitive operator with the presumption of its general usefulness. The
+rank operator (⍤) used throughout this treatment also derives from
+the J traditional, receiving particular interest throughout the APL
+community [1, 17].
+
+Fritz Henglein demonstrated a class of operations, called
+discriminators, of which the Key operator is a member. [16] Namely,
+a discriminator performs the same grouping computation as Key, but
+does not apply a function over these groups with their keys. Henglein
+provides a linear implementation of these operations.
+
+The EigenCFA effort [22] demonstrated significant performance
+improvements of a 0-CFA flow analysis by utilizing similar techniques
+to those demonstrated here. In particular, encoding the AST and using
+accessor functions have a very similar feel to the node coordinates and
+AST encoding given here, though they have a different formulation and
+spend considerable effort understanding the trade-offs of performance
+associated with the different encodings, whereas the encodings here
+were chosen for their clarity and directness, rather than their
+performance.
+
+Mendez-Lojo, et al. implemented a GPU version of Inclusion-based
+Points-to Analysis [20] that also focuses on adapting data structures
+and algorithms to efficiently execute on the GPU. In particular,
+they use similar techniques of prefix sums and sorts to achieve some
+of their adaptation to the GPU, Additionally, they have clever and
+efficient methods of representing graphs on the GPU which enable
 dynamic rewriting of the graph.
 
-The APEX compiler [2] developed vectorized approaches to handling certain analyses to compile 
-traditional APL, including a SIMD tokenizer [4]. It uses a SSA representation, and converts the
-dynamic scope of traditional APL functions into a static form early on. It also uses a matrix 
-format to represent the AST. Traditional APL did not have nested function definitions, however, 
-and thus the APEX compiler does not have any specific approaches to dealing with function lifting.
-
-Bernecky further identified methods of reducing or optimizing the computational complexity or cost 
-of certain array operations, allowing improved performance of easy to understand array expressions [3].
-
-Timothy Budd implemented a compiler [5, 6] for APL which targeted vector processors as well as one 
-for C code. They used a method of lazy evaluation to avoid intermediate data copying. Budd provided 
-thoughts and some ideas on how the compiler might be implemented in parallel as well.
-
-Walter Schwarz implemented an APL to C compiler for the ACORN system targeting the CM-2 machine, 
-demonstrating performance potential for APL as a massively parallel language [21].
-
-W. Ching and D. Ju have spent significant work on the ELI language and other APL-class language
- implementations, especially on parallelized code and optimization. [8, 9, 10, 11, 13, 14, 15]
-
-J. D. Bunda and J. A. Gerth presented a method for doing table driven parsing of APL which 
-suggested a parallel optimization for parsing, but did not elucidate the algorithm [7].
-%[cf]
-%[of]:Conclusion
+The APEX compiler [2] developed vectorized approaches to handling
+certain analyses to compile traditional APL, including a SIMD tokenizer
+[4]. It uses a SSA representation, and converts the dynamic scope of
+traditional APL functions into a static form early on. It also uses
+a matrix format to represent the AST. Traditional APL did not have
+nested function definitions, however, and thus the APEX compiler does
+not have any specific approaches to dealing with function lifting.
+
+Bernecky further identified methods of reducing or optimizing the
+computational complexity or cost of certain array operations, allowing
+improved performance of easy to understand array expressions [3].
+
+Timothy Budd implemented a compiler [5, 6] for APL which targeted
+vector processors as well as one for C code. They used a method of
+lazy evaluation to avoid intermediate data copying. Budd provided
+thoughts and some ideas on how the compiler might be implemented in
+parallel as well.
+
+Walter Schwarz implemented an APL to C compiler for the ACORN system
+targeting the CM-2 machine, demonstrating performance potential for
+APL as a massively parallel language [21].
+
+W. Ching and D. Ju have spent significant work on the ELI language
+and other APL-class language
+ implementations, especially on parallelized code and optimization. [8,
+ 9, 10, 11, 13, 14, 15]
+
+J. D. Bunda and J. A. Gerth presented a method for doing table driven
+parsing of APL which suggested a parallel optimization for parsing,
+but did not elucidate the algorithm [7].
+
 \section{Conclusion}
 
-We have derived a method of performing computation over sub-trees selected on the basis of 
-inter-node properties through the use of the Key (⌸) operator and node coordinates, which enable 
-local computation of these inter-node properties. This method is both general and direct, and when 
-combined with traditional and more mundane array programming, suffices to implement the 
-complete core of a compiler, modulo parsing and code generation. The method requires no special 
-operations or unique special casing primitives in the language. Moreover, it is strictly data-parallel 
-and data-flow, without any complex control flow, which not only results in concise code, but also 
-enables much easier analysis of properties such as runtime complexity and verification of correctness.
-
-We have demonstrated the technique and the core insights behind the data structures involved. It 
-presents a solution to a very old and traditional problem in a very uncommon light, by eschewing 
-the common practices that underlie every other significant and general solution found in modern 
-compilers today and replacing them with an entirely different paradigm centred on parallelism 
-and aggregate operations.
-%[cf]
-%[of]:Acknowledgements
-%[cf]
+We have derived a method of performing computation over sub-trees
+selected on the basis of inter-node properties through the use of the
+Key (⌸) operator and node coordinates, which enable local computation
+of these inter-node properties. This method is both general and direct,
+and when combined with traditional and more mundane array programming,
+suffices to implement the complete core of a compiler, modulo parsing
+and code generation. The method requires no special operations or
+unique special casing primitives in the language. Moreover, it is
+strictly data-parallel and data-flow, without any complex control flow,
+which not only results in concise code, but also enables much easier
+analysis of properties such as runtime complexity and verification
+of correctness.
+
+We have demonstrated the technique and the core insights behind the
+data structures involved. It presents a solution to a very old and
+traditional problem in a very uncommon light, by eschewing the common
+practices that underlie every other significant and general solution
+found in modern compilers today and replacing them with an entirely
+different paradigm centred on parallelism and aggregate operations.
 

### 2016-04-05 2016-spring-array/2016-array-key.tex
Fix up formatting on a lot of lines and get things mostly working.

@@
 line with the rest of the compiler. This integration problem makes it
-difficult to scale the "ad hoc" parallelization effort of individual
+difficult to scale the ``ad hoc'' parallelization effort of individual
 compiler passes to a complete system. This is particularly true in
@@
 
-Furthermore, compilers are increasingly becoming a part of a "live"
+Furthermore, compilers are increasingly becoming a part of a ``live''
 and interactive feedback loop.  Just in time compilation and on
@@
 \texttt{?}               & Roll               & \\
-\texttt{\^}              &                    & And \\
+\texttt{∧}              &                    & And \\
 \texttt{∨}               &                    & Or \\
@@
 \caption{Primitive Monadic/Unary Operators, each takes a single left
-operand and describes a verb operating over one or two arguments}
+operand and describes a function operating over one or two arguments}
 \label{tab:adverbs}
@@
 \texttt{⍣} & Power & Iteration, Limited use only \\
-\texttt{⍤} & Rank & Apply a verb along cells of an array \\
+\texttt{⍤} & Rank & Apply a function along cells of an array \\
 \end{tabular}
 \caption{Primitive Dyadic/Binary Operators, each takes a left
- and right operand and describes a verb operating over one or two arguments}
+ and right operand and describes a function operating over one or two arguments}
 \label{tab:conjunctions}
@@
 
-\subsection{The Key Operator}
-
-The Key operator (written \verb;⌸;), takes a single function
-that expects two arguments and returns a function which takes two
-arguments. The left argument is a set of keys, and the right argument
-is a set of corresponding elements associated with those keys. In our
-case, we provide matrices for both of these arguments, so each pair
-of rows corresponds to a key-value pair. A simple example of the key
-operation at work is to compute a histogram (≢ here can be thought
-of as “tally” or count):
-
-\begin{verbatim}
-       ⊢X←?10⍴5
-1 1 4 0 1 3 1 1 2 1
-      X(⊣,(≢⊢))⌸X
-1 6
-4 1
-0 1
-3 1
-2 1
-\end{verbatim}
-
-To understand a bit better how the Key operator applies its function,
-consider the function {⍺ ⍵} which returns the pair of its right
-and left arguments. If we apply it to the same value as above, we
-get the following:
-
-\begin{verbatim}
-      X{⍺ ⍵}⌸X
-1  1 1 1 1 1 1 
-4  4           
-0  0           
-3  3           
-2  2
-\end{verbatim}
-
-In our case, we use either Fk or Ek as our keys applied to the
-corresponding AST. We also will drop off the first row in each AST
-using 1↓ since this node “contains” everything. In a complete
-AST this is usually the Module boundary node which contains the entire
-set of functions and values in the module.
-
 \subsection{Encoding the AST}
@@
 relationships in the form of a depth vector. The second column is a
-vector of the node types, while the third contains the “value”
+vector of the node types, while the third contains the ``value''
 of the node, such as the name in a variable.
@@
 
-The depth vector for these trees we name Fd and Ed, respectively:
+The depth vector for these trees we name \verb;Fd; and \verb;Ed;, respectively:
 
@@
 
-The node types we call Ft and Et:
+The node types we call \verb;Ft; and \verb;Et;:
 
@@
 
-And finally, we call the values vectors Fv and Ev:
+And finally, we call the values vectors \verb;Fv; and \verb;Ev;:
 
@@
 
-We combine these to form the respective AST matrices. We write A,B
-to catenate arrays A and B along their last axes and ⍪A to turn a
+We combine these to form the respective AST matrices. We write \verb;A,B;
+to catenate arrays \verb;A; and \verb;B; along their last axes and \verb;⍪A; to turn a
 vector into a 1-column matrix. Thus, the two tree matrices are given
@@
 3 N 7
-      Ed,Et,Ev
+      Ed,Et,⍪Ev
 0 E v
@@
 \item Are they siblings?
-\item Does one appear “earlier” in the tree?
+\item Does one appear ``earlier'' in the tree?
 \item Are they at the same depth?
@@
 
-We write f⌿A to reduce the first axis of A using function f. Thus,
-+⌿V is the sum of the elements in vector V. The function x⌈y
+We write \verb;f⌿A; to reduce the first axis of \verb;A; using function \verb;f;. Thus,
+\verb;+⌿V; is the sum of the elements in vector \verb;V;. The function \verb;x⌈y;
 gives the maximum of its two arguments. So, the depth of each tree
@@
 
-The function table or outer product of f over vectors U and V is
-written U ∘.f V giving a U V shaped matrix as a result. Thus,
-(⍳3)∘.×⍳3 gives a small multiplication table:
+The function table or outer product of \verb;f; over vectors \verb;U; and \verb;V; is
+written \verb;U ∘.f V; giving a \verb;U V; shaped matrix as a result. Thus,
+\verb;(⍳3)∘.×⍳3; gives a small multiplication table:
 
@@
 
-If we use = instead, we have a Boolean identity matrix:
+If we use \verb;=; instead, we have a Boolean identity matrix:
 
@@
 
-If we use ∘.= on the depth vector and its set of depths instead,
+If we use \verb;∘.=; on the depth vector and its set of depths instead,
 we see an expanded Boolean representation of the depth vector:
@@
 visually, but also suggest another step. We can compute a sum scan
-with +⍀, also called a prefix sum, along the first axis. Applying
+with \verb;+⍀;, also called a prefix sum, along the first axis. Applying
 this function on the above matrices leads to an interesting result:
@@
 
-The expression V f⍤¯1⊢M applies f to corresponding elements of
-V and rows of M:
+The expression \verb;V f⍤¯1⊢M; applies \verb;f; to corresponding elements of
+\verb;V; and rows of \verb;M;:
 
@@
 
-If n↑V takes the first n elements of V, then we can obtain coordinate
+If \verb;n↑V; takes the first \verb;n; elements of \verb;V;, then we can obtain coordinate
 matrices from the prefix sums by noting that the spurious digits
-all come after column d+1 where d is the depth corresponding to that
+all come after column \verb;d+1; where \verb;d; is the depth corresponding to that
 coordinate. The following gives a complete expression for computing
-a node coordinate matrix from a depth vector, shown using Fd and Fe:
+a node coordinate matrix from a depth vector, shown using \verb;Fd; and \verb;Fe;:
 
@@
 The simplest operation over a node coordinate is to extract the depth
-of the node. The expression C⍳0 finds the first occurrence of 0 in C
+of the node. The expression \verb;C⍳0; finds the first occurrence of \verb;0; in \verb;C;
 and returns the index of that occurrence. (The application of ⍳ here
@@
 
-We write (f g h) to represent the composition of functions f, g,
-and h as a function train. This i s used with another shorthand of
+We write \verb;(f g h); to represent the composition of functions \verb;f;, \verb;g;,
+and \verb;h; as a function train. This is used with another shorthand of
 composition according to the following equivalences:
@@
 A(f g h)B ←→ (A f B) g (A h B)
-A(0 f h)B ←→ 0 f (A h B) ⍝ 0 is a constant
+A(0 f h)B ←→ 0 f (A h B) ⍝ Constant Case
 A(f g)B   ←→ f (A g B)
@@
 
-Here P is the coordinate we wish to check against C, to determine
-whether C is a descendant. The functions are all standard logical
+Here \verb;P; is the coordinate we wish to check against \verb;C;, to determine
+whether \verb;C; is a descendant. The functions are all standard logical
 functions extended to apply point-wise over arrays. The function
-⊢ always returns its right argument. The function (=∨0=⊢)
-reads intuitively as, “equal or right element is zero.” In this
-case to make this a predicate we combine this with the “for all”
-reduction using ∧⌿. This particular pattern is a special case
-of inner product, which we can compute using f.g. Thus, +.× is a
+\verb;⊢; always returns its right argument. The function \verb;(=∨0=⊢);
+reads intuitively as, ``equal or right element is zero.'' In this
+case to make this a predicate we combine this with the ``for all''
+reduction using \verb;∧⌿;. This particular pattern is a special case
+of inner product, which we can compute using \verb;f.g;. Thus, \verb;+.×; is a
 function that computes traditional matrix-matrix multiplication when
@@
 This extends the reduction to allow us to use matrix values for
-C and P instead of vectors, and thus determine whether rows in C
-are descendants of multiple candidates given by P. This becomes
+\verb;C; and \verb;P; instead of vectors, and thus determine whether rows in \verb;C;
+are descendants of multiple candidates given by \verb;P;. This becomes
 important to using the prefix function. In our examples, we want
@@
 we want to determine the nearest ancestor expression node. We can
-extract the rows we care about for each example using ⌷, where i⌷M
-returns the i-th row of M. In the following expressions, we use 0 2
-rather than ¯1 in ⍤ to indicate that ⌷ will receive the entire
+extract the rows we care about for each example using \verb;⌷;, where \verb;i⌷M;
+returns the i-th row of \verb;M;. In the following expressions, we use \verb;0 2;
+rather than \verb;¯1; in \verb;⍤; to indicate that \verb;⌷; will receive the entire
 matrix on the right for each scalar index on the left.
@@
 
-The matrices Fp and Ep correspond to the function and expression
+The matrices \verb;Fp; and \verb;Ep; correspond to the function and expression
 node coordinates in our two examples, respectively. We could now
-use Fp and Ep to compare against Fc and Ec and determine which nodes
+use \verb;Fp; and \verb;Ep; to compare against \verb;Fc; and \verb;Ec; and determine which nodes
 contained which nodes in the tree. However, our prefix function will
@@
 matching against any of its ancestors, but will prevent matching
-against itself. We can use the depth vector with the ↑ function to
+against itself. We can use the depth vector with the \verb;↑; function to
 take all but the last non-zero element. We need to remember to extend
@@
 
-We can now use Fcp and Ecp to determine which Function and Expression
+We can now use \verb;Fcp; and \verb;Ecp; to determine which Function and Expression
 nodes match for each node:
@@
 argument, to ensure that the row and column sizes match up. With the
-above, we can determine the “greatest” match, which is the closest.
+above, we can determine the ``greatest'' match, which is the closest.
 We can replace each 1 in the above matrices by their column numbers
@@
 take the maximum column number in each row to determine an index in
-either Fp or Ep that is the appropriate “parent” coordinate for
-that node. This works because we have made sure to order the Ep and Fp
+either \verb;Fp; or \verb;Ep; that is the appropriate ``parent'' coordinate for
+that node. This works because we have made sure to order the \verb;Ep; and \verb;Fp;
 matrices lexicographically. The following shows this computation for
 the expression example; we elide the function example as redundant
-at this point. We also use ⌈/ instead of ⌈⌿ to reduce along
+at this point. We also use \verb;⌈/; instead of \verb;⌈⌿; to reduce along
 the last, rather than the first, axis.
@@
 
-At this point we have two values, Ek and Fk, which indicate the
+At this point we have two values, \verb;Ek; and \verb;Fk;, which indicate the
 closest containing node that we care about for each node in the tree,
@@
 the necessary information from the node coordinate matrix. Sometimes
-these results might not be obvious, but the example of the “belongs
-to” relationship given above demonstrates a pattern that arises
+these results might not be obvious, but the example of the ``belongs
+to'' relationship given above demonstrates a pattern that arises
 repeatedly throughout our data parallel compiler, and is therefore
@@
 
+\subsection{The Key Operator}
+
+The Key operator (written \verb;⌸;), takes a single function
+that expects two arguments and returns a function which takes two
+arguments. The left argument is a set of keys, and the right argument
+is a set of corresponding elements associated with those keys. In our
+case, we provide matrices for both of these arguments, so each pair
+of rows corresponds to a key-value pair. A simple example of the key
+operation at work is to compute a histogram (\verb;≢; here can be thought
+of as ``tally'' or count):
+
+\begin{verbatim}
+       ⊢X←?10⍴5
+1 1 4 0 1 3 1 1 2 1
+      X(⊣,(≢⊢))⌸X
+1 6
+4 1
+0 1
+3 1
+2 1
+\end{verbatim}
+
+To understand a bit better how the Key operator applies its function,
+consider the function \verb;{⍺ ⍵}; which returns the pair of its right
+and left arguments. If we apply it to the same value as above, we
+get the following:
+
+\begin{verbatim}
+      X{⍺ ⍵}⌸X
+1  1 1 1 1 1 1 
+4  4           
+0  0           
+3  3           
+2  2
+\end{verbatim}
+
+In our case, we use either \verb;Fk; or \verb;Ek; as our keys applied to the
+corresponding AST. We also will drop off the first row in each AST
+using \verb;1↓; since this node ``contains'' everything. In a complete
+AST this is usually the Module boundary node which contains the entire
+set of functions and values in the module.
+
 \section{Case Studies}
@@
 
-If we use the key operator with Fk, we get the following (the monadic
-use of ↓ converts Fc from a matrix to a vector of vectors):
+If we use the key operator with \verb;Fk;, we get the following (the monadic
+use of \verb;↓; converts \verb;Fc; from a matrix to a vector of vectors):
 
@@
 functions after lifting. Refer to the original lifting example in
-the introduction to verify this. Indeed. The second row in the
+the introduction to verify this. Indeed, the second row in the
 above example shows the internal function complete and ready to
@@
 
-If we use Ek as the key for the expression example, we get the
+If we use \verb;Ek; as the key for the expression example, we get the
 following:
@@
 
-The J programming language [18] was the first practical,
+The J programming language \cite{hui2014key} was the first practical,
 general-purpose programming language to introduce the key operator as a
@@
 the J traditional, receiving particular interest throughout the APL
-community [1, 17].
+community \cite{bernecky1987rank,hui1995rank}.
 
 Fritz Henglein demonstrated a class of operations, called
-discriminators, of which the Key operator is a member. [16] Namely,
+discriminators, of which the Key operator is a member.  
+\cite{henglein2013dd}
+Namely,
 a discriminator performs the same grouping computation as Key, but
@@
 
-The EigenCFA effort [22] demonstrated significant performance
+The EigenCFA effort \cite{prabhu2011eigencfa} demonstrated significant performance
 improvements of a 0-CFA flow analysis by utilizing similar techniques
@@
 Mendez-Lojo, et al. implemented a GPU version of Inclusion-based
-Points-to Analysis [20] that also focuses on adapting data structures
+Points-to Analysis \cite{mendez2012inclusion} that also focuses on 
+adapting data structures
 and algorithms to efficiently execute on the GPU. In particular,
@@
 
-The APEX compiler [2] developed vectorized approaches to handling
+The APEX compiler \cite{bernecky1997apex} developed vectorized approaches to handling
 certain analyses to compile traditional APL, including a SIMD tokenizer
-[4]. It uses a SSA representation, and converts the dynamic scope of
+\cite{bernecky2003tokenizer}.
+ It uses a SSA representation, and converts the dynamic scope of
 traditional APL functions into a static form early on. It also uses
@@
 computational complexity or cost of certain array operations, allowing
-improved performance of easy to understand array expressions [3].
-
-Timothy Budd implemented a compiler [5, 6] for APL which targeted
+improved performance of easy to understand array expressions
+\cite{bernecky1999reducing}.
+
+Timothy Budd implemented a compiler \cite{budd1984apl,budd2012apl} for APL which targeted
 vector processors as well as one for C code. They used a method of
@@
 targeting the CM-2 machine, demonstrating performance potential for
-APL as a massively parallel language [21].
+APL as a massively parallel language \cite{schwarz1991acorn}.
 
@@
 and other APL-class language
- implementations, especially on parallelized code and optimization. [8,
- 9, 10, 11, 13, 14, 15]
+ implementations, especially on parallelized code and optimization. 
+\cite{ching2000design,ching1994experimental,ching1993primitive,
+ching1990automatic,hendriks1990sparse,ju1991exploitation,
+ju1991performance}
 
@@
 parsing of APL which suggested a parallel optimization for parsing,
-but did not elucidate the algorithm [7].
+but did not elucidate the algorithm \cite{bunda1984apl}.
 
@@
 
-% Bibliography goes here
+\bibliographystyle{plain}
+\bibliography{biblio}
 

### 2016-04-05 2016-spring-array/2016-array-key.tex
Cleanup the Introduction and Conclusion

@@
 
-Compilers are everywhere, but their design and structure does not scale
-well to modern parallel architectures. Traditional compilers are very
-difficult to parallelize, and the benefits of modern architectures is
-often seen only through the use of distributed build systems which
-invoke the single threaded compilers on multiple ``documents'' at
-the same time. Future CPU architectures promise to be even more data
-parallel, such as the Xeon Phi and the GPGPU architectures. On these
-systems it is even harder to obtain good performance for compilers
-written in the traditional fashion.
-
-Fundamentally, a compiler is a translator that operates over some sort
-of tree and transforms it to another tree of an usually radically
-different shape but equivalent ``semantics.'' These algorithms and
-transformations show up everywhere, from document processing on the
-Web, XML-based publishing workflows, to the programming language
-compilers with which most developers are intimately familar.
-These transformations differ from other widely studied parallel
-graph problems in that they tend to greatly alter the fundamental
-structure of the graph/tree, unlike, say, breadth-first search or
-single-source shortest path algorithms. This rapidly shifting graph
-structure requires different approaches to achieving good parallelism.
-
-There has been some interesting use of GPUs to accelerate certain
-compiler analyses and passes.  Unfortunately, attempts to parallelize
-compilers usually takes the form of a great deal of work spent
-to highly optimize a single compiler pass. The results are often
-fragile and require a great deal of effort to maintain and keep in
-line with the rest of the compiler. This integration problem makes it
-difficult to scale the ``ad hoc'' parallelization effort of individual
-compiler passes to a complete system. This is particularly true in
-situations where the compiler passes may change rapidly, such as with
-XSLT or document publishing, where it is very common for the document
-preparer to write or tweak transformation passes to suit the specific
-class of documents being handled. In these cases, the expertise and
-time involved in parallelizing such transformations makes the ad hoc
-approach mostly intractable.
-
-Furthermore, compilers are increasingly becoming a part of a ``live''
-and interactive feedback loop.  Just in time compilation and on
-the fly static compilation are both used to provide custom binaries
-tuned to specific tasks on the fly. These techniques are used in web
-services and large scale databases, where in-memory databases may
-construct functions on the fly to execute. There is great interest
-in moving such systems on to the GPU or other highly data-parallel
-architectures, but it would significantly hamper the performance of
-the system if the compilation phase of query execution were required
-to occur on the CPU, especially at scale where there may be many more
-GPU accelerators than CPU resources.
-
-We propose a language-driven approach to compiler construction that
-provides a framework for the development of arbitrary compiler
-transformations that are data-parallel by construction. These
-techniques intentionally optimize for the data-parallel case, which
-permits a wide degree of platform independent performance compared to
-the traditional approach to compiler design, which requires significant
-effort and possibly signficiant transformation of the source code to
-achieve good parallel performance on high-parallel architectures.
-
-We use a mathematically oriented, array based language that is
-accessible and popular with information scientists. We restrict
-this language to permit only the composition of data-flow oriented
-array based programs whose core primitives are all data-parallel by
-definition. The permitted compositions of these primitives ensures
-that the resulting programs are both analyzable and parallel
-by construction, permitting a wide array of whole program and
-transformation specific optimizations that have been long studied.
-The notation is intentially close to mathematical notation, and
-as such, the resulting compiler passes more closely resemble the
-compact formal descriptions of the transformations rather than complex
-recursive algorithms for transforming trees.
-
-At the core of such techniques, we must have some method for
-transforming trees in a controlled and predictable manner. We
-build on the Nanopass paradigm for compiler construction, wherein
-a compiler is developed as a series of small, functional tree
-transformations (passes) that have an input AST language and an
-output AST language. Through progressive refinement of the tree, the
-composition of these passes results in the complete compiler. This is a
-highly functional approach, with very little mutation.  Fundamentally,
-each pass must identify regions of the tree upon which it works,
-do some sort of computation on that sub-tree that has been matched
-(often through pattern matching on the nodes) and reassemble the tree
-in some new order that matches the desired structure. In other words,
-a compiler nanopass can be thought of as an arbitrary computation
-over a given selection of sub-trees in an AST, where this computation
-could move, relocate, or change the sub-tree.
-
-The key to our language-drive approach is a set of techniques
-that enable a data-parallel expression of these arbitrary
-transformations. By using a higher-order function called Key, together
-with basic array programming techniques and an encoding for the tree
-connections called a Node Coordinate matrix, we can identify sub-trees
-and select them for computation by their parent-child relationships
-in a data-parallel fashion, and then compute on these sub-trees using
-traditional array programming techniques.
-
-We have constructed a complete compiler for a superset of our language
-called Co-dfns. This compiler is part of a commercial effort to develop
-platform independent performance for array programs. The compiler is
-developed completely in a data-parallel by construction as described in
-this paper, and uses only traditional and well known array primitives
-together with function composition to form the entire core compiler
-passes and generator. As a result, the compiler takes the form of a
-pure data-flow graph without mutation and has an extremely uniform and
-concise definition. It uses no recursion, pattern matching, branching,
-or other forms of explicit looping. The compiler, together with its
-runtime libraries and parser, comes in under 1000 lines of code.
-
-In this paper we fully elucidate three compiler passes from the Co-dfns
-compiler to demonstrate the techniques used and we provide further
-analysis of how these techniques are used in the other compiler
-passes. We further provide a comparative analysis of these passes
-against their counterparts implemented using the Nanopass compiler
-framework, comparing performance both in terms of programmer effort
-and portability/maintainability, as well as runtime performance.
+Compilers represent a peculiar class of tree/graph algorithms that greatly 
+alter the underlying structure of the input graph into something very 
+different in structure, but equivalent by some measure of interpretation. 
+These algorithms are widely applicable, but have stubbornly resisted a 
+general approach to parallelization. Most compilers are single-threaded 
+or make very limited use of parallelism in an ad hoc way. The design and 
+analysis of compilers almost universally deals with compilers as recursive 
+traversals over ASTs. Such formulations usually include heavy reliance on 
+branching, recursion, and sometimes very sophisticated control flow in 
+practice. All of this contributes to make it difficult to effectively 
+parallelize such algorithms onto architectures that are sensitive to 
+branching and recursion, such as GPGPUs or highly vectorized CPUs.
+
+There has been some success in efficiently executing parts of a compiler 
+on the GPU, such as the parser \cite{bunda1984apl}, 
+tokenization \cite{bernecky2003tokenizer}, and certain compiler 
+analyses \cite{prabhu2011eigencfa,mendez2012inclusion}. However, a generalized 
+framework or strategy for parallel compilation remains elusive. One of 
+the major difficulties is the recursive and highly branch dependent nature 
+of the algorithms. These present challenges that must be overcome in 
+vector-centric architectures.
+
+We present a language-driven strategy for creating an inherently parallel 
+compiler. We select a set of well known data-parallel primitives, and then 
+restrict program construction to the composition of these operations into 
+functions, without any forms of non-linear control flow, such as branching, 
+recursion, or pattern matching. Programs constructed in such a style are 
+data-parallel by construction. A key problem to the construction of compilers 
+in such a restricted environment is dealing with the inherently recursive and 
+nested structure of the AST. By working over a linearized representation of 
+the AST as a matrix, combined with a structure we call a Node Coordinate 
+Matrix, and using data-parallel primitives, we are able to tame the recursive 
+and nested nature of the AST into something that can be processed handily 
+using only data-parallel, data-flow programs, without requiring branching 
+or other forms of non-data parallel control flow.
+
+In the following sections we describe the core methods and idioms surrounding 
+the AST encoding, construction of the Node Coordinate Matrix, and the use 
+of these structures in handling nested, recursive AST relationships. This 
+enables us to perform arbitrary computation over arbitrary sub-trees of an 
+AST selected by parent-child relationships in the manner often seen in 
+compiler transformations. These ideas have been further implemented and tested 
+through the implementation of a complete data-parallel by construction 
+compiler, called Co-dfns, that compiles a lexically scoped, functionally 
+oriented dialect of APL with nested functions. The compiler targets both the 
+GPU and the CPU, and its core compiler is implemented in this pure, restricted 
+language that uses only function composition over data-parallel primitives.
 
@@
 
-\item We situate this technique into a broader, language-driven
-compiler construction strategy that enables the development of parallel
-by construction compilers that exhibit desirable programmability
-features in terms of code size, complexity, portability, and
-maintainability.
+\item We situation this technique into a broader, language-driven 
+strategy for compiler construction that enables the development of parallel 
+by construction compilers that are high-level in their implementation and 
+exceptionally concise.
 
 \item We provide analysis of these techniques as used in a commercial
-compiler project called Co-dfns and report on the results.
-
-\item We compare these techniques to the Nanopass approach to
-compiler design in terms of programmer productivity and compiler
-runtime performance.
-
-\item We enumerate a set of programming techniques, patterns, and
-strategies that complement the Key operator and Node Coordinates
-in the construction of tree transformations in general and give some
-intuitions on the use of these techniques on arbitrary transformations.
+compiler project called Co-dfns and report on the results,
+including the overall architecture of the compiler, the uses of these 
+techniques within the compiler, and the demonstration of these techniques 
+applied to two specific compiler passes.
+
 \end{itemize}
@@
 strictly data-parallel and data-flow, without any complex control flow,
-which not only results in concise code, but also enables much easier
-analysis of properties such as runtime complexity and verification
-of correctness.
+which results in exceptionally concise code. These methods permit a 
+general, high-level approach to constructing a compiler that is 
+parallel by construction by constricting the language to only a data parallel 
+subset of a normal array language.
 

### 2016-04-07 2016-spring-array/2016-array-key.tex
Cleanup the main sections

@@
 application may apply a function (such as \verb;+;) to one or two arguments, 
+called monadic and dyadic application respectively,
 which must appear on the right and (optionally) on the left of the function 
@@
 right and left, and will return a function. In the following 
-example, we apply a reduction operator to a function created using the 
-composition operator, as an example:
-
-\begin{verbatim}
+example, we apply the reduction operator (\verb;/;) to a function created using the 
+composition operator (\verb;∘;), as an example:
+
+\begin{verbatim}
+      +/1 2 3 4 5
+15
+      1+-(2+-(3+-(4+-(5))))
+3
+      1 +∘- 2 +∘- 3 +∘- 4 +∘- 5
+3
       +∘-/1 2 3 4 5
 3
-      +/1 2 3 4 5
-15
-\end{verbatim}
-
-Uses of the notation are introduced slowly throughout the treatment, so 
-the reader is encouraged to simply proceed through the treatment, referencing 
-the tables only when required.
-
-\begin{table}
-\centering
-\begin{tabular}{cll}
-\toprule
-Symbol                   & Monadic            & Dyadic \\
-\midrule
-\texttt{+}               & Identity           & Plus (Add) \\
-\texttt{-}               & Negative           & Minus (Subtract) \\
-\texttt{×}               & Direction (Signum) & Times (Multiply) \\
-\texttt{÷}               & Reciprocal         & Divide \\
-\texttt{|}               & Magnitude          & Residue (Modulo) \\
-\texttt{⌊}               & Floor              & Minimum \\
-\texttt{⌈}               & Ceiling            & Maximum \\
-\texttt{*}               & Exponential        & Power \\
-\texttt{⍟}               & Natural Logarithm  & Logarithm \\
-\texttt{○}               & Pi Times           & Circular (Trigonometric) \\
-\texttt{!}               & Factorial          & Binomial \\
-\texttt{\textasciitilde} & Not                & \\
-\texttt{?}               & Roll               & \\
-\texttt{∧}              &                    & And \\
-\texttt{∨}               &                    & Or \\
-\texttt{⍲}               &                    & Nand \\
-\texttt{⍱}               &                    & Nor \\
-\texttt{<}               &                    & Less \\
-\texttt{≤}               &                    & Less Or Equal \\
-\texttt{=}               &                    & Equal \\
-\texttt{≥}               &                    & Greater Or Equal \\
-\texttt{>}               &                    & Greater \\
-\texttt{≠}               &                    & Not Equal \\
-\end{tabular}
-\caption{Scalar Primitives}
-\label{tab:scalarprims}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{cll}
-\toprule
-Symbol     & Monadic       & Dyadic \\
-\midrule
-\texttt{⍴} & Reshape       & \\
-\texttt{,} & Ravel         & Catenate/Laminate \\
-\texttt{⍪} & Table         & Catenate First/Laminate \\
-\texttt{⌽} & Reverse       & Rotate \\
-\texttt{⊖} & Reverse First & Rotate First \\
-\texttt{⍉} & Transpose     & Transpose \\
-\texttt{↑} & Mix           & \\
-\texttt{↓} & Split         & \\
-\texttt{⊂} & Enclose       & Partitioned Enclose \\
-\texttt{∊} & Enlist        & \\
-\end{tabular}
-\caption{Structural Mixed Primitives}
-\label{tab:structprims}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{cll}
-\toprule
-Symbol                   & Monadic  & Dyadic \\
-\midrule
-\texttt{⊃}               & Disclose & Pick\\
-\texttt{↑}               &          & Take \\
-\texttt{↓}               &          & Drop \\
-\texttt{/}               &          & Replicate \\
-\texttt{⌿}               &          & Replicate First \\
-\texttt{\textbackslash}  &          & Expand \\
-\texttt{⍀}               &          & Expand First \\
-\texttt{\textasciitilde} &          & Without (Excluding) \\
-\texttt{∩}               &          & Intersection \\
-\texttt{∪}               & Unique   & Union \\
-\texttt{⊣}               & Same     & Left \\
-\texttt{⊢}               & Identity & Right\\
-\end{tabular}
-\caption{Selection Mixed Primitives}
-\label{tab:selprims}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{cll}
-\toprule
-Symbol     & Monadic         & Dyadic \\
-\midrule
-\texttt{⍳} & Index Generator & Index Of \\
-\texttt{∊} &                 & Membership \\
-\texttt{⍋} & Grade Up        & Grade Up \\
-\texttt{⍒} & Grade Down      & Grade Down \\
-\texttt{?} &                 & Deal \\
-\texttt{⍷} &                 & Find \\
-\end{tabular}
-\caption{Selector Mixed Primitives}
-\label{tab:seltorprims}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{cll}
-\toprule
-Symbol     & Monadic       & Dyadic \\
-\midrule
-\texttt{⍴} & Shape         & \\
-\texttt{≡} & Depth         & Match \\
-\texttt{≢} & Tally         & Not Match \\
-\texttt{⍎} & Execute       & Execute \\
-\texttt{⍕} & Format        & Format \\
-\texttt{⊥} &               & Decode (Base) \\
-\texttt{⊤} &               & Encode (Representation) \\
-\texttt{⌹} & Matrix Divide & Matrix Inverse \\
-\end{tabular}
-\caption{Miscellaneous Mixed Primitives}
-\label{tab:miscprims}
-\end{table}
-
-\begin{table*}
-\centering
-\begin{tabular}{cll}
-\toprule
-Symbol     & Name & Description \\
-\midrule
-\texttt{⍨} & Commute & Swaps arguments or distributes
- right argument to both sides \\
-\texttt{¨} & Each & Applies its operand point-wise over the left/right
- arguments \\
-\texttt{/} & Reduce & Reduce along the last axis \\
-\texttt{⌿} & Reduce First & Reduce along the first axis \\
-\texttt{\textbackslash} & Scan & Scan along the last axis \\
-\texttt{⍀} & Scan First & Scan along the first axis \\
-\texttt{⌸} & Key & Apply operand once for each sub-array grouped by key \\
-\end{tabular}
-\caption{Primitive Monadic/Unary Operators, each takes a single left
-operand and describes a function operating over one or two arguments}
-\label{tab:adverbs}
-\end{table*}
-
-\begin{table*}
-\centering
-\begin{tabular}{cll}
-\toprule
-Symbol     & Name & Description \\
-\midrule
-\texttt{∘} & Compose & Composes two operands as in traditional mathematics \\
-\texttt{.} & Inner Product & Inner product operation, e.g.
- \texttt{+.×} for matrix multiplication \\
-\texttt{∘.} & Outer Product & Cartesian product or ``function table'' \\
-\texttt{⍣} & Power & Iteration, Limited use only \\
-\texttt{⍤} & Rank & Apply a function along cells of an array \\
-\end{tabular}
-\caption{Primitive Dyadic/Binary Operators, each takes a left
- and right operand and describes a function operating over one or two arguments}
-\label{tab:conjunctions}
-\end{table*}
+\end{verbatim}
+
+We introduce notation as needed throughout the exposition, so the reader 
+is encouraged to refer to the tables as necessary to recall particular 
+operations.
 
@@
 
-We divide the work of sub-tree computation into three basic phases. Firstly, 
+We divide the work of sub-tree computation into three basic phases:  
 we maintain a node coordinate with each node that uniquely identifies it 
@@
 work as a single sub-tree, and then we operate over these groupings using 
-the key operator.
+the Key operator (\verb;⌸;).
 
@@
 
-We represent the AST as a matrix of 3 columns and n rows, one for
-each node in the tree. The first column contains the inter-node
+We represent the AST as a 3 column matrix with one row per node in the tree.
+The first column contains the inter-node
 relationships in the form of a depth vector. The second column is a
@@
 
-Parsing the examples above gives the following ASTs:
+We will use the following two running examples throughout. The \verb;F; 
+tree is an example nested function, while the \verb;E; tree is an example 
+nested expression.
 
@@
 
-The depth vector for these trees we name \verb;Fd; and \verb;Ed;, respectively:
-
-\begin{verbatim}
-Fd←0 1 2 3 4 4 4 2 2 3
+The depth vectors for these trees we name \verb;Fd; and \verb;Ed;, respectively:
+
+\begin{verbatim}
+Fd←0 1 2 3 4 4 4 2 3
 Ed←0 1 2 2 2 1 1 2 2 2
@@
 \begin{verbatim}
-Ft←’FEFEVPVAN’
-Et←’EEVPVPEVPV’
+Ft←'FEFEVPVAN'
+Et←'EEVPVPEVPV'
 \end{verbatim}
@@
 \begin{verbatim}
-Fv←’f’ 0 0 0 ‘⍵’ ‘+’ 0 7
-Ev←’v’ 0 ‘a’ ‘+’ ‘b’ ‘÷’ 0 ‘c’ ‘×’ ‘d’
+Fv←'f' 0 0 0 '⍵' '+' '⍵' 0 7
+Ev←'v' 0 'a' '+' 'b' '÷' 0 'c' '×' 'd'
 \end{verbatim}
@@
 to catenate arrays \verb;A; and \verb;B; along their last axes and \verb;⍪A; to turn a
-vector into a 1-column matrix. Thus, the two tree matrices are given
+vector into a 1-column matrix. Thus, the two AST matrices are given
 by the following expressions:
@@
 
-Note especially that all inter-node information lives in the depth
-vector, but that this information requires non-local access as
-described in the previous section to use it. Constructing node
-coordinates fixes this issue.
+The Depth Vector stores all edge information for the tree, but this information 
+requires non-local access to utilize, such as traversing potentially the 
+entire vector to determine the children of a specific node.
+Node Coordinates fix this issue.
 
@@
 
-Every node in an AST has a node coordinate, named because a coordinate
+Every node in an AST has a Node Coordinate, named because a coordinate
 is a precise location in a space.  We can imagine all the nodes
@@
 
+\noindent
 In short, we care about the relative position of each node in a tree
@@
 
-To understand how we construct these coordinates we must consider
-how we represent the AST.
-
 \subsubsection{Constructing Node Coordinates}
 
-In building the node coordinates, our goal is to build a matrix where
-each row is a node coordinate satisfying our previously described
-requirements. We end up with an n d shaped matrix where n is the number
-of rows and d the depth of the tree. As noted above, the depth vector
-contains all the necessary information, but in the wrong form.
+We construct the Node Coordinate Matrix from the depth vector, thereby encoding 
+the depth vector into a more useful format. The matrix has \verb;N; rows and 
+\verb;D; columns, where \verb;N; is the node count and \verb;D; is the depth 
+of the tree.
 
@@
 \verb;+⌿V; is the sum of the elements in vector \verb;V;. The function \verb;x⌈y;
-gives the maximum of its two arguments. So, the depth of each tree
-is given by the following:
+gives the maximum of its two arguments.
+We compute the depth of each tree as follows:
 
@@
 
+\noindent
 We can obtain the ordered sequence  by writing \verb;⍳n;:
@@
 
+\noindent
 So the depths of all nodes that appear in the depth vectors is thus:
@@
 
+\noindent
 The function table or outer product of \verb;f; over vectors \verb;U; and \verb;V; is
@@
 
+\noindent
 If we use \verb;=; instead, we have a Boolean identity matrix:
@@
 
+\noindent
 If we use \verb;∘.=; on the depth vector and its set of depths instead,
@@
 
+\noindent
 These matrices let us see the nesting features of each tree more
@@
 
+\noindent
 These matrices are lexicographically ordered, and each ancestor
@@
 
+\noindent
 If \verb;n↑V; takes the first \verb;n; elements of \verb;V;, then we can obtain coordinate
@@
 The simplest operation over a node coordinate is to extract the depth
-of the node. The expression \verb;C⍳0; finds the first occurrence of \verb;0; in \verb;C;
-and returns the index of that occurrence. (The application of ⍳ here
-is its dyadic form, meaning that it receives two arguments. Previous
-uses of ⍳ have used its monadic form, which generates all indices,
-instead of finding a specific index. Most APL functions have a
-dyadic and monadic form.)  Thus, the depth of a node is given by
-the following:
+of the node.
+The dyadic expression \verb;C⍳0; finds the first occurrence of \verb;0;
+in \verb;C; and returns the index of that occurrence.
+Thus, we compute the depth of a node as follows:
 
@@
 
-The primary calculation we care about in order to do function lifting,
-expression flattening, and other sorts of flattening transformations
-is to determine which nodes are ancestors of another node. The core
-of this computation determines whether one node is a child of another
-based on their node coordinates. This amounts to computing whether
-one node is a prefix of another, ignoring zeros.
-
+\noindent
+In many compiler passes, we primarily care about which nodes are ancestors 
+of other nodes. To determine whether one node is a child of another, we compute 
+whether one node is a prefix of the other, ignoring zeros.
 We write \verb;(f g h); to represent the composition of functions \verb;f;, \verb;g;,
-and \verb;h; as a function train. This is used with another shorthand of
-composition according to the following equivalences:
+and \verb;h; as a function train.
+Function trains obey the following equivalences, where \verb;A; and \verb;B;
+are arrays.
 
@@
 
-With this, we can write a function to compute whether a given
-coordinate is a prefix of another.
+\noindent
+With this, we write a function to compute whether a given coordinate prefixes 
+another.
 
@@
 
-Here \verb;P; is the coordinate we wish to check against \verb;C;, to determine
-whether \verb;C; is a descendant. The functions are all standard logical
-functions extended to apply point-wise over arrays. The function
-\verb;⊢; always returns its right argument. The function \verb;(=∨0=⊢);
-reads intuitively as, ``equal or right element is zero.'' In this
-case to make this a predicate we combine this with the ``for all''
-reduction using \verb;∧⌿;. This particular pattern is a special case
-of inner product, which we can compute using \verb;f.g;. Thus, \verb;+.×; is a
-function that computes traditional matrix-matrix multiplication when
-applied to two matrices. However, we can also use it to compute the
-above reduction and application:
-
+\noindent
+The above determines whether \verb;P; is an ancestor to \verb;C;.
+The logical functions \verb;=;, \verb;∧;, and \verb;∨; are all extended 
+point-wise over arrays.
+The function \verb;⊢; returns its right argument.
+The function \verb;(=∨0=⊢); reads as, ``equal or a zero right argument.'' 
+We take this point-wise operation and reduce it with \verb;∧⌿; (For All).
+This pattern is a special case of inner product, written \verb;f.g;. 
+For example, \verb;+.×; is matrix multiplication. We transform our reduction and 
+boolean function to a single predicate with the inner product operator:
 \begin{verbatim}
@@
 
-This extends the reduction to allow us to use matrix values for
-\verb;C; and \verb;P; instead of vectors, and thus determine whether rows in \verb;C;
-are descendants of multiple candidates given by \verb;P;. This becomes
-important to using the prefix function. In our examples, we want
-to determine the nearest ancestor of a certain node type to which a
-node belongs. In the case of function lifting, we want to determine
-the nearest ancestor function, in the case of expression flattening,
-we want to determine the nearest ancestor expression node. We can
-extract the rows we care about for each example using \verb;⌷;, where \verb;i⌷M;
-returns the i-th row of \verb;M;. In the following expressions, we use \verb;0 2;
-rather than \verb;¯1; in \verb;⍤; to indicate that \verb;⌷; will receive the entire
-matrix on the right for each scalar index on the left.
-
-\begin{verbatim}
-      ⊢Fp←0 2⌷⍤0 2⊢Fc
+The use of inner product extends to cases where \verb;C; or \verb;P; are matrices.
+For our examples, we want to lift the functions and flatten expressions, so we 
+will select the \verb;F; nodes and \verb;E; nodes, respectively, as our 
+parent/ancestor nodes in \verb;Fp; and \verb;Ep;. Given a boolean vector 
+\verb;BV;, the expression \verb;BV⌿M; selects rows from \verb;M; based on the 
+non-zero elements of \verb;BV;.
+
+\begin{verbatim}
+      ⊢Fp←('F'=Ft)⌿Fc
 1 0 0 0 0
 1 1 1 0 0
-      ⊢Ep←0 1 6⌷⍤0 2⊢Ec
+      ⊢Ep←('E'=Et)⌿Ec
 1 0 0
@@
 
-The matrices \verb;Fp; and \verb;Ep; correspond to the function and expression
-node coordinates in our two examples, respectively. We could now
-use \verb;Fp; and \verb;Ep; to compare against \verb;Fc; and \verb;Ec; and determine which nodes
-contained which nodes in the tree. However, our prefix function will
-return 1 when we ask whether a node is a prefix of itself. In this
-case we don’t want that. The solution to this is to drop the last
-non-zero element from the coordinate. This will not prevent it from
-matching against any of its ancestors, but will prevent matching
-against itself. We can use the depth vector with the \verb;↑; function to
-take all but the last non-zero element. We need to remember to extend
-the returned array with an extra zero column, since the shape will
-be too small otherwise.
+\noindent
+We could use \verb;Fp; and \verb;Ep; to compare against \verb;Fc; and \verb;Ec;
+to determine which nodes belong to which parent in \verb;Fp; or \verb;Ep;, 
+but our prefix function returns 1 when \verb;C≡P;. 
+Instead, we drop the last non-zero element from each coordinate.
+This will permit matching against all ancestors, but not against itself.
+We use the depth vector and the Take (\verb;↑;) function to take all but
+the last non-zero element. We extend the resulting array with an extra 
+zero column to ensure we have a compatible shape.
 
@@
 
-We can now use \verb;Fcp; and \verb;Ecp; to determine which Function and Expression
-nodes match for each node:
+\noindent
+We use \verb;Fcp; and \verb;Ecp; to determine the \verb;F; and \verb;E; 
+ancestors for each node (\verb;⍉A; transposes \verb;A;):
 
@@
 
-We use the ⍉ function, which computes the transpose of its right
-argument, to ensure that the row and column sizes match up. With the
-above, we can determine the ``greatest'' match, which is the closest.
-We can replace each 1 in the above matrices by their column numbers
-(the number of the column in which that 1 occurs), and then we can
-take the maximum column number in each row to determine an index in
-either \verb;Fp; or \verb;Ep; that is the appropriate ``parent'' coordinate for
-that node. This works because we have made sure to order the \verb;Ep; and \verb;Fp;
-matrices lexicographically. The following shows this computation for
-the expression example; we elide the function example as redundant
-at this point. We also use \verb;⌈/; instead of \verb;⌈⌿; to reduce along
-the last, rather than the first, axis.
-
-\begin{verbatim}
+\noindent
+In the above, the closest ancestor is the rightmost 1 in each row.
+We can extract the column number of the rightmost 1 by replacing each 1 
+with its column number and selecting the maximum of each row with \verb;⌈/;. 
+For space we will show only the \verb;E; example.
+In the following, \verb;V×⍤1⊢M; gives a matrix where each row is given by 
+\verb;V×i⌷M; where \verb;i⌷M; is the $i$th row of \verb;M;.
+
+\begin{verbatim}
+      (⍳3)×⍤1⊢Ecp∧.(=∨0=⊢)⍉Ep
+0 0 0
+0 0 0
+0 1 0
+0 1 0
+0 1 0
+0 0 0
+0 0 0
+0 0 2
+0 0 2
+0 0 2
       ⊢Ei←⌈/(⍳3)×⍤1⊢Ecp∧.(=∨0=⊢)⍉Ep
 0 0 1 1 1 0 0 2 2 2
+\end{verbatim}
+
+\noindent
+We use these column numbers to index into \verb;Ep; to obtain an ancestor 
+matrix where the node coordinate of the closest ancestor is given for 
+each node, one per row. 
+The expression \verb;I⌷⍤0 2⊢M; gives a matrix, one row per index in \verb;I;, 
+consisting of rows of \verb;M; selected by row indices in \verb;I;.
+
+\begin{verbatim}
       ⊢Ek←Ei⌷⍤0 2⊢Ep 
@@
 
+\noindent
 At this point we have two values, \verb;Ek; and \verb;Fk;, which indicate the
@@
 \begin{verbatim}
-      Fk
+      Fi←⌈/(⍳2)×⍤1⊢Fcp∧.(=∨0=⊢)⍉Fp
+      ⊢Fk←Fi⌷⍤0 2⊢Fp
 1 0 0 0 0
@@
 
-We will use these keys to perform computation over the AST and
-accomplish our tasks of function lifting and expression flattening in
-the next section, but we make a final note here, that we can imagine
-many other operations which might be used throughout the compiler when
-we care about how nodes relate to each other. The use of operators
-like reduction, scan, inner and outer products allow us to obtain
-the necessary information from the node coordinate matrix. Sometimes
-these results might not be obvious, but the example of the ``belongs
-to'' relationship given above demonstrates a pattern that arises
-repeatedly throughout our data parallel compiler, and is therefore
-particularly useful to understand.
+\noindent
+We use these keys to compute over the AST, particularly for function lifting 
+and expression flattening as demonstrated in the next section. The repeating 
+pattern is to leverage node coordinates and array operations to do stackless 
+reasoning about inter-node relationships. This pattern occurs often in our 
+compiler and is particularly useful to enable straightforward data-parallel 
+transformations over an AST.
 
@@
 
-The Key operator (written \verb;⌸;), takes a single function
-that expects two arguments and returns a function which takes two
-arguments. The left argument is a set of keys, and the right argument
-is a set of corresponding elements associated with those keys. In our
-case, we provide matrices for both of these arguments, so each pair
-of rows corresponds to a key-value pair. A simple example of the key
-operation at work is to compute a histogram (\verb;≢; here can be thought
-of as ``tally'' or count):
-
-\begin{verbatim}
+The Key operater (written \verb;⌸;) is central to our strategy for compilation.
+It allows us to use the Node Coordinates to their full effect. 
+The expression \verb;K f⌸ M; groups rows of \verb;M; by their corresponding 
+keys in \verb;K; and computes \verb;k f m; for each unique key \verb;k; in 
+\verb;K; and the matrix \verb;m; of rows of \verb;M; with key \verb;k;. 
+The corresponding rows of matrices \verb;K; and \verb;M; form key-value pairs.
+A simple example of this at work is to compute a histogram using Tally (\verb;≢;):
+
+\begin{verbatim}
+      10⍴5
+5 5 5 5 5 5 5 5 5 5
        ⊢X←?10⍴5
 1 1 4 0 1 3 1 1 2 1
+      ≢⍳5
+5
       X(⊣,(≢⊢))⌸X
@@
 
-The following compiler passes are often found in any suitably complex 
-language being compiled, and represent a difficult problem for anyone attempting 
-to do data-parallel computation. Indeed, they were the primary motivating 
-factor for the development of the node coordinate matrix and its associated 
-techniques. They demonstrate how we can operate on highly nested structures, 
-usually with the intent of eliminating that structure to more closely map 
-the intended output language.
-
-These case studies come from the Co-dfns compiler, which is compiling a 
-lexically scoped language that contains both nested functions and 
-expressions.
+The following compiler passes are found in any compiler from a nested function, 
+nested expression language to a language without nested functions 
+or expressions. They were the chief motivating examples for developing the 
+techniques presented above as a part of developing the Co-dfns compiler, 
+and represent a challenging problem to data-parallel compilation without 
+the above techniques, but fall out almost effortlessly once the above patterns 
+are available.
 
@@
 
-If we use the key operator with \verb;Fk;, we get the following (the monadic
+If we use the Key operator with \verb;Fk;, we get the following (the monadic
 use of \verb;↓; converts \verb;Fc; from a matrix to a vector of vectors):
@@
 a single top level.
+
+This function lifting works even with lexically scoped functions, 
+as is the case for the Co-dfns compiler.
+We can do this because the node coordinates maintain the lexical information 
+for each variable reference, enabling us to construct the appropriate lexical 
+binding for each variable much later than would normally be done in a 
+compiler. 
+In other words, the above lifting operation loses no information that would 
+prevent us from handling lexical scoping, but decouples lexical scoping 
+from the act of lifting functions.
 
@@
 The J programming language \cite{hui2014key} was the first practical,
-general-purpose programming language to introduce the key operator as a
+general-purpose programming language to introduce the Key operator as a
 primitive operator with the presumption of its general usefulness. The
@@
 
+\begin{table}
+\centering
+\begin{tabular}{cll}
+\toprule
+Symbol                   & Monadic            & Dyadic \\
+\midrule
+\texttt{+}               & Identity           & Plus (Add) \\
+\texttt{-}               & Negative           & Minus (Subtract) \\
+\texttt{×}               & Direction (Signum) & Times (Multiply) \\
+\texttt{÷}               & Reciprocal         & Divide \\
+\texttt{|}               & Magnitude          & Residue (Modulo) \\
+\texttt{⌊}               & Floor              & Minimum \\
+\texttt{⌈}               & Ceiling            & Maximum \\
+\texttt{*}               & Exponential        & Power \\
+\texttt{⍟}               & Natural Logarithm  & Logarithm \\
+\texttt{○}               & Pi Times           & Circular (Trigonometric) \\
+\texttt{!}               & Factorial          & Binomial \\
+\texttt{\textasciitilde} & Not                & \\
+\texttt{?}               & Roll               & \\
+\texttt{∧}              &                    & And \\
+\texttt{∨}               &                    & Or \\
+\texttt{⍲}               &                    & Nand \\
+\texttt{⍱}               &                    & Nor \\
+\texttt{<}               &                    & Less \\
+\texttt{≤}               &                    & Less Or Equal \\
+\texttt{=}               &                    & Equal \\
+\texttt{≥}               &                    & Greater Or Equal \\
+\texttt{>}               &                    & Greater \\
+\texttt{≠}               &                    & Not Equal \\
+\end{tabular}
+\caption{Scalar Functions}
+\label{tab:scalarprims}
+\end{table}
+
+\begin{table}
+\centering
+\begin{tabular}{cll}
+\toprule
+Symbol     & Monadic       & Dyadic \\
+\midrule
+\texttt{⍴} & Reshape       & \\
+\texttt{,} & Ravel         & Catenate/Laminate \\
+\texttt{⍪} & Table         & Catenate First/Laminate \\
+\texttt{⌽} & Reverse       & Rotate \\
+\texttt{⊖} & Reverse First & Rotate First \\
+\texttt{⍉} & Transpose     & Transpose \\
+\texttt{↑} & Mix           & \\
+\texttt{↓} & Split         & \\
+\texttt{⊂} & Enclose       & Partitioned Enclose \\
+\texttt{∊} & Enlist        & \\
+\end{tabular}
+\caption{Structural Mixed Functions}
+\label{tab:structprims}
+\end{table}
+
+\begin{table}
+\centering
+\begin{tabular}{cll}
+\toprule
+Symbol                   & Monadic  & Dyadic \\
+\midrule
+\texttt{⊃}               & Disclose & Pick\\
+\texttt{↑}               &          & Take \\
+\texttt{↓}               &          & Drop \\
+\texttt{/}               &          & Replicate \\
+\texttt{⌿}               &          & Replicate First \\
+\texttt{\textbackslash}  &          & Expand \\
+\texttt{⍀}               &          & Expand First \\
+\texttt{\textasciitilde} &          & Without (Excluding) \\
+\texttt{∩}               &          & Intersection \\
+\texttt{∪}               & Unique   & Union \\
+\texttt{⊣}               & Same     & Left \\
+\texttt{⊢}               & Identity & Right\\
+\end{tabular}
+\caption{Selection Mixed Functions}
+\label{tab:selprims}
+\end{table}
+
+\begin{table}
+\centering
+\begin{tabular}{cll}
+\toprule
+Symbol     & Monadic         & Dyadic \\
+\midrule
+\texttt{⍳} & Index Generator & Index Of \\
+\texttt{∊} &                 & Membership \\
+\texttt{⍋} & Grade Up        & Grade Up \\
+\texttt{⍒} & Grade Down      & Grade Down \\
+\texttt{?} &                 & Deal \\
+\texttt{⍷} &                 & Find \\
+\end{tabular}
+\caption{Selector Mixed Functions}
+\label{tab:seltorprims}
+\end{table}
+
+\begin{table}
+\centering
+\begin{tabular}{cll}
+\toprule
+Symbol     & Monadic       & Dyadic \\
+\midrule
+\texttt{⍴} & Shape         & \\
+\texttt{≡} & Depth         & Match \\
+\texttt{≢} & Tally         & Not Match \\
+\texttt{⍎} & Execute       & Execute \\
+\texttt{⍕} & Format        & Format \\
+\texttt{⊥} &               & Decode (Base) \\
+\texttt{⊤} &               & Encode (Representation) \\
+\texttt{⌹} & Matrix Divide & Matrix Inverse \\
+\end{tabular}
+\caption{Miscellaneous Mixed Functions}
+\label{tab:miscprims}
+\end{table}
+
+\begin{table*}
+\centering
+\begin{tabular}{cll}
+\toprule
+Symbol     & Name & Description \\
+\midrule
+\texttt{⍨} & Commute & Swaps arguments or distributes
+ right argument to both sides \\
+\texttt{¨} & Each & Applies its operand point-wise over the left/right
+ arguments \\
+\texttt{/} & Reduce & Reduce along the last axis \\
+\texttt{⌿} & Reduce First & Reduce along the first axis \\
+\texttt{\textbackslash} & Scan & Scan along the last axis \\
+\texttt{⍀} & Scan First & Scan along the first axis \\
+\texttt{⌸} & Key & Apply operand once for each sub-array grouped by key \\
+\end{tabular}
+\caption{Primitive Monadic/Unary Operators, each takes a single left
+operand and describes a function operating over one or two arguments}
+\label{tab:adverbs}
+\end{table*}
+
+\begin{table*}
+\centering
+\begin{tabular}{cll}
+\toprule
+Symbol     & Name & Description \\
+\midrule
+\texttt{∘} & Compose & Composes two operands as in traditional mathematics \\
+\texttt{.} & Inner Product & Inner product operation, e.g.
+ \texttt{+.×} for matrix multiplication \\
+\texttt{∘.} & Outer Product & Cartesian product or ``function table'' \\
+\texttt{⍣} & Power & Iteration, Limited use only \\
+\texttt{⍤} & Rank & Apply a function along cells of an array \\
+\end{tabular}
+\caption{Primitive Dyadic/Binary Operators, each takes a left
+ and right operand and describes a function operating over one or two arguments}
+\label{tab:conjunctions}
+\end{table*}
+
 \end{document}

### 2016-04-07 2016-spring-array/2016-array-key.tex
Add in references to Co-dfns in other places

@@
 compiler, called Co-dfns, that compiles a lexically scoped, functionally 
-oriented dialect of APL with nested functions. The compiler targets both the 
+oriented dialect of APL with nested functions. 
+\cite{hsu2014co,hsu2015accelerating}
+The compiler targets both the
 GPU and the CPU, and its core compiler is implemented in this pure, restricted 
@@
 The Co-dfns compiler is an experimental commercial compiler funded by 
-Dyalog, Ltd. It is intended to provide increased scaling and performance
+Dyalog, Ltd. \cite{hsu2014co,hsu2015accelerating}
+It is intended to provide increased scaling and performance
 for APL programmers using the dfns syntax (lexically scoped, functionally 

### 2016-04-07 2016-spring-array/2016-array-key.tex
Updated with some comments by Jason Hemann

@@
 
-That is, the sample code given indented by 6 spaces, followed by the value 
+That is, we indent the expression by 6 spaces, followed by the value
 of the expression without indent. All expressions are evaluated right to 
-left and all function application is unified and infix; that is, a function 
+left and all function application is infix; that is, a function 
 application may apply a function (such as \verb;+;) to one or two arguments, 
@@
 \toprule
-Pass & Description & Core Patterns & Coordinate? \\
+Pass & Description & Core Traversal Pattern & Uses coordinates? \\
 \midrule
@@
 
-\section{Other Passes}
-
-Table \ref{tab:passes} contains a selection of passes from the
-Co-dfns compiler and breaks down their features based on how they
-traverse the tree and whether or not they use the Key operator or
-Node Coordinates. They are oriented with front-end passes higher in
-the table and passes further down the compiler chain at the bottom
-of the table. These passes give a good idea of how the Key operator
-plays a part in the complete compiler as well as how Node Coordinates
-help to deal with complex sub-tree selections.
-
-We make a few notes on the table and the analysis. The Key operator
-can be used with or without the node coordinates structure. In cases
-where the AST is sufficiently flat and arranged in a sufficiently
-convenient order, the grouping operations are very simple. In these
-cases, the Key operator amounts to a map operation over some selected
-sub-trees selected at a specific depth. These occur later in the
-compiler where the reader will note appearences of the Key operator
-without the use of Node Coordinates.
-
-Likewise, node coordinates find their place early on in the compiler
-where it is necessary to deal with the more highly nested AST. They
-are useful for identifying this information regardless of whether the
-Key operator is used to handle the grouping or not. Thus passes like
-``Drop Unreachable Code'' that remove whole sub-trees from the AST
-may use the Node Coordinates to do the grouping once the correct
-parent node has been identified.
-
-In short, the goal of the structure and design of the compiler passes
-given in table \ref{tab:passes} is to utilize Node Coordinates to
-remove as quickly as possible the nested structure of the AST into
-a flatter form that does not require complex analysis to traverse.
-
-\section{The Co-dfns Compiler}
+\section{The Co-dfns Compiler and Other Passes}
 
@@
 the architecture of the compiler, which is built in the style of a Nanopass 
+\cite{keep2013nanopass}
 compiler, with very small passes chained together through composition, with 
@@
 
+Table \ref{tab:passes} contains a selection of passes from the
+Co-dfns compiler and breaks down their features based on how they
+traverse the tree and whether or not they use the Key operator or
+Node Coordinates.
+The traversal pattern indicates the primary operator used to traverse the AST 
+and perform the transformation, most of which are primitives; the Amend pattern 
+is a ``replacement'' pattern where specific nodes are replaced in the AST with 
+other nodes, and can be thought of as a slightly modified filter and map.
+The table orders the passes with front-end passes higher in
+the table and passes further down the compiler chain at the bottom
+of the table. These passes give a good idea of how the Key operator
+plays a part in the complete compiler as well as how Node Coordinates
+help to deal with complex sub-tree selections.
+
+We make a few notes on the table and the analysis. The Key operator
+can be used with or without the node coordinates structure. In cases
+where the AST is sufficiently flat and arranged in a sufficiently
+convenient order, the grouping operations are very simple. In these
+cases, the Key operator amounts to a map operation over some selected
+sub-trees selected at a specific depth. These occur later in the
+compiler where the reader will note appearences of the Key operator
+without the use of Node Coordinates.
+
+Likewise, node coordinates find their place early on in the compiler
+where it is necessary to deal with the more highly nested AST. They
+are useful for identifying this information regardless of whether the
+Key operator is used to handle the grouping or not. Thus passes like
+``Drop Unreachable Code'' that remove whole sub-trees from the AST
+may use the Node Coordinates to do the grouping once the correct
+parent node has been identified.
+
+In short, the goal of the structure and design of the compiler passes
+given in Table \ref{tab:passes} is to utilize Node Coordinates to
+remove as quickly as possible the nested structure of the AST into
+a flatter form that does not require complex analysis to traverse.
+
 \section{Future Work}
 
-Work remains on expanding the features and capabilities of the compiler, 
-to demonstrate the scalability of these techniques on more sophisticated 
-optimizations. The current compiler represents a non-trivial instance of 
+The current compiler represents a non-trivial instance of
 the successful application of these techniques, but it focuses on an untyped, 
@@
 of the compiler is somewhat naive. More work is required to expand the 
-type inferencer to include more sophisticated type inference. Likewise, 
-the compiler does not currently handle error handling such as signals or 
-exceptions.
-
+type inferencer to include more sophisticated type inference. 
+The compiler does not currently handle errors such as signals or 
+exceptions, nor does it implement some more sophisticated compiler optimizations.
 Further work remains to implement these techniques on an imperative and 
@@
 
-The performance of these techniques has not been a problem in practice, but 
-we do not have a large scale understanding of the performance characteristics 
-of these techniques in real life. Further implementation and analysis is 
+We do not have a large scale understanding of the performance characteristics 
+of these techniques in real life outside of daily use of the compiler.
+Further analysis is
 required to understand what sort of compiler optimizations and analyses are 
 necessary to compile the Co-dfns compiler efficiently for the GPU and CPU. 
-We current lack a good comparison against existing techniques for the same 
+We currently lack a good comparison against existing techniques for the same 
 language, since the current compiler is relatively unique in that no other 

### 2016-04-11 2016-spring-array/2016-array-key.tex
Incorporate David's advice and some spelling corrections

@@
 or other forms of non-data parallel control flow. 
+The methods are general and independent of the language being compiled.
 
@@
 strategy for compiler construction that enables the development of parallel 
-by construction compilers that are high-level in their implementation and 
-exceptionally concise.
+by construction compilers that are high-level in their implementation, 
+exceptionally concise, and independent of the language being compiled.
 
@@
 
-The Key operater (written \verb;⌸;) is central to our strategy for compilation.
+First introduced in the J language \cite{hui2014key},
+the Key operator (written \verb;⌸;) is central to our strategy for compilation.
 It allows us to use the Node Coordinates to their full effect. 
@@
 The corresponding rows of matrices \verb;K; and \verb;M; form key-value pairs.
-A simple example of this at work is to compute a histogram using Tally (\verb;≢;):
+Let's compute a histogram using Tally (\verb;≢;) and Key (\verb;⌸;):
 
@@
 sub-trees selected at a specific depth. These occur later in the
-compiler where the reader will note appearences of the Key operator
+compiler where the reader will note appearances of the Key operator
 without the use of Node Coordinates.
@@
 primitive operator with the presumption of its general usefulness. The
-rank operator (⍤) used throughout this treatment also derives from
+rank operator (\verb;⍤;) used throughout this treatment also derives from
 the J traditional, receiving particular interest throughout the APL
@@
 Fritz Henglein demonstrated a class of operations, called
-discriminators, of which the Key operator is a member.  
+discriminators, of which the Key operator is a member  
 \cite{henglein2013dd}
-Namely,
+, namely,
 a discriminator performs the same grouping computation as Key, but
@@
 selected on the basis of inter-node properties through the use of the
-Key (⌸) operator and node coordinates, which enable local computation
+Key (\verb;⌸;) operator and node coordinates, which enable local computation
 of these inter-node properties. This method is both general and direct,
@@
 found in modern compilers today and replacing them with an entirely
-different paradigm centred on parallelism and aggregate operations.
+different paradigm centered on parallelism and aggregate operations.
 

### 2016-05-08 2016-spring-array/2016-array-key.tex
#1: Incorporate some fixes

@@
 necessary to compile the Co-dfns compiler efficiently for the GPU and CPU. 
-We currently lack a good comparison against existing techniques for the same 
-language, since the current compiler is relatively unique in that no other 
-compilers to our knowledge compile the same language to the same targets 
-with the same sorts of optimizations. We intend to implement other paradigms 
-and other languages which do have existing compiler implementations to compare 
-relative performance on multiple platforms.
+We currently lack a good comparison against existing techniques both for 
+the compiler design and the generated code. Some existing compilers are able 
+to compile variants of APL to target the GPU and the CPU, but there are subtle 
+differences that require care when comparing generated code, and the 
+architecture of these compilers is often quite different. Comparing the 
+relative performance both of the compiler designs (human factors and execution 
+speed) and generated code is important to understanding these techniques more 
+fully.
 
@@
 
+Iverson introduced the idea of the Rank operator while at Sharp Associates 
+\cite{iverson1983rationalized} as a part of ``Rationalized APL.'' 
+The rank operator (\verb;⍤;) used here derived from the J language, a 
+continuation of the rationalized APL idea, with many ideas such as rank 
+percolating back into APL implementations. \cite{bernecky1987rank,hui1995rank}
 The J programming language \cite{hui2014key} was the first practical,
 general-purpose programming language to introduce the Key operator as a
-primitive operator with the presumption of its general usefulness. The
-rank operator (\verb;⍤;) used throughout this treatment also derives from
-the J traditional, receiving particular interest throughout the APL
-community \cite{bernecky1987rank,hui1995rank}.
+primitive operator with the presumption of its general usefulness.
+
+Bernecky \cite{bernecky2015abstract} argues that the increased programmability 
+and desirable human factors of APL-style array programming can be implemented 
+with competitive performance relative to more traditional techniques. This 
+suggests that programs written in this highly abstract style may not suffer the 
+performance gap traditionally assumed to go with their desirable features. 
+This work is bolstered by previous work on the high-performance implemenation 
+of array-oriented languages. 
+\cite{ching1994experimental,ching1993primitive,
+ching1990automatic,ju1991exploitation,
+ju1991performance,bernecky1999reducing,schwarz1991acorn}
 
@@
 \cite{bernecky2003tokenizer}.
- It uses a SSA representation, and converts the dynamic scope of
-traditional APL functions into a static form early on. It also uses
+It also uses 
 a matrix format to represent the AST. Traditional APL did not have
@@
 
-Bernecky further identified methods of reducing or optimizing the
-computational complexity or cost of certain array operations, allowing
-improved performance of easy to understand array expressions
-\cite{bernecky1999reducing}.
-
 Timothy Budd implemented a compiler \cite{budd1984apl,budd2012apl} for APL which targeted
-vector processors as well as one for C code. They used a method of
-lazy evaluation to avoid intermediate data copying. Budd provided
+vector processors as well as C. Budd provided 
 thoughts and some ideas on how the compiler might be implemented in
 parallel as well.
-
-Walter Schwarz implemented an APL to C compiler for the ACORN system
-targeting the CM-2 machine, demonstrating performance potential for
-APL as a massively parallel language \cite{schwarz1991acorn}.
-
-W. Ching and D. Ju have spent significant work on the ELI language
-and other APL-class language
- implementations, especially on parallelized code and optimization. 
-\cite{ching2000design,ching1994experimental,ching1993primitive,
-ching1990automatic,hendriks1990sparse,ju1991exploitation,
-ju1991performance}
 

### 2016-05-08 2016-spring-array/2016-array-key.tex
#1: Fixes identified by ispell.
#1 More fixes, typos, &c.

@@
 compiler, called Co-dfns, that compiles a lexically scoped, functionally 
-oriented dialect of APL with nested functions. 
-\cite{hsu2014co,hsu2015accelerating}
+oriented dialect of APL with nested functions 
+\cite{hsu2014co,hsu2015accelerating}.
 The compiler targets both the
-GPU and the CPU, and its core compiler is implemented in this pure, restricted 
+GPU and the CPU, and its core is implemented in this pure, restricted 
 language that uses only function composition over data-parallel primitives.
@@
 
+\noindent
 That is, we indent the expression by 6 spaces, followed by the value
@@
 
+\noindent
 We introduce notation as needed throughout the exposition, so the reader 
@@
 
+\noindent
 The depth vectors for these trees we name \verb;Fd; and \verb;Ed;, respectively:
@@
 
+\noindent
 The node types we call \verb;Ft; and \verb;Et;:
@@
 
+\noindent
 And finally, we call the values vectors \verb;Fv; and \verb;Ev;:
@@
 
+\noindent
 We combine these to form the respective AST matrices. We write \verb;A,B;
@@
 
-The Depth Vector stores all edge information for the tree, but this information 
+\noindent
+The depth vector stores all edge information for the tree, but this information 
 requires non-local access to utilize, such as traversing potentially the 
@@
 For example, \verb;+.×; is matrix multiplication. We transform our reduction and 
-boolean function to a single predicate with the inner product operator:
+Boolean function to a single predicate with the inner product operator:
 \begin{verbatim}
@@
 Lift Functions & Moves all functions to the top-level. & Key & Yes \\
-Drop Redundant Nodes & Eliminates unnecessary nodes/nesting & Filter & No \\
+Drop Redundant Nodes & Eliminates unnecessary nodes/nesting. & Filter & No \\
 Flatten Expressions & Removes nesting from expressions. & Key & Yes \\
-Compress Atomic Nodes & Atomizes nested atom nodes & Amend & No \\
+Compress Atomic Nodes & Atomizes nested atom nodes. & Amend & No \\
 Propagate Constants & Inlines all references to literal values. &
  Amend/Rank & Yes \\
-Fold Constants & Converts constant expressions to literals & Amend & No \\
+Fold Constants & Converts constant expressions to literals. & Amend & No \\
 Compress Expressions & Converts expression sub-trees into single nodes. & 
@@
  Amend & No \\
-Lift Type-checking & Infers some type information at compile time &
+Lift Type-checking & Infers some type information at compile time. &
  Power Limit/Rank & No \\
-Allocate Value Slots & Does a form of frame allocation for variables &
+Allocate Value Slots & Does a form of frame allocation for variables. &
  Key & No \\
-Anchor Variables & Resolves lexically scoped variables &
+Anchor Variables & Resolves lexically scoped variables. &
  Key & Yes \\
@@
  execution. & Key & No \\
-Fuse Scalar Loops & Identifies Fusion opportunities and fuses expressions &
+Fuse Scalar Loops & Identifies Fusion opportunities and fuses expressions. &
  Key & No \\
@@
 
+\noindent
 To understand a bit better how the Key operator applies its function,
@@
 
+\noindent
 In our case, we use either \verb;Fk; or \verb;Ek; as our keys applied to the
@@
 body before lifting is to shift the depths in the depth vectors
-to correspond to those of a function lifted to the top-level, and
+to correspond to those of a function lifted to the top-level and
 to attach a function node to the top of each of the bodies. At that
@@
 problem. The compiler currently uses parsing combinators to do its parsing, 
-though we plan to implement the Two-by-Two parser given in the literature. 
+though we plan to implement the Two-by-Two parser given in the literature
+\cite{bunda1984apl}. 
 The code generator itself is a simple parallel map over each node, as 
@@
 functionally oriented language. The type inference in the current version 
-of the compiler is somewhat naive. More work is required to expand the 
+of the compiler is somewhat na\"ive. More work is required to expand the 
 type inferencer to include more sophisticated type inference. 
@@
 performance gap traditionally assumed to go with their desirable features. 
-This work is bolstered by previous work on the high-performance implemenation 
+This work is bolstered by previous work on the high-performance implementation 
 of array-oriented languages. 

### 2016-05-09 2016-spring-array/2016-array-key.tex
Fix tables for camera ready to fit on a single page.
More stuff for the camera ready

@@
-\documentclass[numbers,preprint]{sigplanconf}
+\documentclass[numbers,9pt]{sigplanconf}
 % \usepackage[utf8]{inputenc}
@@
 \usepackage{enumitem}      % adjust spacing in enums
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
+\usepackage[colorlinks=false,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
 % known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
@@
 
-\conferenceinfo{ARRAY '16}{June, 2016, Santa Barbara, CA, USA}
-\copyrightyear{2016}
+\toappear{}
 
@@
 \end{abstract}
+
+\category{D.3.4}{Programming Languages}{Processors}[Compilers]
+
+\keywords
+APL, Parallel, Compilers, Key, Node Coordinates
 
@@
 of both these algorithms and this notation in the large. 
-
 All code examples are given in the following form:
@@
 \begin{verbatim}
-      Fd,Ft,⍪Fv
-0 F f
-1 E 0
-2 F 0
-3 E 0
-4 V ⍵
-4 P +
-4 V ⍵
-2 A 0
-3 N 7
-      Ed,Et,⍪Ev
-0 E v
-1 E 0
-2 V a
-2 P +
-2 V b
-1 P ÷
-1 E 0
-2 V c
-2 P ×
-2 V d
+      Fd,Ft,⍪Fv  │        Ed,Et,⍪Ev
+0 F f            │  0 E v
+1 E 0            │  1 E 0
+2 F 0            │  2 V a
+3 E 0            │  2 P +
+4 V ⍵            │  2 V b
+4 P +            │  1 P ÷
+4 V ⍵            │  1 E 0
+2 A 0            │  2 V c
+3 N 7            │  2 P ×
+                 │  2 V d
 \end{verbatim}
@@
 
-\bibliographystyle{plain}
+\balancecolumns
+
+\bibliographystyle{abbrvnat}
 \bibliography{biblio}
 
-\begin{table}
+\begin{table*}
 \centering
-\begin{tabular}{cll}
+\begin{tabular}{cllcll}
 \toprule
+Symbol                   & Monadic            & Dyadic  &
 Symbol                   & Monadic            & Dyadic  \\
 \midrule
-\texttt{+}               & Identity           & Plus (Add) \\
-\texttt{-}               & Negative           & Minus (Subtract) \\
-\texttt{×}               & Direction (Signum) & Times (Multiply) \\
-\texttt{÷}               & Reciprocal         & Divide \\
-\texttt{|}               & Magnitude          & Residue (Modulo) \\
-\texttt{⌊}               & Floor              & Minimum \\
-\texttt{⌈}               & Ceiling            & Maximum \\
-\texttt{*}               & Exponential        & Power \\
-\texttt{⍟}               & Natural Logarithm  & Logarithm \\
-\texttt{○}               & Pi Times           & Circular (Trigonometric) \\
-\texttt{!}               & Factorial          & Binomial \\
+\multicolumn{6}{c}{Scalar Functions} \\ 
+\midrule
+\texttt{+}               & Identity           & Plus (Add) &
 \texttt{\textasciitilde} & Not                & \\
+\texttt{-}               & Negative           & Minus (Subtract) &
 \texttt{?}               & Roll               & \\
+\texttt{×}               & Direction (Signum) & Times (Multiply) &
 \texttt{∧}               &                    & And \\
+\texttt{÷}               & Reciprocal         & Divide &
 \texttt{∨}               &                    & Or \\
+\texttt{|}               & Magnitude          & Residue (Modulo) &
 \texttt{⍲}               &                    & Nand \\
+\texttt{⌊}               & Floor              & Minimum &
 \texttt{⍱}               &                    & Nor \\
+\texttt{⌈}               & Ceiling            & Maximum &
 \texttt{<}               &                    & Less \\
+\texttt{*}               & Exponential        & Power &
 \texttt{≤}               &                    & Less Or Equal \\
+\texttt{⍟}               & Natural Logarithm  & Logarithm &
 \texttt{=}               &                    & Equal \\
+\texttt{○}               & Pi Times           & Circular (Trigonometric) &
 \texttt{≥}               &                    & Greater Or Equal \\
+\texttt{!}               & Factorial          & Binomial &
 \texttt{>}               &                    & Greater \\
 \texttt{≠}               &                    & Not Equal \\
-\end{tabular}
-\caption{Scalar Functions}
-\label{tab:scalarprims}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{cll}
-\toprule
-Symbol     & Monadic       & Dyadic \\
 \midrule
+\multicolumn{3}{c}{Selection Mixed Functions} &
+\multicolumn{3}{c}{Structural Mixed Functions} \\
+\cmidrule(r){1-3} \cmidrule(l){4-6}
+\texttt{⊃}               & Disclose & Pick &
 \texttt{⍴} & Reshape       & \\
+\texttt{↑}               &          & Take &
 \texttt{,} & Ravel         & Catenate/Laminate \\
+\texttt{↓}               &          & Drop &
 \texttt{⍪} & Table         & Catenate First/Laminate \\
+\texttt{/}               &          & Replicate &
 \texttt{⌽} & Reverse       & Rotate \\
+\texttt{⌿}               &          & Replicate First &
 \texttt{⊖} & Reverse First & Rotate First \\
+\texttt{\textbackslash}  &          & Expand &
 \texttt{⍉} & Transpose     & Transpose \\
+\texttt{⍀}               &          & Expand First &
 \texttt{↑} & Mix           & \\
+\texttt{\textasciitilde} &          & Without (Excluding) &
 \texttt{↓} & Split         & \\
+\texttt{∩}               &          & Intersection &
 \texttt{⊂} & Enclose       & Partitioned Enclose \\
+\texttt{∪}               & Unique   & Union &
 \texttt{∊} & Enlist        & \\
-\end{tabular}
-\caption{Structural Mixed Functions}
-\label{tab:structprims}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{cll}
-\toprule
-Symbol                   & Monadic  & Dyadic \\
-\midrule
-\texttt{⊃}               & Disclose & Pick\\
-\texttt{↑}               &          & Take \\
-\texttt{↓}               &          & Drop \\
-\texttt{/}               &          & Replicate \\
-\texttt{⌿}               &          & Replicate First \\
-\texttt{\textbackslash}  &          & Expand \\
-\texttt{⍀}               &          & Expand First \\
-\texttt{\textasciitilde} &          & Without (Excluding) \\
-\texttt{∩}               &          & Intersection \\
-\texttt{∪}               & Unique   & Union \\
 \texttt{⊣}               & Same     & Left \\
 \texttt{⊢}               & Identity & Right\\
+\midrule
+\multicolumn{3}{c}{Selector Mixed Functions} &
+\multicolumn{3}{c}{Miscellaneous Mixed Functions} \\
+\cmidrule(r){1-3} \cmidrule(l){4-6}
+\texttt{⍳} & Index Generator & Index Of &
+\texttt{⍴} & Shape         & \\
+\texttt{∊} &                 & Membership &
+\texttt{≡} & Depth         & Match \\
+\texttt{⍋} & Grade Up        & Grade Up &
+\texttt{≢} & Tally         & Not Match \\
+\texttt{⍒} & Grade Down      & Grade Down &
+\texttt{⍎} & Execute       & Execute \\
+\texttt{?} &                 & Deal &
+\texttt{⍕} & Format        & Format \\
+\texttt{⍷} &                 & Find &
+\texttt{⊥} &               & Decode (Base) \\
+& & & \texttt{⊤} &               & Encode (Representation) \\
+& & & \texttt{⌹} & Matrix Divide & Matrix Inverse \\
 \end{tabular}
-\caption{Selection Mixed Functions}
-\label{tab:selprims}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{cll}
-\toprule
-Symbol     & Monadic         & Dyadic \\
-\midrule
-\texttt{⍳} & Index Generator & Index Of \\
-\texttt{∊} &                 & Membership \\
-\texttt{⍋} & Grade Up        & Grade Up \\
-\texttt{⍒} & Grade Down      & Grade Down \\
-\texttt{?} &                 & Deal \\
-\texttt{⍷} &                 & Find \\
-\end{tabular}
-\caption{Selector Mixed Functions}
-\label{tab:seltorprims}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{cll}
-\toprule
-Symbol     & Monadic       & Dyadic \\
-\midrule
-\texttt{⍴} & Shape         & \\
-\texttt{≡} & Depth         & Match \\
-\texttt{≢} & Tally         & Not Match \\
-\texttt{⍎} & Execute       & Execute \\
-\texttt{⍕} & Format        & Format \\
-\texttt{⊥} &               & Decode (Base) \\
-\texttt{⊤} &               & Encode (Representation) \\
-\texttt{⌹} & Matrix Divide & Matrix Inverse \\
-\end{tabular}
-\caption{Miscellaneous Mixed Functions}
-\label{tab:miscprims}
-\end{table}
+\caption{Primitive Functions}
+\label{tab:scalarprims}
+\end{table*}
 

### 2016-05-10 2016-spring-array/2016-array-key.tex
More fixes from Camera Ready Audit.
Copyright information

@@
 
-\toappear{}
-
-\authorinfo{Aaron W. Hsu}{Indiana University}{awhsu@indiana.edu}
+\toappear{Permission to make digital or hard copies of all or part of this
+work for personal or classroom use is granted without fee provided that copies
+are not made or distributed for profit or commercial advantage and that copies
+bear this notice and the full citation on the first page. Copyrights for
+components of this work owned by others than the author(s) must be honored.
+Abstracting with credit is permitted. To copy otherwise, or republish, to post
+on servers or to redistribute to lists, requires prior specific permission
+and/or a fee. Request permissions from Permissions@acm.org.
+
+\emph{ARRAY'16}, June 13-17 2016, Santa Barbara, CA, USA
+
+Copyright is held by the owner/author(s). Publication rights licensed to ACM.
+
+ACM 978-1-4503-4384-8/16/06…\$15.00
+
+DOI: http://dx.doi.org/10.1145/2935323.2935331}
+
+\authorinfo{Aaron W. Hsu}{Indiana University, USA}{awhsu@indiana.edu}
 
@@
 different paradigm centered on parallelism and aggregate operations.
-
 \balancecolumns

### 2016-11-04 2017-spring-pldi/benchmark.c
Basic CPU Flattening

@@ -1,0 +1,34 @@
+#include <stdio.h>
+#include <stdlib.h>
+typedef int I;typedef void V;typedef size_t B;typedef char C;
+typedef struct cn{B i;I t;B c;struct cn*k[];}*CN;
+#define R return
+#define P printf
+#define DO(n,x) for(I i=0;i<(n);i++){x;}
+B id=0;
+V pi(I i){P("%d",i);}V nl(V){P("\n");}
+V pp(B d,B w,C*s){P("Benchmarking %s algorithm (Depth: %d, Width: %d)...\n",s,d,w);}
+V help(V){P("benchmark <cpu|gpu> [<depth> <width>]\n");exit(1);}
+V merr(V){P("Out of Memory.\n");exit(2);}
+
+CN mcn(I i,I t,B c){CN n=malloc(sizeof(struct cn)+c*sizeof(CN));if(!n)merr();
+ n->i=i;n->t=t;n->c=c;R n;}
+CN mct(B d,B w){CN n;
+ if(d){n=mcn(id++,0,w);DO(w,n->k[i]=mct(d-1,w));}else n=mcn(id++,1,0);R n;}
+V fct(CN t){if(t->c){DO(t->c,fct(t->k[i]));free(t);}else free(t);}
+V pct(CN t,I d){DO(d,P(" "));P("%d %d\n",t->t,t->i);DO(t->c,pct(t->k[i],d+1));}
+B fltc1(CN t){B x=0;if(!t->t){x++;DO(t->c,x+=fltc1(t->k[i]));};R x;}
+B fltc2(CN r,CN t,B c){
+ if(!t->t){r->k[c++]=t;
+  DO(t->c,{CN k=t->k[i];t->k[i]=k->t?k:mcn(k->i,1,0);c=fltc2(r,k,c);})}
+ R c;}
+CN fltc(CN t){CN r=mcn(id++,0,fltc1(t));fltc2(r,t,0);R r;}
+V cpu(B d,B w){pp(d,w,"CPU");P("Before\n");CN t=mct(d,w);pct(t,1);
+ P("After\n");t=fltc(t);pct(t,1);fct(t);}
+
+V gpu(B d,B w){pp(d,w,"GPU");}
+
+I main(I c,C*v[]){B d=3,w=2;if(c!=2&&c!=4)help();
+ if(c==4){d=atoi(v[2]);w=atoi(v[3]);}
[…]

### 2016-11-04 2017-spring-pldi/benchmark.c
Make the presentation nicer and add timing

@@ -1,14 +1,18 @@
 #include <stdio.h>
 #include <stdlib.h>
-typedef int I;typedef void V;typedef size_t B;typedef char C;
-typedef struct cn{B i;I t;B c;struct cn*k[];}*CN;
+#include <time.h>
+typedef int I;typedef void V;typedef size_t B;typedef char C;typedef double D;
+typedef struct cn{B i;I t;B c;struct cn*k[];}*CN;typedef clock_t CT;
 #define R return
 #define P printf
+#define EQ(s1,s2) (!strcmp((s1),(s2)))
 #define DO(n,x) for(I i=0;i<(n);i++){x;}
+#define TIME(x) {CT s=clock();x;CT e=clock();\
+ P("Took %0.3f secs.\n",(D)(e-s)/CLOCKS_PER_SEC);}
 B id=0;
 V pi(I i){P("%d",i);}V nl(V){P("\n");}
 V pp(B d,B w,C*s){P("Benchmarking %s algorithm (Depth: %d, Width: %d)...\n",s,d,w);}
-V help(V){P("benchmark <cpu|gpu> [<depth> <width>]\n");exit(1);}
+V help(V){P("benchmark <cpu|gpu> <print|quiet> [<depth> <width>]\n");exit(1);}
 V merr(V){P("Out of Memory.\n");exit(2);}
 
 CN mcn(I i,I t,B c){CN n=malloc(sizeof(struct cn)+c*sizeof(CN));if(!n)merr();
@@ -23,12 +27,13 @@ B fltc2(CN r,CN t,B c){
   DO(t->c,{CN k=t->k[i];t->k[i]=k->t?k:mcn(k->i,1,0);c=fltc2(r,k,c);})}
  R c;}
 CN fltc(CN t){CN r=mcn(id++,0,fltc1(t));fltc2(r,t,0);R r;}
-V cpu(B d,B w){pp(d,w,"CPU");P("Before\n");CN t=mct(d,w);pct(t,1);
- P("After\n");t=fltc(t);pct(t,1);fct(t);}
+V cpu(B d,B w,I p){pp(d,w,"CPU");CN t=mct(d,w);if(p){P("Before\n");pct(t,1);}
+ TIME(t=fltc(t));if(p){P("After\n");pct(t,1);}fct(t);}
 
-V gpu(B d,B w){pp(d,w,"GPU");}
[…]

### 2016-11-13 2017-spring-pldi/benchmark.c
Remove unnecessary old benchmark code

@@ -1,65 +1,0 @@
-#include <math.h>
-#include <stdio.h>
-#include <stdlib.h>
-#include <time.h>
-typedef int I;typedef void V;typedef size_t B;typedef char C;typedef double D;
-typedef struct cn{B i;C t;B c;struct cn*k[];}*CN;typedef clock_t CT;
-typedef struct gn{B d;B w;B c;B*id;B*dv;C*tv;B*rf;}*GN;
-#define R return
-#define P printf
-#define EQ(s1,s2) (!strcmp((s1),(s2)))
-#define DO(n,x) for(I i=0;i<(n);i++){x;}
-#define TIME(x) {CT s=clock();x;CT e=clock();\
- P("Took %0.3f secs.\n",(D)(e-s)/CLOCKS_PER_SEC);}
-#define ga(c,T) ga_(c,sizeof(T))
-#define gf(b,c,T) gf_((C*)b,c,sizeof(T))
-B id=0;
-V pi(I i){P("%d",i);}V nl(V){P("\n");}
-V pp(B d,B w,C*s){P("Benchmarking %s algorithm (Depth: %d, Width: %d)...\n",s,d,w);}
-V help(V){P("benchmark <cpu|gpu> <print|quiet> [<depth> <width>]\n");exit(1);}
-V merr(V){P("Out of Memory.\n");exit(2);}
-V*ga_(B c,B sz){B cz=c*sz;C*b=malloc(cz);if(!b)merr();
- #pragma acc enter data create(b[:cz])
- R b;}
-V gf_(C*b,B c,B sz){B cz=c*sz;
- #pragma acc exit data delete(b[:cz])
- free(b);}
-
-CN mcn(I i,I t,B c){CN n=malloc(sizeof(struct cn)+c*sizeof(CN));if(!n)merr();
- n->i=i;n->t=t;n->c=c;R n;}
-CN mct(B d,B w){CN n;
- if(d){n=mcn(id++,0,w);DO(w,n->k[i]=mct(d-1,w));}else n=mcn(id++,1,0);R n;}
-V fct(CN t){if(t->c){DO(t->c,fct(t->k[i]));free(t);}else free(t);}
[…]

### 2016-11-13 2017-spring-pldi/Benchmark/Benchmark/main.cu
Almost working reduce_by_key version

@@ -10,11 +10,16 @@
 #include <thrust/device_vector.h>
 #include <thrust/for_each.h>
 #include <thrust/iterator/zip_iterator.h>
+#include <thrust/iterator/transform_iterator.h>
 #include <thrust/remove.h>
 #include <thrust/sequence.h>
 #include <thrust/inner_product.h>
 #include <thrust/functional.h>
 #include <thrust/sort.h>
+#include <thrust/fill.h>
+#include <thrust/execution_policy.h>
+#include <thrust/reduce.h>
+#include <thrust/tuple.h>
 
 using namespace std;
 
@@ -128,11 +133,10 @@ void benchmark_cpu(int depth, int width, bool print)
 struct GPUNode {
 	int depth;
 	int width;
-	int count;
-	thrust::host_vector<int> depths;
-	thrust::host_vector<long long> ids;
-	thrust::host_vector<short> types;
-	thrust::host_vector<vector<int>> coords;
+	long long count;
+	thrust::device_vector<int> depths;
+	thrust::device_vector<short> types;
+	thrust::device_vector<int> coords;
 	GPUNode(int depth, int width);
 };
 
@@ -141,13 +145,12 @@ GPUNode::GPUNode(int depth, int width)
 {
 	count = 0;
 
-	for (int i = 0; i <= depth; i++)
-		count += pow(width, i);
+	for (int i = 0; i < depth; i++)
+		count += (long long)pow(width, i);
 
-	depths = thrust::host_vector<int>(count);
-	ids = thrust::host_vector<long long>(count);
-	types = thrust::host_vector<short>(count);
-	coords = thrust::host_vector<vector<int>>(count);
+	thrust::host_vector<int> host_depths(count);
+	thrust::host_vector<short> host_types(count);
+	thrust::host_vector<int> host_coords(count * depth, 0);
 
 	int cur_width = 0;
 	vector<int> cur_coord(1, 0);
@@ -160,33 +163,48 @@ GPUNode::GPUNode(int depth, int width)
 			continue;
 		}
 
-		depths[i] = cur_coord.size() - 1;
-		types[i] = cur_coord.size() > depth ? 1 : 0;
+		host_depths[i] = (int)cur_coord.size() - 1;
+		host_types[i] = cur_coord.size() >= depth ? 1 : 0;
 		cur_coord.back()++;
-		coords[i] = cur_coord;
+		for (int j = 0; j < cur_coord.size(); j++)
[…]

### 2016-11-15 2017-spring-pldi/Benchmark/Benchmark/main.cu
Working Benchmarks

@@ -1,4 +1,4 @@
-#include <algorithm>
+﻿#include <algorithm>
 #include <chrono>
 #include <cmath>
 #include <ctime>
@@ -20,6 +20,8 @@
 #include <thrust/execution_policy.h>
 #include <thrust/reduce.h>
 #include <thrust/tuple.h>
+#include <thrust/equal.h>
+#include <thrust/transform.h>
 
 using namespace std;
 
@@ -74,34 +76,36 @@ void print_cpu_tree(CPUNode& node, int depth)
 
 void help(void) 
 {
-	cout << "Benchmark <cpu|gpu> <print|quiet> [<depth> <width>]" << endl;
+	cout << "Benchmark <fnc|mut|gpu> <print|quiet> [<depth> <width>|<depth_start> <depth_end> <width_start> <width_end>]" << endl;
 }
 
-void cpu_flatten_helper(CPUNode& node, vector<CPUNode>& lifted)
+CPUNode functional_flatten(CPUNode& node)
 {
+	vector<CPUNode> nodes;
+
 	if (!node.type) {
-		for_each(node.kids.rbegin(), node.kids.rend(), [&lifted](CPUNode& kid) {
-			cpu_flatten_helper(kid, lifted);
-		});
+		vector<CPUNode> kids(node.kids.size());
 
-		lifted.push_back(node);
-		node.type = 1;
-		node.kids = vector<CPUNode>();
+		for (int i = 0; i < node.kids.size(); i++) {
+			kids[i].id = node.kids[i].id;
+			kids[i].type = 1;
+			kids[i].kids = vector<CPUNode>();
+		}
+
+		nodes.push_back(CPUNode(node.id, node.type, kids));
+
+		for (int i = 0; i < node.kids.size(); i++) {
+			auto f = functional_flatten(node.kids[i]);
+			for (auto k : f.kids) nodes.push_back(k);
+		}
 	}
+
+	return CPUNode(-1, 0, nodes);
 }
 
-void cpu_flatten(CPUNode& node)
+void benchmark_functional(int depth, int width, bool print)
 {
-	vector<CPUNode> lifted;
-
-	cpu_flatten_helper(node, lifted);
-
-	node = CPUNode{ -1, 0, lifted };
-}
-
-void benchmark_cpu(int depth, int width, bool print)
-{
-	cout << "Benchmarking CPU algorithm (Depth: " << depth << " Width: " << width << ")" << endl;
[…]

### 2016-11-15 2017-spring-pldi/hsu-2017-key_data_parallel_compiler.tex
We have to use LaTeX I guess.

--- 2016-05-09 2016-spring-array/2016-array-key.tex
@@
-\documentclass[numbers,9pt]{sigplanconf}
-% \usepackage[utf8]{inputenc}
+\documentclass[numbers,10pt,preprint]{sigplanconf}
 \usepackage{array}
 \usepackage{fontspec}
-\setmainfont[Ligatures=TeX]{STIX}
+\setmainfont[Ligatures=TeX]{Times New Roman}
 \setmonofont{APL385 Unicode}
@@
 
-\title{The Key to a Data Parallel Compiler}
-
-\toappear{}
-
-\authorinfo{Aaron W. Hsu}{Indiana University}{awhsu@indiana.edu}
-
+\title{The Key to a Performant Data Parallel Compiler}
+\authorinfo{Aaron W. Hsu}{Indiana University, USA}{awhsu@indiana.edu}
 \maketitle
+
 \begin{abstract}
-We present a language-driven strategy for the construction of 
-compilers that are inherently data-parallel in their design and 
-implementation. Using an encoding of the inter-node relationships 
-between nodes in an AST called a Node Coordinate Matrix, we demonstrate how 
-an operator called the Key operator, that applies a function over groupings 
-of array cells grouped and ordered by their keys, when used in conjunction 
-with a Node Coordinate Matrix, permits arbitrary computation over sub-trees 
-of an AST using purely data-parallel array programming techniques. We discuss 
-the use of these techniques when applied to an experimental commercial 
-compiler called Co-dfns, which is a fully data-parallel compiler developed 
-using the techniques discussed here.
+We present a language-driven strategy for the construction of compilers that are inherently data-parallel in their design and implementation.
+Using an encoding of the inter-node relationships between nodes in an AST called a Node Co-ordinate Matrix, we demonstrate how an operator called the Key operator, that applies a function over groupings of array cells grouped and ordered by their keys, when used in conjunction with a Node Coordinate Matrix, permits arbi-trary computation over sub-trees of an AST using purely data-parallel array programming techniques.
+We discuss the use of these techniques when applied to an experi-mental commercial compiler called Co-dfns, which is a fully data-parallel compiler developed using the techniques discussed here.
 \end{abstract}
@@
 \category{D.3.4}{Programming Languages}{Processors}[Compilers]
-
-\keywords
-APL, Parallel, Compilers, Key, Node Coordinates
-
-% Proposed new outline:
-% 
-% I. 	Generalization on manipulations over Sub-tree computation
-%   A.	Key + Encoding → Data parallel, Array-based computation
-%   B.	Easy to write/read
-%   C.	Uniform across passes
-% II.	Case Studies
-%   A.	Function Lifting
-%   B.	Expression Flattening
-%   C.	Lexical Scope Resolution
-%   D.	Illustrates the problem space
-% III.	Other passes/Issues
-%   A.	Discussion of other passes
-%   B.	Beyond the Key Operator
-% IV.	In practice/Co-dfns Compiler
-% V.	Performance/Results
-% 
-% Questions that must be addressed:
-% 
-% 1. Why the formalization in APL?
-% 2. Why parallelize a compiler?
-% 3. How well does it work/scale/generalize
+\terms Algorithms, Performance, Design, Languages
+\keywords APL, Parallel, Compilers, Key, Node Coordinates
 
@@
 
-Compilers represent a peculiar class of tree/graph algorithms that greatly 
-alter the underlying structure of the input graph into something very 
-different in structure, but equivalent by some measure of interpretation. 
-These algorithms are widely applicable, but have stubbornly resisted a 
-general approach to parallelization. Most compilers are single-threaded 
-or make very limited use of parallelism in an ad hoc way. The design and 
-analysis of compilers almost universally deals with compilers as recursive 
-traversals over ASTs. Such formulations usually include heavy reliance on 
-branching, recursion, and sometimes very sophisticated control flow in 
-practice. All of this contributes to make it difficult to effectively 
-parallelize such algorithms onto architectures that are sensitive to 
-branching and recursion, such as GPGPUs or highly vectorized CPUs.
-
-There has been some success in efficiently executing parts of a compiler 
-on the GPU, such as the parser \cite{bunda1984apl}, 
-tokenization \cite{bernecky2003tokenizer}, and certain compiler 
-analyses \cite{prabhu2011eigencfa,mendez2012inclusion}. However, a generalized 
-framework or strategy for parallel compilation remains elusive. One of 
-the major difficulties is the recursive and highly branch dependent nature 
-of the algorithms. These present challenges that must be overcome in 
-vector-centric architectures.
-
-We present a language-driven strategy for creating an inherently parallel 
-compiler. We select a set of well known data-parallel primitives, and then 
-restrict program construction to the composition of these operations into 
-functions, without any forms of non-linear control flow, such as branching, 
-recursion, or pattern matching. Programs constructed in such a style are 
-data-parallel by construction. A key problem to the construction of compilers 
-in such a restricted environment is dealing with the inherently recursive and 
-nested structure of the AST. By working over a linearized representation of 
-the AST as a matrix, combined with a structure we call a Node Coordinate 
-Matrix, and using data-parallel primitives, we are able to tame the recursive 
-and nested nature of the AST into something that can be processed handily 
-using only data-parallel, data-flow programs, without requiring branching 
-or other forms of non-data parallel control flow. 
+Compilers represent a peculiar class of tree/graph algo-rithms that greatly alter the underlying structure of the input graph into something very different in structure, but equivalent by some measure of interpretation.
+These algo-rithms are widely applicable, but have stubbornly resisted a general approach to parallelization.
+Most compilers are single-threaded or make very limited use of parallelism in an ad hoc way.
+The design and analysis of compilers al-most universally deals with compilers as recursive travers-als over ASTs.
+Such formulations usually include heavy reliance on branching, recursion, and sometimes very so-phisticated control flow in practice.
+All of this contributes to make it difficult to effectively parallelize such algo-rithms onto architectures that are sensitive to branching and recursion, such as GPGPUs or highly vectorised CPUs.
+
+There has been some success in efficiently executing parts of a compiler on the GPU, such as the parser \cite{bunda1984apl}, tokenization \cite{bernecky2003tokenizer}, and certain compiler analyses \cite{prabhu2011eigencfa,mendez2012inclusion}.
+However, a generalized framework or strategy for parallel compilation remains elusive.
+One of the major difficulties is the recursive and highly branch dependent nature of the algorithms.
+These present challenges that must be overcome in vector-centric architectures.
+
+We present a language-driven strategy for creating an inherently parallel compiler.
+We select a set of well known data-parallel primitives, and then restrict program construction to the composition of these operations into functions, without any forms of non-linear control flow, such as branching, recursion, or pattern matching.
+Programs constructed in such a style are data-parallel by construction.
+A key problem to the construction of compilers in such a restricted environment is dealing with the inherently recursive and nested structure of the AST.
+By working over a linearized representation of the AST as a matrix, combined with a structure we call a Node Coordinate Matrix, and using data-parallel primitives, we are able to tame the recursive and nested nature of the AST into something that can be processed handily using only data-parallel, data-flow programs, without requiring branching or other forms of non-data parallel control flow.
 The methods are general and independent of the language being compiled.
 
-In the following sections we describe the core methods and idioms surrounding 
-the AST encoding, construction of the Node Coordinate Matrix, and the use 
-of these structures in handling nested, recursive AST relationships. This 
-enables us to perform arbitrary computation over arbitrary sub-trees of an 
-AST selected by parent-child relationships in the manner often seen in 
-compiler transformations. These ideas have been further implemented and tested 
-through the implementation of a complete data-parallel by construction 
-compiler, called Co-dfns, that compiles a lexically scoped, functionally 
-oriented dialect of APL with nested functions 
-\cite{hsu2014co,hsu2015accelerating}.
-The compiler targets both the
-GPU and the CPU, and its core is implemented in this pure, restricted 
-language that uses only function composition over data-parallel primitives.
+In the following sections we describe the core methods and idioms surrounding the AST encoding, construction of the Node Coordinate Matrix, and the use of these structures in handling nested, recursive AST relationships.
+This enables us to perform arbitrary computation over arbitrary sub-trees of an AST selected by parent-child relationships in the manner often seen in compiler transformations.
+These ideas have been further implemented and tested through the implementation of a complete data-parallel by construction compiler, called Co-dfns, that compiles a lexically scoped, functionally oriented dialect of APL with nested functions \cite{hsu2014co,hsu2015accelerating}.
+The compiler targets both the GPU and the CPU, and its core is implemented in this pure, restricted language that uses only function composition over data-parallel primitives.
 
@@
 
-\item We describe a method of computing over arbitrary sub-trees
-selected by their parent-child relationships in a data-parallel manner
-using the Key operator and Node Coordinates.
-
-\item We situation this technique into a broader, language-driven 
-strategy for compiler construction that enables the development of parallel 
-by construction compilers that are high-level in their implementation, 
-exceptionally concise, and independent of the language being compiled.
-
-\item We provide analysis of these techniques as used in a commercial
-compiler project called Co-dfns and report on the results,
-including the overall architecture of the compiler, the uses of these 
-techniques within the compiler, and the demonstration of these techniques 
-applied to two specific compiler passes.
+\item We describe a method of computing over arbitrary sub-trees selected by their parent-child relationships in a data-parallel manner using the Key operator and Node Coordinates.
+
+\item We situation this technique into a broader, language-driven strategy for compiler construction that enables the development of parallel by construction compilers that are high-level in their implementation, exceptionally concise, and independent of the language being compiled.
+
+\item We provide analysis of these techniques as used in a commercial compiler project called Co-dfns and report on the results, including the overall architecture of the compiler, the uses of these techniques within the compiler, and the demonstration of these techniques applied to two specific compiler passes.
 
@@
 
-For space and convenience, we use a concise array notational convention to 
-describe our techniques and approach. The notation is executable and well 
-established in the array community (it is a limited subset of APL). This is 
-in fact the same notation used within the Co-dfns compiler itself, whose 
-source code is available online and provides a complete example of the use 
-of both these algorithms and this notation in the large. 
+For space and convenience, we use a concise array notational convention to describe our techniques and approach.
+The notation is executable and well established in the array community (it is a limited subset of APL).
+This is in fact the same notation used within the Co-dfns compiler itself, whose source code is available online and provides a complete example of the use of both these algorithms and this notation in the large.
 All code examples are given in the following form:
@@
 
-\noindent
-That is, we indent the expression by 6 spaces, followed by the value
-of the expression without indent. All expressions are evaluated right to 
-left and all function application is infix; that is, a function 
-application may apply a function (such as \verb;+;) to one or two arguments, 
-called monadic and dyadic application respectively,
-which must appear on the right and (optionally) on the left of the function 
-name, as in the above example. Tables \ref{tab:scalarprims} to 
-\ref{tab:conjunctions}
-provide a listing of all of the primitives used in the Co-dfns compiler, 
-a selection of which are used here.
-
-We separate out higher-order functions called operators, from functions that 
-operate over arrays that we call simply functions. Operators take a function 
-argument or two, and bind more strongly to the left than the right. An 
-operator may take one or two operands, either on the left, or on the 
-right and left, and will return a function. In the following 
-example, we apply the reduction operator (\verb;/;) to a function created using the 
-composition operator (\verb;∘;), as an example:
+\noindent That is, we indent the expression by 6 spaces, followed by the value of the expression without indent.
+All expressions are evaluated right to left and all function application is infix; that is, a function application may apply a function (such as \verb;+;) to one or two arguments, called monadic and dyadic application respectively, which must appear on the right and (optionally) on the left of the function name, as in the above example.
+Tables \ref{tab:scalarprims} to \ref{tab:conjunctions} provide a listing of all of the primitives used in the Co-dfns compiler, a selection of which are used here.
+
+We separate out higher-order functions called operators, from functions that operate over arrays that we call simply functions.
+Operators take a function argument or two, and bind more strongly to the left than the right.
+An operator may take one or two operands, either on the left, or on the right and left, and will return a function.
+In the following example, we apply the reduction operator (\verb;/;) to a function created using the composition operator (\verb;∘;), as an example:
 
@@
 
-\noindent
-We introduce notation as needed throughout the exposition, so the reader 
-is encouraged to refer to the tables as necessary to recall particular 
-operations.
+\noindent We introduce notation as needed throughout the exposition, so the reader is encouraged to refer to the tables as necessary to recall particular operations.
 
@@
 
-Computing over sub-trees normally involves recursing down the structure of 
-the AST, identifying the root nodes of each sub-tree, and then dispatching
-to a handler for that sub-tree, which will continue the recursion at that 
-point and return the modified tree, which will replace the previous 
-sub-tree. More complicated transformations may involved moving the root 
-nodes around in the tree or other large, distant structural modifications.
-
-We divide the work of sub-tree computation into three basic phases:  
-we maintain a node coordinate with each node that uniquely identifies it 
-relative to others, we use these coordinates to select and group nodes for 
-work as a single sub-tree, and then we operate over these groupings using 
-the Key operator (\verb;⌸;).
+Computing over sub-trees normally involves recursing down the structure of the AST, identifying the root nodes of each sub-tree, and then dispatching to a handler for that sub-tree, which will continue the recursion at that point and return the modified tree, which will replace the previous sub-tree.
+More complicated transformations may involved moving the root nodes around in the tree or other large, distant structural modifications.
+
+We divide the work of sub-tree computation into three basic phases: we maintain a node coordinate with each node that uniquely identifies it relative to others, we use these coordinates to select and group nodes for work as a single sub-tree, and then we operate over these groupings using the Key operator (\verb;⌸;).
 
@@
 We represent the AST as a 3 column matrix with one row per node in the tree.
-The first column contains the inter-node
-relationships in the form of a depth vector. The second column is a
-vector of the node types, while the third contains the ``value''
-of the node, such as the name in a variable.
-
-We will use the following two running examples throughout. The \verb;F; 
-tree is an example nested function, while the \verb;E; tree is an example 
-nested expression.
+The first column contains the inter-node relationships in the form of a depth vector.
+The second column is a vector of the node types, while the third contains the ``value'' of the node, such as the name in a variable.
+
+We will use the following two running examples throughout. The \verb;F; tree is an example nested function, while the \verb;E; tree is an example nested expression.
 
@@
 
-\noindent
-We combine these to form the respective AST matrices. We write \verb;A,B;
-to catenate arrays \verb;A; and \verb;B; along their last axes and \verb;⍪A; to turn a
-vector into a 1-column matrix. Thus, the two AST matrices are given
-by the following expressions:
+\noindent We combine these to form the respective AST matrices.
+We write \verb;A,B; to catenate arrays \verb;A; and \verb;B; along their last axes and \verb;⍪A; to turn a vector into a 1-column matrix.
+Thus, the two AST matrices are given by the following expressions:
 
@@
 
-\noindent
-The depth vector stores all edge information for the tree, but this information 
-requires non-local access to utilize, such as traversing potentially the 
-entire vector to determine the children of a specific node.
+\noindent The depth vector stores all edge information for the tree, but this information requires non-local access to utilize, such as traversing potentially the entire vector to determine the children of a specific node.
 Node Coordinates fix this issue.
@@
 
-Every node in an AST has a Node Coordinate, named because a coordinate
-is a precise location in a space.  We can imagine all the nodes
-arranged inside some multi-dimensional space, leading to a specific
-set of node coordinates. Many such arrangements exist, of varying
-usefulness. In our case, any arrangement should allow us to answer
-the following questions about any two arbitrary nodes in a tree:
+Every node in an AST has a Node Coordinate, named because a coordinate is a precise location in a space.
+We can imagine all the nodes arranged inside some multi-dimensional space, leading to a specific set of node coordinates.
+Many such arrangements exist, of varying usefulness.
+In our case, any arrangement should allow us to answer the following questions about any two arbitrary nodes in a tree:
 
@@
 
-\noindent
-In short, we care about the relative position of each node in a tree
-relative to any other. We define a node coordinate as follows.
-
-A node coordinate is a vector whose length is the depth of the
-tree. Its elements are natural numbers.  The count of non-zero elements
-in the vector is equal to the depth of that node in the tree. All zero
-elements appear after the non-zero elements. That is, a coordinate
-is zero-padded on the right. When ordered lexicographically, the
-nodes for each coordinate appear in order according to a depth-first
-pre-order traversal of the tree. Each coordinate uniquely identifies a
-single node. Every ancestor’s coordinate is a prefix of any child’s
-coordinate, ignoring zeros. From the above it follows that every node
-is lexicographically greater than its left sibling and differs from
-it by exactly one non-zero element, and that this element is the last
-non-zero element in the coordinate.
+\noindent In short, we care about the relative position of each node in a tree relative to any other. We define a node coordinate as follows.
+
+A node coordinate is a vector whose length is the depth of the tree.
+Its elements are natural numbers.
+The count of non-zero elements in the vector is equal to the depth of that node in the tree.
+All zero elements appear after the non-zero elements.
+That is, a coordinate is zero-padded on the right.
+When ordered lexicographically, the nodes for each coordinate appear in order according to a depth-first pre-order traversal of the tree.
+Each coordinate uniquely identifies a single node.
+Every ancestor’s coordinate is a prefix of any child’s coordinate, ignoring zeros.
+From the above it follows that every node is lexicographically greater than its left sibling and differs from it by exactly one non-zero element, and that this element is the last non-zero element in the coordinate.
 
@@
 
-We construct the Node Coordinate Matrix from the depth vector, thereby encoding 
-the depth vector into a more useful format. The matrix has \verb;N; rows and 
-\verb;D; columns, where \verb;N; is the node count and \verb;D; is the depth 
-of the tree.
-
-We write \verb;f⌿A; to reduce the first axis of \verb;A; using function \verb;f;. Thus,
-\verb;+⌿V; is the sum of the elements in vector \verb;V;. The function \verb;x⌈y;
-gives the maximum of its two arguments.
+We construct the Node Coordinate Matrix from the depth vector, thereby encoding the depth vector into a more useful format.
+The matrix has \verb;N; rows and \verb;D; columns, where \verb;N; is the node count and \verb;D; is the depth of the tree.
+
+We write \verb;f⌿A; to reduce the first axis of \verb;A; using function \verb;f;.
+Thus, \verb;+⌿V; is the sum of the elements in vector \verb;V;.
+The function \verb;x⌈y; gives the maximum of its two arguments.
 We compute the depth of each tree as follows:
@@
 \begin{verbatim}
-      1+⌈⌿Fd
-5
-      1+⌈⌿Ed
-3
-\end{verbatim}
-
-\noindent
-We can obtain the ordered sequence  by writing \verb;⍳n;:
+      1+⌈⌿Fd  │        1+⌈⌿Ed
+5             │  3
+\end{verbatim}
+
+\noindent We can obtain the ordered sequence  by writing \verb;⍳n;:
 
@@
 \begin{verbatim}
-      ⍳1+⌈⌿Fd
-0 1 2 3 4 5
-      ⍳1+⌈⌿Ed
-0 1 2
-\end{verbatim}
-
-\noindent
-The function table or outer product of \verb;f; over vectors \verb;U; and \verb;V; is
-written \verb;U ∘.f V; giving a \verb;U V; shaped matrix as a result. Thus,
-\verb;(⍳3)∘.×⍳3; gives a small multiplication table:
-
-\begin{verbatim}
-      (⍳3)∘.×⍳3
-0 0 0
-0 1 2
-0 2 4
-\end{verbatim}
-
-\noindent
-If we use \verb;=; instead, we have a Boolean identity matrix:
+      ⍳1+⌈⌿Fd │        ⍳1+⌈⌿Ed
+0 1 2 3 4 5   │  0 1 2
+\end{verbatim}
+
+\noindent The function table or outer product of \verb;f; over vectors \verb;U; and \verb;V; is written \verb;U ∘.f V; giving a \verb;U V; shaped matrix as a result.
+Thus, \verb;(⍳3)∘.×⍳3; gives a small multiplication table:
+
+\begin{verbatim}
+      (⍳3)∘.×⍳3  │        ⍳3
+0 0 0            │  0 1 2
+0 1 2            │        ⍴⍳3
+0 2 4            │  3
+\end{verbatim}
+
+\noindent If we use \verb;=; instead, we have a Boolean identity matrix:
 
@@
 
-\noindent
-If we use \verb;∘.=; on the depth vector and its set of depths instead,
-we see an expanded Boolean representation of the depth vector:
-
-\begin{verbatim}
-      Fd∘.=⍳1+⌈⌿Fd
-1 0 0 0 0
-0 1 0 0 0
-0 0 1 0 0
-0 0 0 1 0
-0 0 0 0 1
-0 0 0 0 1
-0 0 0 0 1
-0 0 1 0 0
-0 0 0 1 0
-      Ed∘.=⍳1+⌈⌿Ed
-1 0 0
-0 1 0
-0 0 1
-0 0 1
-0 0 1
-0 1 0
-0 1 0
-0 0 1
-0 0 1
-0 0 1
+\noindent If we use \verb;∘.=; on the depth vector and its set of depths instead, we see an expanded Boolean representation of the depth vector:
+
+\begin{verbatim}
+      Fd∘.=⍳1+⌈⌿Fd │      Ed∘.=⍳1+⌈⌿Ed
+1 0 0 0 0          │ 1 0 0
+0 1 0 0 0          │ 0 1 0
+0 0 1 0 0          │ 0 0 1
+0 0 0 1 0          │ 0 0 1
+0 0 0 0 1          │ 0 0 1
+0 0 0 0 1          │ 0 1 0
+0 0 0 0 1          │ 0 1 0
+0 0 1 0 0          │ 0 0 1
+0 0 0 1 0          │ 0 0 1
+                   │ 0 0 1
 \end{verbatim}

### 2016-11-15 2017-spring-pldi/hsu-2017-key_data_parallel_compiler.tex
Conversion of existing work into the LaTeX format

@@
 
-\noindent
-These matrices let us see the nesting features of each tree more
-visually, but also suggest another step. We can compute a sum scan
-with \verb;+⍀;, also called a prefix sum, along the first axis. Applying
-this function on the above matrices leads to an interesting result:
-
-\begin{verbatim}
-      +⍀Fd∘.=⍳1+⌈⌿Fd
-1 0 0 0 0
-1 1 0 0 0
-1 1 1 0 0
-1 1 1 1 0
-1 1 1 1 1
-1 1 1 1 2
-1 1 1 1 3
-1 1 2 1 3
-1 1 2 2 3
-      +⍀Ed∘.=⍳1+⌈⌿Ed
-1 0 0
-1 1 0
-1 1 1
-1 1 2
-1 1 3
-1 2 3
-1 3 3
-1 3 4
-1 3 5
-1 3 6
-\end{verbatim}
-
-\noindent
-These matrices are lexicographically ordered, and each ancestor
-shares a common prefix with its descendants. They are also unique
-coordinates. Only the spurious digits at the end of each coordinate
-prevent these matrices from meeting all our requirements for valid
-node coordinates.
-
-The expression \verb;V f⍤¯1⊢M; applies \verb;f; to corresponding elements of
-\verb;V; and rows of \verb;M;:
-
-\begin{verbatim}
-      3 3⍴⍳9
-0 1 2
-3 4 5
-6 7 8
-      (⍳3)+⍤¯1⊢3 3⍴⍳9
-0 1  2
-4 5  6
-8 9 10
-\end{verbatim}
-
-\noindent
-If \verb;n↑V; takes the first \verb;n; elements of \verb;V;, then we can obtain coordinate
-matrices from the prefix sums by noting that the spurious digits
-all come after column \verb;d+1; where \verb;d; is the depth corresponding to that
-coordinate. The following gives a complete expression for computing
-a node coordinate matrix from a depth vector, shown using \verb;Fd; and \verb;Fe;:
-
-\begin{verbatim}
-      ⊢Fc←(1+Fd)↑⍤¯1⊢+⍀Fd∘.=⍳1+⌈⌿Fd
-1 0 0 0 0
-1 1 0 0 0
-1 1 1 0 0
-1 1 1 1 0
-1 1 1 1 1
-1 1 1 1 2
-1 1 1 1 3
-1 1 2 0 0
-1 1 2 2 0
-      ⊢Ec←(1+Ed)↑⍤¯1⊢+⍀Ed∘.=⍳1+⌈⌿Ed
-1 0 0
-1 1 0
-1 1 1
-1 1 2
-1 1 3
-1 2 0
-1 3 0
-1 3 4
-1 3 5
-1 3 6
-\end{verbatim}
-
-A careful study of the definition of a node coordinate and the
-above construction should reveal why this works. Intuitively, we are
-creating a multi-dimensional space or a number system in which each
-digit place or dimension contains or circumscribes a smaller space
-in which are contained all the descendant nodes that appear lower in
-the tree. Each coordinate is a sort of special path through the tree
-encoded to have desirable properties relative to other paths.
+\noindent These matrices let us see the nesting features of each tree more visually, but also suggest another step.
+We can compute a sum scan with \verb;+⍀;, also called a prefix sum, along the first axis.
+Applying this function on the above matrices leads to an interesting result:
+
+\begin{verbatim}
+    +⍀Fd∘.=⍳1+⌈⌿Fd  │     +⍀Ed∘.=⍳1+⌈⌿Ed
+1 0 0 0 0           │ 1 0 0
+1 1 0 0 0           │ 1 1 0
+1 1 1 0 0           │ 1 1 1
+1 1 1 1 0           │ 1 1 2
+1 1 1 1 1           │ 1 1 3
+1 1 1 1 2           │ 1 2 3
+1 1 1 1 3           │ 1 3 3
+1 1 2 1 3           │ 1 3 4
+1 1 2 2 3           │ 1 3 5
+                    │ 1 3 6
+\end{verbatim}
+
+\noindent These matrices are lexicographically ordered, and each ancestor shares a common prefix with its descendants.
+They are also unique coordinates.
+Only the spurious digits at the end of each coordinate prevent these matrices from meeting all our requirements for valid node coordinates.
+
+The expression \verb;V f⍤¯1⊢M; applies \verb;f; to corresponding elements of \verb;V; and rows of \verb;M;:
+
+\begin{verbatim}
+      3 3⍴⍳9 │       (⍳3)+⍤¯1⊢3 3⍴⍳9
+0 1 2        │ 0 1  2
+3 4 5        │ 4 5  6
+6 7 8        │ 8 9 10
+\end{verbatim}
+
+\noindent If \verb;n↑V; takes the first \verb;n; elements of \verb;V;, then we can obtain coordinate matrices from the prefix sums by noting that the spurious digits all come after column \verb;d+1; where \verb;d; is the depth corresponding to that coordinate.
+The following gives a complete expression for computing a node coordinate matrix from a depth vector, shown using \verb;Fd; and \verb;Fe;:
+
+\begin{verbatim}
+      Fc←(1+Fd)↑⍤¯1⊢+⍀Fd∘.=⍳1+⌈⌿Fd
+      Ec←(1+Ed)↑⍤¯1⊢+⍀Ed∘.=⍳1+⌈⌿Ed
+      Fc   │        Ec
+1 0 0 0 0  │  1 0 0
+1 1 0 0 0  │  1 1 0
+1 1 1 0 0  │  1 1 1
+1 1 1 1 0  │  1 1 2
+1 1 1 1 1  │  1 1 3
+1 1 1 1 2  │  1 2 0
+1 1 1 1 3  │  1 3 0
+1 1 2 0 0  │  1 3 4
+1 1 2 2 0  │  1 3 5
+           │  1 3 6
+\end{verbatim}
+
+A careful study of the definition of a node coordinate and the above construction should reveal why this works.
+Intuitively, we are creating a multi-dimensional space or a number system in which each digit place or dimension contains or circumscribes a smaller space in which are contained all the descendant nodes that appear lower in the tree.
+Each coordinate is a sort of special path through the tree encoded to have desirable properties relative to other paths.
 
@@
 
-The simplest operation over a node coordinate is to extract the depth
-of the node.
-The dyadic expression \verb;C⍳0; finds the first occurrence of \verb;0;
-in \verb;C; and returns the index of that occurrence.
+The simplest operation over a node coordinate is to extract the depth of the node.
+The dyadic expression \verb;C⍳0; finds the first occurrence of \verb;0; in \verb;C; and returns the index of that occurrence.
 Thus, we compute the depth of a node as follows:
@@
 
-\noindent
-In many compiler passes, we primarily care about which nodes are ancestors 
-of other nodes. To determine whether one node is a child of another, we compute 
-whether one node is a prefix of the other, ignoring zeros.
-We write \verb;(f g h); to represent the composition of functions \verb;f;, \verb;g;,
-and \verb;h; as a function train.
-Function trains obey the following equivalences, where \verb;A; and \verb;B;
-are arrays.
+\noindent In many compiler passes, we primarily care about which nodes are ancestors of other nodes.
+To determine whether one node is a child of another, we compute whether one node is a prefix of the other, ignoring zeros.
+We write \verb;(f g h); to represent the composition of functions \verb;f;, \verb;g;, and \verb;h; as a function train.
+Function trains obey the following equivalences, where \verb;A; and \verb;B; are arrays.
 
@@
 
-\noindent
-With this, we write a function to compute whether a given coordinate prefixes 
-another.
+\noindent With this, we write a function to compute whether a given coordinate prefixes another.
 
@@
 
-\noindent
-The above determines whether \verb;P; is an ancestor to \verb;C;.
-The logical functions \verb;=;, \verb;∧;, and \verb;∨; are all extended 
-point-wise over arrays.
+\noindent The above determines whether \verb;P; is an ancestor to \verb;C;.
+The logical functions \verb;=;, \verb;∧;, and \verb;∨; are all extended point-wise over arrays.
 The function \verb;⊢; returns its right argument.
-The function \verb;(=∨0=⊢); reads as, ``equal or a zero right argument.'' 
-We take this point-wise operation and reduce it with \verb;∧⌿; (For All).
+The function \verb;(=∨0=⊢); reads as, ``equal or a zero right argument.'' We take this point-wise operation and reduce it with \verb;∧⌿; (For All).
 This pattern is a special case of inner product, written \verb;f.g;.
-For example, \verb;+.×; is matrix multiplication. We transform our reduction and 
-Boolean function to a single predicate with the inner product operator:
+For example, \verb;+.×; is matrix multiplication.
+We transform our reduction and Boolean function to a single predicate with the inner product operator:
+
 \begin{verbatim}
@@
 The use of inner product extends to cases where \verb;C; or \verb;P; are matrices.
-For our examples, we want to lift the functions and flatten expressions, so we 
-will select the \verb;F; nodes and \verb;E; nodes, respectively, as our 
-parent/ancestor nodes in \verb;Fp; and \verb;Ep;. Given a boolean vector 
-\verb;BV;, the expression \verb;BV⌿M; selects rows from \verb;M; based on the 
-non-zero elements of \verb;BV;.
-
-\begin{verbatim}
-      ⊢Fp←('F'=Ft)⌿Fc
-1 0 0 0 0
-1 1 1 0 0
-      ⊢Ep←('E'=Et)⌿Ec
-1 0 0
-1 1 0
-1 3 0
-\end{verbatim}
-
-\noindent
-We could use \verb;Fp; and \verb;Ep; to compare against \verb;Fc; and \verb;Ec;
-to determine which nodes belong to which parent in \verb;Fp; or \verb;Ep;, 
-but our prefix function returns 1 when \verb;C≡P;. 
+For our examples, we want to lift the functions and flatten expressions, so we will select the \verb;F; nodes and \verb;E; nodes, respectively, as our parent/ancestor nodes in \verb;Fp; and \verb;Ep;.
+Given a boolean vector \verb;BV;, the expression \verb;BV⌿M; selects rows from \verb;M; based on the non-zero elements of \verb;BV;.
+
+\begin{verbatim}
+      Fp←('F'=Ft)⌿Fc
+      Ep←('E'=Et)⌿Ec
+      Fp   │        Ep
+1 0 0 0 0  │  1 0 0
+1 1 1 0 0  │  1 1 0
+           │  1 3 0
+\end{verbatim}
+
+\noindent We could use \verb;Fp; and \verb;Ep; to compare against \verb;Fc; and \verb;Ec; to determine which nodes belong to which parent in \verb;Fp; or \verb;Ep;, but our prefix function returns 1 when \verb;C≡P;.
 Instead, we drop the last non-zero element from each coordinate.
 This will permit matching against all ancestors, but not against itself.
-We use the depth vector and the Take (\verb;↑;) function to take all but
-the last non-zero element. We extend the resulting array with an extra 
-zero column to ensure we have a compatible shape.
-
-\begin{verbatim}
-      ⊢Fcp←(Fd↑⍤¯1⊢Fc),0
-0 0 0 0 0
-1 0 0 0 0
-1 1 0 0 0
-1 1 1 0 0
-1 1 1 1 0
-1 1 1 1 0
-1 1 1 1 0
-1 1 0 0 0
-1 1 2 0 0
-      ⊢Ecp←(Ed↑⍤¯1⊢Ec),0
-0 0 0
-1 0 0
-1 1 0
-1 1 0
-1 1 0
-1 0 0
-1 0 0
-1 3 0
-1 3 0
-1 3 0
-\end{verbatim}
-
-\noindent
-We use \verb;Fcp; and \verb;Ecp; to determine the \verb;F; and \verb;E; 
-ancestors for each node (\verb;⍉A; transposes \verb;A;):
-
-\begin{verbatim}
-      Fcp∧.(=∨0=⊢)⍉Fp
-0 0
-1 0
-1 0
-1 1
-1 1
-1 1
-1 1
-1 0
-1 0
-      Ecp∧.(=∨0=⊢)⍉Ep
-0 0 0
-1 0 0
-1 1 0
-1 1 0
-1 1 0
-1 0 0
-1 0 0
-1 0 1
-1 0 1
-1 0 1
-\end{verbatim}
-
-\noindent
-In the above, the closest ancestor is the rightmost 1 in each row.
-We can extract the column number of the rightmost 1 by replacing each 1 
-with its column number and selecting the maximum of each row with \verb;⌈/;. 
+We use the depth vector and the Take (\verb;↑;) function to take all but the last non-zero element.
+We extend the resulting array with an extra zero column to ensure we have a compatible shape.
+
+\begin{verbatim}
+      Fcp←(Fd↑⍤¯1⊢Fc),0
+      Ecp←(Ed↑⍤¯1⊢Ec),0
+      Fcp  │        Ecp
+0 0 0 0 0  │  0 0 0
+1 0 0 0 0  │  1 0 0
+1 1 0 0 0  │  1 1 0
+1 1 1 0 0  │  1 1 0
+1 1 1 1 0  │  1 1 0
+1 1 1 1 0  │  1 0 0
+1 1 1 1 0  │  1 0 0
+1 1 0 0 0  │  1 3 0
+1 1 2 0 0  │  1 3 0
+           │  1 3 0
+\end{verbatim}
+
+\noindent We use \verb;Fcp; and \verb;Ecp; to determine the \verb;F; and \verb;E; ancestors for each node (\verb;⍉A; transposes \verb;A;):
+
+\begin{verbatim}
+   Fcp∧.(=∨0=⊢)⍉Fp │    Ecp∧.(=∨0=⊢)⍉Ep
+0 0                │ 0 0 0
+1 0                │ 1 0 0
+1 0                │ 1 1 0
+1 1                │ 1 1 0
+1 1                │ 1 1 0
+1 1                │ 1 0 0
+1 1                │ 1 0 0
+1 0                │ 1 0 1
+1 0                │ 1 0 1
+                   │ 1 0 1
+\end{verbatim}
+
+\noindent In the above, the closest ancestor is the rightmost 1 in each row.
+We can extract the column number of the rightmost 1 by replacing each 1 with its column number and selecting the maximum of each row with \verb;⌈/;.
 For space we will show only the \verb;E; example.
-In the following, \verb;V×⍤1⊢M; gives a matrix where each row is given by 
-\verb;V×i⌷M; where \verb;i⌷M; is the $i$th row of \verb;M;.
+In the following, \verb;V×⍤1⊢M; gives a matrix where each row is given by \verb;V×i⌷M; where \verb;i⌷M; is the $i$th row of \verb;M;.
 
@@
 0 0 1 1 1 0 0 2 2 2
-\end{verbatim}
-
-\noindent
-We use these column numbers to index into \verb;Ep; to obtain an ancestor 
-matrix where the node coordinate of the closest ancestor is given for 
-each node, one per row. 
-The expression \verb;I⌷⍤0 2⊢M; gives a matrix, one row per index in \verb;I;, 
-consisting of rows of \verb;M; selected by row indices in \verb;I;.
-
-\begin{verbatim}
-      ⊢Ek←Ei⌷⍤0 2⊢Ep 
-1 0 0
-1 0 0
-1 1 0
-1 1 0
-1 1 0
-1 0 0
-1 0 0
-1 3 0
-1 3 0
-1 3 0
-\end{verbatim}
-
-\noindent
-At this point we have two values, \verb;Ek; and \verb;Fk;, which indicate the
-closest containing node that we care about for each node in the tree,
-using its node coordinate.
-
-\begin{verbatim}
-      Fi←⌈/(⍳2)×⍤1⊢Fcp∧.(=∨0=⊢)⍉Fp
-      ⊢Fk←Fi⌷⍤0 2⊢Fp
-1 0 0 0 0
-1 0 0 0 0
-1 0 0 0 0
-1 1 1 0 0
-1 1 1 0 0
-1 1 1 0 0
-1 1 1 0 0
-1 0 0 0 0
-1 0 0 0 0
-\end{verbatim}
-
-\noindent
-We use these keys to compute over the AST, particularly for function lifting 
-and expression flattening as demonstrated in the next section. The repeating 
-pattern is to leverage node coordinates and array operations to do stackless 
-reasoning about inter-node relationships. This pattern occurs often in our 
-compiler and is particularly useful to enable straightforward data-parallel 
-transformations over an AST.
+      ⊢Fi←⌈/(⍳2)×⍤1⊢Fcp∧.(=∨0=⊢)⍉Fp
+0 0 0 1 1 1 1 0 0
+\end{verbatim}
+
+\noindent We use these column numbers to index into \verb;Ep; to obtain an ancestor matrix where the node coordinate of the closest ancestor is given for each node, one per row.
+The expression \verb;I⌷⍤0 2⊢M; gives a matrix, one row per index in \verb;I;, consisting of rows of \verb;M; selected by row indices in \verb;I;.
+
+\begin{verbatim}
+      Fk←Fi⌷⍤0 2⊢Fp
+      Ek←Ei⌷⍤0 2⊢Ep 
+      Fk   │        Ek
+1 0 0 0 0  │  1 0 0
+1 0 0 0 0  │  1 0 0
+1 0 0 0 0  │  1 1 0
+1 1 1 0 0  │  1 1 0
+1 1 1 0 0  │  1 1 0
+1 1 1 0 0  │  1 0 0
+1 1 1 0 0  │  1 0 0
+1 0 0 0 0  │  1 3 0
+1 0 0 0 0  │  1 3 0
+           │  1 3 0
+\end{verbatim}
+
+\noindent At this point we have two values, \verb;Ek; and \verb;Fk;, which indicate the closest containing node that we care about for each node in the tree, using its node coordinate.
+We use these keys to compute over the AST, particularly for function lifting and expression flattening as demonstrated in the next section.
+The repeating pattern is to leverage node coordinates and array operations to do stackless reasoning about inter-node relationships.
+This pattern occurs often in our compiler and is particularly useful to enable straightforward data-parallel transformations over an AST.
 
@@
 
-First introduced in the J language \cite{hui2014key},
-the Key operator (written \verb;⌸;) is central to our strategy for compilation.
+First introduced in the J language \cite{hui2014key}, the Key operator (written \verb;⌸;) is central to our strategy for compilation.
 It allows us to use the Node Coordinates to their full effect.
-The expression \verb;K f⌸ M; groups rows of \verb;M; by their corresponding 
-keys in \verb;K; and computes \verb;k f m; for each unique key \verb;k; in 
-\verb;K; and the matrix \verb;m; of rows of \verb;M; with key \verb;k;. 
+The expression \verb;K f⌸ M; groups rows of \verb;M; by their corresponding keys in \verb;K; and computes \verb;k f m; for each unique key \verb;k; in \verb;K; and the matrix \verb;m; of rows of \verb;M; with key \verb;k;.
 The corresponding rows of matrices \verb;K; and \verb;M; form key-value pairs.
@@
 
-\noindent
-To understand a bit better how the Key operator applies its function,
-consider the function \verb;{⍺ ⍵}; which returns the pair of its right
-and left arguments. If we apply it to the same value as above, we
-get the following:
+\noindent To understand a bit better how the Key operator applies its function, consider the function \verb;{⍺ ⍵}; which returns the pair of its right and left arguments.
+If we apply it to the same value as above, we get the following:
 
@@
 
-\noindent
-In our case, we use either \verb;Fk; or \verb;Ek; as our keys applied to the
-corresponding AST. We also will drop off the first row in each AST
-using \verb;1↓; since this node ``contains'' everything. In a complete
-AST this is usually the Module boundary node which contains the entire
-set of functions and values in the module.
+\noindent In our case, we use either \verb;Fk; or \verb;Ek; as our keys applied to the corresponding AST.
+We also will drop off the first row in each AST using \verb;1↓; since this node ``contains'' everything.
+In a complete AST this is usually the Module boundary node which contains the entire set of functions and values in the module.
 
@@
 
-The following compiler passes are found in any compiler from a nested function, 
-nested expression language to a language without nested functions 
-or expressions. They were the chief motivating examples for developing the 
-techniques presented above as a part of developing the Co-dfns compiler, 
-and represent a challenging problem to data-parallel compilation without 
-the above techniques, but fall out almost effortlessly once the above patterns 
-are available.
+The following compiler passes are found in any compiler from a nested function, nested expression language to a language without nested functions or expressions.
+They were the chief motivating examples for developing the techniques presented above as a part of developing the Co-dfns compiler, and represent a challenging problem to data-parallel compilation without the above techniques, but fall out almost effortlessly once the above patterns are available.
+
+Each of the following passes operates over an AST that we represent as the catenation of the field vectors above including the Node Coordinates.
+
+\begin{verbatim}
+Fast←Fd,Ft,Fv,⍪↓Fc
+East←Ed,Et,Ev,⍪↓Ec
+\end{verbatim}
+
+\noindent The function \verb;↓; converts the coordinate matrices from matrices to vectors of vectors, so that the coordinates can be catenated to the other fields, leading to a 4-column matrix for the ASTs, whose fourth column is a vector of vectors.
 
@@
 
-If we use the Key operator with \verb;Fk;, we get the following (the monadic
-use of \verb;↓; converts \verb;Fc; from a matrix to a vector of vectors):
-
-\begin{verbatim}
-      (1↓Fk){⍺ ⍵}⌸1↓Fd,Ft,Fv,⍪↓Fc
+Function lifting takes an AST with nested functions (func-tions appearing inside of other functions) and lifts these functions out of their local scopes and puts them all at the top level.
+Normally, function lifting must consider both the core operation of lifting function as well as the lexical scop-ing of the functions and variables that appear inside of a function.
+As seen below, our treatment uses a slightly dif-ferent method.
+
+If we use the Key operator (\verb;⌸;) with the pair function (\verb;{⍺ ⍵};) using \verb;Fk; as the keys and \verb;Fast; as the AST, we get the following:
+
+\begin{verbatim}
+      (1↓Fk){⍺ ⍵}⌸1↓Fast
 ┌─────────┬─────────────────┐
@@
 
-Notice that we have now grouped all of the relevant parts of the
-tree according to which nodes would appear in their respective
-functions after lifting. Refer to the original lifting example in
-the introduction to verify this. Indeed, the second row in the
-above example shows the internal function complete and ready to
-name. Each element in the second column corresponds to the body of
-one of our lifted functions. In the case of the first function, the
-outer function, we have a spurious function node in the body. This
-is intentional. When we lift these functions, we will replace
-each spurious function node with a variable node referring to the
-function’s generated name.
-
-Each of these function bodies has a specific coordinate associated
-with it. Because these coordinates are uniquely identifying, we can
-use these as input into a name generator to generate names that we
-know are unique for each function body. Furthermore, because we
-retain this information in the corresponding function nodes that
-appear in the body of each function to be lifted, we know exactly
-what name that function has been given, and we can replace the
-function node with a variable node referencing that name instead,
-without referring to any state outside of the immediate information
-given to the function lifter. Indeed, each row in the above matrix
-represents a function lifting task that can be completed without any
-additional information. That is, there are no dependencies between
-rows to perform lifting. This gives us a straightforward parallel
-execution of function lifting.
-
-The final task to complete function lifting of each function
-body before lifting is to shift the depths in the depth vectors
-to correspond to those of a function lifted to the top-level and
-to attach a function node to the top of each of the bodies. At that
-point, we simply recombine all of the newly created functions into
-a single top level.
-
-This function lifting works even with lexically scoped functions, 
-as is the case for the Co-dfns compiler.
-We can do this because the node coordinates maintain the lexical information 
-for each variable reference, enabling us to construct the appropriate lexical 
-binding for each variable much later than would normally be done in a 
-compiler. 
-In other words, the above lifting operation loses no information that would 
-prevent us from handling lexical scoping, but decouples lexical scoping 
-from the act of lifting functions.
+Notice that we have now grouped all of the relevant parts of the tree according to which nodes would appear in their respective functions after lifting.
+Refer to the original lifting example in the introduction to verify this.
+Indeed, the second row in the above example shows the internal function complete and ready to name.
+Each element in the second column corresponds to the body of one of our lifted functions.
+In the case of the first function, the outer function, we have a spurious function node in the body.
+This is intentional.
+When we lift these functions, we will replace each spurious function node with a variable node referring to the function’s generated name.
+
+\begin{verbatim}
+      (1↓Fk){⊂(0 'F' 0 ⍺)⍪⍵}⌸1↓Fast
+┌─────────────────┬─────────────────┐
+│┌─┬─┬─┬─────────┐│┌─┬─┬─┬─────────┐│
+││0│F│0│1 0 0 0 0│││0│F│0│1 1 1 0 0││
+│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
+││1│E│0│1 1 0 0 0│││3│E│0│1 1 1 1 0││
+│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
+││2│F│0│1 1 1 0 0│││4│V│⍵│1 1 1 1 1││
+│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
+││2│A│0│1 1 2 0 0│││4│P│+│1 1 1 1 2││
+│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
+││3│N│7│1 1 2 2 0│││4│V│⍵│1 1 1 1 3││
+│└─┴─┴─┴─────────┘│└─┴─┴─┴─────────┘│
+└─────────────────┴─────────────────┘ 
+      hd←0 'F' 0,∘⊂⊣      ⍝ Func. Header
+      dp←(1+⊢-⊃)0⌷⍤1⊢     ⍝ New Depths
+      rf←'V'@('F'∘=)1⌷⍤1⊢ ⍝ F → V
+      (1↓Fk)(⊂hd⍪dp,rf,2↓⍤1⊢)⌸1↓Fast
+┌─────────────────┬─────────────────┐
+│┌─┬─┬─┬─────────┐│┌─┬─┬─┬─────────┐│
+││0│F│0│1 0 0 0 0│││0│F│0│1 1 1 0 0││
+│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
+││1│E│0│1 1 0 0 0│││1│E│0│1 1 1 1 0││
+│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
+││2│V│0│1 1 1 0 0│││2│V│⍵│1 1 1 1 1││
+│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
+││2│A│0│1 1 2 0 0│││2│P│+│1 1 1 1 2││
+│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
+││3│N│7│1 1 2 2 0│││2│V│⍵│1 1 1 1 3││
+│└─┴─┴─┴─────────┘│└─┴─┴─┴─────────┘│
+└─────────────────┴─────────────────┘
+\end{verbatim}
+
+Each of these function bodies has a specific coordinate associated with it.
+Because these coordinates are uniquely identifying, we can use these as input into a name generator to generate names that we know are unique for each function body.
+Furthermore, because we retain this information in the corresponding function nodes that appear in the body of each function to be lifted, we know exactly what name that function has been given, and we can replace the function node with a variable node referencing that name instead, without referring to any state outside of the immediate information given to the function lifter.
+Indeed, each row in the above matrix represents a function lifting task that can be completed without any additional information.
+That is, there are no dependencies between rows to perform lifting.
+This gives us a straightforward parallel execution of function lifting.
+
+The final task to complete function lifting of each function body before lifting is to shift the depths in the depth vectors to correspond to those of a function lifted to the top-level and to attach a function node to the top of each of the bodies.
+At that point, we simply recombine all of the newly created functions into a single top level.
+
+This function lifting works even with lexically scoped functions, as is the case for the Co-dfns compiler.
+We can do this because the node coordinates maintain the lexical information for each variable reference, enabling us to construct the appropriate lexical binding for each variable much later than would normally be done in a compiler.
+In other words, the above lifting operation loses no information that would prevent us from handling lexical scoping, but decouples lexical scoping from the act of lifting functions.
 
@@
 
-If we use \verb;Ek; as the key for the expression example, we get the
-following:
-
-\begin{verbatim}
-      (1↓Ek){⍺ ⍵}⌸1↓Ed,Et,Ev,⍪↓Ec
+If we use \verb;Ek; as the key for the expression example, we get the following:
+
+\begin{verbatim}
+      (1↓Ek){⍺ ⍵}⌸1↓East
 ┌─────┬─────────────┐
@@
 └─────┴─────────────┘
-\end{verbatim}
-
-Again, we can see immediately that we have grouped each set of nodes
-according to the expressions that are to be lifted. Just as in the
-case of function lifting, we can adjust the depths of each expression
-to the correct depth and we can replace each expression node with a
-variable reference based on that node’s coordinate. Each expression
-can be given a unique name based on its coordinate. A later compiler
-pass can reduce these names down to the minimum actually required to
-represent the expression if desired.
-
-The only extra issue involved here is to ensure that the order of
-evaluation matches. In our case, we are assuming that the order of
-evaluation is right to left, which means that the above order is
-actually backwards of the desired order. During recombination, we
-simply reverse these orders and this fixes that problem. More work
-would be required to take into consideration a specific precedence
-hierarchy.
+      hd←0'E'0,∘⊂⊣
+      rf←'V'@('E'∘=)1⌷⍤1⊢
+      ⊖(1↓Ek)(hd⍪dp,rf,2↓⍤1⊢)⌸1↓East
+┌─┬─┬─┬─────┐
+│0│E│0│1 3 0│
+├─┼─┼─┼─────┤
+│1│V│c│1 3 4│
+├─┼─┼─┼─────┤
+│1│P│×│1 3 5│
+├─┼─┼─┼─────┤
+│1│V│d│1 3 6│
+└─┴─┴─┴─────┘
+┌─┬─┬─┬─────┐
+│0│E│0│1 1 0│
+├─┼─┼─┼─────┤
+│1│V│a│1 1 1│
+├─┼─┼─┼─────┤
+│1│P│+│1 1 2│
+├─┼─┼─┼─────┤
+│1│V│b│1 1 3│
+└─┴─┴─┴─────┘
+┌─┬─┬─┬─────┐
+│0│E│0│1 0 0│
+├─┼─┼─┼─────┤
+│1│V│0│1 1 0│
+├─┼─┼─┼─────┤
+│1│P│÷│1 2 0│
+├─┼─┼─┼─────┤
+│1│V│0│1 3 0│
+└─┴─┴─┴─────┘
+\end{verbatim}
+
+Again, we can see immediately that we have grouped each set of nodes according to the expressions that are to be lifted.
+Just as in the case of function lifting, we can adjust the depths of each expression to the correct depth and we can replace each expression node with a variable reference based on that node’s coordinate.
+Each expression can be given a unique name based on its coordinate.
+A later compiler pass can reduce these names down to the minimum actually required to represent the expression if desired.
+
+The only extra issue involved here is to ensure that the order of evaluation matches.
+In our case, we are assuming that the order of evaluation is right to left, which means that the above order is actually backwards of the desired order.
+During recombination, we simply reverse these orders and this fixes that problem.
+More work would be required to take into consideration a specific precedence hierarchy.
 
@@
 
-The Co-dfns compiler is an experimental commercial compiler funded by 
-Dyalog, Ltd. \cite{hsu2014co,hsu2015accelerating}
-It is intended to provide increased scaling and performance
-for APL programmers using the dfns syntax (lexically scoped, functionally 
-oriented APL). These techniques form one of the foundational elements to 
-the architecture of the compiler, which is built in the style of a Nanopass 
-\cite{keep2013nanopass}
-compiler, with very small passes chained together through composition, with 
-the added requirements that each compiler pass must be fully data parallel.
-
-The compiler produces good code that performs closely with hand-written 
-C code for the same programs on the GPU and CPU. It produces platform 
-independent code, meaning that the compiler is able to target either the 
-GPU or CPU from the same program source and produce reasonable performance 
-on both.
-
-The compiler itself is composed of three main parts: the core compiler, 
-the parser, and the code generator. The code generator includes the runtime 
-library of the compiler. Parsing APL code in parallel is a well-studied 
-problem. The compiler currently uses parsing combinators to do its parsing, 
-though we plan to implement the Two-by-Two parser given in the literature
-\cite{bunda1984apl}. 
-The code generator itself is a simple parallel map over each node, as 
-each node may be generated independent from the other nodes, leading to a 
-fully parallel generator, though there are some branching elements in the 
-generator that still remain, currently.
-
-The core of the Co-dfns compiler is written entirely using function trains 
-and parallel operations in the style described above. We use these techniques 
-extensively throughout the compiler, and the code is publicly available 
-at our repository. The core of the compiler, including some of the large 
-tables comes in at around 250 lines of code. Without the tables, the core 
-logic comes in at around 150 lines of code.
-
-The entire compiler, together with its runtime libraries, parser, generator, 
-and core, comes in at around 1500 lines of commented code.
-
-Table \ref{tab:passes} contains a selection of passes from the
-Co-dfns compiler and breaks down their features based on how they
-traverse the tree and whether or not they use the Key operator or
-Node Coordinates.
-The traversal pattern indicates the primary operator used to traverse the AST 
-and perform the transformation, most of which are primitives; the Amend pattern 
-is a ``replacement'' pattern where specific nodes are replaced in the AST with 
-other nodes, and can be thought of as a slightly modified filter and map.
-The table orders the passes with front-end passes higher in
-the table and passes further down the compiler chain at the bottom
-of the table. These passes give a good idea of how the Key operator
-plays a part in the complete compiler as well as how Node Coordinates
-help to deal with complex sub-tree selections.
-
-We make a few notes on the table and the analysis. The Key operator
-can be used with or without the node coordinates structure. In cases
-where the AST is sufficiently flat and arranged in a sufficiently
-convenient order, the grouping operations are very simple. In these
-cases, the Key operator amounts to a map operation over some selected
-sub-trees selected at a specific depth. These occur later in the
-compiler where the reader will note appearances of the Key operator
-without the use of Node Coordinates.
-
-Likewise, node coordinates find their place early on in the compiler
-where it is necessary to deal with the more highly nested AST. They
-are useful for identifying this information regardless of whether the
-Key operator is used to handle the grouping or not. Thus passes like
-``Drop Unreachable Code'' that remove whole sub-trees from the AST
-may use the Node Coordinates to do the grouping once the correct
-parent node has been identified.
-
-In short, the goal of the structure and design of the compiler passes
-given in Table \ref{tab:passes} is to utilize Node Coordinates to
-remove as quickly as possible the nested structure of the AST into
-a flatter form that does not require complex analysis to traverse.
+The Co-dfns compiler is an experimental commercial compiler funded by Dyalog, Ltd.
+\cite{hsu2014co,hsu2015accelerating} It is intended to provide increased scaling and performance for APL programmers using the dfns syntax (lexically scoped, functionally oriented APL).
+These techniques form one of the foundational elements to the architecture of the compiler, which is built in the style of a Nanopass \cite{keep2013nanopass} compiler, with very small passes chained together through composition, with the added requirements that each compiler pass must be fully data parallel.
+
+The compiler produces good code that performs closely with hand-written C code for the same programs on the GPU and CPU.
+It produces platform independent code, meaning that the compiler is able to target either the GPU or CPU from the same program source and produce reasonable performance on both.
+
+The compiler itself is composed of three main parts: the core compiler, the parser, and the code generator.
+The code generator includes the runtime library of the compiler.
+Parsing APL code in parallel is a well-studied problem.
+The compiler currently uses parsing combinators to do its parsing, though we plan to implement the Two-by-Two parser given in the literature \cite{bunda1984apl}.
+The code generator itself is a simple parallel map over each node, as each node may be generated independent from the other nodes, leading to a fully parallel generator, though there are some branching elements in the generator that still remain, currently.
+
+The core of the Co-dfns compiler is written entirely using function trains and parallel operations in the style described above.
+We use these techniques extensively throughout the compiler, and the code is publicly available at our repository.
+The core of the compiler, including some of the large tables comes in at around 250 lines of code.
+Without the tables, the core logic comes in at around 150 lines of code.
+
+The entire compiler, together with its runtime libraries, parser, generator, and core, comes in at around 1500 lines of commented code.
+
+Table \ref{tab:passes} contains a selection of passes from the Co-dfns compiler and breaks down their features based on how they traverse the tree and whether or not they use the Key operator or Node Coordinates.
+The traversal pattern indicates the primary operator used to traverse the AST and perform the transformation, most of which are primitives; the Amend pattern is a ``replacement'' pattern where specific nodes are replaced in the AST with other nodes, and can be thought of as a slightly modified filter and map.
+The table orders the passes with front-end passes higher in the table and passes further down the compiler chain at the bottom of the table.
+These passes give a good idea of how the Key operator plays a part in the complete compiler as well as how Node Coordinates help to deal with complex sub-tree selections.
+
+We make a few notes on the table and the analysis.
+The Key operator can be used with or without the node coordinates structure.
+In cases where the AST is sufficiently flat and arranged in a sufficiently convenient order, the grouping operations are very simple.
+In these cases, the Key operator amounts to a map operation over some selected sub-trees selected at a specific depth.
+These occur later in the compiler where the reader will note appearances of the Key operator without the use of Node Coordinates.
+
+Likewise, node coordinates find their place early on in the compiler where it is necessary to deal with the more highly nested AST.
+They are useful for identifying this information regardless of whether the Key operator is used to handle the grouping or not.
+Thus passes like ``Drop Unreachable Code'' that remove whole sub-trees from the AST may use the Node Coordinates to do the grouping once the correct parent node has been identified.
+
+In short, the goal of the structure and design of the compiler passes given in Table \ref{tab:passes} is to utilize Node Coordinates to remove as quickly as possible the nested structure of the AST into a flatter form that does not require complex analysis to traverse.
+
+\section{Performance}
+
+\section{Discussion}
 
@@
 
-The current compiler represents a non-trivial instance of
-the successful application of these techniques, but it focuses on an untyped, 
-functionally oriented language. The type inference in the current version 
-of the compiler is somewhat na\"ive. More work is required to expand the 
-type inferencer to include more sophisticated type inference. 
-The compiler does not currently handle errors such as signals or 
-exceptions, nor does it implement some more sophisticated compiler optimizations.
-Further work remains to implement these techniques on an imperative and 
-object-oriented language instead of a functional one. It would also be 
-educational to implement a version for a lazy language.
-
-We do not have a large scale understanding of the performance characteristics 
-of these techniques in real life outside of daily use of the compiler.
-Further analysis is
-required to understand what sort of compiler optimizations and analyses are 
-necessary to compile the Co-dfns compiler efficiently for the GPU and CPU. 
-We currently lack a good comparison against existing techniques both for 
-the compiler design and the generated code. Some existing compilers are able 
-to compile variants of APL to target the GPU and the CPU, but there are subtle 
-differences that require care when comparing generated code, and the 
-architecture of these compilers is often quite different. Comparing the 
-relative performance both of the compiler designs (human factors and execution 
-speed) and generated code is important to understanding these techniques more 
-fully.
-
-While we have found this method of compiler construction to be very compact 
-and to permit working with a data-parallel compiler as easily or nearly as 
-easily as with a standard Nanopass compiler, we have not done extensive 
-studies to determine just how comparable in programmability, ease of 
-maintenance, or extensibility these techniques are with established 
-compiler construction methods, such as OOP Visitors, Nanopass, or the 
-type-directed functional style.
+The current compiler represents a non-trivial instance of the successful application of these techniques, but it focuses on an untyped, functionally oriented language.
+The type inference in the current version of the compiler is somewhat na\"ive.
+More work is required to expand the type inferencer to include more sophisticated type inference.
+The compiler does not currently handle errors such as signals or exceptions, nor does it implement some more sophisticated compiler optimizations.
+Further work remains to implement these techniques on an imperative and object-oriented language instead of a functional one.
+It would also be educational to implement a version for a lazy language.
+
+We do not have a large scale understanding of the performance characteristics of these techniques in real life outside of daily use of the compiler.
+Further analysis is required to understand what sort of compiler optimizations and analyses are necessary to compile the Co-dfns compiler efficiently for the GPU and CPU.
+We currently lack a good comparison against existing techniques both for the compiler design and the generated code.
+Some existing compilers are able to compile variants of APL to target the GPU and the CPU, but there are subtle differences that require care when comparing generated code, and the architecture of these compilers is often quite different.
+Comparing the relative performance both of the compiler designs (human factors and execution speed) and generated code is important to understanding these techniques more fully.
+
+While we have found this method of compiler construction to be very compact and to permit working with a data-parallel compiler as easily or nearly as easily as with a standard Nanopass compiler, we have not done extensive studies to determine just how comparable in programmability, ease of maintenance, or extensibility these techniques are with established compiler construction methods, such as OOP Visitors, Nanopass, or the type-directed functional style.
 
@@
 
-Iverson introduced the idea of the Rank operator while at Sharp Associates 
-\cite{iverson1983rationalized} as a part of ``Rationalized APL.'' 
-The rank operator (\verb;⍤;) used here derived from the J language, a 
-continuation of the rationalized APL idea, with many ideas such as rank 
-percolating back into APL implementations. \cite{bernecky1987rank,hui1995rank}
-The J programming language \cite{hui2014key} was the first practical,
-general-purpose programming language to introduce the Key operator as a
-primitive operator with the presumption of its general usefulness.
-
-Bernecky \cite{bernecky2015abstract} argues that the increased programmability 
-and desirable human factors of APL-style array programming can be implemented 
-with competitive performance relative to more traditional techniques. This 
-suggests that programs written in this highly abstract style may not suffer the 
-performance gap traditionally assumed to go with their desirable features. 
-This work is bolstered by previous work on the high-performance implementation 
-of array-oriented languages. 
-\cite{ching1994experimental,ching1993primitive,
-ching1990automatic,ju1991exploitation,
-ju1991performance,bernecky1999reducing,schwarz1991acorn}
-
-Fritz Henglein demonstrated a class of operations, called
-discriminators, of which the Key operator is a member  
-\cite{henglein2013dd}
-, namely,
-a discriminator performs the same grouping computation as Key, but
-does not apply a function over these groups with their keys. Henglein
-provides a linear implementation of these operations.
-
-The EigenCFA effort \cite{prabhu2011eigencfa} demonstrated significant performance
-improvements of a 0-CFA flow analysis by utilizing similar techniques
-to those demonstrated here. In particular, encoding the AST and using
-accessor functions have a very similar feel to the node coordinates and
-AST encoding given here, though they have a different formulation and
-spend considerable effort understanding the trade-offs of performance
-associated with the different encodings, whereas the encodings here
-were chosen for their clarity and directness, rather than their
-performance.
-
-Mendez-Lojo, et al. implemented a GPU version of Inclusion-based
-Points-to Analysis \cite{mendez2012inclusion} that also focuses on 
-adapting data structures
-and algorithms to efficiently execute on the GPU. In particular,
-they use similar techniques of prefix sums and sorts to achieve some
-of their adaptation to the GPU, Additionally, they have clever and
-efficient methods of representing graphs on the GPU which enable
-dynamic rewriting of the graph.
-
-The APEX compiler \cite{bernecky1997apex} developed vectorized approaches to handling
-certain analyses to compile traditional APL, including a SIMD tokenizer
-\cite{bernecky2003tokenizer}.
-It also uses 
-a matrix format to represent the AST. Traditional APL did not have
-nested function definitions, however, and thus the APEX compiler does
-not have any specific approaches to dealing with function lifting.
-
-Timothy Budd implemented a compiler \cite{budd1984apl,budd2012apl} for APL which targeted
-vector processors as well as C. Budd provided 
-thoughts and some ideas on how the compiler might be implemented in
-parallel as well.
-
-J. D. Bunda and J. A. Gerth presented a method for doing table driven
-parsing of APL which suggested a parallel optimization for parsing,
-but did not elucidate the algorithm \cite{bunda1984apl}.
+Iverson introduced the idea of the Rank operator while at Sharp Associates \cite{iverson1983rationalized} as a part of ``Rationalized APL.'' The rank operator (\verb;⍤;) used here derived from the J language, a continuation of the rationalized APL idea, with many ideas such as rank percolating back into APL implementations.
+\cite{bernecky1987rank,hui1995rank} The J programming language \cite{hui2014key} was the first practical, general-purpose programming language to introduce the Key operator as a primitive operator with the presumption of its general usefulness.
+
+Bernecky \cite{bernecky2015abstract} argues that the increased programmability and desirable human factors of APL-style array programming can be implemented with competitive performance relative to more traditional techniques.
+This suggests that programs written in this highly abstract style may not suffer the performance gap traditionally assumed to go with their desirable features.
+This work is bolstered by previous work on the high-performance implementation of array-oriented languages.
+\cite{ching1994experimental,ching1993primitive, ching1990automatic,ju1991exploitation, ju1991performance,bernecky1999reducing,schwarz1991acorn}
+
+Fritz Henglein demonstrated a class of operations, called discriminators, of which the Key operator is a member \cite{henglein2013dd} , namely, a discriminator performs the same grouping computation as Key, but does not apply a function over these groups with their keys.
+Henglein provides a linear implementation of these operations.
+
+The EigenCFA effort \cite{prabhu2011eigencfa} demonstrated significant performance improvements of a 0-CFA flow analysis by utilizing similar techniques to those demonstrated here.
+In particular, encoding the AST and using accessor functions have a very similar feel to the node coordinates and AST encoding given here, though they have a different formulation and spend considerable effort understanding the trade-offs of performance associated with the different encodings, whereas the encodings here were chosen for their clarity and directness, rather than their performance.
+
+Mendez-Lojo, et al.
+implemented a GPU version of Inclusion-based Points-to Analysis \cite{mendez2012inclusion} that also focuses on adapting data structures and algorithms to efficiently execute on the GPU.
+In particular, they use similar techniques of prefix sums and sorts to achieve some of their adaptation to the GPU, Additionally, they have clever and efficient methods of representing graphs on the GPU which enable dynamic rewriting of the graph.
+
+The APEX compiler \cite{bernecky1997apex} developed vectorized approaches to handling certain analyses to compile traditional APL, including a SIMD tokenizer \cite{bernecky2003tokenizer}.
+It also uses a matrix format to represent the AST.
+Traditional APL did not have nested function definitions, however, and thus the APEX compiler does not have any specific approaches to dealing with function lifting.
+
+Timothy Budd implemented a compiler \cite{budd1984apl,budd2012apl} for APL which targeted vector processors as well as C.
+Budd provided thoughts and some ideas on how the compiler might be implemented in parallel as well.
+
+J. D. Bunda and J. A. Gerth presented a method for doing table driven parsing of APL which suggested a parallel optimization for parsing, but did not elucidate the algorithm \cite{bunda1984apl}.
 
@@
 
-We have derived a method of performing computation over sub-trees
-selected on the basis of inter-node properties through the use of the
-Key (\verb;⌸;) operator and node coordinates, which enable local computation
-of these inter-node properties. This method is both general and direct,
-and when combined with traditional and more mundane array programming,
-suffices to implement the complete core of a compiler, modulo parsing
-and code generation. The method requires no special operations or
-unique special casing primitives in the language. Moreover, it is
-strictly data-parallel and data-flow, without any complex control flow,
-which results in exceptionally concise code. These methods permit a 
-general, high-level approach to constructing a compiler that is 
-parallel by construction by constricting the language to only a data parallel 
-subset of a normal array language.
-
-We have demonstrated the technique and the core insights behind the
-data structures involved. It presents a solution to a very old and
-traditional problem in a very uncommon light, by eschewing the common
-practices that underlie every other significant and general solution
-found in modern compilers today and replacing them with an entirely
-different paradigm centered on parallelism and aggregate operations.
+We have derived a method of performing computation over sub-trees selected on the basis of inter-node properties through the use of the Key (\verb;⌸;) operator and node coordinates, which enable local computation of these inter-node properties.
+This method is both general and direct, and when combined with traditional and more mundane array programming, suffices to implement the complete core of a compiler, modulo parsing and code generation.
+The method requires no special operations or unique special casing primitives in the language.
+Moreover, it is strictly data-parallel and data-flow, without any complex control flow, which results in exceptionally concise code.
+These methods permit a general, high-level approach to constructing a compiler that is parallel by construction by constricting the language to only a data parallel subset of a normal array language.
+
+We have demonstrated the technique and the core insights behind the data structures involved.
+It presents a solution to a very old and traditional problem in a very uncommon light, by eschewing the common practices that underlie every other significant and general solution found in modern compilers today and replacing them with an entirely different paradigm centered on parallelism and aggregate operations.
 \balancecolumns

### 2016-11-15 2017-spring-pldi/hsu-2017-key_data_parallel_compiler.tex
Add in the graphics to the paper
Add in a performance section

@@
 \setmonofont{APL385 Unicode}
+\usepackage{graphicx}
 \usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
@@
 
+\begin{figure*}
+\begin{center}
+\includegraphics[width=3.3in]{fnc_gpu_3d}
+\includegraphics[width=3.3in]{mut_gpu_3d}
+\includegraphics[width=3.3in]{fnc_gpu_2d}
+\includegraphics[width=3.3in]{mut_gpu_2d}
+\end{center}
+\caption{Performance of Key-based Expression Flattening compared to traditional methods}
+\label{fig-bench}
+\end{figure*}
+
 \section{Performance}
+
+To examine the performance characteristics of the above techniques, we prototyped a micro-benchmark that flattens expressions on an AST.
+This mimics the core functionality of both the function lifting and expression flattening passes described in more detail above and used in the Co-dfns compiler.
+It also serves to compare the performance of using Key (\verb;⌸;) based algorithms to perform tree transformation compared to recursively defined algorithms.
+We implemented the above lifting algorithm using CUDA 8.0 and Thrust primitives.
+
+We implemented the same algorithm using two traditional recursive methods.
+The first is a ``Functional'' analogue, that avoids mutation of the result and instead uses a pure-functional interface to pass nodes to be lifted back up the call stack.
+This is a very common method, taught in compiler courses that teach flattening passes, as well as used in various compilers.
+The benefit of the functional algorithm is the concision with which it may be written in most higher-level languages, particularly those which eschew mutatoin and favor a pure coding style.
+The disadvantage is that it has a higher memory usage requirement and may copy data as it moves up through the call stack to reassemble various containers for child nodes.
+
+The second algorithm is a common optimization on the functional recursive style that makes use of an accumulator container that threads through the computation.
+The traversal functions will then mutate this accumulator as nodes are discovered that require lifting.
+The advantage over the functional method is that you can avoid excess copying overheads and in some languages that are more favorable to mutation, it may even be easier to write.
+A disadvantage is that you must be carefully aware of the order of traversal through the AST, since a change in the ordering of traversal can affect the order in which traversal functions write to the accumulator.
+This makes it less friendly to multi-threading.
+It can also be cumbersome or less desirable to write in this style in some programming languages in which the functional algorithm finds natural expression.
+
+Normally, one would test GPU oriented benchmarks based on their input sizes.
+However, in this case, the traditional algorithms and the GPU algorithms are written quite differently.
+Thus, we introduce two dimensions of input size in order to examine differences with more precision.
+Particularly, we benchmarked the GPU algorithm against the traditional methods by depth of the AST and Branching Factor.
+We chose to examine depth as a specific factor because the path matrix is inherently dependent on the depth.
+Furthermore, the mutation algorithm is optimized in such a way that fewer expression nodes may result in fewer writes to the accumulator, while our algorithm has no such optimization, and traverses the entire tree independent of the number of expression nodes.
+
+We ran the benchmark to compute the time to flatten an AST on all combinations of depth and branching factor along the interval $[0,32)$ that were of a suitable size (that is, fit within the machine constraints).
+The GPU was an NVIDIA Tesla K40c, and the CPU was an Intel Xeon .....
+
+Figure \ref{fig-bench} shows the results of this experiment.
+It is interesting to note the difference in shapes between the functional and mutation algorithms relative to the GPU algorithms when mapped along both the depth and branching factor.
+As expected, the speedups against the functional algorithm are higher on the whole because of the higher memory demands of that algorithm.
+All these algorithms should be memory bound, so a reduction in overall memory traffic benefits the mutation algorithm quite a bit.
+
+Our peak speedups are around 58× comparing against the functional algorithm, and around 9× when comparing against the mutation algorithm.
+However, the ``performance'' zone for the GPU algorithm is fairly similar in shape between the two.
+Particularly, this "curved mountain peak" represents the boundaries where the number of AST elements in the whole AST becomes large enough to attain sufficient parallelism on the GPU.
+
+The second row of graphs in Figure \ref{fig-bench} highlights a single plane of the mountain where we fix the branching factor or the depth.
+This gives a better picture of the behavior of the algorithm along the most pronounced areas of the area graph.
+These are also of particular interest because they represent two common cases found in real code.
+When the branching factor is 2, we have a reasonable analogue of common expressions found in code such as APL or other mathematically intense code which may have long strings of dense expressions, leading to a higher depth, but a low branching factor (since most operators of this sort take one or two arguments).
+When the Depth is fixed at 5, but we increase the branching factor, this more closely simulates code that may have functions nested throughout the code that need to be lifted which contain many child statements, but may not be particularly deep.
+In such cases the branching factor would increase, but the ASTs themselves may be relatively low depth.
 

### 2016-11-16 2017-spring-pldi/hsu-2017-key_data_parallel_compiler.tex
Add a discussion section

@@
 We ran the benchmark to compute the time to flatten an AST on all combinations of depth and branching factor along the interval $[0,32)$ that were of a suitable size (that is, fit within the machine constraints).
-The GPU was an NVIDIA Tesla K40c, and the CPU was an Intel Xeon .....
+The GPU was an NVIDIA Tesla K40c, and the CPU was an Intel Xeon  E5-1620v3.
 
@@
 \section{Discussion}
+
+There are many options for representing graphs on data-parallel architecture.
+Most of them tend to optimize traversal instead of transformation.
+Because of the transformation heavy nature of compiler-style algorithms, these other representations can introduce work that complicates both the algorithms themselves, as well as the work necessary to maintain those structures over the lifetime of the compilation phase.
+
+The use of a path matrix provides benefits to locality, which allows for increased parallelism for transformation passes such as the flattening passes mentioned above.
+While this can be used to increase the performance of a single compiler pass, the above treatment is not intended primarily for highly optimizing a single pass in a compiler.
+Instead, the design of the above algorithms and techniques are designed to improve the data parallelism and ease of expression across the entire compiler, not just a single pass.
+
+To illustrate the difference, consider a few other possible representations.
+One possibility is the use of a single parent vector for the entire AST.
+We could also work directly off of the depth vector or the traditional method of listing pointers to the children and/or parents.
+In the case of the expression flattening above, we could likely achieve equal or better performance from using a single parent vector for this one computation.
+However, would be increasing costs elsewhere, particularly in maintenance.
+A path matrix provides both parent information, as well as information about the depth of the node, the siblings, position in the tree, and a unique identifier in a single structure that is cache/memory friendly.
+Since many operations on trees are memory bound, this memory optimization becomes important.
+In the Co-dfns compiler, we are able to generate the path matrix once, at the beginning of the compilation, and pass them around freely, and even duplicate them when appropriate, to communicate information without requiring cross thread or non-local memory access patterns.
+
+This means that we never need to recompute the matrix, and indeed, doing so would defeat the purpose.
+In contrast, if you require a parent vector to organize your AST, then that parent vector may need to be shuffled, extended, and permuted each time the AST changes, and fails to maintain the original edge information of the graph.
+Keeping the original parent vector but maintaining a set of links would be a potential solution, but removes the locality of the access patterns, making it impossible to do a coalesced read of all of edge information needed for a single computation as a single access.
+Using a path matrix allows us to avoid that and increase the amount of stride-1 regular accesses and reduces memory indirections.
+
+The cost of a path matrix is its size.
+Many other representations are size efficient, depending only on the number of nodes in the AST.
+A path matrix on the other hand, in the worst case, is dependent both on the depth of the AST and the number of nodes in the AST.
+Depending on the specific application, it is possible to choose different path matrix representations that optimize certain cases.
+For instance, one may not require that traversing the parent path be fast, and thus, we may reduce the cost of generating the path matrix by removing that requirement.
+Removing this requirement may also allow a path matrix to use a smaller sized integer representation for the matrix, further reducing the size at the cost of increased computation for certain operations.
+We have shown general path matrix techniques above that will work regardless of these particular choices in creating the path matrix.
 

### 2016-11-16 2017-spring-pldi/hsu-2017-key_data_parallel_compiler.tex
Clean up the body, intro, and notation

@@
 We present a language-driven strategy for the construction of compilers that are inherently data-parallel in their design and implementation.
-Using an encoding of the inter-node relationships between nodes in an AST called a Node Co-ordinate Matrix, we demonstrate how an operator called the Key operator, that applies a function over groupings of array cells grouped and ordered by their keys, when used in conjunction with a Node Coordinate Matrix, permits arbi-trary computation over sub-trees of an AST using purely data-parallel array programming techniques.
-We discuss the use of these techniques when applied to an experi-mental commercial compiler called Co-dfns, which is a fully data-parallel compiler developed using the techniques discussed here.
+Using an encoding of the inter-node relationships between nodes in an AST called a Path Matrix, we demonstrate how an operator called the Key operator, that applies a function over groupings of array cells grouped and ordered by their keys, when used in conjunction with a Path Matrix, permits arbitrary computation over sub-trees of an AST using purely data-parallel array programming techniques.
+We discuss the use of these techniques when applied to an experimental commercial compiler called Co-dfns, a fully data-parallel compiler developed using the techniques discussed here.
+On transformation heavy operations such as expression flattening or function lifting, these techniques, when compared against pure and effectful recursive implementations, outperforms the CPU-based algorithms by a factor of 10 - 58 on the GPU.
 \end{abstract}
@@
 
-Compilers represent a peculiar class of tree/graph algo-rithms that greatly alter the underlying structure of the input graph into something very different in structure, but equivalent by some measure of interpretation.
-These algo-rithms are widely applicable, but have stubbornly resisted a general approach to parallelization.
+Compilers represent a peculiar class of tree/graph algorithms that greatly alter the underlying structure of the input graph into something very different in structure, but equivalent by some semantic interpretation.
+These algorithms are widely applicable, but have stubbornly resisted a general approach to parallelization.
 Most compilers are single-threaded or make very limited use of parallelism in an ad hoc way.
-The design and analysis of compilers al-most universally deals with compilers as recursive travers-als over ASTs.
-Such formulations usually include heavy reliance on branching, recursion, and sometimes very so-phisticated control flow in practice.
+The design and analysis of compilers almost universally deals with compilers as recursive traversals over ASTs.
+Such formulations usually include heavy reliance on branching, recursion, and sometimes very sophisticated control flow.
 All of this contributes to make it difficult to effectively parallelize such algo-rithms onto architectures that are sensitive to branching and recursion, such as GPGPUs or highly vectorised CPUs.
@@
 A key problem to the construction of compilers in such a restricted environment is dealing with the inherently recursive and nested structure of the AST.
-By working over a linearized representation of the AST as a matrix, combined with a structure we call a Node Coordinate Matrix, and using data-parallel primitives, we are able to tame the recursive and nested nature of the AST into something that can be processed handily using only data-parallel, data-flow programs, without requiring branching or other forms of non-data parallel control flow.
+By working over a linearized representation of the AST as a matrix, combined with a structure we call a Path Matrix, and using data-parallel primitives, we are able to tame the recursive and nested nature of the AST into something that can be processed handily using only data-parallel, data-flow programs, without requiring branching or other forms of non-data parallel control flow.
 The methods are general and independent of the language being compiled.
 
-In the following sections we describe the core methods and idioms surrounding the AST encoding, construction of the Node Coordinate Matrix, and the use of these structures in handling nested, recursive AST relationships.
+In the following sections we describe the core methods and idioms surrounding the AST encoding, construction of the Path Matrix, and the use of these structures in handling nested, recursive AST relationships.
 This enables us to perform arbitrary computation over arbitrary sub-trees of an AST selected by parent-child relationships in the manner often seen in compiler transformations.
@@
 
+To illustrate the performance of these techniques, we micro-benchmark a transformation heavy expression lifting algorithm that is found almost universally in compilers.
+This benchmark shows performance speedups of up to a factor of 58 when compared against pure, functionally oriented algorithms.
+
 \paragraph{Contributions}
@@
 
-\item We describe a method of computing over arbitrary sub-trees selected by their parent-child relationships in a data-parallel manner using the Key operator and Node Coordinates.
-
-\item We situation this technique into a broader, language-driven strategy for compiler construction that enables the development of parallel by construction compilers that are high-level in their implementation, exceptionally concise, and independent of the language being compiled.
+\item We describe a method of computing over arbitrary sub-trees selected by their parent-child relationships in a data-parallel manner using the Key operator and a Path Matrix.
+
+\item We situate this technique into a broader, language-driven strategy for compiler construction that enables the development of parallel by construction compilers that are high-level in their implementation, exceptionally concise, and independent of the language being compiled.
 
 \item We provide analysis of these techniques as used in a commercial compiler project called Co-dfns and report on the results, including the overall architecture of the compiler, the uses of these techniques within the compiler, and the demonstration of these techniques applied to two specific compiler passes.
+
+\item We compare the performance of the Key-based algorithms against traditional recursive methods for the task of lifting/flattening expressions, such as commonly done for function lifting and expression flattening.
+Our results demonstrate consistent speedups over the traditional methods.
 
@@
 All expressions are evaluated right to left and all function application is infix; that is, a function application may apply a function (such as \verb;+;) to one or two arguments, called monadic and dyadic application respectively, which must appear on the right and (optionally) on the left of the function name, as in the above example.
-Tables \ref{tab:scalarprims} to \ref{tab:conjunctions} provide a listing of all of the primitives used in the Co-dfns compiler, a selection of which are used here.
-
-We separate out higher-order functions called operators, from functions that operate over arrays that we call simply functions.
+The Appendix provides a listing of all of the primitives used in the Co-dfns compiler, a selection of which are used here.
+
+We differentiate higher-order functions called operators, from functions that operate over arrays that we call simply functions.
 Operators take a function argument or two, and bind more strongly to the left than the right.
@@
 
-\noindent We introduce notation as needed throughout the exposition, so the reader is encouraged to refer to the tables as necessary to recall particular operations.
+\noindent We introduce notation as needed throughout the exposition, so the reader is encouraged to refer to the appendix as necessary to recall particular operations.
 
@@
 
+Each of these grouped sub-trees will be given a new local root \verb;F; node that is at the appropriate depth (0) and uses the key as its path vector (\verb;⊣;).
+
 \begin{verbatim}
@@
 └─────────────────┴─────────────────┘ 
+\end{verbatim}
+
+\noindent At this point, the internal function nodes refer by path vector to the newly added local roots.
+This was down without traversing or maintaining these connections, but falls out as a natural artifact of using the Key (\verb;⌸;) operator.
+At this point we are ready to fix up the sub-tree to be correct.
+Firstly, we need a function to encapsulate the function header information.
+
+\begin{verbatim}
       hd←0 'F' 0,∘⊂⊣      ⍝ Func. Header
+\end{verbatim}
+
+\noindent Next, we write a function to adjust the depths, which is one more than the difference between each node's depth and the depth of the first node of the sub-tree.
+
+\begin{verbatim}
       dp←(1+⊢-⊃)0⌷⍤1⊢     ⍝ New Depths
+\end{verbatim}
+
+\noindent Finally, we use the \verb;@; operator to replace occurances of \verb;F; in the Type column with \verb;V; elements.
+
+\begin{verbatim}
       rf←'V'@('F'∘=)1⌷⍤1⊢ ⍝ F → V
+\end{verbatim}
+
+\noindent Once we put all this together, we can construct the new sub-trees with the appropriate depths, types, references, and links with the following expression.
+
+\begin{verbatim}
       (1↓Fk)(⊂hd⍪dp,rf,2↓⍤1⊢)⌸1↓Fast
@@
 
-The final task to complete function lifting of each function body before lifting is to shift the depths in the depth vectors to correspond to those of a function lifted to the top-level and to attach a function node to the top of each of the bodies.
-At that point, we simply recombine all of the newly created functions into a single top level.
-
 This function lifting works even with lexically scoped functions, as is the case for the Co-dfns compiler.
@@
 \subsection{Expression Flattening}
+
+Flattening expressions is almost identical to lifting functions, in part because we have delayed the lexical scope resolution aspect of lifting functions to a later pass that will handle this explicitly.
+In addition to lifting the expressions themselves, however, we will also take the opportunity to handle precedence and ordering.
+In our examples, we are assuming that all expressions evaluate right to left.
+Because the Key (\verb;⌸;) operator is a stable operation, and we require that nodes have a depth-first pre-order traversal pattern, then we can trivially handle precedence order of this type, including handling parentheses, by reversing the results of flattening.
 
@@
 └─────┴─────────────┘
+\end{verbatim}
+
+\noindent Again, we can see immediately that we have grouped each set of nodes according to the expressions that are to be lifted.
+Just as in the case of function lifting, we can adjust the depths of each expression to the correct depth and we can replace each expression node with a variable reference based on that node’s coordinate.
+Each expression can be given a unique name based on its coordinate.
+A later compiler pass can reduce these names down to the minimum actually required to represent the expression if desired.
+As with the function lifting case, we need some functions to help abstract each new field.
+In our case, we use the same functions as with function lifting but alter the \verb;hd; and \verb;rf; functions to use the right node types:
+
+\begin{verbatim}
       hd←0'E'0,∘⊂⊣
       rf←'V'@('E'∘=)1⌷⍤1⊢
+\end{verbatim}
+
+\noindent At this point, the expression lifting operation is then almost identical to that of function lifting.
+Notice that we reverse the 3-D matrix result at the end along the first access to ensure that order of evaluation is preserved correctly.
+
+\begin{verbatim}
       ⊖(1↓Ek)(hd⍪dp,rf,2↓⍤1⊢)⌸1↓East
@@
 
-Again, we can see immediately that we have grouped each set of nodes according to the expressions that are to be lifted.
-Just as in the case of function lifting, we can adjust the depths of each expression to the correct depth and we can replace each expression node with a variable reference based on that node’s coordinate.
-Each expression can be given a unique name based on its coordinate.
-A later compiler pass can reduce these names down to the minimum actually required to represent the expression if desired.
-
-The only extra issue involved here is to ensure that the order of evaluation matches.
-In our case, we are assuming that the order of evaluation is right to left, which means that the above order is actually backwards of the desired order.
+The only extra issue involved here over function lifting is to ensure that the order of evaluation matches.
+In our case, we are assuming that the order of evaluation is right to left, which means that the above order given by \verb;⌸; is actually backwards of the desired order.
 During recombination, we simply reverse these orders and this fixes that problem.
@@
 In such cases the branching factor would increase, but the ASTs themselves may be relatively low depth.
+
+It should also be noted that while the mutation-oriented algorithm is obviously faster than the functional algorithm on the CPU, it comes at the cost of stylistic clarity and the capability to reason locally about computation.
+Our Key-based algorithm remains pure and functional while still being fast.
+It is also exceptionally concise relative to the other two algorithms.
 

### 2016-11-16 2017-spring-pldi/hsu-2017-key_data_parallel_compiler.tex
Improve clarity and treatment of computation sections

@@
 Computing over sub-trees normally involves recursing down the structure of the AST, identifying the root nodes of each sub-tree, and then dispatching to a handler for that sub-tree, which will continue the recursion at that point and return the modified tree, which will replace the previous sub-tree.
-More complicated transformations may involved moving the root nodes around in the tree or other large, distant structural modifications.
-
-We divide the work of sub-tree computation into three basic phases: we maintain a node coordinate with each node that uniquely identifies it relative to others, we use these coordinates to select and group nodes for work as a single sub-tree, and then we operate over these groupings using the Key operator (\verb;⌸;).
+More complicated transformations may involve moving the root nodes around in the tree or other large, distant structural modifications.
+
+We divide the work of sub-tree computation into three basic phases: we maintain a Path Vector with each node that uniquely identifies it relative to others, we use these Paths to select and group nodes for work as a single sub-tree, and then we operate over these groupings using the Key operator (\verb;⌸;).
 
@@
 \noindent The depth vector stores all edge information for the tree, but this information requires non-local access to utilize, such as traversing potentially the entire vector to determine the children of a specific node.
-Node Coordinates fix this issue.
-
-\subsection{Node Coordinates}
-
-Every node in an AST has a Node Coordinate, named because a coordinate is a precise location in a space.
-We can imagine all the nodes arranged inside some multi-dimensional space, leading to a specific set of node coordinates.
+Path Vectors fix this issue.
+
+\subsection{Path Matrix}
+
+Every node in an AST has a Path Vector, together forming a matrix.
+We can imagine all the nodes arranged inside some multi-dimensional space, leading to a specific set of Paths or coordinates.
 Many such arrangements exist, of varying usefulness.
@@
 
-\noindent In short, we care about the relative position of each node in a tree relative to any other. We define a node coordinate as follows.
-
-A node coordinate is a vector whose length is the depth of the tree.
+\noindent In short, we care about the relative position of each node in a tree relative to any other.
+We define a path vector as a vector whose length is the depth of the tree.
 Its elements are natural numbers.
@@
 All zero elements appear after the non-zero elements.
-That is, a coordinate is zero-padded on the right.
-When ordered lexicographically, the nodes for each coordinate appear in order according to a depth-first pre-order traversal of the tree.
-Each coordinate uniquely identifies a single node.
-Every ancestor’s coordinate is a prefix of any child’s coordinate, ignoring zeros.
-From the above it follows that every node is lexicographically greater than its left sibling and differs from it by exactly one non-zero element, and that this element is the last non-zero element in the coordinate.
-
-\subsubsection{Constructing Node Coordinates}
-
-We construct the Node Coordinate Matrix from the depth vector, thereby encoding the depth vector into a more useful format.
+That is, a path is zero-padded on the right.
+When ordered lexicographically, the nodes for each path appear in order according to a depth-first pre-order traversal of the tree.
+Each path uniquely identifies a single node.
+Every ancestor’s path is a prefix of any child’s path, ignoring zeros.
+From the above it follows that every node is lexicographically greater than its left sibling and differs from it by exactly one non-zero element, and that this element is the last non-zero element in the path.
+
+\subsubsection{Constructing the Path Matrix}
+
+We construct the Path Matrix from the depth vector, thereby encoding the depth vector into a more useful format.
 The matrix has \verb;N; rows and \verb;D; columns, where \verb;N; is the node count and \verb;D; is the depth of the tree.
@@
 
-\noindent We can obtain the ordered sequence  by writing \verb;⍳n;:
+\noindent We can obtain an ordered sequence  by writing \verb;⍳n;:
 
@@
 \begin{verbatim}
-      Fd∘.=⍳1+⌈⌿Fd │      Ed∘.=⍳1+⌈⌿Ed
+   Fb←Fd∘.=⍳1+⌈⌿Fd │   Eb←Ed∘.=⍳1+⌈⌿Ed
 1 0 0 0 0          │ 1 0 0
@@
 \begin{verbatim}
-    +⍀Fd∘.=⍳1+⌈⌿Fd  │     +⍀Ed∘.=⍳1+⌈⌿Ed
+    +⍀Fb   │     +⍀Fb
 1 0 0 0 0  │ 1 0 0
@@
 Only the spurious digits at the end of each coordinate prevent these matrices from meeting all our requirements for valid node coordinates.
-
-The expression \verb;V f⍤¯1⊢M; applies \verb;f; to corresponding elements of \verb;V; and rows of \verb;M;:
-
-\begin{verbatim}
-      3 3⍴⍳9 │       (⍳3)+⍤¯1⊢3 3⍴⍳9
-0 1 2        │ 0 1  2
-3 4 5        │ 4 5  6
-6 7 8        │ 8 9 10
-\end{verbatim}
-
-\noindent If \verb;n↑V; takes the first \verb;n; elements of \verb;V;, then we can obtain coordinate matrices from the prefix sums by noting that the spurious digits all come after column \verb;d+1; where \verb;d; is the depth corresponding to that coordinate.
-The following gives a complete expression for computing a node coordinate matrix from a depth vector, shown using \verb;Fd; and \verb;Fe;:
-
-\begin{verbatim}
-      Fc←(1+Fd)↑⍤¯1⊢+⍀Fd∘.=⍳1+⌈⌿Fd
-      Ec←(1+Ed)↑⍤¯1⊢+⍀Ed∘.=⍳1+⌈⌿Ed
-      Fc   │        Ec
+We generate a mask to remove these spurious digits by using a different outer product on our original depth vector:
+
+\begin{verbatim}
+    Fm←Fd∘.≥⍳1+⌈⌿Fd │   Em←Ed∘.≥⍳1+⌈⌿Ed
+1 0 0 0 0           │  1 0 0
+1 1 0 0 0           │  1 1 0
+1 1 1 0 0           │  1 1 1
+1 1 1 1 0           │  1 1 1
+1 1 1 1 1           │  1 1 1
+1 1 1 1 1           │  1 1 0
+1 1 1 1 1           │  1 1 0
+1 1 1 0 0           │  1 1 1
+1 1 1 1 0           │  1 1 1
+                    │  1 1 1
+\end{verbatim}
+
+%The expression \verb;V f⍤¯1⊢M; applies \verb;f; to corresponding elements of \verb;V; and rows of \verb;M;:
+%
+%\begin{verbatim}
+%      3 3⍴⍳9 │       (⍳3)+⍤¯1⊢3 3⍴⍳9
+%0 1 2        │ 0 1  2
+%3 4 5        │ 4 5  6
+%6 7 8        │ 8 9 10
+%\end{verbatim}
+
+\noindent We use this mask with the \verb;×; function to get a complete expression for computing a path matrix from a depth vector, shown using \verb;Fd; and \verb;Fe;:
+
+\begin{verbatim}
+      Fc←Fm×+⍀Fb │        Ec←Fm×+⍀Fb
 1 0 0 0 0        │  1 0 0
@@
 
-A careful study of the definition of a node coordinate and the above construction should reveal why this works.
+A careful study of the definition of a path vector and the above construction should reveal why this works.
 Intuitively, we are creating a multi-dimensional space or a number system in which each digit place or dimension contains or circumscribes a smaller space in which are contained all the descendant nodes that appear lower in the tree.
-Each coordinate is a sort of special path through the tree encoded to have desirable properties relative to other paths.
-
-\subsubsection{Operations on Node Coordinates}
-
-The simplest operation over a node coordinate is to extract the depth of the node.
+Each vector is a path through the tree encoded to have desirable properties relative to other paths.
+
+Note that there are other valid path matrix encodings.
+The one above is used throughout most of the rest of this paper because of its simplicity and easy of exposition.
+We use it to demonstrate the full generality of the structure.
+However, certain more complex matrix encodings permit certain operations at much lower cost while preserving the full generality.
+
+\subsubsection{Operations on Path Vectors}
+
+The simplest operation over a Path Vector is to extract the depth of the node.
 The dyadic expression \verb;C⍳0; finds the first occurrence of \verb;0; in \verb;C; and returns the index of that occurrence.
@@
 
-\noindent With this, we write a function to compute whether a given coordinate prefixes another.
+\noindent With this, we write a function to compute whether a given path vector prefixes another.
 
@@
 \noindent We could use \verb;Fp; and \verb;Ep; to compare against \verb;Fc; and \verb;Ec; to determine which nodes belong to which parent in \verb;Fp; or \verb;Ep;, but our prefix function returns 1 when \verb;C≡P;.
-Instead, we drop the last non-zero element from each coordinate.
+Instead, we drop the last non-zero element from each path.
 This will permit matching against all ancestors, but not against itself.
-We use the depth vector and the Take (\verb;↑;) function to take all but the last non-zero element.
-We extend the resulting array with an extra zero column to ensure we have a compatible shape.
-
-\begin{verbatim}
-      Fcp←(Fd↑⍤¯1⊢Fc),0
-      Ecp←(Ed↑⍤¯1⊢Ec),0
+We use the same masking technique as above, but adjust the comparison to use \verb;>; instead of \verb;≥; to leave off the last non-zero element of \verb;Fc; and \verb;Ec;:
+
+\begin{verbatim}
+      Fcp←Fc×Fd∘.>⍳1+⌈⌿Fd
+      Ecp←Ec×Ed∘.>⍳1+⌈⌿Ed
       Fcp  │        Ecp
@@
 \noindent In the above, the closest ancestor is the rightmost 1 in each row.
-We can extract the column number of the rightmost 1 by replacing each 1 with its column number and selecting the maximum of each row with \verb;⌈/;.
-For space we will show only the \verb;E; example.
-In the following, \verb;V×⍤1⊢M; gives a matrix where each row is given by \verb;V×i⌷M; where \verb;i⌷M; is the $i$th row of \verb;M;.
-
-\begin{verbatim}
-      (⍳3)×⍤1⊢Ecp∧.(=∨0=⊢)⍉Ep
-0 0 0
-0 0 0
-0 1 0
-0 1 0
-0 1 0
-0 0 0
-0 0 0
-0 0 2
-0 0 2
-0 0 2
-      ⊢Ei←⌈/(⍳3)×⍤1⊢Ecp∧.(=∨0=⊢)⍉Ep
+We can extract the column number of the rightmost 1 by replacing each 1 with its column number and selecting the maximum of each row with \verb;⌈.×;.
+The function \verb;≢; gives the number of rows of its argument.
+
+\begin{verbatim}
+      ⊢Ei←(Ecp∧.(=∨0=⊢)⍉Ep)⌈.×⍳≢Ep
 0 0 1 1 1 0 0 2 2 2
-      ⊢Fi←⌈/(⍳2)×⍤1⊢Fcp∧.(=∨0=⊢)⍉Fp
+      ⊢Fi←(Fcp∧.(=∨0=⊢)⍉Fp)⌈.×⍳≢Fp
 0 0 0 1 1 1 1 0 0
@@
 
-\noindent We use these column numbers to index into \verb;Ep; to obtain an ancestor matrix where the node coordinate of the closest ancestor is given for each node, one per row.
-The expression \verb;I⌷⍤0 2⊢M; gives a matrix, one row per index in \verb;I;, consisting of rows of \verb;M; selected by row indices in \verb;I;.
-
-\begin{verbatim}
-      Fk←Fi⌷⍤0 2⊢Fp
-      Ek←Ei⌷⍤0 2⊢Ep 
-      Fk   │        Ek
+\noindent We use these column numbers to index into \verb;Ep; to obtain an ancestor matrix where the path vector of the closest ancestor is given for each node, one per row.
+
+\begin{verbatim}
+      Fk←Fp[Fi;]   │        Ek←Ep[Ei;]
 1 0 0 0 0          │  1 0 0
@@
 
-\noindent At this point we have two values, \verb;Ek; and \verb;Fk;, which indicate the closest containing node that we care about for each node in the tree, using its node coordinate.
+\noindent At this point we have two values, \verb;Ek; and \verb;Fk;, which indicate the closest containing node that we care about for each node in the tree, using its path vector.
 We use these keys to compute over the AST, particularly for function lifting and expression flattening as demonstrated in the next section.
-The repeating pattern is to leverage node coordinates and array operations to do stackless reasoning about inter-node relationships.
+The repeating pattern is to leverage path vectors and array operations to do stackless reasoning about inter-node relationships.
 This pattern occurs often in our compiler and is particularly useful to enable straightforward data-parallel transformations over an AST.

### 2016-11-16 2017-spring-pldi/hsu-2017-key_data_parallel_compiler.tex
More node coordinate reference removal

@@
 First introduced in the J language \cite{hui2014key}, the Key operator (written \verb;⌸;) is central to our strategy for compilation.
-It allows us to use the Node Coordinates to their full effect.
+It allows us to use the path vectors to their full effect.
 The expression \verb;K f⌸ M; groups rows of \verb;M; by their corresponding keys in \verb;K; and computes \verb;k f m; for each unique key \verb;k; in \verb;K; and the matrix \verb;m; of rows of \verb;M; with key \verb;k;.
@@
 
-Function lifting takes an AST with nested functions (func-tions appearing inside of other functions) and lifts these functions out of their local scopes and puts them all at the top level.
+Function lifting takes an AST with nested functions (functions appearing inside of other functions) and lifts these functions out of their local scopes and puts them all at the top level.
 Normally, function lifting must consider both the core operation of lifting function as well as the lexical scop-ing of the functions and variables that appear inside of a function.
-As seen below, our treatment uses a slightly dif-ferent method.
+As seen below, our treatment uses a slightly different method.
 
@@
 
-Each of these function bodies has a specific coordinate associated with it.
-Because these coordinates are uniquely identifying, we can use these as input into a name generator to generate names that we know are unique for each function body.
+Each of these function bodies has a specific path vector associated with it.
+Because these paths are uniquely identifying, we can use these as input into a name generator to generate names that we know are unique for each function body.
 Furthermore, because we retain this information in the corresponding function nodes that appear in the body of each function to be lifted, we know exactly what name that function has been given, and we can replace the function node with a variable node referencing that name instead, without referring to any state outside of the immediate information given to the function lifter.

### 2016-11-16 2017-spring-pldi/hsu-2017-key_data_parallel_compiler.tex
More clean-up and fixes

@@
 \title{The Key to a Performant Data Parallel Compiler}
-\authorinfo{Aaron W. Hsu}{Indiana University, USA}{awhsu@indiana.edu}
+%\authorinfo{Aaron W. Hsu}{Indiana University, USA}{awhsu@indiana.edu}
+\authorinfo{}{}{}
+
 \maketitle
@@
 We discuss the use of these techniques when applied to an experimental commercial compiler called Co-dfns, a fully data-parallel compiler developed using the techniques discussed here.
-On transformation heavy operations such as expression flattening or function lifting, these techniques, when compared against pure and effectful recursive implementations, outperforms the CPU-based algorithms by a factor of 10 - 58 on the GPU.
+On transformation heavy operations such as expression flattening or function lifting, these techniques, when compared against pure and effectful recursive implementations, outperform the CPU-based algorithms by a factor of 10 - 58 on the GPU.
 \end{abstract}
@@
 \terms Algorithms, Performance, Design, Languages
-\keywords APL, Parallel, Compilers, Key, Node Coordinates
+\keywords APL, Parallel, Compilers, Key, Paths
 
@@
 Such formulations usually include heavy reliance on branching, recursion, and sometimes very sophisticated control flow.
-All of this contributes to make it difficult to effectively parallelize such algo-rithms onto architectures that are sensitive to branching and recursion, such as GPGPUs or highly vectorised CPUs.
-
-There has been some success in efficiently executing parts of a compiler on the GPU, such as the parser \cite{bunda1984apl}, tokenization \cite{bernecky2003tokenizer}, and certain compiler analyses \cite{prabhu2011eigencfa,mendez2012inclusion}.
+All of this contributes to make it difficult to effectively parallelize such algorithms onto architectures that are sensitive to branching and recursion, such as GPGPUs or highly vectorised CPUs.
+
+There has been some success in efficiently executing parts of a compiler on the GPU, such as parsing \cite{bunda1984apl}, tokenization \cite{bernecky2003tokenizer}, and certain compiler analyses \cite{prabhu2011eigencfa,mendez2012inclusion}.
 However, a generalized framework or strategy for parallel compilation remains elusive.
@@
 By working over a linearized representation of the AST as a matrix, combined with a structure we call a Path Matrix, and using data-parallel primitives, we are able to tame the recursive and nested nature of the AST into something that can be processed handily using only data-parallel, data-flow programs, without requiring branching or other forms of non-data parallel control flow.
-The methods are general and independent of the language being compiled.
+The methods are general and independent of the transformation or computation.
 
@@
 All expressions are evaluated right to left and all function application is infix; that is, a function application may apply a function (such as \verb;+;) to one or two arguments, called monadic and dyadic application respectively, which must appear on the right and (optionally) on the left of the function name, as in the above example.
-The Appendix provides a listing of all of the primitives used in the Co-dfns compiler, a selection of which are used here.
+The appendix provides a listing of all of the primitives used in the Co-dfns compiler, a selection of which are used here.
 
@@
 \noindent We combine these to form the respective AST matrices.
-We write \verb;A,B; to catenate arrays \verb;A; and \verb;B; along their last axes and \verb;⍪A; to turn a vector into a 1-column matrix.
+We write \verb;A,B; to catenate arrays \verb;A; and \verb;B; along the last axis and \verb;⍪A; to turn a vector into a 1-column matrix.
 Thus, the two AST matrices are given by the following expressions:
@@
 Every ancestor’s path is a prefix of any child’s path, ignoring zeros.
-From the above it follows that every node is lexicographically greater than its left sibling and differs from it by exactly one non-zero element, and that this element is the last non-zero element in the path.
+From the above it follows that every node is lexicographically greater than its left sibling and differs only in the last non-zero element.
 
@@
 
+\subsection{The Key Operator}
+
+First introduced in the J language \cite{hui2014key}, the Key operator (written \verb;⌸;) is central to our strategy for compilation.
+It allows us to use the path vectors to their full effect.
+The expression \verb;K f⌸ M; groups rows of \verb;M; by their corresponding keys in \verb;K; and computes \verb;k f m; for each unique key \verb;k; in \verb;K; and the matrix \verb;m; of rows of \verb;M; with key \verb;k;.
+The corresponding rows of matrices \verb;K; and \verb;M; form key-value pairs.
+Let's compute a histogram using Tally (\verb;≢;) and Key (\verb;⌸;):
+
+\begin{verbatim}
+      10⍴5
+5 5 5 5 5 5 5 5 5 5
+      ⊢X←?10⍴5
+1 1 4 0 1 3 1 1 2 1
+      ≢⍳5
+5
+      X(⊣,(≢⊢))⌸X
+1 6
+4 1
+0 1
+3 1
+2 1
+\end{verbatim}
+
+\noindent To understand a bit better how the Key operator applies its function, consider the function \verb;{⍺ ⍵}; which returns the pair of its right and left arguments.
+If we apply it to the same value as above, we get the following:
+
+\begin{verbatim}
+      X{⍺ ⍵}⌸X
+1  1 1 1 1 1 1 
+4  4           
+0  0           
+3  3           
+2  2
+\end{verbatim}
+
+\noindent In our case, we use either \verb;Fk; or \verb;Ek; as our keys applied to the corresponding AST.
+We also will drop off the first row in each AST using \verb;1↓; since this node ``contains'' everything.
+In a complete AST this is usually the Module boundary node which contains the entire set of functions and values in the module.
+
+\section{Case Studies}
+
+The following compiler passes are found in any compiler from a nested function, nested expression language to a language without nested functions or expressions.
+They were the chief motivating examples for developing the techniques presented above as a part of developing the Co-dfns compiler, and represent a challenging problem to data-parallel compilation without the above techniques, but fall out almost effortlessly once the above patterns are available.
+
+Each of the following passes operates over an AST represented as the catenation of the field vectors above including the Path Matrix.
+
+\begin{verbatim}
+Fast←Fd,Ft,Fv,⍪↓Fc
+East←Ed,Et,Ev,⍪↓Ec
+\end{verbatim}
+
+\noindent The function \verb;↓; converts the path matrices from matrices to vectors of vectors, so that the coordinates can be catenated to the other fields, leading to a 4-column matrix for the ASTs, whose fourth column is a vector of vectors.
+
+\subsection{Function Lifting}
+
+Function lifting takes an AST with nested functions (functions appearing inside of other functions) and lifts these functions out of their local scopes and puts them all at the top level.
+Normally, function lifting must consider both the core operation of lifting function as well as the lexical scoping of the functions and variables that appear inside of a function.
+As seen below, our treatment uses a slightly different method.
+By using the path matrix, we are able to maintain the relevant information throughout the compiler and avoid resolving scope until around the same time that we handle other variable resolution tasks.
+
+If we use the Key operator (\verb;⌸;) with the pair function (\verb;{⍺ ⍵};) using \verb;Fk; as the keys and \verb;Fast; as the AST, we group all of the relevant parts of the tree according to which nodes would appear in their respective functions after lifting.
+Refer to the original AST given above to verify this.
+The second column in the following result shows the body of each function, complete and ready to name.
+Each element in the second column corresponds to the body of one of our to be lifted functions.
+
+\begin{verbatim}
+      (1↓Fk){⍺ ⍵}⌸1↓Fast
+┌─────────┬─────────────────┐
+│1 0 0 0 0│┌─┬─┬─┬─────────┐│
+│         ││1│E│0│1 1 0 0 0││
+│         │├─┼─┼─┼─────────┤│
+│         ││2│F│0│1 1 1 0 0││
+│         │├─┼─┼─┼─────────┤│
+│         ││2│A│0│1 1 2 0 0││
+│         │├─┼─┼─┼─────────┤│
+│         ││3│N│7│1 1 2 2 0││
+│         │└─┴─┴─┴─────────┘│
+├─────────┼─────────────────┤
+│1 1 1 0 0│┌─┬─┬─┬─────────┐│
+│         ││3│E│0│1 1 1 1 0││
+│         │├─┼─┼─┼─────────┤│
+│         ││4│V│⍵│1 1 1 1 1││
+│         │├─┼─┼─┼─────────┤│
+│         ││4│P│+│1 1 1 1 2││
+│         │├─┼─┼─┼─────────┤│
+│         ││4│V│⍵│1 1 1 1 3││
+│         │└─┴─┴─┴─────────┘│
+└─────────┴─────────────────┘
+\end{verbatim}
+
+In the case of the first function, the outer function, notice the spurious function node in the body.
+This is intentional.
+When we lift these functions, we will replace each spurious function node with a variable node referring to the function’s generated name.
+
+Each of these grouped sub-trees will be given a new local root \verb;F; node that is at the appropriate depth (0) and uses the key as its path vector (\verb;⊣; or \verb;⍺;).
+
+\begin{verbatim}
+      (1↓Fk){⊂(0 'F' 0 ⍺)⍪⍵}⌸1↓Fast
+┌─────────────────┬─────────────────┐
+│┌─┬─┬─┬─────────┐│┌─┬─┬─┬─────────┐│
+││0│F│0│1 0 0 0 0│││0│F│0│1 1 1 0 0││
+│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
+││1│E│0│1 1 0 0 0│││3│E│0│1 1 1 1 0││
+│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
+││2│F│0│1 1 1 0 0│││4│V│⍵│1 1 1 1 1││
+│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
+││2│A│0│1 1 2 0 0│││4│P│+│1 1 1 1 2││
+│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
+││3│N│7│1 1 2 2 0│││4│V│⍵│1 1 1 1 3││
+│└─┴─┴─┴─────────┘│└─┴─┴─┴─────────┘│
+└─────────────────┴─────────────────┘ 
+\end{verbatim}
+
+\noindent At this point, the internal function nodes refer by path vector to the newly added local roots.
+This does not require traversing or maintaining these connections, but falls out as a natural artifact of using the Key (\verb;⌸;) operator.
+At this point we are ready to fix up the sub-tree to be correct.
+Firstly, we need a function to encapsulate the function header information.
+
+\begin{verbatim}
+      hd←0 'F' 0,∘⊂⊣      ⍝ Func. Header
+\end{verbatim}
+
+\noindent Next, we write a function to adjust the depths, which is one more than the difference between each node's depth and the depth of the first node of the sub-tree (\verb;n⌷⍤1; gives the $n$th column).
+
+\begin{verbatim}
+      dp←(1+⊢-⊃)0⌷⍤1⊢     ⍝ New Depths
+\end{verbatim}
+
+\noindent Finally, we use the \verb;@; operator to replace occurances of \verb;F; in the Type column with \verb;V; elements.
+
+\begin{verbatim}
+      rf←'V'@('F'∘=)1⌷⍤1⊢ ⍝ F → V
+\end{verbatim}
+
+\noindent Once we put all this together, we can construct the new sub-trees with the appropriate depths, types, references, and links with the following expression.
+
+\begin{verbatim}
+      (1↓Fk)(⊂hd⍪dp,rf,2↓⍤1⊢)⌸1↓Fast
+┌─────────────────┬─────────────────┐
+│┌─┬─┬─┬─────────┐│┌─┬─┬─┬─────────┐│
+││0│F│0│1 0 0 0 0│││0│F│0│1 1 1 0 0││
+│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
+││1│E│0│1 1 0 0 0│││1│E│0│1 1 1 1 0││
+│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
+││2│V│0│1 1 1 0 0│││2│V│⍵│1 1 1 1 1││
+│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
+││2│A│0│1 1 2 0 0│││2│P│+│1 1 1 1 2││
+│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
+││3│N│7│1 1 2 2 0│││2│V│⍵│1 1 1 1 3││
+│└─┴─┴─┴─────────┘│└─┴─┴─┴─────────┘│
+└─────────────────┴─────────────────┘
+\end{verbatim}
+
+Each of these function bodies has a specific path vector associated with it.
+Because these paths are uniquely identifying, we can use these as input into a name generator to generate names that we know are unique for each function body.
+Furthermore, because we retain this information in the corresponding function nodes that appear in the body of each function to be lifted, we know exactly what name that function has been given, and we can replace the function node with a variable node referencing that name instead, without referring to any state outside of the immediate information given to the function lifter.
+Indeed, each row in the above matrix represents a function lifting task that can be completed without any additional information.
+That is, there are no dependencies between rows to perform lifting.
+This gives us a straightforward parallel execution of function lifting.
+
+This function lifting works even with lexically scoped functions, as is the case for the Co-dfns compiler.
+We can do this because the node coordinates maintain the lexical information for each variable reference, enabling us to construct the appropriate lexical binding for each variable much later than would normally be done in a compiler.
+In other words, the above lifting operation loses no information that would prevent us from handling lexical scoping, but decouples lexical scoping from the act of lifting functions.
+
+\subsection{Expression Flattening}
+
+Flattening expressions is almost identical to lifting functions, in part because we have delayed the lexical scope resolution aspect of lifting functions to a later pass that will handle this explicitly.
+In addition to lifting the expressions themselves, however, we will also take the opportunity to handle precedence and ordering.
+In our examples, we are assuming that all expressions evaluate right to left.
+Because the Key (\verb;⌸;) operator is a stable operation, and we require that nodes have a depth-first pre-order traversal pattern, then we can trivially handle precedence order of this type, including handling parentheses, by reversing the results of flattening.
+
+If we use \verb;Ek; as the key for the expression example, we get the following:
+
+\begin{verbatim}
+      (1↓Ek){⍺ ⍵}⌸1↓East
+┌─────┬─────────────┐
+│1 0 0│┌─┬─┬─┬─────┐│
+│     ││1│E│0│1 1 0││
+│     │├─┼─┼─┼─────┤│
+│     ││1│P│÷│1 2 0││
+│     │├─┼─┼─┼─────┤│
+│     ││1│E│0│1 3 0││
+│     │└─┴─┴─┴─────┘│
+├─────┼─────────────┤
+│1 1 0│┌─┬─┬─┬─────┐│
+│     ││2│V│a│1 1 1││
+│     │├─┼─┼─┼─────┤│
+│     ││2│P│+│1 1 2││
+│     │├─┼─┼─┼─────┤│
+│     ││2│V│b│1 1 3││
+│     │└─┴─┴─┴─────┘│
+├─────┼─────────────┤
+│1 3 0│┌─┬─┬─┬─────┐│
+│     ││2│V│c│1 3 4││
+│     │├─┼─┼─┼─────┤│
+│     ││2│P│×│1 3 5││
+│     │├─┼─┼─┼─────┤│
+│     ││2│V│d│1 3 6││
+│     │└─┴─┴─┴─────┘│
+└─────┴─────────────┘
+\end{verbatim}
+
+\noindent Again, we can see immediately that we have grouped each set of nodes according to the expressions that are to be lifted.
+Just as in the case of function lifting, we can adjust the depths of each expression to the correct depth and we can replace each expression node with a reference based on that node’s path.
+Each expression has a unique name based on its path.
+A later compiler pass minimizes the names used in expressions (like register allocation).
+As with the function lifting case, we need some functions to help abstract each new field.
+In our case, we use the same functions as with function lifting but alter the \verb;hd; and \verb;rf; functions to use the right node types:
+
+\begin{verbatim}
+      hd←0'E'0,∘⊂⊣
+      rf←'V'@('E'∘=)1⌷⍤1⊢
+\end{verbatim}
+
+\noindent The resulting expression lifting operation is almost identical to that of function lifting.
+Notice that we reverse the 3-D matrix result at the end along the first axis to preserve order of evaluation correctly.
+
+\begin{verbatim}
+      ⊖(1↓Ek)(hd⍪dp,rf,2↓⍤1⊢)⌸1↓East
+┌─┬─┬─┬─────┐
+│0│E│0│1 3 0│
+├─┼─┼─┼─────┤
+│1│V│c│1 3 4│
+├─┼─┼─┼─────┤
+│1│P│×│1 3 5│
+├─┼─┼─┼─────┤
+│1│V│d│1 3 6│
+└─┴─┴─┴─────┘
+┌─┬─┬─┬─────┐
+│0│E│0│1 1 0│
+├─┼─┼─┼─────┤
+│1│V│a│1 1 1│
+├─┼─┼─┼─────┤
+│1│P│+│1 1 2│
+├─┼─┼─┼─────┤
+│1│V│b│1 1 3│
+└─┴─┴─┴─────┘
+┌─┬─┬─┬─────┐
+│0│E│0│1 0 0│
+├─┼─┼─┼─────┤
+│1│V│0│1 1 0│
+├─┼─┼─┼─────┤
+│1│P│÷│1 2 0│
+├─┼─┼─┼─────┤
+│1│V│0│1 3 0│
+└─┴─┴─┴─────┘
+\end{verbatim}
+
+The only extra issue involved here over function lifting is to ensure that the order of evaluation matches.
+In our case, we are assuming that the order of evaluation is right to left, which means that the above order given by \verb;⌸; is actually backwards of the desired order.
+During recombination, we simply reverse these orders and this fixes that problem.
+More work would be required to take into consideration a specific precedence hierarchy.
+
 \begin{table*}
@@
 \toprule
-Pass & Description & Core Traversal Pattern & Uses coordinates? \\
+Pass & Description & Core Traversal Pattern & Uses paths? \\
 \midrule
-Record Node Coordinates & Adds a new field to each node containing that node's 
- node coordinate. & Outer Product/Scan & Yes \\
+Record Path Vectors & Adds a new field to each node containing that node's 
+ path vector. & Outer Product/Scan & Yes \\
 Record Function Depths & Adds a new field to each node recording how many 
@@
 
-\subsection{The Key Operator}
-
-First introduced in the J language \cite{hui2014key}, the Key operator (written \verb;⌸;) is central to our strategy for compilation.
-It allows us to use the path vectors to their full effect.
-The expression \verb;K f⌸ M; groups rows of \verb;M; by their corresponding keys in \verb;K; and computes \verb;k f m; for each unique key \verb;k; in \verb;K; and the matrix \verb;m; of rows of \verb;M; with key \verb;k;.
-The corresponding rows of matrices \verb;K; and \verb;M; form key-value pairs.
-Let's compute a histogram using Tally (\verb;≢;) and Key (\verb;⌸;):
-
-\begin{verbatim}
-      10⍴5
-5 5 5 5 5 5 5 5 5 5
-      ⊢X←?10⍴5
-1 1 4 0 1 3 1 1 2 1
-      ≢⍳5
-5
-      X(⊣,(≢⊢))⌸X
-1 6
-4 1
-0 1
-3 1
-2 1
-\end{verbatim}
-
-\noindent To understand a bit better how the Key operator applies its function, consider the function \verb;{⍺ ⍵}; which returns the pair of its right and left arguments.
-If we apply it to the same value as above, we get the following:
-
-\begin{verbatim}
-      X{⍺ ⍵}⌸X
-1  1 1 1 1 1 1 
-4  4           
-0  0           
-3  3           
-2  2
-\end{verbatim}
-
-\noindent In our case, we use either \verb;Fk; or \verb;Ek; as our keys applied to the corresponding AST.
-We also will drop off the first row in each AST using \verb;1↓; since this node ``contains'' everything.
-In a complete AST this is usually the Module boundary node which contains the entire set of functions and values in the module.
-
-\section{Case Studies}
-
-The following compiler passes are found in any compiler from a nested function, nested expression language to a language without nested functions or expressions.
-They were the chief motivating examples for developing the techniques presented above as a part of developing the Co-dfns compiler, and represent a challenging problem to data-parallel compilation without the above techniques, but fall out almost effortlessly once the above patterns are available.
-
-Each of the following passes operates over an AST that we represent as the catenation of the field vectors above including the Node Coordinates.
-
-\begin{verbatim}
-Fast←Fd,Ft,Fv,⍪↓Fc
-East←Ed,Et,Ev,⍪↓Ec
-\end{verbatim}
-
-\noindent The function \verb;↓; converts the coordinate matrices from matrices to vectors of vectors, so that the coordinates can be catenated to the other fields, leading to a 4-column matrix for the ASTs, whose fourth column is a vector of vectors.
-
-\subsection{Function Lifting}
-
-Function lifting takes an AST with nested functions (functions appearing inside of other functions) and lifts these functions out of their local scopes and puts them all at the top level.
-Normally, function lifting must consider both the core operation of lifting function as well as the lexical scop-ing of the functions and variables that appear inside of a function.
-As seen below, our treatment uses a slightly different method.
-
-If we use the Key operator (\verb;⌸;) with the pair function (\verb;{⍺ ⍵};) using \verb;Fk; as the keys and \verb;Fast; as the AST, we get the following:
-
-\begin{verbatim}
-      (1↓Fk){⍺ ⍵}⌸1↓Fast
-┌─────────┬─────────────────┐
-│1 0 0 0 0│┌─┬─┬─┬─────────┐│
-│         ││1│E│0│1 1 0 0 0││
-│         │├─┼─┼─┼─────────┤│
-│         ││2│F│0│1 1 1 0 0││
-│         │├─┼─┼─┼─────────┤│
-│         ││2│A│0│1 1 2 0 0││
-│         │├─┼─┼─┼─────────┤│
-│         ││3│N│7│1 1 2 2 0││
-│         │└─┴─┴─┴─────────┘│
-├─────────┼─────────────────┤
-│1 1 1 0 0│┌─┬─┬─┬─────────┐│
-│         ││3│E│0│1 1 1 1 0││
-│         │├─┼─┼─┼─────────┤│
-│         ││4│V│⍵│1 1 1 1 1││
-│         │├─┼─┼─┼─────────┤│
-│         ││4│P│+│1 1 1 1 2││
-│         │├─┼─┼─┼─────────┤│
-│         ││4│V│⍵│1 1 1 1 3││
-│         │└─┴─┴─┴─────────┘│
-└─────────┴─────────────────┘
-\end{verbatim}
-
-Notice that we have now grouped all of the relevant parts of the tree according to which nodes would appear in their respective functions after lifting.
-Refer to the original lifting example in the introduction to verify this.
-Indeed, the second row in the above example shows the internal function complete and ready to name.
-Each element in the second column corresponds to the body of one of our lifted functions.
-In the case of the first function, the outer function, we have a spurious function node in the body.
-This is intentional.
-When we lift these functions, we will replace each spurious function node with a variable node referring to the function’s generated name.
-
-Each of these grouped sub-trees will be given a new local root \verb;F; node that is at the appropriate depth (0) and uses the key as its path vector (\verb;⊣;).
-
-\begin{verbatim}
-      (1↓Fk){⊂(0 'F' 0 ⍺)⍪⍵}⌸1↓Fast
-┌─────────────────┬─────────────────┐
-│┌─┬─┬─┬─────────┐│┌─┬─┬─┬─────────┐│
-││0│F│0│1 0 0 0 0│││0│F│0│1 1 1 0 0││
-│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
-││1│E│0│1 1 0 0 0│││3│E│0│1 1 1 1 0││
-│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
-││2│F│0│1 1 1 0 0│││4│V│⍵│1 1 1 1 1││
-│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
-││2│A│0│1 1 2 0 0│││4│P│+│1 1 1 1 2││
-│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
-││3│N│7│1 1 2 2 0│││4│V│⍵│1 1 1 1 3││
-│└─┴─┴─┴─────────┘│└─┴─┴─┴─────────┘│
-└─────────────────┴─────────────────┘ 
-\end{verbatim}
-
-\noindent At this point, the internal function nodes refer by path vector to the newly added local roots.
-This was down without traversing or maintaining these connections, but falls out as a natural artifact of using the Key (\verb;⌸;) operator.
-At this point we are ready to fix up the sub-tree to be correct.
-Firstly, we need a function to encapsulate the function header information.
-
-\begin{verbatim}
-      hd←0 'F' 0,∘⊂⊣      ⍝ Func. Header
-\end{verbatim}
-
-\noindent Next, we write a function to adjust the depths, which is one more than the difference between each node's depth and the depth of the first node of the sub-tree.
-
-\begin{verbatim}
-      dp←(1+⊢-⊃)0⌷⍤1⊢     ⍝ New Depths
-\end{verbatim}
-
-\noindent Finally, we use the \verb;@; operator to replace occurances of \verb;F; in the Type column with \verb;V; elements.
-
-\begin{verbatim}
-      rf←'V'@('F'∘=)1⌷⍤1⊢ ⍝ F → V
-\end{verbatim}
-
-\noindent Once we put all this together, we can construct the new sub-trees with the appropriate depths, types, references, and links with the following expression.
-
-\begin{verbatim}
-      (1↓Fk)(⊂hd⍪dp,rf,2↓⍤1⊢)⌸1↓Fast
-┌─────────────────┬─────────────────┐
-│┌─┬─┬─┬─────────┐│┌─┬─┬─┬─────────┐│
-││0│F│0│1 0 0 0 0│││0│F│0│1 1 1 0 0││
-│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
-││1│E│0│1 1 0 0 0│││1│E│0│1 1 1 1 0││
-│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
-││2│V│0│1 1 1 0 0│││2│V│⍵│1 1 1 1 1││
-│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
-││2│A│0│1 1 2 0 0│││2│P│+│1 1 1 1 2││
-│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
-││3│N│7│1 1 2 2 0│││2│V│⍵│1 1 1 1 3││
-│└─┴─┴─┴─────────┘│└─┴─┴─┴─────────┘│
-└─────────────────┴─────────────────┘
-\end{verbatim}
-
-Each of these function bodies has a specific path vector associated with it.
-Because these paths are uniquely identifying, we can use these as input into a name generator to generate names that we know are unique for each function body.
-Furthermore, because we retain this information in the corresponding function nodes that appear in the body of each function to be lifted, we know exactly what name that function has been given, and we can replace the function node with a variable node referencing that name instead, without referring to any state outside of the immediate information given to the function lifter.
-Indeed, each row in the above matrix represents a function lifting task that can be completed without any additional information.
-That is, there are no dependencies between rows to perform lifting.
-This gives us a straightforward parallel execution of function lifting.
-
-This function lifting works even with lexically scoped functions, as is the case for the Co-dfns compiler.
-We can do this because the node coordinates maintain the lexical information for each variable reference, enabling us to construct the appropriate lexical binding for each variable much later than would normally be done in a compiler.
-In other words, the above lifting operation loses no information that would prevent us from handling lexical scoping, but decouples lexical scoping from the act of lifting functions.
-
-\subsection{Expression Flattening}
-
-Flattening expressions is almost identical to lifting functions, in part because we have delayed the lexical scope resolution aspect of lifting functions to a later pass that will handle this explicitly.
-In addition to lifting the expressions themselves, however, we will also take the opportunity to handle precedence and ordering.
-In our examples, we are assuming that all expressions evaluate right to left.
-Because the Key (\verb;⌸;) operator is a stable operation, and we require that nodes have a depth-first pre-order traversal pattern, then we can trivially handle precedence order of this type, including handling parentheses, by reversing the results of flattening.
-
-If we use \verb;Ek; as the key for the expression example, we get the following:
-
-\begin{verbatim}
-      (1↓Ek){⍺ ⍵}⌸1↓East
-┌─────┬─────────────┐
-│1 0 0│┌─┬─┬─┬─────┐│
-│     ││1│E│0│1 1 0││
-│     │├─┼─┼─┼─────┤│
-│     ││1│P│÷│1 2 0││
-│     │├─┼─┼─┼─────┤│
-│     ││1│E│0│1 3 0││
-│     │└─┴─┴─┴─────┘│
-├─────┼─────────────┤
-│1 1 0│┌─┬─┬─┬─────┐│
-│     ││2│V│a│1 1 1││
-│     │├─┼─┼─┼─────┤│
-│     ││2│P│+│1 1 2││
-│     │├─┼─┼─┼─────┤│
-│     ││2│V│b│1 1 3││
-│     │└─┴─┴─┴─────┘│
-├─────┼─────────────┤
-│1 3 0│┌─┬─┬─┬─────┐│
-│     ││2│V│c│1 3 4││
-│     │├─┼─┼─┼─────┤│
-│     ││2│P│×│1 3 5││
-│     │├─┼─┼─┼─────┤│
-│     ││2│V│d│1 3 6││
-│     │└─┴─┴─┴─────┘│
-└─────┴─────────────┘
-\end{verbatim}
-
-\noindent Again, we can see immediately that we have grouped each set of nodes according to the expressions that are to be lifted.
-Just as in the case of function lifting, we can adjust the depths of each expression to the correct depth and we can replace each expression node with a variable reference based on that node’s coordinate.
-Each expression can be given a unique name based on its coordinate.
-A later compiler pass can reduce these names down to the minimum actually required to represent the expression if desired.
-As with the function lifting case, we need some functions to help abstract each new field.
-In our case, we use the same functions as with function lifting but alter the \verb;hd; and \verb;rf; functions to use the right node types:
-
-\begin{verbatim}
-      hd←0'E'0,∘⊂⊣
-      rf←'V'@('E'∘=)1⌷⍤1⊢
-\end{verbatim}
-
-\noindent At this point, the expression lifting operation is then almost identical to that of function lifting.
-Notice that we reverse the 3-D matrix result at the end along the first access to ensure that order of evaluation is preserved correctly.
-
-\begin{verbatim}
-      ⊖(1↓Ek)(hd⍪dp,rf,2↓⍤1⊢)⌸1↓East
-┌─┬─┬─┬─────┐
-│0│E│0│1 3 0│
-├─┼─┼─┼─────┤
-│1│V│c│1 3 4│
-├─┼─┼─┼─────┤
-│1│P│×│1 3 5│
-├─┼─┼─┼─────┤
-│1│V│d│1 3 6│
-└─┴─┴─┴─────┘
-┌─┬─┬─┬─────┐
-│0│E│0│1 1 0│
-├─┼─┼─┼─────┤
-│1│V│a│1 1 1│
-├─┼─┼─┼─────┤
-│1│P│+│1 1 2│
-├─┼─┼─┼─────┤
-│1│V│b│1 1 3│
-└─┴─┴─┴─────┘
-┌─┬─┬─┬─────┐
-│0│E│0│1 0 0│
-├─┼─┼─┼─────┤
-│1│V│0│1 1 0│
-├─┼─┼─┼─────┤
-│1│P│÷│1 2 0│
-├─┼─┼─┼─────┤
-│1│V│0│1 3 0│
-└─┴─┴─┴─────┘
-\end{verbatim}
-
-The only extra issue involved here over function lifting is to ensure that the order of evaluation matches.
-In our case, we are assuming that the order of evaluation is right to left, which means that the above order given by \verb;⌸; is actually backwards of the desired order.
-During recombination, we simply reverse these orders and this fixes that problem.
-More work would be required to take into consideration a specific precedence hierarchy.
-
 \section{The Co-dfns Compiler and Other Passes}
 
-The Co-dfns compiler is an experimental commercial compiler funded by Dyalog, Ltd.
-\cite{hsu2014co,hsu2015accelerating} It is intended to provide increased scaling and performance for APL programmers using the dfns syntax (lexically scoped, functionally oriented APL).
+The Co-dfns compiler is an experimental commercial compiler \cite{hsu2014co,hsu2015accelerating}.
+It is intended to provide increased scaling and performance for APL programmers using the dfns syntax (lexically scoped, functionally oriented APL).
 These techniques form one of the foundational elements to the architecture of the compiler, which is built in the style of a Nanopass \cite{keep2013nanopass} compiler, with very small passes chained together through composition, with the added requirements that each compiler pass must be fully data parallel.
@@
 
-Table \ref{tab:passes} contains a selection of passes from the Co-dfns compiler and breaks down their features based on how they traverse the tree and whether or not they use the Key operator or Node Coordinates.
+Table \ref{tab:passes} contains a selection of passes from the Co-dfns compiler and breaks down their features based on how they traverse the tree and whether or not they use the Key operator or Path Matrix.
 The traversal pattern indicates the primary operator used to traverse the AST and perform the transformation, most of which are primitives; the Amend pattern is a ``replacement'' pattern where specific nodes are replaced in the AST with other nodes, and can be thought of as a slightly modified filter and map.
 The table orders the passes with front-end passes higher in the table and passes further down the compiler chain at the bottom of the table.
-These passes give a good idea of how the Key operator plays a part in the complete compiler as well as how Node Coordinates help to deal with complex sub-tree selections.
+These passes give a good idea of how the Key operator plays a part in the complete compiler as well as how Path vectors help to deal with complex sub-tree selections.
 
 We make a few notes on the table and the analysis.
-The Key operator can be used with or without the node coordinates structure.
+The Key operator can be used with or without the path matrix.
 In cases where the AST is sufficiently flat and arranged in a sufficiently convenient order, the grouping operations are very simple.
 In these cases, the Key operator amounts to a map operation over some selected sub-trees selected at a specific depth.
-These occur later in the compiler where the reader will note appearances of the Key operator without the use of Node Coordinates.
-
-Likewise, node coordinates find their place early on in the compiler where it is necessary to deal with the more highly nested AST.
+These occur later in the compiler where the reader will note appearances of the Key operator without the use of Path Vectors.
+
+Likewise, path vectors find their place early on in the compiler where it is necessary to deal with the more highly nested AST.
 They are useful for identifying this information regardless of whether the Key operator is used to handle the grouping or not.
-Thus passes like ``Drop Unreachable Code'' that remove whole sub-trees from the AST may use the Node Coordinates to do the grouping once the correct parent node has been identified.
-
-In short, the goal of the structure and design of the compiler passes given in Table \ref{tab:passes} is to utilize Node Coordinates to remove as quickly as possible the nested structure of the AST into a flatter form that does not require complex analysis to traverse.
+Thus passes like ``Drop Unreachable Code'' that remove whole sub-trees from the AST may use the Path Matrix to do the grouping once the correct parent node has been identified.
+
+In short, the goal of the structure and design of the compiler passes given in Table \ref{tab:passes} is to utilize Path Vectors to remove as quickly as possible the nested structure of the AST into a flatter form that does not require complex analysis to traverse.
 
@@
 This is a very common method, taught in compiler courses that teach flattening passes, as well as used in various compilers.
-The benefit of the functional algorithm is the concision with which it may be written in most higher-level languages, particularly those which eschew mutatoin and favor a pure coding style.
+The benefit of the functional algorithm is the concision with which it may be written in most higher-level languages, particularly those which eschew mutation and favor a pure coding style.
 The disadvantage is that it has a higher memory usage requirement and may copy data as it moves up through the call stack to reassemble various containers for child nodes.
@@
 
-Normally, one would test GPU oriented benchmarks based on their input sizes.
-However, in this case, the traditional algorithms and the GPU algorithms are written quite differently.
-Thus, we introduce two dimensions of input size in order to examine differences with more precision.
-Particularly, we benchmarked the GPU algorithm against the traditional methods by depth of the AST and Branching Factor.
+We use a slightly more complex, but optimized path matrix than the one given above.
+In particular, we require that each non-zero element of a path vector be a valid row index into the path matrix, and that this index be a reference to the corresponding ancestor of the given depth for the node corresponding to the given path vector.
+By ensuring that each element is a valid index, we are able to optimize the question of whether a given parent has a given node type.
+Instead of the more expensive  \verb;∧.(=∨0=⊢); based general solution, we can use the expression \verb;⌈/C×t=T[C]; to acquire suitable keys to group our sub-trees for a type \verb;t;, path matrix \verb;C;, and type vector \verb;T;.
+This optimized expression is not as general, but it covers a common case.
+The critical path for computing the nearest parent of a given type for the AST is less than depth of the AST, depending on how one implements the above expression.
+
+We benchmarked the GPU algorithm against the traditional methods by depth of the AST and branching factor.
 We chose to examine depth as a specific factor because the path matrix is inherently dependent on the depth.
 Furthermore, the mutation algorithm is optimized in such a way that fewer expression nodes may result in fewer writes to the accumulator, while our algorithm has no such optimization, and traverses the entire tree independent of the number of expression nodes.
-
 We ran the benchmark to compute the time to flatten an AST on all combinations of depth and branching factor along the interval $[0,32)$ that were of a suitable size (that is, fit within the machine constraints).
@@
 
-There are many options for representing graphs on data-parallel architecture.
+There are many options for representing graphs on data-parallel architectures.
 Most of them tend to optimize traversal instead of transformation.
@@
 In the case of the expression flattening above, we could likely achieve equal or better performance from using a single parent vector for this one computation.
-However, would be increasing costs elsewhere, particularly in maintenance.
+However, we would increase costs elsewhere, particularly in maintenance.
 A path matrix provides both parent information, as well as information about the depth of the node, the siblings, position in the tree, and a unique identifier in a single structure that is cache/memory friendly.
@@
 
-We do not have a large scale understanding of the performance characteristics of these techniques in real life outside of daily use of the compiler.
-Further analysis is required to understand what sort of compiler optimizations and analyses are necessary to compile the Co-dfns compiler efficiently for the GPU and CPU.
-We currently lack a good comparison against existing techniques both for the compiler design and the generated code.
-Some existing compilers are able to compile variants of APL to target the GPU and the CPU, but there are subtle differences that require care when comparing generated code, and the architecture of these compilers is often quite different.
-Comparing the relative performance both of the compiler designs (human factors and execution speed) and generated code is important to understanding these techniques more fully.
-
 While we have found this method of compiler construction to be very compact and to permit working with a data-parallel compiler as easily or nearly as easily as with a standard Nanopass compiler, we have not done extensive studies to determine just how comparable in programmability, ease of maintenance, or extensibility these techniques are with established compiler construction methods, such as OOP Visitors, Nanopass, or the type-directed functional style.
@@
 
-We have derived a method of performing computation over sub-trees selected on the basis of inter-node properties through the use of the Key (\verb;⌸;) operator and node coordinates, which enable local computation of these inter-node properties.
+We have derived a method of performing computation over sub-trees selected on the basis of inter-node properties through the use of the Key (\verb;⌸;) operator and path vectors, which enable local computation about these inter-node properties.
 This method is both general and direct, and when combined with traditional and more mundane array programming, suffices to implement the complete core of a compiler, modulo parsing and code generation.
@@
 It presents a solution to a very old and traditional problem in a very uncommon light, by eschewing the common practices that underlie every other significant and general solution found in modern compilers today and replacing them with an entirely different paradigm centered on parallelism and aggregate operations.
+This new paradigm shows improved performance for transformation heavy algorithms which form the backbone of any non-trivial compiler on data-parallel or vector hardware such as GPUs.
 \balancecolumns
@@
 
-\begin{table*}
-\centering
-\begin{tabular}{cllcll}
-\toprule
-Symbol                   & Monadic            & Dyadic  &
-Symbol                   & Monadic            & Dyadic  \\
-\midrule
-\multicolumn{6}{c}{Scalar Functions} \\ 
-\midrule
-\texttt{+}               & Identity           & Plus (Add) &
-\texttt{\textasciitilde} & Not                & \\
-\texttt{-}               & Negative           & Minus (Subtract) &
-\texttt{?}               & Roll               & \\
-\texttt{×}               & Direction (Signum) & Times (Multiply) &
-\texttt{∧}               &                    & And \\
-\texttt{÷}               & Reciprocal         & Divide &
-\texttt{∨}               &                    & Or \\
-\texttt{|}               & Magnitude          & Residue (Modulo) &
-\texttt{⍲}               &                    & Nand \\
-\texttt{⌊}               & Floor              & Minimum &
-\texttt{⍱}               &                    & Nor \\
-\texttt{⌈}               & Ceiling            & Maximum &
-\texttt{<}               &                    & Less \\
-\texttt{*}               & Exponential        & Power &
-\texttt{≤}               &                    & Less Or Equal \\
-\texttt{⍟}               & Natural Logarithm  & Logarithm &
-\texttt{=}               &                    & Equal \\
-\texttt{○}               & Pi Times           & Circular (Trigonometric) &
-\texttt{≥}               &                    & Greater Or Equal \\
-\texttt{!}               & Factorial          & Binomial &
-\texttt{>}               &                    & Greater \\
-\texttt{≠}               &                    & Not Equal \\
-\midrule
-\multicolumn{3}{c}{Selection Mixed Functions} &
-\multicolumn{3}{c}{Structural Mixed Functions} \\
-\cmidrule(r){1-3} \cmidrule(l){4-6}
-\texttt{⊃}               & Disclose & Pick &
-\texttt{⍴} & Reshape       & \\
-\texttt{↑}               &          & Take &
-\texttt{,} & Ravel         & Catenate/Laminate \\
-\texttt{↓}               &          & Drop &
-\texttt{⍪} & Table         & Catenate First/Laminate \\
-\texttt{/}               &          & Replicate &
-\texttt{⌽} & Reverse       & Rotate \\
-\texttt{⌿}               &          & Replicate First &
-\texttt{⊖} & Reverse First & Rotate First \\
-\texttt{\textbackslash}  &          & Expand &
-\texttt{⍉} & Transpose     & Transpose \\
-\texttt{⍀}               &          & Expand First &
-\texttt{↑} & Mix           & \\
-\texttt{\textasciitilde} &          & Without (Excluding) &
-\texttt{↓} & Split         & \\
-\texttt{∩}               &          & Intersection &
-\texttt{⊂} & Enclose       & Partitioned Enclose \\
-\texttt{∪}               & Unique   & Union &
-\texttt{∊} & Enlist        & \\
-\texttt{⊣}               & Same     & Left \\
-\texttt{⊢}               & Identity & Right\\
-\midrule
-\multicolumn{3}{c}{Selector Mixed Functions} &
-\multicolumn{3}{c}{Miscellaneous Mixed Functions} \\
-\cmidrule(r){1-3} \cmidrule(l){4-6}
-\texttt{⍳} & Index Generator & Index Of &
-\texttt{⍴} & Shape         & \\
-\texttt{∊} &                 & Membership &
-\texttt{≡} & Depth         & Match \\
-\texttt{⍋} & Grade Up        & Grade Up &
-\texttt{≢} & Tally         & Not Match \\
-\texttt{⍒} & Grade Down      & Grade Down &
-\texttt{⍎} & Execute       & Execute \\
-\texttt{?} &                 & Deal &
-\texttt{⍕} & Format        & Format \\
-\texttt{⍷} &                 & Find &
-\texttt{⊥} &               & Decode (Base) \\
-& & & \texttt{⊤} &               & Encode (Representation) \\
-& & & \texttt{⌹} & Matrix Divide & Matrix Inverse \\
-\end{tabular}
-\caption{Primitive Functions}
-\label{tab:scalarprims}
-\end{table*}
-
-\begin{table*}
-\centering
-\begin{tabular}{cll}
-\toprule
-Symbol     & Name & Description \\
-\midrule
-\texttt{⍨} & Commute & Swaps arguments or distributes
- right argument to both sides \\
-\texttt{¨} & Each & Applies its operand point-wise over the left/right
- arguments \\
-\texttt{/} & Reduce & Reduce along the last axis \\
-\texttt{⌿} & Reduce First & Reduce along the first axis \\
-\texttt{\textbackslash} & Scan & Scan along the last axis \\
-\texttt{⍀} & Scan First & Scan along the first axis \\
-\texttt{⌸} & Key & Apply operand once for each sub-array grouped by key \\
-\end{tabular}
-\caption{Primitive Monadic/Unary Operators, each takes a single left
-operand and describes a function operating over one or two arguments}
-\label{tab:adverbs}
-\end{table*}
-
-\begin{table*}
-\centering
-\begin{tabular}{cll}
-\toprule
-Symbol     & Name & Description \\
-\midrule
-\texttt{∘} & Compose & Composes two operands as in traditional mathematics \\
-\texttt{.} & Inner Product & Inner product operation, e.g.
- \texttt{+.×} for matrix multiplication \\
-\texttt{∘.} & Outer Product & Cartesian product or ``function table'' \\
-\texttt{⍣} & Power & Iteration, Limited use only \\
-\texttt{⍤} & Rank & Apply a function along cells of an array \\
-\end{tabular}
-\caption{Primitive Dyadic/Binary Operators, each takes a left
- and right operand and describes a function operating over one or two arguments}
-\label{tab:conjunctions}
-\end{table*}
-
 \end{document}

### 2016-11-16 2017-spring-pldi/hsu-2017-key_data_parallel_compiler.tex
Republication Policy conformance

@@
 
-In the following sections we describe the core methods and idioms surrounding the AST encoding, construction of the Path Matrix, and the use of these structures in handling nested, recursive AST relationships.
-This enables us to perform arbitrary computation over arbitrary sub-trees of an AST selected by parent-child relationships in the manner often seen in compiler transformations.
+In the following sections we extend, expound, and clarify on existing work \cite{hsu2016key} to describe the core methods and idioms surrounding the AST encoding, construction of the Path Matrix, and the use of these structures in handling nested, recursive AST relationships, which enable us to perform arbitrary computation over arbitrary sub-trees of an AST selected by parent-child relationships in the manner often seen in compiler transformations.
 These ideas have been further implemented and tested through the implementation of a complete data-parallel by construction compiler, called Co-dfns, that compiles a lexically scoped, functionally oriented dialect of APL with nested functions \cite{hsu2014co,hsu2015accelerating}.
 The compiler targets both the GPU and the CPU, and its core is implemented in this pure, restricted language that uses only function composition over data-parallel primitives.
-
-To illustrate the performance of these techniques, we micro-benchmark a transformation heavy expression lifting algorithm that is found almost universally in compilers.
+To illustrate the performance of these techniques, we benchmark a transformation heavy expression lifting algorithm that is found almost universally in compilers.
 This benchmark shows performance speedups of up to a factor of 58 when compared against pure, functionally oriented algorithms.
@@
 
-\item We describe a method of computing over arbitrary sub-trees selected by their parent-child relationships in a data-parallel manner using the Key operator and a Path Matrix.
+\item We expound on a method \cite{hsu2016key} of computing over arbitrary sub-trees selected by their parent-child relationships in a data-parallel manner using the Key operator and a Path Matrix.
 

### 2016-11-17 2017-spring-pldi/hsu-2017-key_data_parallel_compiler.tex
More adjustment to the Republication policy stuff.

@@
 
-In the following sections we extend, expound, and clarify on existing work \cite{hsu2016key} to describe the core methods and idioms surrounding the AST encoding, construction of the Path Matrix, and the use of these structures in handling nested, recursive AST relationships, which enable us to perform arbitrary computation over arbitrary sub-trees of an AST selected by parent-child relationships in the manner often seen in compiler transformations.
 These ideas have been further implemented and tested through the implementation of a complete data-parallel by construction compiler, called Co-dfns, that compiles a lexically scoped, functionally oriented dialect of APL with nested functions \cite{hsu2014co,hsu2015accelerating}.
 The compiler targets both the GPU and the CPU, and its core is implemented in this pure, restricted language that uses only function composition over data-parallel primitives.
-To illustrate the performance of these techniques, we benchmark a transformation heavy expression lifting algorithm that is found almost universally in compilers.
+We extend and improve on prior work \cite{hsu2016key} by simplifying the core expressions and examining specific case studies in greater detail than previously done.
+We extend the methodology with performance-centered optimizations of the Path Matrix and benchmark a transformation heavy expression lifting algorithm that is found almost universally in compilers.
 This benchmark shows performance speedups of up to a factor of 58 when compared against pure, functionally oriented algorithms.
@@
 
-\item We provide analysis of these techniques as used in a commercial compiler project called Co-dfns and report on the results, including the overall architecture of the compiler, the uses of these techniques within the compiler, and the demonstration of these techniques applied to two specific compiler passes.
+\item We provide analysis of these techniques as used in a commercial compiler project called Co-dfns and report on the results, including the overall architecture of the compiler, the uses of these techniques within the compiler, and the demonstration of these techniques applied to two specific compiler passes in full detail.
 

### 2022-01-15 2017-spring-pldi/response.txt
2017 PLDI responses

Thank you to all of the reviewers. There are significant misunderstandings to clarify, importantly, the intent and scope of this paper. We are not writing a compiler for a data parallel language, we are using an existing data parallel language as a means of writing any arbitrary compiler, indeed, any arbitrary set of synergistic tree transformation (as opposed to traversals). To wit, what would it look like to have an entire compiler natively expressed as a data parallel program such that it ran natively and entirely on the GPU? We focus on transformations over traversals since the literature does very little to provide concrete techniques for implementing any arbitrary tree transformation natively on the GPU in a concise form. Our techniques scale to compilers of programming languages, document processing, web processing, database query optimization; we focus our scope on the easily delineated area of classic compiler transformations.

The entire Co-dfns compiler is written in this style, with Table 1 in the paper giving the primary compiler passes. Every pass has already been implemented in this data parallel style, summarized in the Table. Importantly, having implemented the entire compiler in this manner, distinguish code size between it and a compiler of similar complexity written in a traditional compiler-writing DSL (Nanopass). The Co-dfns compiler is 150 LoC compared to the published Nanopass compiler at 2,000 - 4,000 LoC. This is a significant result. Usually, the opposite is true. In the case of Scan, the smallest, cleanest code example given in the literature for a parallel prefix sum (which is expressed in 1 line of C for sequential/CPU versions) is 100 LoC. This is a 100 times increase in code size to move to the GPU. In this paper, we not only can natively express arbitrary compiler transformations on the GPU, but we find an order of magnitude *decrease* in code size when doing so, which is in stark contrast to the status quo for GPU implementations of non-trivial algorithms.

Regarding notation, a critical element is that the notation is not just a mathematical abstraction over the actual implementation. The expressions in the paper are literally the code and full implementation of these ideas in executable form. The entire code is available and explicated in the paper, requiring no additional code. The expressions can be copied into a session (tryapl.org) and executed without modification. The implementations given for expression flattening and function lifting are the actual implementation of these compiler passes. The reader has no need for additional implementation. The notation is a precise template for implementing these algorithms in any programming language from C and beyond.

Full code over traditional mathematical notation avoids any ambiguity for the uninitiated. Readers may not be familiar with data parallel array programming as a "style" for implementing tree transformations. Readers may struggle to replace their traditional techniques with these without the benefit of full and complete code available with copious examples. The development of the path matrix is an example. We would not expect most developers to readily conceive of a data parallel algorithm for generating the path matrix by a simple description alone. With the exposition, it becomes obvious (even seemingly trivial, which is the goal). Readers need not question whether some detail is omitted. It is trivial, almost rote, to convert this code to the reader's own language of choice. We are not aware of any other standard and readily available notation which is simultaneously concise, executable, and semantically generic enough to serve this purpose within the space constraints we face in the paper.

Real world benchmarks are future work, but the synthetic benchmark is important given the scope. These techniques are not specific or limited to traditional compiler transformations and scale to any range of arbitrary tree transformations. A real world benchmark does not clarify behavior across the range of inputs. The synthetic benchmark distills the comparison of pattern matching/recursion vs. key/path matrix to give a clear idea of how they compare. Real world depends on the local domain. We did not have space to provide enough detail on the various passes in a real compiler to provide an adequate and clear picture of what would be benchmarked if we used real APL code as our input.

The benchmark results drop to zero as we increase the depth/branching factor because we did not test that combination due to the time limitations on running the code of that size (the time on the CPU grew too excessive).

A reviewer requested a plain English description of a path matrix, which can be found in the second full paragraph of section 3.2. The description given by the reviewer is close to the description given in the paper already, but is too specific. We intentionally allow for a wide range of path matrix implementations to suit specific domain requirements. The last sentence in the proposed description about the depth node and the left node is true in the path matrix version given in the paper, but not true in other versions of a path matrix optimized for different cases.

We are not sure whether reviewer B referring to the performance of "other benchmarks" refers to other compiler passes written in the style given in the paper or the traditional methods. If the reviewer is speaking of the other passes in the Co-dfns compiler, we should note that the entire compiler is written in the same style as the given passes here, and there is no "local" view style programming in the compiler. It is all fully data parallel style.

A reviewer remarks that we are taking a local view regarding compile time being not a primary concern, over programmer productivity and high performance. We are actually concerned with productivity and performance, and are taking a broad, instead of local view. We are presenting a general suite of techniques for the productive and performant implementation of any sort of tree transformation in a natively data parallel, vector processor friendly way that actually reduces code size and provides improved performance.

### 2022-01-15 2019-jio/mapper_benchmark-master/src/main/java/com/xnsio/MapperMain.java
2019 JIO

@@ -1,0 +1,58 @@
+package com.xnsio;
+
+import com.xnsio.filter.BucketCustomers;
+import generated.CustomerRecords;
+import generated.QueryResult;
+
+import javax.xml.bind.JAXBContext;
+import javax.xml.bind.JAXBException;
+import javax.xml.bind.Unmarshaller;
+import java.io.File;
+import java.io.IOException;
+import java.net.URISyntaxException;
+import java.util.ArrayList;
+import java.util.List;
+
+import static java.lang.Runtime.getRuntime;
+import static java.lang.System.nanoTime;
+
+public class MapperMain {
+
+    public static final int ITERATIONS = 10_000;
+
+    public static void main(String[] args) throws URISyntaxException, IOException, JAXBException, InterruptedException {
+
+        QueryResult result = buildInput();
+
+        execute(result);
+    }
+
+    private static List<CustomerRecords> execute(QueryResult result) throws InterruptedException {
+        Runtime runtime = getRuntime();
+        System.gc();
+        Thread.sleep(2000);
+        long startTime = nanoTime();
+        long initialMemoryUsed = memUsed(runtime);
+        List<CustomerRecords> perfResults = new ArrayList<>(ITERATIONS);
+        for (int i = 0; i < ITERATIONS; i++) {
+            perfResults.add(new BucketCustomers().segregate(result));
+        }
+        long timeElapsed = nanoTime() - startTime;
+        long afterMemoryUsed = memUsed(runtime);
+        System.out.print("Mem consumed " + (afterMemoryUsed - initialMemoryUsed) / (1024 * 1024) + " MB");
+        System.out.println(" & Exec-time: " + (timeElapsed / 1000000) + " milli-secs");
+        return perfResults;
+    }
+
+    private static long memUsed(Runtime runtime) {
+        return runtime.totalMemory() - runtime.freeMemory();
+    }
+
+    private static QueryResult buildInput() throws URISyntaxException, JAXBException {
+        JAXBContext jaxbContext = JAXBContext.newInstance(QueryResult.class);
+
+        Unmarshaller unmarshaller = jaxbContext.createUnmarshaller();
+
+        return (QueryResult) unmarshaller.unmarshal(MapperMain.class.getClassLoader().getResource("input.xml"));
+    }
+}

### 2022-01-15 2019-jio/mapper_benchmark-master/src/main/java/com/xnsio/MyBenchmark.java

@@ -1,0 +1,37 @@
+/*
+ * Copyright (c) 2005, 2014, Oracle and/or its affiliates. All rights reserved.
+ * DO NOT ALTER OR REMOVE COPYRIGHT NOTICES OR THIS FILE HEADER.
+ *
+ * This code is free software; you can redistribute it and/or modify it
+ * under the terms of the GNU General Public License version 2 only, as
+ * published by the Free Software Foundation.  Oracle designates this
+ * particular file as subject to the "Classpath" exception as provided
+ * by Oracle in the LICENSE file that accompanied this code.
+ *
+ * This code is distributed in the hope that it will be useful, but WITHOUT
+ * ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+ * FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+ * version 2 for more details (a copy is included in the LICENSE file that
+ * accompanied this code).
+ *
+ * You should have received a copy of the GNU General Public License version
+ * 2 along with this work; if not, write to the Free Software Foundation,
+ * Inc., 51 Franklin St, Fifth Floor, Boston, MA 02110-1301 USA.
+ *
+ * Please contact Oracle, 500 Oracle Parkway, Redwood Shores, CA 94065 USA
+ * or visit www.oracle.com if you need additional information or have any
+ * questions.
+ */
+
+package com.xnsio;
+
+import org.openjdk.jmh.annotations.Benchmark;
+
+public class MyBenchmark {
+
+    @Benchmark
+    public void testMethod() {
+        // place your benchmarked code here
+    }
+
+}

### 2022-01-15 2019-jio/mapper_benchmark-master/src/main/java/com/xnsio/filter/BucketCustomers.java

@@ -1,0 +1,53 @@
+package com.xnsio.filter;
+
+import generated.CustomerRecords;
+import generated.QueryResult;
+
+import java.util.List;
+
+public class BucketCustomers {
+    public CustomerRecords segregate(QueryResult queryResult) {
+        CustomerRecords customerRecords = new CustomerRecords();
+        CustomerRecords.EvenData evenData = new CustomerRecords.EvenData();
+        CustomerRecords.OddData oddData = new CustomerRecords.OddData();
+        for (QueryResult.Record record: queryResult.getRecord()) {
+            List<CustomerRecords.EvenData.Customer> evenCustomers = evenData.getCustomer();
+            CustomerRecords.EvenData.Customer evenCustomer = new CustomerRecords.EvenData.Customer();
+
+            List<CustomerRecords.OddData.Customer> oddCustomers = oddData.getCustomer();
+
+            CustomerRecords.OddData.Customer oddCustomer = new CustomerRecords.OddData.Customer();
+            for (QueryResult.Record.Property property :
+                    record.getProperty()) {
+                if (property.getKey() % 2 == 0) {
+                    List<CustomerRecords.EvenData.Customer.Key> keys = evenCustomer.getKey();
+                    CustomerRecords.EvenData.Customer.Key even = new CustomerRecords.EvenData.Customer.Key();
+                    even.setId(property.getKey());
+                    even.setValue(property.getValue());
+                    keys.add(even);
+                } else {
+                    List<CustomerRecords.OddData.Customer.Key> keys = oddCustomer.getKey();
+                    CustomerRecords.OddData.Customer.Key odd = new CustomerRecords.OddData.Customer.Key();
+                    odd.setId(property.getKey());
+                    odd.setValue(property.getValue());
+                    keys.add(odd);
+                }
+            }
+            if (!evenCustomer.getKey().isEmpty()) {
+                evenCustomer.setId(record.getId());
+                evenCustomers.add(evenCustomer);
+            }
+            if(!oddCustomer.getKey().isEmpty()) {
+                oddCustomer.setId(record.getId());
+                oddCustomers.add(oddCustomer);
+            }
+        }
+        if (!evenData.getCustomer().isEmpty()) {
+            customerRecords.setEvenData(evenData);
+        }
+        if (!oddData.getCustomer().isEmpty()) {
+            customerRecords.setOddData(oddData);
+        }
+        return customerRecords;
+    }
+}

### 2022-01-15 2019-jio/mapper_benchmark-master/src/test/java/com/xnsio/filter/BucketCustomersTest.java

@@ -1,0 +1,49 @@
+package com.xnsio.filter;
+
+import generated.CustomerRecords;
+import generated.QueryResult;
+import org.junit.jupiter.api.Test;
+
+import java.util.List;
+
+import static org.junit.jupiter.api.Assertions.*;
+
+class BucketCustomersTest {
+
+    @Test
+    void even_customer_exists() {
+        QueryResult queryResult = new QueryResult();
+        List<QueryResult.Record> records = queryResult.getRecord();
+        QueryResult.Record someCustomer = new QueryResult.Record();
+        someCustomer.setId("WEBDW");
+        List<QueryResult.Record.Property> properties = someCustomer.getProperty();
+        QueryResult.Record.Property p = new QueryResult.Record.Property();
+        p.setKey(2);
+        p.setValue("BSEFQPINLICBHSHO");
+        properties.add(p);
+        records.add(someCustomer);
+
+        CustomerRecords segregatedRecords = new BucketCustomers().segregate(queryResult);
+        assertNotNull(segregatedRecords.getEvenData());
+        assertEquals("WEBDW", segregatedRecords.getEvenData().getCustomer().get(0).getId());
+        assertEquals("BSEFQPINLICBHSHO", segregatedRecords.getEvenData().getCustomer().get(0).getKey().get(0).getValue());
+    }
+    @Test
+    void odd_customer_exists() {
+        QueryResult queryResult = new QueryResult();
+        List<QueryResult.Record> records = queryResult.getRecord();
+        QueryResult.Record someCustomer = new QueryResult.Record();
+        someCustomer.setId("WEBDW");
+        List<QueryResult.Record.Property> properties = someCustomer.getProperty();
+        QueryResult.Record.Property p = new QueryResult.Record.Property();
+        p.setKey(1);
+        p.setValue("ECIVFZLEQIWSZBRD");
+        properties.add(p);
+        records.add(someCustomer);
+
+        CustomerRecords segregatedRecords = new BucketCustomers().segregate(queryResult);
+        assertNotNull(segregatedRecords.getOddData());
+        assertEquals("WEBDW", segregatedRecords.getOddData().getCustomer().get(0).getId());
+        assertEquals("ECIVFZLEQIWSZBRD", segregatedRecords.getOddData().getCustomer().get(0).getKey().get(0).getValue());
+    }
+}

### 2022-05-04 2022-fnconf-architecture/weenote.js
FnConf 2022

@@ -1,0 +1,71 @@
+onload = function() {
+  var body = document.body
+  var slides = {}
+  var slide
+
+  function fit(el) {
+    var style = el.style
+    var i = 1000
+    var top
+    var left
+
+    style.display  = "inline"
+    style.fontSize = i + "px"
+    style.position = "absolute"
+
+    while (1) {
+      left = innerWidth - el.offsetWidth
+      top  = innerHeight - el.offsetHeight
+
+      if (top > 0 && left > 0) break
+
+      style.fontSize = (i -= i * 0.05) + "px"
+    }
+
+    style.display = "block"
+    style.top     = top / 2 + "px"
+    style.left    = left / 2 + "px"
+  }
+
+  for (var el, count = 0; el = body.firstChild;) {
+    if (el.nodeType == 1) slides[++count] = el
+    body.removeChild(el)
+  }
+
+  body.appendChild(document.createComment(""))
+
+  !function sync() {
+    setTimeout(sync, 50)
+
+    var next = 0 | location.hash.match(/\d+/)
+
+    if (slide == next) return
+
+    next = Math.max(Math.min(count, next), 1)
+    next = slides[location.hash = slide = next]
+
+    body.replaceChild(next, body.firstChild)
+    fit(next)
+  }()
+
+  const FORWARD = 1
+  const BACKWARD = -1
+  const keyToSlideDirection = {
+    "ArrowRight": FORWARD, "ArrowLeft": BACKWARD,
+    "PageDown": FORWARD, "PageUp": BACKWARD
+  }
+
+  document.onkeydown = function(e) {
+    var i = slide + keyToSlideDirection[e.key]
+
+    if (i in slides) location.hash = i
+  }
+
+  document.ontouchstart = function(e) {
+    if (e.target.href) return
+
+    var i = slide + (e.touches[0].pageX > innerWidth / 2 ? 1 : -1)
+
+    if (i in slides) location.hash = i
+  }
+}

### 2022-05-04 2022-icfp-unet/Paper/Response.txt
ICFP 2022 Reviews and initial response draft

The authors would like to express their gratitude for the reviewers thoughtful responses. We first would like to acknowledge the alien appearance of APL expressions and the challenge that any alien paradigm can present to a reader. We do not wish to claim either that the expressions given in the paper should be considered trivial or obvious to the uninitiated. Much as standard mathematical notation or the use of laziness, static typing, and recursion can present a serious potential challenge to a reader who is unpracticed and unfamiliar with their use, the APL language presents a challenge precisely because it delivers a novel and different approach to solving problems.

We believe, however, that this novelty, and its associated challenges, creates such remarkably unexpected and surprising results that the challenge is worth undertaking. Indeed, the results are remarkable enough that even without taking the time to acquire a complete understanding of the APL expressions included in our paper, the very nature of their construction, the very fact that they can be put into a paper of this size while expressing what they express, makes them worthy of, at the very least, an increased awareness and publication. That is to say, we believe that a reader can derive significant value from our paper simply by becoming aware of what can be done by approaching problems in this manner, even if the exact details remain somewhat unclear without more effort, much in the same way that one could appreciate the potential of algebraic notation after seeing what is possible with the notation even if the only form of reasoning known at that point has been longform prose and considering the associated time it would take to learn mathematical notation.

The reviewers have compared the perceived challenge of our APL expressions against the simpler to call PyTorch API. We contest that this strongly misses the point, and is in fact the wrong comparison. They are comparing the "wiring" code for a neural network, which is no more than the threading of data through pre-defined library functions of specific, limited, and constrained scope, against a fully defined, complete, from scratch, zero-library, zero-framework implementation of that same neural network in a programming language that has no notion of neural networks, no notion of machine learning, and was not designed with machine learning in mind. Our implementation more accurately represents a reimplementation of large portions of the PyTorch functionality. It would be more accurate to imagine what it would be like to reimplement PyTorch's functionality from scratch, with the same performance, and then compare that implementation to the APL version.

We include PyTorch as a reference, but were unable to include a directly comparable reimplementation of PyTorch's functionality in another programming language due to space constraints and difficulty of implementation. The authors have some experience doing such things in various languages, including leveraging "supporting tools" like numpy, to implement even simpler machine learning projects than the U-net architecture described here, but even such simpler implementations with the full support of a system like numpy are far too complex, verbose, and difficult, to say nothing of its performance, for inclusion literally and in full within the body of a conference paper. Even less would be the feasability of including a performant reimplementation of PyTorch's functionality as used in this paper in its totality within the body of a single paper. Nobody would even attempt to do this.

Yet, with APL, we are able to include 2 copies of every expression, a typeset and a source-style formatting version, in the paper, in full detail, and these expressions implement the entire semantics of a neural network from the ground up in complete detail. Even more, when we compared these APL expressions to the mathematical expressions used in various online resources on neural network programming to describe the same things, our APL expressions were actually shorter and structurally simpler than the standard mathematical expressions for the same concepts!

Thus, these single APL expressions are sufficiently expressive, complete, and concise that they can subsume both the mathematical expressions associated with the operations as well as libraries like PyTorch, while also achieving performance that is "shockingly" close to that of highly tuned and labored over domain-specific libraries such as PyTorch, without a single domain-specific optimization anywhere.

These results become even more worthy of note by the fact that other than the use of the APL paradigm and its notation there are zero new novel concepts applied. The compiler itself uses no special or novel optimizations. There are no special implementation details, no novel data structures or other programming language semantics. There is no special code generation. The only novel thing is the use of APL as the language/semantics. There are no domain-specific features optimized for neural networks anywhere in the programming stack.

In a paper of this sort, we cannot hope to provide a full treatment of CNNs, so we have chosen a network of sufficient complexity as well as sufficient history that a reader may find ample discussion of the architecture and possible variations elsewhere, and mimicked its implementation as closely to the standard as possible. Likewise, we cannot hope to turn the reader into an expert at APL within the space of a single paper; the language is too alien for that.

What we can do is provide a demonstration of the power of the notation and inspire the reader to pursue such ideas in more depth. We intend to provide extended resources that would not fit in the paper for understanding these ideas, but these will not fit within the space limitations of the paper.

With all of that said, the reviewers have highlighted important points of clarity that will permit us to more fully and simply explain some of the pain points that tripped up the reviewers, and this should greatly simplify the parsing of some of the expressions, though the intuition of the array programming model will still be foreign to some. A few careful simplifications and removals will eliminate many of the specific points brought up by reviewers and simplify the overall paper, while also giving us more space to dedicate to explicating the expressions in more detail.

...

### 2022-05-05 2022-icfp-unet/Paper/Response.txt
Update Response.txt

@@
-The authors would like to express their gratitude for the reviewers thoughtful responses.
-We first would like to acknowledge the alien appearance of APL expressions and the challenge that any alien paradigm can present to a reader.
-We do not wish to claim either that the expressions given in the paper should be considered trivial or obvious to the uninitiated.
-Much as standard mathematical notation or the use of laziness, static typing, and recursion can present a serious potential challenge to a reader who is unpracticed and unfamiliar with their use, the APL language presents a challenge precisely because it delivers a novel and different approach to solving problems.
+The authors thank the reviewers for their thoughtful responses.
 
-We believe, however, that this novelty, and its associated challenges, creates such remarkably unexpected and surprising results that the challenge is worth undertaking.
-Indeed, the results are remarkable enough that even without taking the time to acquire a complete understanding of the APL expressions included in our paper, the very nature of their construction, the very fact that they can be put into a paper of this size while expressing what they express, makes them worthy of, at the very least, an increased awareness and publication.
-That is to say, we believe that a reader can derive significant value from our paper simply by becoming aware of what can be done by approaching problems in this manner, even if the exact details remain somewhat unclear without more effort, much in the same way that one could appreciate the potential of algebraic notation after seeing what is possible with the notation even if the only form of reasoning known at that point has been longform prose and considering the associated time it would take to learn mathematical notation.
+Much like traditional math notation, recursion, static typing, or laziness, APL is cryptic and challenging to the uninitiated.
+We do not expect uninitiated readers to easily apprehend the APL expressions.
+We *do* expect readers to perceive the novelty and impact of our results.
+Ironically, while APL appears complex, PyTorch *is* complex.
 
-The reviewers have compared the perceived challenge of our APL expressions against the simpler to call PyTorch API.
-We contest that this strongly misses the point, and is in fact the wrong comparison.
-They are comparing the "wiring" code for a neural network, which is no more than the threading of data through pre-defined library functions of specific, limited, and constrained scope, against a fully defined, complete, from scratch, zero-library, zero-framework implementation of that same neural network in a programming language that has no notion of neural networks, no notion of machine learning, and was not designed with machine learning in mind.
-Our implementation more accurately represents a reimplementation of large portions of the PyTorch functionality.
-It would be more accurate to imagine what it would be like to reimplement PyTorch's functionality from scratch, with the same performance, and then compare that implementation to the APL version.
+Calling the PyTorch API is easier than using APL, but it is not the comparison we make and we will clarify this point in the final paper.
+The better comparison is our APL implementation vs.
+PyTorch's underlying code.
+Our paper contains the complete, from scratch APL code developed almost entirely by a single developer.
+APL itself has no domain-specifc support for machine learning or neural networks.
+PyTorch contains over 200,000 lines of code from over 2,000 contributors specifically targeting ML.
+We omitted the equivalent implementation of PyTorch for space concerns.
 
-We include PyTorch as a reference, but were unable to include a directly comparable reimplementation of PyTorch's functionality in another programming language due to space constraints and difficulty of implementation.
-The authors have some experience doing such things in various languages, including leveraging "supporting tools" like numpy, to implement even simpler machine learning projects than the U-net architecture described here, but even such simpler implementations with the full support of a system like numpy are far too complex, verbose, and difficult, to say nothing of its performance, for inclusion literally and in full within the body of a conference paper.
-Even less would be the feasability of including a performant reimplementation of PyTorch's functionality as used in this paper in its totality within the body of a single paper.
-Nobody would even attempt to do this.
+Likewise, the APL code is more simple, concise, and well-defined than the equivalent traditional mathematical equations.
+We do not believe APL to be more cryptic to the uninitiated than the same expressions written in traditional math would be to someone uninitiated in traditional math notation.
 
-Yet, with APL, we are able to include 2 copies of every expression, a typeset and a source-style formatting version, in the paper, in full detail, and these expressions implement the entire semantics of a neural network from the ground up in complete detail.
-Even more, when we compared these APL expressions to the mathematical expressions used in various online resources on neural network programming to describe the same things, our APL expressions were actually shorter and structurally simpler than the standard mathematical expressions for the same concepts!
+APL is the only novel technique in our entire computational stack.
+The compiler uses no special optimizations and the runtime depends on no specialized machine learning code in order to see the runtime performance that we do.
 
-Thus, these single APL expressions are sufficiently expressive, complete, and concise that they can subsume both the mathematical expressions associated with the operations as well as libraries like PyTorch, while also achieving performance that is "shockingly" close to that of highly tuned and labored over domain-specific libraries such as PyTorch, without a single domain-specific optimization anywhere.
+These results are surprising and remarkable; the reader can grasp these results without understanding the details of the APL expressions.
+That we did not use any novelty other than APL makes the novelty and impact of our results that much greater.
+We hope that such results inspire and motivate readers to spend the time and effort to dive more deeply into the APL with the help of external resources that we could not fit into the paper but intend to provide.
 
-These results become even more worthy of note by the fact that other than the use of the APL paradigm and its notation there are zero new novel concepts applied.
-The compiler itself uses no special or novel optimizations.
-There are no special implementation details, no novel data structures or other programming language semantics.
-There is no special code generation.
-The only novel thing is the use of APL as the language/semantics.
-There are no domain-specific features optimized for neural networks anywhere in the programming stack.
-
-In a paper of this sort, we cannot hope to provide a full treatment of CNNs, so we have chosen a network of sufficient complexity as well as sufficient history that a reader may find ample discussion of the architecture and possible variations elsewhere, and mimicked its implementation as closely to the standard as possible.
-Likewise, we cannot hope to turn the reader into an expert at APL within the space of a single paper; the language is too alien for that.
-
-What we can do is provide a demonstration of the power of the notation and inspire the reader to pursue such ideas in more depth.
-We intend to provide extended resources that would not fit in the paper for understanding these ideas, but these will not fit within the space limitations of the paper.
-
-With all of that said, the reviewers have highlighted important points of clarity that will permit us to more fully and simply explain some of the pain points that tripped up the reviewers, and this should greatly simplify the parsing of some of the expressions, though the intuition of the array programming model will still be foreign to some.
-A few careful simplifications and removals will eliminate many of the specific points brought up by reviewers and simplify the overall paper, while also giving us more space to dedicate to explicating the expressions in more detail.
-
-...
+As the reviewers note, a paper of this size cannot do everything.
+It can neither teach the reader to rapidly comprehend APL nor elucidate the nuances of CNNs to those uninitiated in APL.
+We can, however, make the contribution of the paper and the exposition of our code as clear as space permits.
+Embracing the reviewers' salutary comments, we propose making the following changes in the final paper to improve the clarity of the paper's contribution and exposition:

### 2022-05-05 2022-icfp-unet/Paper/Response.txt
Update Response.txt

@@
+# RESPONSE
+
 The authors thank the reviewers for their thoughtful responses.
@@
 
-APL is the only novel technique in our entire computational stack.
+APL is the sole novelty in our entire stack.
 The compiler uses no special optimizations and the runtime depends on no specialized machine learning code in order to see the runtime performance that we do.
+Our key result is as much about what we didn't do as what we did.
 
@@
 Embracing the reviewers' salutary comments, we propose making the following changes in the final paper to improve the clarity of the paper's contribution and exposition:
+
+## PROPOSED CHANGES
+
+1.
+Remove all sections and discussions related to different implementations of the stencil function and operator.
+This is unnecessary to understand the main point of the paper and it is less accessible or interesting to a general audience.
+
+2. Simplify the discussion of Rank since we do not use Rank in its full generality.
+
+3. Include fuller descriptions of the APL primitives used include those mentioned by the reviewers such as the turnstile ⊢.
+
+4. We will include type signatures for APL primitives to help assist in building an intuition of the APL code.
+
+5. Discuss our verification of the correctness the APL version, which includes verifying the output of the APL code against the output of the PyTorch implementation.
+
+6. Provide references to external resources that provide a more expanded treatment of the APL code.
+
+7. Clarify the introduction and discussion to focus specifically on our key results and remove discussion around tangential topics.
+
+8. Add an appendix with our APL code transliterated into a Python-esque pseudocode as an aid to parsing the expressions
+
+9. Make it clear that our primary contribution is not comparing the usability of PyTorch and its performance against APL.
+
+10. Make it clear that the compiler is not doing any special optimizations behind the scenes.
+
+## ADDITIONAL, SPECIFIC REVIEWER COMMENTS RESPONSES
+
+Additional comments on direct points made by reviewers:
+
+* Our back-propagation code was written manually, without auto-differentiation.
+
+* We avoided deviating from the U-net architecture and training practices found in the origional paper in order to maximize the applicability of external resources in understanding our architecture and to avoid introducing any novel CNN strategies that might conflate the point of the paper.
+
+* The use of bilinear upsampling would not materially alter the design and appearance of the transposed convolutional layers except to have a little more mathematical operations.
+In APL at least, they don't represent a fundamentally different computational flavor.
+
+* The rules and code are duplicates of one another with the rules describing the pure computation of the layer modulo any threading of values in and out of a layer-store.
+The source code given is the same expression but with the code that gets and sets the stored layers in the network included within the code
+
+* Re: Debugging.
+Dyalog APL includes a complete line-oriented debugger with traps, stops, restarting, stack inspection, frame-localized REPLs, and so forth.
+There are also specific display and graphics utilities for rendering the layers as bitmaps or graphs for tracking their progress during the debugging of the training process
+
+* We chose not to optimize the APL code because we were primarily interested in establishing a baseline of performance to see how the most direct and naive implemenations of the neural network in APL would look and perform without doing any sophisticated work to alter that behavior.
+
+* Re: line 1019, this is part of the sections that we propose deleting in their entirety from the paper
+
+* re: line 1104, We will clarify that we are not trying to equate the APL code to calling PyTorch's API, but rather, comparing APL against what the underlying implementation of PyTorch looks like and its complexity.
+
+* "What is needed is more detail on the algorithm, for non-neural-network
+wizards. Then a *much* more detailed development of how the
+convolutions and other computational elements of the application are
+implemented in APL." 
+  --- We will include references to detailed expositions of the U-net architecture, since a detailed explanation of the algorithm if one is not familiar with neural network programming would not fit within the paper.
+Likewise, we will reference a much more detailed development of the APL code, too.
+Neither of these are appropriate within the paper itself, at the very least because of the space constraints.
+
+* "My dissatisfaction with this paper is that the heart of the paper,
+which is the development of the CNN application in APL, is pretty
+incomprehensible. APL is a tough language to explicate with clarity."
+  --- The final paper will make the results and impact more accessible to those who do not fully grasp the APL.
+A full understanding of the APL involved is not something that the reader should expect to be fast or easy, and we recognize that it can take significant time to do so, more than would be expected of the time to read a paper in which one were more familiar with the languages and ideas.
+We think there is an inherent element of difficulty in this problem because of the nature of the language and programming paradigm that we cannot escape: motivated readers who wish to learn the APL will have to accept that doing so will take a longer time than they might first expect.
+We believe that the reader can still benefit without this heavy investment if they choose not to dive deeply into the APL.

### 2023-05-23 2023-array-unet/Paper/U-net_in_APL.tex
Final stuff for the camera ready requests
Boy that is ugly.

\batchmode
\makeatletter
\def\input@path{{C:/Users/arcfi/OneDrive/Documents/Co-dfns/Papers/2023-array-unet/Paper/}}
\makeatother
\documentclass[10pt,twocolumn,english,format=sigplan,screen,balance]{acmart}
\usepackage{fontspec}
\setmonofont{APL385 Unicode}
\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}
\usepackage{array}
\usepackage{prettyref}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{rotating}
\ifx\hypersetup\undefined
  \AtBeginDocument{%
    \hypersetup{xetex,unicode=true}
  }
\else
  \hypersetup{xetex,unicode=true}
\fi

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.

\setcopyright{rightsretained}

\acmDOI{10.1145/3589246.3595371}

\acmYear{2023}

\copyrightyear{2023}

\acmSubmissionID{pldiws23arraymain-p5-p}

\acmISBN{979-8-4007-0169-6/23/06}

\acmConference[ARRAY '23]{Proceedings of the 9th ACM SIGPLAN International Workshop on Libraries,
Languages and Compilers for Array Programming}{June 18, 2023}{Orlando, FL, USA}{}

\received{2023-03-31}

\received[accepted]{2023-04-21}

\newcommand{\noun}[1]{\textsc{#1}}
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

\makeatother

\usepackage{polyglossia}
\setdefaultlanguage[variant=american]{english}
\begin{document}
\acmBooktitle{Proceedings of the 9th ACM SIGPLAN International Workshop on Libraries, Languages and Compilers for Array Programming (ARRAY '23), June 18, 2023, Orlando, FL, USA}
\begin{CCSXML}
<ccs2012>
  <concept>
    <concept_id>10011007.10011006.10011008.10011009.10010175</concept_id>
    <concept_desc>Software and its engineering~Parallel programming languages</concept_desc> 
    <concept_significance>500</concept_significance>
  </concept>
  <concept>
    <concept_id>10011007.10011006.10011041.10011048</concept_id>
    <concept_desc>Software and its engineering~Runtime environments</concept_desc>
    <concept_significance>500</concept_significance>
  </concept>
</ccs2012>
\end{CCSXML}
\ccsdesc{Software and its engineering~Parallel programming languages}

\ccsdesc{Software and its engineering~Runtime environments}

\keywords{APL, Compilers, Neural Networks, Machine Learning, GPU, Co-dfns}
\title{U-Net CNN in APL}
\subtitle{Exploring Zero-Framework, Zero-Library Machine Learning}
\author{Aaron W. Hsu}
\email{aaron@dyalog.com}
\orcid{0000-0001-9292-7783}
\affiliation{\institution{Dyalog, Ltd.}\city{Bloomington, IN}\country{United States}}
\author{Rodrigo Girão Serrão}
\email{rodrigo@dyalog.com}
\orcid{0009-0009-3361-3835}
\affiliation{\institution{Dyalog, Ltd.}\city{Bramley}\country{United Kingdom}}
\begin{abstract}
The APL notation would appear to be a clear match for convolutional
neural networks, but traditional implementations of APL have lagged
behind the performance of highly tuned, specialized frameworks designed
to execute CNNs on the GPU. Moreover, most demonstrations of APL for
neural networking have involved relatively small examples. We explore
a more complex example in the U-net architecture and utilize a modern
APL compiler with GPU support, Co-dfns, to compare the state of the
art of APL against the current crop of specialized neural network
frameworks in the form of PyTorch. We compare performance as well
as the language design of APL for neural network programming and the
clarity and transparency of the resulting code.

We found that the complete “from scratch” APL source was on
par with the complexity of the PyTorch reference implementation, albeit
more foreign, while being more concise and complete. We also found
that when compiled with Co-dfns, despite the naïve implementation
both of Co-dfns and our own code, performance on the GPU and the CPU
were within a factor of 2.2 - 2.4 times that of the PyTorch implementation.
We believe this suggests significant avenues of future exploration
for machine learning language design, pedagogy, and implementation,
both inside and outside of the APL community. 
\end{abstract}
\maketitle

\section{Introduction}

Specialized machine learning frameworks dominate present deep learning
applications. A wide number of highly specialized, optimized libraries
exist. These systems are more complex than typical libraries and are
better thought of as domain-specific languages (DSLs). While these
libraries bolster the current machine learning explosion, because
of their highly specialized nature, users tend to become specialists
around a very specific framework for computation. This creates a sharp
wall that impedes skills transference, where experts can use frameworks
effectively but may be under-prepared to handle situations that benefit
from broader or more flexible skill-sets.\footnote{To someone who only has a hammer, everything looks like a nail.}
This specialization makes sense in the context of traditional programming,
where simple networks are accessible, but tedious, while non-trivial
ML architectures quickly strain the performance and usability of a
“from scratch” implementation. This creates a sharp contrast
between simple systems programmed “by hand” and opaque, complex
frameworks that work as inflexible black boxes. A lack of profound
and intuitive understanding of the underlying mechanics of deep learning
systems fosters a type of “programming by knob turning” where
networks are programmed via trial and error rather than intentionality,
further exacerbating the danger of unintended consequences, already
an inherent difficulty in statistical ML \citep{domingos2012few}.
Machine learning frameworks are often highly vendor-specific; even
vendor-neutral systems encode greater specificity than warranted for
long-running code. This demands higher levels of programmer investment
in system maintenance over a product lifetime. Despite this, specialist
frameworks prove highly effective, in part because of the performance
demands of ML. However, in recent years, reemerging general-purpose
array programming languages have presented an alternative potential
direction. Such languages were popular during early neural network
programming in the 20th century \citep{alfonseca-1990-aplnn}, but
performance of then-current hardware prevented further development.

APL, a general-purpose array programming language, created by Kenneth
Iverson as an improved mathematical notation \citep{apl}, has risen
in popularity over the past decades, in part because of the renewed
interest in parallel computation and a wider acceptance of the use
of a variety of programming languages. Recently, new research into
APL as a machine learning language has begun. APL's origins in education
and its reputation for direct algorithmic expression \citep{knuth1993computer,knuth2007computer}
address some of the above concerns around general-purpose languages
and ML. As a linguistically stable language that is both high performance
and high level \citep{hsu2019data}, it is suitable for long lived
code as well as rapid prototyping, and the data-parallel by default
semantics suggests obvious applications to GPU programming. While
these advantages might recommend APL for machine learning, the majority
of implementations have run on the CPU only, usually via interpreter.
While challenging to compile, recent innovations to the language (particularly
those with a functional programming focus) make compilation more tractable,
and the Co-dfns compiler now exists as an APL implementation with
native GPU support \citep{hsu2019data}.

Given the available APL technology and the paucity of existing research
on modern ML in APL, we explored APL as a ML language in terms of
language design and runtime performance by implementing and benchmarking
the U-net convolutional neural network \citep{unet}. This is a popular
image segmentation architecture with a particularly interesting U-shaped
design. It makes use of a range of popular CNN vocabularies and functions
while having a clear, non-trivial architecture. This makes an ideal
candidate for exploring APL's capabilities.

We make the following contributions:
\begin{itemize}
\item A from scratch implementation of the U-net convolutional neural network
in APL
\item Our U-net implementation is exceptionally simple, concise, transparent,
and direct
\item Our implementation consists of pure APL and no dependencies, frameworks,
libraries, or other supporting code
\item Performance results of two modern APL implementations, one compiled
and the other interpreted, on CPU and GPU hardware against a reference
PyTorch implementation for U-net
\end{itemize}

\section{Background}

\subsection{Convolutional Neural Networks}

The experiment this paper uses to produce its benchmarks is the reproduction
of a well-known convolutional neural network architecture. The use
of CNNs in machine learning was widely popularized with the publication
of a paper \citep{cnns-imagenet} that used CNNs to achieve state-of-the-art
performance in labeling pictures of the ImageNet \citep{imagenet}
challenge. However, a prominent paper from 1998 \citep{cnns-lecun-doc-recognition}
shows that the modern use of CNNs can be dated farther back.

The use of convolutional neural networks, as we know them today, builds
on top of the convolutional layer \citep{intro-to-cnn}. Convolutional
layers receive 3-dimensional tensors as input and produce three-dimensional
tensors as output. These inputs have a fixed number of channels\footnote{“channel” typically refers to the leading dimension of these
inputs/outputs, a nomenclature that is derived from the fact that
CNNs were popularized in the context of image processing.} $n_{in}$ which are then transformed into $n_{out}$ channels through
means of discrete convolutions with a total of $n_{in}\times n_{out}$
kernels, the learnable parameters of the convolutional layer. One
of the advantages of CNNs is that, although the total number of kernels
$n_{in}\times n_{out}$ depends on the number of input and output
channels, the sizes of the kernels are independent of the size of
the other two dimensions of the inputs. Despite the fact that the
main dynamics of a convolutional layer is governed by discrete convolution
with the learnable kernels, the exact behavior of a convolutional
layer depends on layer parameters like the padding and the stride
used \citep{conv-arithmetic-guide}.

Since CNNs were primarily used in image recognition-related tasks,
convolutional layers were often paired with pooling layers to ease
the recognition of features over small neighborhoods \citep{pooling}
by aggregating low-level information to recognize larger features
of interest, with the rationale that typical image features (e.g.,
the recognition or segmentation of objects, or image labeling) consist
of regions of pixels and not individual pixels.

\subsection{Original U-Net Architecture}

\begin{figure*}
\includegraphics[width=0.6\paperwidth]{0C__Users_arcfi_OneDrive_Documents_Co-dfns_Papers_2023-array-unet_Paper_u-net-architecture.png}

\caption{Original u-net architecture, as seen in the original paper \citep{unet}.
Arrows represent operations between the multi-channel feature maps
represented by the rectangles. The number on top of each rectangle
is its number of channels and the numbers in the lower-left corner
are the $x$ and $y$ dimensions of the feature maps.}
\label{fig:unet-architecture}
\end{figure*}
\citet{unet} introduced the U-net architecture, a novel CNN architecture
that won several biomedical image segmentation challenges. Since the
original implementation in Caffe \citep{caffe}, others have implemented
U-net in numerous other frameworks and systems\footnote{Numbers by \href{https://paperswithcode.com/paper/u-net-convolutional-networks-for-biomedical}{Papers with Code}
as of March, 2022.}, including PyTorch \citep{pytorch}.

Figure \ref{fig:unet-architecture} shows the original diagram that
represents the U-net architecture \citep{unet}. The blue right arrows,
labeled “conv 3x3, ReLU”, represent unpadded convolutions with
$3\times3$ kernels. After each of these convolutions, the size of
the feature map decreases by $2$, from which it can be inferred that
the stride \citep{conv-arithmetic-guide} is $1$. A rectified linear
unit (ReLU) \citep{relu} activation function follows each convolution.
After two such convolutions, max-pooling operations represented by
the red down arrows downsample each feature map to half the size using
a $2\times2$ region with stride $2$.\footnote{This makes the input size somewhat brittle. Specifically, the input
dimensions must be congruent to $12\mod16$} After every downsampling step, the first convolution doubles the
number of channels. The pattern of two convolutions (with ReLUs) followed
by downsampling via max-pooling happens four times and makes up the
contracting path of the network, on the left of the diagram.

After the contracting path (at the diagram bottom) completes, a corresponding
expanding path reduces the channels while increasing the map size.
Expansion uses unpadded $3\times3$ convolutions preceded by an upsampling
operation (green up arrows) that doubles the map size while halving
the channels. We infer that the original authors used a transposed
convolution (with $2\times2$ kernels) of stride $2$ \citep{conv-arithmetic-guide}\footnote{See \citep{up-transposed} for an informal discussion of this inference.}.
Half of the channels to the initial convolution after upsampling come
from the corresponding map on the contracting side of the architecture,
as represented by the gray right long arrows in the middle of the
diagram. The copied maps are center cropped to ensure matching dimensions.
At the end of the contracting path, a $1\times1$ unpadded convolution
reduces the $64$ feature maps to $2$ feature maps (one per class).

\subsection{APL Notation}

APL, introduced by Turing award winner Kenneth E. Iverson in the '60s
\citep{apl}, is an executable mathematical notation \citep{apl-since-78}.
This section introduces the basics of the notation used in this paper,
but \citet{mdapl} more fully describes the language. Online interactive
systems allow for experimentation with the language without needing
to install anything.\footnote{\url{https://tryapl.org}} All notation
used here is compatible with Dyalog APL 18.0\footnote{\url{https://dyalog.com/}}.

\subsubsection{Arrays}

\begin{table}
\caption{Array names according to \emph{rank}.\label{tab:rank-names}}
\begin{tabular}{ll}
\emph{Rank} &
\emph{Name}\tabularnewline
\midrule 
0 &
Scalar\tabularnewline
1 &
Vector\tabularnewline
2 &
Matrix\tabularnewline
3 &
Cuboid\tabularnewline
4 &
Hypercube\tabularnewline
5+ &
Noble\tabularnewline
\bottomrule
\end{tabular}
\end{table}

The primary datatype of an APL system is the array.\footnote{Other types may exist for commercial systems, but these are not treated
here.} In APL, an array is an $N$-dimensional rectangular arrangement of
one or more elements in row-major order according to a list of zero
or more orthogonal dimensions (axes) called its \noun{shape}, with
the number of axes called its \noun{rank}. Table \ref{tab:rank-names}
gives the common names for sub-classes of arrays by rank. Scalars
have exactly one element. Higher ranked, non-empty arrays contain
as many elements as the product of its axes, while those with at least
one zero length axis, called \noun{empty} arrays, have a single element
called the \noun{prototype}, or \noun{fill}, element. For all numeric
arrays, such as those used in this paper, the fill element is \texttt{0},
and it is used by some primitives to extend or resize arrays to larger
shapes. Indices begin at 0, rather than 1.\footnote{Dyalog APL defaults to 1-based indexing.}

The elements of an array consist solely of other arrays, allowing
APL to have a single, uniform datatype.\footnote{This historically contentious design choice will be taken for granted.
Industrious readers may explore alternative designs elsewhere.} To make computation meaningful, APL defines a set of scalar arrays
that correspond to primitive values in the system, such as numbers
and characters. Thus, when we write a number like \texttt{1}, \texttt{3.5},
or \texttt{1e¯8}, this is the scalar array corresponding to that number
in the APL system.

\subsubsection{Expressions}

Expressions consist of a series of literals/variables, applications,
bindings, or assignments. Literal expressions describe an array. A
single numeric value is a literal. Adjacent values, separated by whitespace,
form a vector of those values. For example:
\begin{verbatim}
¯5    ⍝ Scalar negative 5
5 3 1 ⍝ Vector of 3 elements
\end{verbatim}
Note the use of a high minus to indicate a literal negative number
instead of \texttt{-5}, which is the application of the negate function
to \texttt{5}.

Functions can be first or second order. First order functions, called
\noun{functions}, take arrays as arguments and return a single array
value as a result, while second order functions, called \noun{operators,
}take arrays or \noun{functions} as arguments and return a \noun{function
}result. From now on, we use the term \noun{function} to refer to
first-order functions only. Application takes one of the following
infix forms:
\begin{verbatim}
   fn ⍵  ⍝ Monadic function application
 ⍺ fn ⍵  ⍝ Dyadic  function application
⍺⍺ op    ⍝ Monadic operator application
⍺⍺ op ⍵⍵ ⍝ Dyadic  operator application
\end{verbatim}
The terms \noun{monadic} and \noun{dyadic} refer to arity, not monads
in the functional programming sense. We use \texttt{⍺}, \texttt{⍵},
\texttt{⍺⍺}, and \texttt{⍵⍵} to refer to the left and right arguments
(for function application) and operands (for operator application),
respectively. Operators have higher binding precedence than functions,
and operators always associate to the left while functions always
associate to the right. This applies even in cases where traditional
math primitives are used. Consider the following example, where \texttt{⌿}
is a monadic operator, and \texttt{.} a dyadic:
\begin{verbatim}
   +.×⌿-2×3-4+3×5
←→ (((+.×)⌿)(-(2×(3-(4+(3×5))))))
\end{verbatim}
Notice that \texttt{-} is applied both monadically and dyadically.

\subsubsection{Primitives}

APL is known for its primitives: built-in functions and operators
assigned special symbols instead of the standard alphanumeric variable
names given to user-defined values. The examples used in this paper
are expressions of mostly primitive functions and operators.

The largest class of primitives are the \noun{scalar functions}, defined
over a single scalar value or values and extended to all arrays by
point-wise, also called element-wise, extension. The left and right
arguments of a scalar function must share the same shape, with scalar
arguments extended to the same shape as the other argument. Consider
the following example using addition:
\begin{verbatim}
      5+1 3 7 10 ⍝ Scalar extension
6 8 12 15
\end{verbatim}
Scalar functions return arrays the same shape as their inputs. The
set of scalar functions used in this paper include the standard arithmetic
functions \texttt{+}, \texttt{×}, \texttt{÷}, \texttt{-}, \texttt{⌈}
(ceiling or maximum), \texttt{⌊} (floor/minimum), \texttt{|} (absolute
value), and \texttt{{*}} (exponent), as well as the Boolean relations
\texttt{<}, \texttt{≤}, \texttt{=}, \texttt{≠}, \texttt{≥}, \texttt{>},
and \texttt{\textasciitilde} (not) which return \texttt{1} for \noun{true}
and \texttt{0} for \noun{false}.

The core primitive \texttt{⍴} is used for the shape of an array. The
monadic application \texttt{⍴⍵} returns a vector of the axes of \texttt{⍵}.
The dyadic application \texttt{⍺⍴⍵} returns an array whose shape is
\texttt{⍺} and whose elements are taken in wrapping order from \texttt{⍵}.
For example:
\begin{verbatim}
      3 5⍴1 2 3 4 ⍝ A 3×5 matrix
1 2 3 4 1
2 3 4 1 2
3 4 1 2 3
\end{verbatim}
The function \texttt{≢⍵} applied monadically returns the length of
the first axis of \texttt{⍵}, that is, the first element of \texttt{⍴⍵}. 
\begin{verbatim}
      ≢7 5 6⍴0
7
\end{verbatim}
Thus, the rank of an array is given by \texttt{≢⍴⍵}.
\begin{verbatim}
      ≢⍴7 5 6⍴0
3
\end{verbatim}
The expression \texttt{⍳N} gives the natural numbers up to \texttt{N}.
\begin{verbatim}
      ⍳7
0 1 2 3 4 5 6
\end{verbatim}
The primitives \texttt{⊖} and \texttt{⌽} rotate arrays along the first
and last axis, respectively. Applied monadically, they reverse/flip
an array along the appropriate axis. When applied dyadically, they
shift an array by the specified amount. Thus \texttt{¯2⌽⍵} rotates
(with wrapping) \texttt{⍵} two positions to the right along the last
axis. Writing \texttt{⌽{[}x{]}} or \texttt{⊖{[}x{]}}specifies rotation
along axis \texttt{x}. 
\begin{verbatim}
      ¯1⊖3 5⍴1 2 3 4 ⍝ Shift down 1 row
3 4 1 2 3
1 2 3 4 1
2 3 4 1 2
\end{verbatim}
The transpose primitive \texttt{⍉} permutes its argument's axes into
different positions. Monadic \texttt{⍉⍵} reverses the shape of \texttt{⍵},
while giving \texttt{⍉} a left argument indicates how to permute the
dimensions. Thus, \texttt{0 1 3 2⍉⍵} transposes a hypercube \texttt{⍵}
by exchanging the last two axes; likewise, \texttt{3 0 1 2⍉⍵} transposes
\texttt{⍵} so that its first axis becomes its last, while all other
dimensions retain the same relative order. The following transposes
two matrix slices (sub-arrays) of a cube:
\begin{verbatim}
      2 3 3⍴0 1 2 3 4
0 1 2
3 4 0
1 2 3

4 0 1
2 3 4
0 1 2
      0 2 1⍉2 3 3⍴0 1 2 3 4
0 3 1
1 4 2
2 0 3

4 2 0
0 3 1
1 4 2
\end{verbatim}
The functions \texttt{↑} and \texttt{↓} resize \texttt{⍵} by taking
or dropping elements. The expression \texttt{2 ¯3↑⍵} will give a \texttt{2×3}
matrix from \texttt{⍵} with rows taken from the top and columns taken
from the right, since a negative size takes from the tail instead
of the head. Drop (\texttt{↓}) works the same way, but eliminates
instead of taking items. The following example centers a \texttt{2×2}
matrix into a \texttt{4×4} matrix. Notice how the matrix is extended
using \texttt{0} as the \noun{fill} element.
\begin{verbatim}
      2 2⍴1 2 3 4
1 2
3 4
      3 3↑2 2⍴1 2 3 4
1 2 0
3 4 0
0 0 0
      ¯4 ¯4↑3 3↑2 2⍴1 2 3 4
0 0 0 0
0 1 2 0
0 3 4 0
0 0 0 0
\end{verbatim}
The expression \texttt{⍺,⍵} catenates \texttt{⍺} and \texttt{⍵} together
as a single array along the last axis. However, in this paper we primarily
use \texttt{,} monadically. The expression \texttt{,⍵} gives a vector
of the elements of \texttt{⍵}, collapsing the axes of \texttt{⍵} into
a single axis whose size is the product of the axes of \texttt{⍵}.
We write \texttt{,{[}d{]}⍵} to collapse the specified axes \texttt{d}
to a single axis. The following example collapses the trailing two
axes of a cube, creating a matrix:
\begin{verbatim}
      2 3 3⍴0 1 2 3 4
0 1 2
3 4 0
1 2 3

4 0 1
2 3 4
0 1 2
      ,[1 2]2 3 3⍴0 1 2 3 4
0 1 2 3 4 0 1 2 3
4 0 1 2 3 4 0 1 2
\end{verbatim}
We write \texttt{N⌿⍵} or \texttt{N/⍵} to mean the array \texttt{⍵}
with each element duplicated \texttt{N} times along the first or last
axis, respectively:
\begin{verbatim}
      2 2⍴1 2 3 4
1 2
3 4
      2⌿2/2 2⍴1 2 3 4
1 1 2 2
1 1 2 2
3 3 4 4
3 3 4 4
\end{verbatim}
The expressions \texttt{f⌿⍵} and \texttt{f/⍵} compute the reduction
of \texttt{⍵} using function \texttt{f} over the first or last axis,
respectively. This is often called a right fold in many functional
languages.
\begin{verbatim}
      +⌿1 2 3 4 5
15
\end{verbatim}
The dyadic application \texttt{+.×} is the function for matrix multiplication
extended to higher dimensions.

We can mutate the elements of an array or bind the result of an expression
to a name. The form \texttt{V←expr} binds \texttt{V} to the result
of \texttt{expr}. The expression \texttt{A{[}i{]}}←\texttt{expr} assigns
the elements of \texttt{expr} to the corresponding elements of \texttt{A}
according to indices \texttt{i}. If we have a nested array \texttt{A}
in which we want to store a single non-scalar array \texttt{X} at
index \texttt{i}, we box the value \texttt{X} into a scalar first,
using \texttt{⊂}, resulting in the expression \texttt{A{[}i{]}←⊂X}.
We write \texttt{i⊃A} to extract the same array \texttt{X} from \texttt{A}.

The syntax \texttt{\{expr+ ...\}} returns a function. Within the body
of the function, consisting of a set of expressions, \texttt{⍺} is
the left argument to the function and \texttt{⍵} is the right, while
\texttt{∇} refers to the function itself, permitting anonymous recursion.
Functions execute expressions in order and return the first unbound
expression or the last bound expression.
\begin{verbatim}
      add1←{1+⍵}
      add1 3 4 5
4 5 6
\end{verbatim}

\section{Implementation}

\subsection{Overview}

Our implementation of U-net divides into two significant considerations:
the implementation of the fundamental vocabulary of neural networks
and the wiring of those operations into the actual U-net architecture.
We leverage features of APL to implement both aspects of the system,
and so we treat each in its own sub-section.

Because Co-dfns does not yet support the complete array of Dyalog
primitives and their semantics, we could enhance some implementation
techniques through the use of the richer feature-set. The effect of
these features would be increased concision and clarity, but such
improvements would not significantly affect the overall performance
of the code, either positively or negatively. The overall structure
of the code is clear and simple enough in its current state to warrant
inclusion almost verbatim, but some adjustments were made to reduce
the total number of primitives used, which makes the code slightly
less elegant while reducing the burden to the reader to learn more
APL primitives.

One area deserving particular attention is the design of APL as a
language itself and the features that present themselves as well-suited
to expressing neural network computations. Our exploration of these
features uncovered design tensions worth discussing in detail. A complete
copy of the code discussed in this paper is included in the appendices.

\subsection{Design of APL Primitives for Neural Networks}

The majority of APL primitives find fundamental use in computing neural
networks, which isn't surprising given the array-oriented and numerical
nature of the domain. However, the stencil operator, introduced in
Dyalog APL \citep{stencil-lives}, stands out as the most obviously
aligned with convolutional neural networks. The J programming language
introduced an alternative implementation of the stencil operator earlier
\citep{stencil-in-J}, from which Dyalog derived inspiration for the
implementation of their own stencil operator. We propose an alternative
that was first suggested to us by the late Roger Hui, the stencil
\noun{function} \citep{SF-hui-blog}. The stencil function is a function
whose left argument is the same as the right operand of the stencil
operator, and which receives the same right argument as the right
argument to the function returned by the stencil operator. A reasonable
definition (padded) of the stencil function using the Dyalog stencil
operator might be:
\begin{verbatim}
SF←{({⍵}⌺⍺)⍵}
\end{verbatim}
A naïve implementation of the stencil function that did not pad its
results was implemented in Co-dfns and used in U-net implementation.
We subsequently use \texttt{⌺} to mean the stencil \noun{function}
without padding and not the stencil \noun{operator} as it appears
in Dyalog APL. Given a specification \texttt{s} as the left argument,
the stencil function, written \texttt{s⌺a}, returns an array of each
sliding window slice of \texttt{a} specified by \texttt{s}. The first
row of the specification indicates the window sizes while the second
row indicates the step size, or 1 if omitted. The window is conceptually
“slid” along the leading axes of the input array. The two most
common sliding window sizes for \texttt{⌺} in U-net are \texttt{3
3} for convolutions, corresponding to a window size of \texttt{3×3}
and a step of \texttt{1} for each axis, and \texttt{2 2⍴2} (a 2 by
2 matrix of 2's) for the max pooling layers and up convolutions, corresponding
to a \texttt{2×2} window size and a step of \texttt{2} for each axis.

For example, a \texttt{3 3} sliding window of step 1 on a \texttt{6
6} matrix returns an array of shape \texttt{4 4 3 3} consisting of
16 \texttt{3 3} matrix slices arranged in a \texttt{4 4} grid. The
dimensions of the two leading axes will shrink by 2 each time because
we implemented stencil without padding.
\begin{verbatim}
      6 6⍴⍳36
 0  1  2  3  4  5
 6  7  8  9 10 11
12 13 14 15 16 17
18 19 20 21 22 23
24 25 26 27 28 29
30 31 32 33 34 35
      ⍴3 3⌺6 6⍴⍳36
4 4 3 3
\end{verbatim}
In the following, we box each \texttt{3 3} matrix slice using \texttt{⊂{[}2
3{]}} to show the slices in a more compact form as a nested matrix
of matrices, but the result of \texttt{⌺} is actually a rank 4 hypercube
in this case.
\begin{verbatim}
      ⊂[2 3]3 3⌺6 6⍴⍳36
┌────────┬────────┬────────┬────────┐
│ 0  1  2│ 1  2  3│ 2  3  4│ 3  4  5│
│ 6  7  8│ 7  8  9│ 8  9 10│ 9 10 11│
│12 13 14│13 14 15│14 15 16│15 16 17│
├────────┼────────┼────────┼────────┤
│ 6  7  8│ 7  8  9│ 8  9 10│ 9 10 11│
│12 13 14│13 14 15│14 15 16│15 16 17│
│18 19 20│19 20 21│20 21 22│21 22 23│
├────────┼────────┼────────┼────────┤
│12 13 14│13 14 15│14 15 16│15 16 17│
│18 19 20│19 20 21│20 21 22│21 22 23│
│24 25 26│25 26 27│26 27 28│27 28 29│
├────────┼────────┼────────┼────────┤
│18 19 20│19 20 21│20 21 22│21 22 23│
│24 25 26│25 26 27│26 27 28│27 28 29│
│30 31 32│31 32 33│32 33 34│33 34 35│
└────────┴────────┴────────┴────────┘
\end{verbatim}
Notice that the window slides along the leading axes, so each slice
captures all trailing axes for each slice.

\subsection{Neural Network Vocabulary\label{subsec:Neural-Network-Vocabulary}}

The original U-net paper uses five distinct operations to describe
the network (see \prettyref{fig:unet-architecture}):
\begin{enumerate}
\item A \texttt{3×3} convolution with a ReLU activation function is used
as the primary operation
\item A copy and crop operation is used to transfer data across one row
of the network
\item Max pooling layers on a \texttt{2×2} window are used to compute “down”
the network
\item A \texttt{2×2} transposed convolution goes back “up” the network
\item The final output has a single \texttt{1×1} convolution with a soft-max
layer
\end{enumerate}
In our implementation, we mirror this vocabulary by implementing the
forward and back functions for each of these layers, one for each
of the above operations. This results in a total of 10 functions grouped
into 5 pairs, which we will take in turn.

Each of our operations works over a given network “layer,” which
is a cube of shape \texttt{N M C} where \texttt{N M} are the layer
dimensions and \texttt{C} is the number of layer channels.

\subsubsection{Convolution (3×3) with ReLU}

The primary U-net convolutional kernel is a 3×3 convolution with a
ReLU activation function. The convolution in the paper uses “valid”
convolutions, meaning that no padding is used. This implies that the
convolution dimensions of the output array shrink by 2 for each dimension
compared to the input. We define the forward propagation function
\texttt{CV} as a function over a kernel \texttt{⍺} of shape \texttt{3
3 I O} of \texttt{I} input channels and \texttt{O} output channels
with a layer \texttt{⍵} of shape \texttt{N M I} that obeys the following
shape invariant:
\begin{verbatim}
⍴⍺ CV ⍵ ←→ (¯2+2↑⍴⍵),¯1↑⍴⍺
\end{verbatim}
This says that the size of the first two axes of \texttt{⍺ CV ⍵} will
be two less than the first two axes of \texttt{⍵}, while the remaining
axis, corresponding to the output channels, will match the last axis
of \texttt{⍺}. In other words, given the kernel of shape \texttt{3
3 I O} and layer of shape \texttt{N M I}, the result of \texttt{⍺
CV ⍵} will have shape \texttt{(N-2) (M-2) O}. Using the stencil function,
we define \texttt{CV} as follows for rank 4 kernel inputs and rank
3 layer inputs:
\begin{verbatim}
CV←{0⌈(,[2 3 4]3 3⌺⍵)+.×,[0 1 2]⍺}
\end{verbatim}
The result of the expression \texttt{3 3⌺⍵} will have a rank of 5
and shape \texttt{(N-2) (M-2) 3 3 I}, so raveling the trailing dimensions
with the expression \texttt{,{[}2 3 4{]}3 3⌺⍵} gives us a cube of
shape \texttt{(N-2) (M-2) (3×3×I)} while the expression \texttt{,{[}0
1 2{]}⍺} gives us a matrix of shape \texttt{(3×3×I) O}. We use the
extended matrix multiplication \texttt{+.×} to compute the convolution,
resulting in a cube of shape \texttt{N M O}. The ReLU function \texttt{0⌈⍵}
follows as the final operation.

Computing backpropagation uses very similar approaches. Given the
output layer \texttt{z}, input layer \texttt{x}, activation layer
\texttt{a}, weights \texttt{⍺}, and the gradient backpropagated so
far \texttt{⍵}, we compute the transposed weights \texttt{w}, the
derivative output layer \texttt{∆z}, the weight gradient \texttt{∆w},
padded output layer \texttt{∆Z}, and the resulting back gradient \texttt{∆x}
as follows:
\begin{verbatim}
∆CV←{
     w←⊖⌽[1]0 1 3 2⍉⍺
    ∆z←⍵×0<a
    ∆w←3 0 1 2⍉(⍉,[0 1]∆z)+.×,[0 1]3 3⌺x
    ∆Z←¯2⊖¯2⌽[1](4+2↑⍴∆z)↑∆z
    ∆x←(,[2 3 4]3 3⌺∆Z)+.×,[0 1 2]w
}
\end{verbatim}
Since our stencil function does not pad its results, the expression
\texttt{¯2⊖¯2⌽{[}1{]}(4+2↑⍴∆z)↑∆z} expands the shape of \texttt{∆z}
to ensure that the output convolution dimensions are 2 greater than
those of \texttt{∆z}.

\subsubsection{Copy and Crop}

Conceptually, the copy and crop operation is the simplest of the functions
in U-net. Its sole job is to take the output from one side of the
U-shaped net and move it over to the other side, adjusting the dimensions
to ensure that it fits. In the forward direction, the input layer
will have a greater dimension than the output layer, so we crop as
evenly as possible around the edges and then catenate the result at
the head of the layer coming “up” from the network to form the
output layer with twice the channels of the “up” layer. The
following function \texttt{CC} computes the crop of \texttt{⍺} catenated
with \texttt{⍵} using \texttt{↓}:
\begin{verbatim}
CC←{
    p←((⍴⍺)-⍴⍵)÷2),⍵
    ((⌊p)↓(-⌈p)↓⍺
}
\end{verbatim}
For dimensions not evenly divisible by two, we choose to round up
on the right and bottom sides and round down on the left and upper
sides of the layer. Computing the backpropagation of \texttt{CC} given
the input \texttt{⍺} and output gradient \texttt{⍵} simply reverses
this operation and expands the shape back to the original input size.
This result is then added to the appropriate layer in the U-net architecture
described in \prettyref{subsec:U-net-Architecture}. 
\begin{verbatim}
∆CC←{
    n m←-⌊(2↑(⍴⍺)-⍴⍵)÷2
    ∆x←n⊖m⌽[1](⍴⍺)↑⍵
}
\end{verbatim}

\subsubsection{Max Pooling}

Max pooling is a shrinking convolution that computes the maximum value
in a non-overlapping sliding window. Given the stencil function, the
max pool over a layer is given by the following definition for \texttt{MX}:
\begin{verbatim}
MX←{⌈⌿[2],[2 3](2 2⍴2)⌺⍵}
\end{verbatim}
Here we write \texttt{⌈⌿{[}2{]},{[}2 3{]}⍵} to describe an array where
we have collapsed dimensions 2 and 3 and computed the maximum value
reduction over the resulting dimension. For example, given an input
layer \texttt{⍵} of shape \texttt{N M C}, the result of \texttt{(2
2⍴2)⌺⍵} is a rank 5 array of shape \texttt{(N÷2) (M÷2) 2 2 C}. We
then collapse axes 2 and 3 to form an array of shape \texttt{(N÷2)
(M÷2) 4 C} and subsequently find the maximum value for each vector
along axis 2, resulting in an array of shape \texttt{(N÷2) (M÷2) C}.

Computing the backpropagation of this involves replicating each of
the stencil dimensions, which are the two leading axes in our implementation.
Given an input \texttt{⍺} and output layer \texttt{⍵} the following
function \texttt{∆MX} computes the backpropagation:
\begin{verbatim}
∆MX←{
    y←(⍴⍺)↑2⌿2/[1]⍵
    y×⍺=y
}
\end{verbatim}

\subsubsection{Transposed Convolution (2×2)}

The upsampling computation with convolution proved to be the most
subtle and challenging, mostly due to the opaqueness of implementations.
The U-net paper was not immediately transparent regarding the exact
operations used for this layer and there were a number of potential
design decisions that could have been made. Moreover, for users reading
about upsampling through convolutions, the descriptions are also the
furthest removed from a reasonable implementation of the same. However,
once the intuition hits that an upsampling convolution matches the
shape and form of a non-overlapping sliding window in the output layer,
the computation becomes much clearer.\footnote{\href{https://mathspp.com/blog/til/033}{https://mathspp.com/blog/til/033}}

For this convolution, we change the anticipated kernel shape from
that used for \texttt{CV} above. Whereas \texttt{CV} expects kernels
of shape \texttt{3 3 I O}, our transposed convolutions expect kernels
of shape \texttt{I 2 2 O} for input channels \texttt{I} and output
channels \texttt{O}. With a layer of standard shape \texttt{N M I},
this gives the following definition for the upsampling pass:
\begin{verbatim}
UP←{,[0 1],[2 3]0 2 1 3 4⍉⍵+.×⍺}
\end{verbatim}
The key change here from the reliance on \texttt{+.×} with \texttt{CV}
is the use of a dyadic transpose. The input into the \texttt{⍉} transpose
is the result of \texttt{⍵+.×⍺} which has shape \texttt{N M 2 2 O}.
Thus, the expression \texttt{0 2 1 3 4⍉⍵+.×⍺} returns an array of
shape \texttt{N 2 M 2 O}. As the final operation, we collapse the
first two pairs of leading dimensions, giving a final output array
of shape \texttt{(N×2) (M×2) O}.

To compute the backpropagation pass, we compute the convolutions on
a \texttt{2×2} sliding window with stride 2.
\begin{verbatim}
∆UP←{
    ∆w←(⍉,[0 1]x)+.×,[0 1](2 2⍴2)⌺∆z
    ∆x←(,[2 3 4](2 2⍴2)⌺∆z)+.×⍉,[1 2 3]⍺
}
\end{verbatim}

\subsubsection{Final 1×1 Convolution}

The final convolution is a \texttt{1×1} convolution with 2 output
channels, which means that it collapses the final incoming channels
into an output layer with only two channels. This gives the trivial
simplification of our convolution code over layer \texttt{⍵} and kernel
\texttt{⍺} of \texttt{⍵+.×⍺}, which we combine with the soft-max layer
described in the paper:
\begin{verbatim}
C1←{
    z←⍵+.×⍺
    z←*z-[0 1]⌈/z
    1E¯8+z÷[0 1] +/z
}
\end{verbatim}
Computing the backpropagation is likewise a simplification of the
more complex \texttt{CV} code:
\begin{verbatim}
∆C1←{
    ∆w←(⍉,[0 1]x)+.×,[0 1]∆z
    ∆x←∆z+.×⍉w
}
\end{verbatim}
\begin{table*}
\caption{A rectangular arrangement of the U-net network\label{tab:A-rectangular-arrangement}}

\begin{tabular}{crrrlrrrlrr}
 &
 &
\multicolumn{7}{c}{\noun{Operation}} &
 &
\tabularnewline
\cmidrule{3-9} \cmidrule{4-9} \cmidrule{5-9} \cmidrule{6-9} \cmidrule{7-9} \cmidrule{8-9} \cmidrule{9-9} 
 &
 &
\multicolumn{1}{l}{$CV$} &
\multicolumn{1}{l}{$CV$} &
$MX$ &
 &
\multicolumn{1}{l}{$CV$} &
\multicolumn{1}{l}{$CV$} &
$UP$ &
 &
\tabularnewline
\cmidrule{3-5} \cmidrule{4-5} \cmidrule{5-5} \cmidrule{7-9} \cmidrule{8-9} \cmidrule{9-9} 
\multirow{4}{*}{\begin{turn}{90}
\noun{Depth}
\end{turn}} &
{\tiny{}0} &
\texttt{\tiny{}$\texttt{3\ 3\ \ \ 1\ \ 64}$} &
{\tiny{}$\texttt{3 3 \ 64 \ 64}$} &
{\tiny{}$\texttt{0 0 \ 64 \ 64}$} &
 &
{\tiny{}$\texttt{3 3 \ 256 \ 128}$} &
{\tiny{}$\texttt{3 3 \ 128 \ 128}$} &
{\tiny{}$\texttt{\ 128 2 2 \ 64}$} &
 &
\tabularnewline
 & {\tiny{}1} &
\multirow{1}{*}{{\tiny{}$\texttt{3\ 3\ \ 64\ 128}$}} &
{\tiny{}$\texttt{3 3 128 128}$} &
{\tiny{}$\texttt{0 0 128 128}$} &
 &
{\tiny{}$\texttt{3 3 \ 512 \ 256}$} &
{\tiny{}$\texttt{3 3 \ 256 \ 256}$} &
{\tiny{}$\texttt{\ 256 2 2 128}$} &
 &
\tabularnewline
 & {\tiny{}2} &
{\tiny{}$\texttt{3 3 128 256}$} &
{\tiny{}$\texttt{3 3 256 256}$} &
{\tiny{}$\texttt{0 0 256 256}$} &
 &
{\tiny{}$\texttt{3 3 1024 \ 512}$} &
{\tiny{}$\texttt{3 3 \ 512 \ 512}$} &
{\tiny{}$\texttt{\ 512 2 2 256}$} &
 &
\tabularnewline
 & {\tiny{}3} &
{\tiny{}$\texttt{3 3 256 512}$} &
{\tiny{}$\texttt{3 3 512 512}$} &
{\tiny{}$\texttt{0 0 512 512}$} &
 &
{\tiny{}$\texttt{3 3 \ 512 1024}$} &
{\tiny{}$\texttt{3 3 1024 1024}$} &
{\tiny{}$\texttt{1024 2 2 512}$} &
 &
\tabularnewline
\cmidrule{3-5} \cmidrule{4-5} \cmidrule{5-5} \cmidrule{7-9} \cmidrule{8-9} \cmidrule{9-9} 
 &
 &
\multicolumn{3}{c}{\emph{Downward Pass}} &
 &
\multicolumn{3}{c}{\emph{Upward Pass}} &
 &
\tabularnewline
\cmidrule{3-5} \cmidrule{4-5} \cmidrule{5-5} \cmidrule{7-9} \cmidrule{8-9} \cmidrule{9-9} 
\end{tabular}
\end{table*}

\subsection{U-Net Architecture\label{subsec:U-net-Architecture}}

Given the core vocabularies defined in \prettyref{subsec:Neural-Network-Vocabulary},
the remaining challenge with implementing U-net is to link together
the appropriate layers and compositions to form the complete network
as described by \prettyref{fig:unet-architecture}. To do this, we
observe that the structure of the U-net diagram is an almost symmetric
pattern. The output layer computations form three operations which
are not part of the pattern, but the rest of the pattern decomposes
into 4 depths, each with 6 operations each. \prettyref{tab:A-rectangular-arrangement}
contains a visual arrangement of the kernel shapes used in our architecture
mirroring the overall structure of \prettyref{fig:unet-architecture}.

Additionally, we note that the U-shaped structure also mimics the
down and up nature of a recursive program call-tree. Thus, our overall
strategy is to implement a recursive function \texttt{LA} that receives
an index identifying a particular depth of the network, computes the
appropriate “downward pass” operations before recurring deeper
into the network and finally computing the upwards passes on the return
of its recursive call. We likewise implement backpropagation in same
way, but in the opposite direction. Assuming that \texttt{⍺} contains
the computed depth offset for the network layer, we write \texttt{⍺+i}
to access the $i$th column of the network described in \prettyref{tab:A-rectangular-arrangement}
at the depth \texttt{⍺÷6}.

Our architecture is agnostic to any specific size of kernels or depth
of the “U”; using greater depth with larger resolutions is possible
simply by providing a different network of kernels as input. Many
systems typically allocate networks statically, including our own
PyTorch implementation. Our implementation instead assumes a network
defined by a set of kernels \texttt{W} that match the expected shape.
We mirror the \texttt{W} network with our intermediate results in
\texttt{Z} so that we can do backpropagation. The full code for our
layers extract their values from these \texttt{Z} and \texttt{W} vectors
directly using expressions like \texttt{⍺⊃W} to extract the weights
for the layer at index \texttt{⍺}. We likewise store intermediate
results with expressions like \texttt{Z{[}⍺{]}←⊂⍵}.

Our forward pass function is responsible for initializing \texttt{Z}
to hold the intermediate results produced by forward propagation for
use by the backpropagation function. After the recursive computation,
there are the final three operations, \texttt{C1} and two \texttt{CV}
operations, that must be called before returning. We assume that we
may receive a rank 2 matrix instead of a rank 3 layer as input, and
so we reshape the input to ensure that we always have a rank 3 input
to \texttt{LA}. This gives us the following function definition:
\begin{verbatim}
FWD←{
  Z⊢←(≢W)⍴⊂⍬
  ⍝ Forward propagation layers ...
  LA←{
    ⍺≥≢Z:⍵
    down←(⍺+6)∇(⍺+2)MX(⍺+1)CV(⍺+0)CV ⍵
    (⍺+2)CC(⍺+5)UP(⍺+4)CV(⍺+3)CV down
  }
  2 C1 1 CV 0 CV 3 LA (3↑(⍴⍵),1)⍴⍵
}
\end{verbatim}
The backwards computation mirrors this pattern, except that it proceeds
in the opposite direction and also defines an updater function \texttt{∆}
that will update the network weights in \texttt{W} and the velocities
in \texttt{V} based on a given gradient \texttt{⍵} and index \texttt{⍺}
pointing to a specific location in the network. 
\begin{verbatim}
BCK←{
  Y←⍺ ⋄ Y∆←⍵
  ∆←{
    V[⍺]←⊂⍵+MO×(⍴⍵)⍴⍺⊃V
    W[⍺]←⊂(⍺⊃W)-LR×⍺⊃V
  }
  ⍝ Back propagation layers ...
  ∆LA←{
    ⍺≥≢Z:⍵
    up←(⍺+5)∆UP(-(≢⍉⍵)÷2)↑[2]⍵
    down←(⍺+6)∇(⍺+3)∆CV(⍺+4)∆CV up
    mx←(⍺+2)∆MX down
    (⍺+0)∆CV(⍺+1)∆CV mx+(⍺+2)∆CC ⍵
  }
  diff←Y∆-(~Y),[1.5]Y
  3 ∆LA 0 ∆CV 1 ∆CV 2 ∆C1 diff
}
\end{verbatim}

\section{Performance}

We primarily focused on comparing our U-net implementation against
a reference implemented in PyTorch \citep{pytorch}, which is an easy
to use Python framework with good performance.

\begin{figure*}
\includegraphics[width=1\textwidth]{1C__Users_arcfi_OneDrive_Documents_Co-dfns_Papers_2023-array-unet_Paper_benchmark-comp.pdf}

\caption{Performance results for U-net across a range of platforms\label{fig:Performance-results-for}}
\end{figure*}
We were primarily interested in the costs of executing a single run
of the U-net over a single image source in both the forwards and backwards
directions. We compared performance over the following platforms:
\begin{itemize}
\item Dyalog APL 18.0 64-bit Windows interpreter
\item Co-dfns (v4.1.0 master branch) using the CUDA backend (AF v3.8.1)
\item Co-dfns (v4.1.0 master branch) using the CPU backend (AF v3.8.1)
\item PyTorch v1.10.2 with the CUDA GPU backend
\item PyTorch v1.10.2 with the multi-threaded CPU backend
\item PyTorch v1.10.2 with the single-threaded CPU backend
\end{itemize}
The results of the execution can be seen in \prettyref{fig:Performance-results-for}.
The timings do not include the cost of reading the image data from
disk, but they do include the costs of transferring the image input
data and the resulting error and forward propagation results back
to the CPU main memory. In our testing, data transfer costs in Co-dfns
accounted for significantly less than 5\% of the total runtime.

The hardware used was an NVIDIA GeForce RTX 3070 Laptop GPU with 8GB
of dedicated graphics memory. We used NVIDIA driver version 511.65.
The CPU was an Intel Core i7-10870H with 16 logical cores @ 2.2GHz.
Main system memory was 32GB of DDR4 RAM. The system was running an
up to date release of Microsoft Windows 11.

As input we used the original image data from the ISBI benchmark referenced
in the U-net paper \citep{unet,isbi-data}. These images are $512\times512$
images in grayscale with a binary mask for training. Each run took
one of these images and associated training mask and computed the
result of forward and backwards propagation and the error as well
as updating the weights for the network.

When working on the network, APL implementations generally do not
have a concept of small floating point values. Rather, their default
is to always use 64-bit floating point values when floats are called
for. In order to try to mimic this behavior as closely as possible,
we attempted to feed 64-bit data into the PyTorch models. However,
because of the opaqueness of the PyTorch implementation, we were not
able to fully verify that 64-bit values are used throughout the PyTorch
computational network. On the other hand, the reliance on 64-bit only
floating points, while a boon to convenience and user-friendliness
for non-computer science programmers, creates well-defined performance
issues for an application like this.

When running the benchmark, we computed the average of 10 runs, ensuring
that we discarded the first run each time, since these runs often
contained significant setup and bootstrapping code (PyTorch's optimizer,
the JIT optimization in Co-dfns, and so forth). The figure includes
information about the variance of the individual runs as well as the
average run time in seconds.

Examining the data, it is clear why traditional APL implementations
were relatively unsuited to extensive use within the machine learning
space. Dyalog's interpreter preformed the slowest by a very large
magnitude. After this, the single-threaded CPU implementations in
Co-dfns and PyTorch are predictably the next slowest, with the Co-dfns
implementation running about a factor of 2.2 times slower than the
equivalent PyTorch implementation.

When acceleration techniques are employed, the differences in execution
speed begin to shrink, with PyTorch's multi-threaded and GPU-based
implementations coming in fasted, and Co-dfns' GPU backend running
at roughly 2.4 times slower than the PyTorch GPU execution.

We observed the widest variance in performance results in the Co-dfns
CPU and Dyalog interpreter-based runs, and very little variance in
the GPU-based runs or in PyTorch itself.

The stencil function was modeled in APL and used to conduct the above
benchmark. The model, written in APL, is a naïve implementation of
the stencil function and contains no special optimizations other than
to distinguish between sliding windows of step 1 and non-overlapping,
adjacent windows (such as used for the max pooling layers). Additionally,
no specialized code was used within Co-dfns that was specific or specialized
to neural network programming.

The above benchmark therefore represents a comparison of the PyTorch
implementation against a naïve and unspecialized implementation in
APL executed with the general-purpose runtime used in Co-dfns that
provides generalized GPU computation but does not include domain-specific
optimizations such as those available in PyTorch.

\section{Discussion}

\subsection{Performance}

Clearly, specialized frameworks for deep neural networks are still
the best way to go in order to achieve the absolute maximum in performance
at present. However, our results indicate that the gap between reliance
on specialized frameworks and the freedom to use more general purpose
and transferable programming languages while still achieving competitive
performance is not nearly as large as might have been the case even
a few years ago.

Given that almost zero special optimization is taking place for the
APL implementation executed under the Co-dfns runtime, it is impressive
that we are able to see results that come close to a factor of 2 of
the specialized frameworks. Given some of the obvious design issues
that would contribute to slower performance, it seems more reasonable
to be able to expect more general purpose languages like APL to be
able to reach performance parity with specialized frameworks, without
the requirement that the user learn a special API, or import specialized
dependencies. In more complex applications that leverage APL for other
domain-intensive work, this suggests that APL might facilitate scaling
such applications to integrate machine learning algorithms more easily
and with less programmer effort than might be required to integrate
a separate framework like PyTorch.

\subsection{APL vs. Frameworks}

We have demonstrated that APL itself, without libraries or additional
dependencies, is exceptionally well suited to expressing neural network
computations, at a level of inherent complexity that is arguably equal
or possibly even less than that of the reference PyTorch implementation.
At the very least, it is less code with less underlying background
code and layers. This comes at the admittedly heavy cost of being
completely foreign and alien to most programmers who are more familiar
with languages like Python. This certainly creates a fundamental and
immediate learning cost to APL over other frameworks, since other
frameworks can assume a wider range of foreknowledge around their
chosen language implementation.

It remains unclear, however, whether, if this foreknowledge were taken
away, APL would represent a compelling value proposition for such
programming tasks or not. Indeed, it is exceptionally challenging
to divorce the present reality of prior background knowledge from
such a question. Even fundamental knowledge like what it means to
do array programming and how to structure problems in an array-style
are rarely if ever taught at universities, whereas most classes spend
significant amounts of time teaching students how to utilize the Python-style
programming model of PyTorch.

The argument that APL may be used more widely and broadly than PyTorch
on a wider range of problems using the same skill-set may not matter
to users who are only interested in deep learning algorithms.

APL presently has a higher barrier to entry, but rewards the user
with full and effortless control over what's being done in a way that
other systems do not. This may present itself as a distinct advantage
to users who are looking to expand “off the beaten track” and
utilize novel approaches that do not easily fit within existing frameworks.

We encountered significant difficulties in identifying exactly what
the original authors did based on their paper alone because of many
implementation details that were omitted. On the other hand, APL enables
us to express our entire implementation in a way that makes every
implementation detail clear, improving the ability of others to reproduce
our work.

Finally, in the implementation of U-net in APL, we gained insights
into the architecture that had a direct and material influence on
the PyTorch reference implementation that would not have emerged without
first having studied the APL implementation. Thus, we gained significant
insight simply from doing the APL implementation, even if we were
to re-implement that code in PyTorch.

\section{Related Work}

In addition to the Co-dfns compiler, \citet{bernecky-cnn} have explored
alternative implementations to CNNs, though not U-net specifically.
They focused on APL as a productivity enhancement for CNN development,
and only benchmarked the APL implementation on the CPU using the Dyalog
APL interpreter. They make a case for the usability of APL even with
the performance numbers they achieved. Research in SaC demonstrates
implementations competitive with the current state of the art \citep{sac-cnn}.
While their code exhibits some material differences to that given
here, there are nonetheless some similarities that demonstrate some
level of convergence around implementing CNNs in APL.

Another approach to GPU-based array programming with an APL focus
is the TAIL/Futhark system \citep{futhark}, which is a compiler chain
taking APL to the TAIL (Typed Array Intermediate Language) and then
compiling TAIL code using the Futhark GPU compiler backend. Futhark
has been used to implement deep learning primitives \citep{futhark-deeplearning},
and it represents an interesting approach to compilation of APL via
typed intermediate languages, which have the potential to enhance
the fusion that can be done with an operation like the stencil function.

Other programming environments that are often categorized as array
programming environments, such as Matlab \citep{matlab}, Julia \citep{julia},
and Python/ Numpy \citep{python,numpy}, are not exceptionally performant
on their own for machine learning, but often wrap libraries to do
so. Unfortunately, many of these languages use a syntax that much
more closely mirrors that of Python than APL. In our perspective,
this reduces the value proposition of such languages over using specialized
frameworks, since one does not obtain the particular clarity and productivity
benefits associated with the APL notation.

\section{Future Work}

One of the most obvious questions to answer in future work is the
performance potential of the specialized convolution functions against
our naïve implementation when using the same backend in Co-dfns; initial
tests showed worse results and were omitted here.

There are a number of design elements of the current crop of APL implementations,
including Co-dfns, which hamper performance for machine learning.
Especially, the use of 64-bit floating points without any feature
to reduce their size makes memory usage a concern. Additionally, no
optimization on the design of stencil has been done, while optimizations
related to lazy indexing, batch processing, and a number of other
features seem readily accessible.

Additionally, we would like to explore the potential of using such
systems to improve machine learning pedagogy by encouraging students
to have access to high-performance, but also transparent, implementations
of foundational machine learning concepts. There are still some challenges
to recommending this approach at scale for a large number of educational
institutions, but we believe work on understanding the pedagogical
benefits of APL warrants further research in addition to exploring
APL in the professional space.

\section{Conclusion}

\balance

Given the notational advantages of APL and the concision and clarity
of expression that one can obtain, we explored the potential impact
of using APL as a language for implementing convolutional neural networks
of reasonable complexity. We found that, though the traditional implementations
of APL suffer from performance issues that would prevent widespread
use in either academic, educational, or industrial contexts, compilers
such as Co-dfns are capable of compiling complete neural network programs
(in our case, the U-net architecture) and producing much more competitive
performance results (within a factor of 2.2 - 2.4 times of our reference
PyTorch implementation). This is despite the naïve nature of our implementation
and the naïve optimization support for neural networks on the part
of the Co-dfns compiler.

Furthermore, we found that our effort to implement U-net in APL resulted
in a concise but fully unambiguous and completely transparent implementation,
without any frameworks or library dependencies. Despite being a complete
“by hand” implementation, its complexity of expression is on
par with that of PyTorch and other specialized frameworks, or even
better, particularly in cases where more exploration and novel implementation
is required, or when customized integrations may be called for. The
insights that we gained from implementing U-net in APL affected our
implementation of a reference implementation in PyTorch directly,
suggesting that APL may have significant pedagogical advantages for
teaching neural network programming and machine learning in general.

\appendix

\section{Reference Code and Timings}

\begin{figure*}
\begin{verbatim}
:Namespace UNET

 W←⍬ ⋄ V←⍬ ⋄ Z←⍬ ⋄ LR←1e¯9 ⋄ MO←0.99

 FWD←{Z⊢←(≢W)⍴⊂⍬
  CV←{0⌈z⊣Z[⍺]←⊂Z[⍺],⊂z←(,[2+⍳3]3 3⌺⊃Z[⍺]←⊂⍵)+.×,[⍳3]⍺⊃W}
  CC←{⍵,⍨(⌊p)↓(-⌈p)↓(⍺⊃Z)⊣p←2÷⍨(⍴⍺⊃Z)-⍴⍵}
  MX←{⌈⌿[2],[2 3](2 2⍴2)⌺⊃Z[⍺]←⊂⍵}
  UP←{((2×¯1↓⍴⍵),¯1↑⍴⍺⊃W)⍴0 2 1 3 4⍉⍵+.×⍺⊃W⊣Z[⍺]←⊂⍵}
  C1←{1E¯8+z÷[⍳2]+/z←*z-[⍳2]⌈/z←⍵+.×⍺⊃W⊣Z[⍺]←⊂⍵}
  LA←{⍺≥≢Z:⍵
    down←(⍺+6)∇(⍺+2)MX(⍺+1)CV(⍺+0)CV ⍵
    (⍺+2)CC(⍺+5)UP(⍺+4)CV(⍺+3)CV down}
  2 C1 1 CV 0 CV 3 LA ⍵⍴⍨3↑1,⍨⍴⍵}

 BCK←{Y←⍺ ⋄ Y∆←⍵
  ∆←{0⊣W[⍺]←⊂(⍺⊃W)-LR×⊃V[⍺]←⊂⍵+MO×(⍴⍵)⍴⍺⊃V}
  ∆CV←{w←,[⍳3]⊖⌽[1]0 1 3 2⍉⍺⊃W ⋄ x←⊃⍺⊃Z ⋄ ∆z←⍵×0<1⊃⍺⊃Z
   ∆Z←¯2⊖¯2⌽[1](4+2↑⍴∆z)↑∆z
   _←⍺ ∆ 3 0 1 2⍉(⍉,[⍳2]∆z)+.×,[⍳2]3 3⌺x
   w+.×⍨,[2+⍳3]3 3⌺∆Z}
  ∆CC←{x←⍺⊃Z ⋄ ∆z←⍵ ⋄ d←-⌊2÷⍨2↑(⍴x)-⍴∆z ⋄ (⊃d)⊖(1⊃d)⌽[1](⍴x)↑∆z}
  ∆MX←{x←⍺⊃Z ⋄ ∆z←⍵ ⋄ y×x=y←(⍴x)↑2⌿2/[1]∆z}
  ∆UP←{w←⍺⊃W ⋄ x←⍺⊃Z ⋄ ∆z←⍵ ⋄ cz←(2 2⍴2)⌺∆z
   _←⍺ ∆(⍉,[⍳2]x)+.×,[⍳2]cz
   (,[2+⍳3]cz)+.×⍉⍪w}
  ∆C1←{w←⍺⊃W ⋄ x←⍺⊃Z ⋄ ∆z←⍵ ⋄ _←⍺ ∆(⍉,[⍳2]x)+.×,[⍳2]∆z ⋄ ∆z+.×⍉w}
  ∆LA←{⍺≥≢Z:⍵
   down←(⍺+6)∇(⍺+3)∆CV(⍺+4)∆CV(⍺+5)∆UP ⍵↑[2]⍨-2÷⍨⊃⌽⍴⍵
   (⍺+0)∆CV(⍺+1)∆CV(⍵ ∆CC⍨⍺+2)+(⍺+2)∆MX down}
  3 ∆LA 0 ∆CV 1 ∆CV 2 ∆C1 Y∆-(~Y),[1.5]Y}

 E←{-+⌿,⍟(⍺×⍵[;;1])+(~⍺)×⍵[;;0]}

 RUN←{Y Y∆(Y E Y∆)⊣(Y←⌊0.5+nm↑⍵↓⍨2÷⍨(⍴⍵)-nm←2↑⍴Y∆)BCK⊢Y∆←FWD ⍺}

:EndNamespace
\end{verbatim}
\caption{Complete APL U-Net Implementation}

\end{figure*}

\begin{figure*}
\begin{verbatim}
import torch
import torch.nn as nn
import torchvision
import torchvision.transforms.functional

class TwoConv(nn.Module):

   def __init__(self, in_channels, out_channels):
        super().__init__()

        self.path = nn.Sequential(
            nn.Conv2d(in_channels, out_channels, 
                kernel_size=(3, 3), bias=False),
            nn.ReLU(inplace=True),
            nn.Conv2d(out_channels, out_channels, 
                kernel_size=(3, 3), bias=False),
            nn.ReLU(inplace=True),
        )

    def forward(self, x):
        return self.path(x)
\end{verbatim}
\caption{PyTorch Reference Implementation, Preamble and Helpers}

\end{figure*}
\begin{figure*}
\begin{verbatim}
class Down(nn.Module):

    def __init__(self, in_channels):
        super().__init__()

        self.path = nn.Sequential(
            nn.MaxPool2d(kernel_size=(2, 2), stride=2),
            TwoConv(in_channels, 2 * in_channels),
        )

    def forward(self, x):
        return self.path(x)

class Up(nn.Module):

    def __init__(self, in_channels):
        super().__init__()

        self.upsampling = nn.ConvTranspose2d(
            in_channels,
            in_channels // 2,
            kernel_size=(2, 2),
            stride=2,
            bias=False,
        )
        self.convolutions = 
            TwoConv(in_channels, in_channels // 2)

    def forward(self, x_to_crop, x_in):

        upped = self.upsampling(x_in)
        cropped = torchvision.transforms.functional.center_crop(
            x_to_crop, upped.shape[-2:]
        )
        x = torch.cat([cropped, upped], dim=1)
        return self.convolutions(x)
\end{verbatim}
\caption{PyTorch Reference Implementation, Down and Up passes}
\end{figure*}

\begin{figure*}
\begin{verbatim}
class USegment(nn.Module):

    def __init__(self, in_channels, bottom_u=None):
        super().__init__()

        # Default value for the bottom U.
        if bottom_u is None:
            bottom_u = lambda x: x

        self.down = Down(in_channels)
        self.bottom_u = bottom_u
        self.up = Up(2 * in_channels)

    def forward(self, x):
        return self.up(x, self.bottom_u(self.down(x)))

class UNet(nn.Module):

    def __init__(self):
        super().__init__()

        self.u = USegment(512)
        self.u = USegment(256, self.u)
        self.u = USegment(128, self.u)
        self.u = USegment(64, self.u)
        self.path = nn.Sequential(
            TwoConv(1, 64),
            self.u,
            nn.Conv2d(64, 2, kernel_size=1, bias=False),
        )

    def forward(self, x):
        return self.path(x)
\end{verbatim}
\caption{PyTorch Reference Implementation, Wiring}

\end{figure*}

\begin{table*}
\begin{centering}
\begin{tabular}{rr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}l}
 &
\multicolumn{2}{c}{0} &
\multicolumn{2}{c}{1} &
\multicolumn{2}{c}{2} &
\multicolumn{2}{c}{3} &
\multicolumn{2}{c}{4} &
\multicolumn{2}{c}{5} &
\multicolumn{2}{c}{6} &
\multicolumn{2}{c}{7} &
\multicolumn{2}{c}{8} &
\multicolumn{2}{c}{9} &
\multicolumn{2}{c}{} &
\multicolumn{2}{c}{\emph{Avg}}\tabularnewline
\cmidrule{2-25} \cmidrule{4-25} \cmidrule{6-25} \cmidrule{8-25} \cmidrule{10-25} \cmidrule{12-25} \cmidrule{14-25} \cmidrule{16-25} \cmidrule{18-25} \cmidrule{20-25} \cmidrule{22-25} \cmidrule{24-25} 
\noun{\footnotesize{}Dyalog} &
{\footnotesize{}80}&{\footnotesize{}90} &
{\footnotesize{}77}&{\footnotesize{}56} &
{\footnotesize{}78}&{\footnotesize{}32} &
{\footnotesize{}81}&{\footnotesize{}33} &
{\footnotesize{}77}&{\footnotesize{}95} &
{\footnotesize{}80}&{\footnotesize{}51} &
{\footnotesize{}81}&{\footnotesize{}08} &
{\footnotesize{}81}&{\footnotesize{}87} &
{\footnotesize{}80}&{\footnotesize{}23} &
{\footnotesize{}79}&{\footnotesize{}42} &
\multicolumn{2}{c}{} &
{\footnotesize{}79}&{\footnotesize{}92}\tabularnewline
\noun{\footnotesize{}Co-dfns (CPU)} &
{\footnotesize{}34}&{\footnotesize{}11} &
{\footnotesize{}33}&{\footnotesize{}66} &
{\footnotesize{}35}&{\footnotesize{}98} &
{\footnotesize{}37}&{\footnotesize{}72} &
{\footnotesize{}38}&{\footnotesize{}20} &
{\footnotesize{}38}&{\footnotesize{}64} &
{\footnotesize{}38}&{\footnotesize{}11} &
{\footnotesize{}38}&{\footnotesize{}46} &
{\footnotesize{}38}&{\footnotesize{}57} &
{\footnotesize{}38}&{\footnotesize{}76} &
\multicolumn{2}{c}{} &
{\footnotesize{}37}&{\footnotesize{}22}\tabularnewline
\noun{\footnotesize{}Co-dfns (GPU)} &
{\footnotesize{}8}&{\footnotesize{}39} &
{\footnotesize{}8}&{\footnotesize{}20} &
{\footnotesize{}8}&{\footnotesize{}62} &
{\footnotesize{}8}&{\footnotesize{}63} &
{\footnotesize{}8}&{\footnotesize{}49} &
{\footnotesize{}8}&{\footnotesize{}39} &
{\footnotesize{}8}&{\footnotesize{}19} &
{\footnotesize{}8}&{\footnotesize{}63} &
{\footnotesize{}8}&{\footnotesize{}51} &
{\footnotesize{}8}&{\footnotesize{}59} &
\multicolumn{2}{c}{} &
{\footnotesize{}8}&{\footnotesize{}46}\tabularnewline
\noun{\footnotesize{}PyTorch (CPU)} &
{\footnotesize{}17}&{\footnotesize{}27} &
{\footnotesize{}17}&{\footnotesize{}34} &
{\footnotesize{}17}&{\footnotesize{}21} &
{\footnotesize{}16}&{\footnotesize{}60} &
{\footnotesize{}17}&{\footnotesize{}46} &
{\footnotesize{}17}&{\footnotesize{}01} &
{\footnotesize{}17}&{\footnotesize{}33} &
{\footnotesize{}17}&{\footnotesize{}04} &
{\footnotesize{}17}&{\footnotesize{}36} &
{\footnotesize{}17}&{\footnotesize{}33} &
\multicolumn{2}{c}{} &
{\footnotesize{}17}&{\footnotesize{}22}\tabularnewline
\noun{\footnotesize{}PyTorch (SMP)} &
{\footnotesize{}6}&{\footnotesize{}31} &
{\footnotesize{}6}&{\footnotesize{}42} &
{\footnotesize{}6}&{\footnotesize{}45} &
{\footnotesize{}6}&{\footnotesize{}19} &
{\footnotesize{}6}&{\footnotesize{}67} &
{\footnotesize{}6}&{\footnotesize{}49} &
{\footnotesize{}6}&{\footnotesize{}43} &
{\footnotesize{}6}&{\footnotesize{}38} &
{\footnotesize{}6}&{\footnotesize{}36} &
{\footnotesize{}6}&{\footnotesize{}55} &
\multicolumn{2}{c}{} &
{\footnotesize{}6}&{\footnotesize{}42}\tabularnewline
\noun{\footnotesize{}PyTorch (GPU)} &
{\footnotesize{}3}&{\footnotesize{}50} &
{\footnotesize{}3}&{\footnotesize{}50} &
{\footnotesize{}3}&{\footnotesize{}47} &
{\footnotesize{}3}&{\footnotesize{}44} &
{\footnotesize{}3}&{\footnotesize{}44} &
{\footnotesize{}3}&{\footnotesize{}44} &
{\footnotesize{}3}&{\footnotesize{}44} &
{\footnotesize{}3}&{\footnotesize{}44} &
{\footnotesize{}3}&{\footnotesize{}46} &
{\footnotesize{}3}&{\footnotesize{}46} &
\multicolumn{2}{c}{} &
{\footnotesize{}3}&{\footnotesize{}46}\tabularnewline
\cmidrule{2-25} \cmidrule{4-25} \cmidrule{6-25} \cmidrule{8-25} \cmidrule{10-25} \cmidrule{12-25} \cmidrule{14-25} \cmidrule{16-25} \cmidrule{18-25} \cmidrule{20-25} \cmidrule{22-25} \cmidrule{24-25} 
\end{tabular}
\par\end{centering}
\caption{Raw Performance Timings (Secs)}

\end{table*}

\bibliographystyle{ACM-Reference-Format}
\bibliography{2C__Users_arcfi_OneDrive_Documents_Co-dfns_Papers_2023-array-unet_Paper_bibliography}

\end{document}

### 2023-05-29 2023-array-unet/Paper/U-net in APL.tex
I think this version will finally satisfy the publication requirements

--- 2023-05-23 2023-array-unet/Paper/U-net_in_APL.tex
@@
-\batchmode
-\makeatletter
-\def\input@path{{C:/Users/arcfi/OneDrive/Documents/Co-dfns/Papers/2023-array-unet/Paper/}}
-\makeatother
+%% LyX 2.3.7 created this file.  For more info, see http://www.lyx.org/.
+%% Do not edit unless you really know what you are doing.
 \documentclass[10pt,twocolumn,english,format=sigplan,screen,balance]{acmart}
@@
   \AtBeginDocument{%
-    \hypersetup{xetex,unicode=true}
+    \hypersetup{unicode=true}
   }
 \else
-  \hypersetup{xetex,unicode=true}
+  \hypersetup{unicode=true}
 \fi
@@
 
-\acmConference[ARRAY '23]{Proceedings of the 9th ACM SIGPLAN International Workshop on Libraries,
-Languages and Compilers for Array Programming}{June 18, 2023}{Orlando, FL, USA}{}
+\acmConference[ARRAY '23]{Proceedings of the 9th ACM SIGPLAN International Workshop on Libraries, Languages and Compilers for Array Programming}{June 18, 2023}{Orlando, FL, USA}
 
@@
 \providecommand{\tabularnewline}{\\}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
+
+\usepackage{dblfloatfix}
 
@@
 \acmBooktitle{Proceedings of the 9th ACM SIGPLAN International Workshop on Libraries, Languages and Compilers for Array Programming (ARRAY '23), June 18, 2023, Orlando, FL, USA}
+
+\acmPrice{}
 \begin{CCSXML}
@@
 \begin{figure*}
-\includegraphics[width=0.6\paperwidth]{0C__Users_arcfi_OneDrive_Documents_Co-dfns_Papers_2023-array-unet_Paper_u-net-architecture.png}
+\includegraphics[width=0.6\paperwidth]{u-net-architecture}
 
@@
 \begin{tabular}{ll}
-\emph{Rank} &
-\emph{Name}\tabularnewline
+\emph{Rank} & \emph{Name}\tabularnewline
 \midrule 
-0 &
-Scalar\tabularnewline
-1 &
-Vector\tabularnewline
-2 &
-Matrix\tabularnewline
-3 &
-Cuboid\tabularnewline
-4 &
-Hypercube\tabularnewline
-5+ &
-Noble\tabularnewline
+0 & Scalar\tabularnewline
+1 & Vector\tabularnewline
+2 & Matrix\tabularnewline
+3 & Cuboid\tabularnewline
+4 & Hypercube\tabularnewline
+5+ & Noble\tabularnewline
 \bottomrule
@@
 \begin{tabular}{crrrlrrrlrr}
- &
- &
-\multicolumn{7}{c}{\noun{Operation}} &
- &
-\tabularnewline
+ &  & \multicolumn{7}{c}{\noun{Operation}} &  & \tabularnewline
 \cmidrule{3-9} \cmidrule{4-9} \cmidrule{5-9} \cmidrule{6-9} \cmidrule{7-9} \cmidrule{8-9} \cmidrule{9-9} 
- &
- &
-\multicolumn{1}{l}{$CV$} &
-\multicolumn{1}{l}{$CV$} &
-$MX$ &
- &
-\multicolumn{1}{l}{$CV$} &
-\multicolumn{1}{l}{$CV$} &
-$UP$ &
- &
-\tabularnewline
+ &  & \multicolumn{1}{l}{$CV$} & \multicolumn{1}{l}{$CV$} & $MX$ &  & \multicolumn{1}{l}{$CV$} & \multicolumn{1}{l}{$CV$} & $UP$ &  & \tabularnewline
 \cmidrule{3-5} \cmidrule{4-5} \cmidrule{5-5} \cmidrule{7-9} \cmidrule{8-9} \cmidrule{9-9} 
@@
 \noun{Depth}
-\end{turn}} &
-{\tiny{}0} &
-\texttt{\tiny{}$\texttt{3\ 3\ \ \ 1\ \ 64}$} &
-{\tiny{}$\texttt{3 3 \ 64 \ 64}$} &
-{\tiny{}$\texttt{0 0 \ 64 \ 64}$} &
- &
-{\tiny{}$\texttt{3 3 \ 256 \ 128}$} &
-{\tiny{}$\texttt{3 3 \ 128 \ 128}$} &
-{\tiny{}$\texttt{\ 128 2 2 \ 64}$} &
- &
-\tabularnewline
- & {\tiny{}1} &
-\multirow{1}{*}{{\tiny{}$\texttt{3\ 3\ \ 64\ 128}$}} &
-{\tiny{}$\texttt{3 3 128 128}$} &
-{\tiny{}$\texttt{0 0 128 128}$} &
- &
-{\tiny{}$\texttt{3 3 \ 512 \ 256}$} &
-{\tiny{}$\texttt{3 3 \ 256 \ 256}$} &
-{\tiny{}$\texttt{\ 256 2 2 128}$} &
- &
-\tabularnewline
- & {\tiny{}2} &
-{\tiny{}$\texttt{3 3 128 256}$} &
-{\tiny{}$\texttt{3 3 256 256}$} &
-{\tiny{}$\texttt{0 0 256 256}$} &
- &
-{\tiny{}$\texttt{3 3 1024 \ 512}$} &
-{\tiny{}$\texttt{3 3 \ 512 \ 512}$} &
-{\tiny{}$\texttt{\ 512 2 2 256}$} &
- &
-\tabularnewline
- & {\tiny{}3} &
-{\tiny{}$\texttt{3 3 256 512}$} &
-{\tiny{}$\texttt{3 3 512 512}$} &
-{\tiny{}$\texttt{0 0 512 512}$} &
- &
-{\tiny{}$\texttt{3 3 \ 512 1024}$} &
-{\tiny{}$\texttt{3 3 1024 1024}$} &
-{\tiny{}$\texttt{1024 2 2 512}$} &
- &
-\tabularnewline
+\end{turn}} & {\tiny{}0} & \texttt{\tiny{}$\texttt{3\ 3\ \ \ 1\ \ 64}$} & {\tiny{}$\texttt{3 3 \ 64 \ 64}$} & {\tiny{}$\texttt{0 0 \ 64 \ 64}$} &  & {\tiny{}$\texttt{3 3 \ 256 \ 128}$} & {\tiny{}$\texttt{3 3 \ 128 \ 128}$} & {\tiny{}$\texttt{\ 128 2 2 \ 64}$} &  & \tabularnewline
+ & {\tiny{}1} & \multirow{1}{*}{{\tiny{}$\texttt{3\ 3\ \ 64\ 128}$}} & {\tiny{}$\texttt{3 3 128 128}$} & {\tiny{}$\texttt{0 0 128 128}$} &  & {\tiny{}$\texttt{3 3 \ 512 \ 256}$} & {\tiny{}$\texttt{3 3 \ 256 \ 256}$} & {\tiny{}$\texttt{\ 256 2 2 128}$} &  & \tabularnewline
+ & {\tiny{}2} & {\tiny{}$\texttt{3 3 128 256}$} & {\tiny{}$\texttt{3 3 256 256}$} & {\tiny{}$\texttt{0 0 256 256}$} &  & {\tiny{}$\texttt{3 3 1024 \ 512}$} & {\tiny{}$\texttt{3 3 \ 512 \ 512}$} & {\tiny{}$\texttt{\ 512 2 2 256}$} &  & \tabularnewline
+ & {\tiny{}3} & {\tiny{}$\texttt{3 3 256 512}$} & {\tiny{}$\texttt{3 3 512 512}$} & {\tiny{}$\texttt{0 0 512 512}$} &  & {\tiny{}$\texttt{3 3 \ 512 1024}$} & {\tiny{}$\texttt{3 3 1024 1024}$} & {\tiny{}$\texttt{1024 2 2 512}$} &  & \tabularnewline
 \cmidrule{3-5} \cmidrule{4-5} \cmidrule{5-5} \cmidrule{7-9} \cmidrule{8-9} \cmidrule{9-9} 
- &
- &
-\multicolumn{3}{c}{\emph{Downward Pass}} &
- &
-\multicolumn{3}{c}{\emph{Upward Pass}} &
- &
-\tabularnewline
+ &  & \multicolumn{3}{c}{\emph{Downward Pass}} &  & \multicolumn{3}{c}{\emph{Upward Pass}} &  & \tabularnewline
 \cmidrule{3-5} \cmidrule{4-5} \cmidrule{5-5} \cmidrule{7-9} \cmidrule{8-9} \cmidrule{9-9} 
@@
 \begin{figure*}
-\includegraphics[width=1\textwidth]{1C__Users_arcfi_OneDrive_Documents_Co-dfns_Papers_2023-array-unet_Paper_benchmark-comp.pdf}
+\includegraphics[width=1\textwidth]{benchmark-comp}
 
@@
 
+\begin{table*}
+\caption{Raw Performance Timings (Secs)}
+
+\centering{}%
+\begin{tabular}{rr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}l}
+ & \multicolumn{2}{c}{0} & \multicolumn{2}{c}{1} & \multicolumn{2}{c}{2} & \multicolumn{2}{c}{3} & \multicolumn{2}{c}{4} & \multicolumn{2}{c}{5} & \multicolumn{2}{c}{6} & \multicolumn{2}{c}{7} & \multicolumn{2}{c}{8} & \multicolumn{2}{c}{9} & \multicolumn{2}{c}{} & \multicolumn{2}{c}{\emph{Avg}}\tabularnewline
+\cmidrule{2-25} \cmidrule{4-25} \cmidrule{6-25} \cmidrule{8-25} \cmidrule{10-25} \cmidrule{12-25} \cmidrule{14-25} \cmidrule{16-25} \cmidrule{18-25} \cmidrule{20-25} \cmidrule{22-25} \cmidrule{24-25} 
+\noun{\footnotesize{}Dyalog} & {\footnotesize{}80}&{\footnotesize{}90} & {\footnotesize{}77}&{\footnotesize{}56} & {\footnotesize{}78}&{\footnotesize{}32} & {\footnotesize{}81}&{\footnotesize{}33} & {\footnotesize{}77}&{\footnotesize{}95} & {\footnotesize{}80}&{\footnotesize{}51} & {\footnotesize{}81}&{\footnotesize{}08} & {\footnotesize{}81}&{\footnotesize{}87} & {\footnotesize{}80}&{\footnotesize{}23} & {\footnotesize{}79}&{\footnotesize{}42} & \multicolumn{2}{c}{} & {\footnotesize{}79}&{\footnotesize{}92}\tabularnewline
+\noun{\footnotesize{}Co-dfns (CPU)} & {\footnotesize{}34}&{\footnotesize{}11} & {\footnotesize{}33}&{\footnotesize{}66} & {\footnotesize{}35}&{\footnotesize{}98} & {\footnotesize{}37}&{\footnotesize{}72} & {\footnotesize{}38}&{\footnotesize{}20} & {\footnotesize{}38}&{\footnotesize{}64} & {\footnotesize{}38}&{\footnotesize{}11} & {\footnotesize{}38}&{\footnotesize{}46} & {\footnotesize{}38}&{\footnotesize{}57} & {\footnotesize{}38}&{\footnotesize{}76} & \multicolumn{2}{c}{} & {\footnotesize{}37}&{\footnotesize{}22}\tabularnewline
+\noun{\footnotesize{}Co-dfns (GPU)} & {\footnotesize{}8}&{\footnotesize{}39} & {\footnotesize{}8}&{\footnotesize{}20} & {\footnotesize{}8}&{\footnotesize{}62} & {\footnotesize{}8}&{\footnotesize{}63} & {\footnotesize{}8}&{\footnotesize{}49} & {\footnotesize{}8}&{\footnotesize{}39} & {\footnotesize{}8}&{\footnotesize{}19} & {\footnotesize{}8}&{\footnotesize{}63} & {\footnotesize{}8}&{\footnotesize{}51} & {\footnotesize{}8}&{\footnotesize{}59} & \multicolumn{2}{c}{} & {\footnotesize{}8}&{\footnotesize{}46}\tabularnewline
+\noun{\footnotesize{}PyTorch (CPU)} & {\footnotesize{}17}&{\footnotesize{}27} & {\footnotesize{}17}&{\footnotesize{}34} & {\footnotesize{}17}&{\footnotesize{}21} & {\footnotesize{}16}&{\footnotesize{}60} & {\footnotesize{}17}&{\footnotesize{}46} & {\footnotesize{}17}&{\footnotesize{}01} & {\footnotesize{}17}&{\footnotesize{}33} & {\footnotesize{}17}&{\footnotesize{}04} & {\footnotesize{}17}&{\footnotesize{}36} & {\footnotesize{}17}&{\footnotesize{}33} & \multicolumn{2}{c}{} & {\footnotesize{}17}&{\footnotesize{}22}\tabularnewline
+\noun{\footnotesize{}PyTorch (SMP)} & {\footnotesize{}6}&{\footnotesize{}31} & {\footnotesize{}6}&{\footnotesize{}42} & {\footnotesize{}6}&{\footnotesize{}45} & {\footnotesize{}6}&{\footnotesize{}19} & {\footnotesize{}6}&{\footnotesize{}67} & {\footnotesize{}6}&{\footnotesize{}49} & {\footnotesize{}6}&{\footnotesize{}43} & {\footnotesize{}6}&{\footnotesize{}38} & {\footnotesize{}6}&{\footnotesize{}36} & {\footnotesize{}6}&{\footnotesize{}55} & \multicolumn{2}{c}{} & {\footnotesize{}6}&{\footnotesize{}42}\tabularnewline
+\noun{\footnotesize{}PyTorch (GPU)} & {\footnotesize{}3}&{\footnotesize{}50} & {\footnotesize{}3}&{\footnotesize{}50} & {\footnotesize{}3}&{\footnotesize{}47} & {\footnotesize{}3}&{\footnotesize{}44} & {\footnotesize{}3}&{\footnotesize{}44} & {\footnotesize{}3}&{\footnotesize{}44} & {\footnotesize{}3}&{\footnotesize{}44} & {\footnotesize{}3}&{\footnotesize{}44} & {\footnotesize{}3}&{\footnotesize{}46} & {\footnotesize{}3}&{\footnotesize{}46} & \multicolumn{2}{c}{} & {\footnotesize{}3}&{\footnotesize{}46}\tabularnewline
+\cmidrule{2-25} \cmidrule{4-25} \cmidrule{6-25} \cmidrule{8-25} \cmidrule{10-25} \cmidrule{12-25} \cmidrule{14-25} \cmidrule{16-25} \cmidrule{18-25} \cmidrule{20-25} \cmidrule{22-25} \cmidrule{24-25} 
+\end{tabular}
+\end{table*}
+
 Clearly, specialized frameworks for deep neural networks are still
@@
 
-\balance
-
 Given the notational advantages of APL and the concision and clarity
@@
 
-\section{Reference Code and Timings}
-
-\begin{figure*}
-\begin{verbatim}
-:Namespace UNET
-
+\section{Complete APL U-Net Implementation}
+\begin{verbatim}
 W←⍬ ⋄ V←⍬ ⋄ Z←⍬ ⋄ LR←1e¯9 ⋄ MO←0.99
@@
 FWD←{Z⊢←(≢W)⍴⊂⍬
-  CV←{0⌈z⊣Z[⍺]←⊂Z[⍺],⊂z←(,[2+⍳3]3 3⌺⊃Z[⍺]←⊂⍵)+.×,[⍳3]⍺⊃W}
-  CC←{⍵,⍨(⌊p)↓(-⌈p)↓(⍺⊃Z)⊣p←2÷⍨(⍴⍺⊃Z)-⍴⍵}
+ CV←{Z[⍺]←⊂⍵
+  z←(,[2+⍳3]3 3⌺⍺⊃Z)+.×,[⍳3]⍺⊃W
+  0⌈z⊣Z[⍺]←⊂Z[⍺],⊂z}
+ CC←{p←2÷⍨(⍴⍺⊃Z)-⍴⍵
+  ⍵,⍨(⌊p)↓(-⌈p)↓(⍺⊃Z)}
  MX←{⌈⌿[2],[2 3](2 2⍴2)⌺⊃Z[⍺]←⊂⍵}
-  UP←{((2×¯1↓⍴⍵),¯1↑⍴⍺⊃W)⍴0 2 1 3 4⍉⍵+.×⍺⊃W⊣Z[⍺]←⊂⍵}
-  C1←{1E¯8+z÷[⍳2]+/z←*z-[⍳2]⌈/z←⍵+.×⍺⊃W⊣Z[⍺]←⊂⍵}
+ UP←{Z[⍺]←⊂⍵
+  s←(2×¯1↓⍴⍵),¯1↑⍴⍺⊃W
+  s⍴0 2 1 3 4⍉⍵+.×⍺⊃W}
+ C1←{Z[⍺]←⊂⍵
+  1E¯8+z÷[⍳2]+/z←*z-[⍳2]⌈/z←⍵+.×⍺⊃W}
  LA←{⍺≥≢Z:⍵
@@
 BCK←{Y←⍺ ⋄ Y∆←⍵
-  ∆←{0⊣W[⍺]←⊂(⍺⊃W)-LR×⊃V[⍺]←⊂⍵+MO×(⍴⍵)⍴⍺⊃V}
-  ∆CV←{w←,[⍳3]⊖⌽[1]0 1 3 2⍉⍺⊃W ⋄ x←⊃⍺⊃Z ⋄ ∆z←⍵×0<1⊃⍺⊃Z
+ ∆←{V[⍺]←⊂⍵+MO×(⍴⍵)⍴⍺⊃V
+  W[⍺]←⊂(⍺⊃W)-LR×⍺⊃V}
+ ∆CV←{w←,[⍳3]⊖⌽[1]0 1 3 2⍉⍺⊃W ⋄ x←⊃⍺⊃Z
+  ∆z←⍵×0<1⊃⍺⊃Z
   ∆Z←¯2⊖¯2⌽[1](4+2↑⍴∆z)↑∆z
@@
   w+.×⍨,[2+⍳3]3 3⌺∆Z}
-  ∆CC←{x←⍺⊃Z ⋄ ∆z←⍵ ⋄ d←-⌊2÷⍨2↑(⍴x)-⍴∆z ⋄ (⊃d)⊖(1⊃d)⌽[1](⍴x)↑∆z}
-  ∆MX←{x←⍺⊃Z ⋄ ∆z←⍵ ⋄ y×x=y←(⍴x)↑2⌿2/[1]∆z}
-  ∆UP←{w←⍺⊃W ⋄ x←⍺⊃Z ⋄ ∆z←⍵ ⋄ cz←(2 2⍴2)⌺∆z
-   _←⍺ ∆(⍉,[⍳2]x)+.×,[⍳2]cz
+ ∆CC←{x←⍺⊃Z ⋄ ∆z←⍵ ⋄ d←-⌊2÷⍨2↑(⍴x)-⍴∆z
+  (⊃d)⊖(1⊃d)⌽[1](⍴x)↑∆z}
+ ∆MX←{x←⍺⊃Z ⋄ ∆z←⍵
+  y×x=y←(⍴x)↑2⌿2/[1]∆z}
+ ∆UP←{w←⍺⊃W ⋄ x←⍺⊃Z ⋄ ∆z←⍵
+  _←⍺ ∆(⍉,[⍳2]x)+.×,[⍳2]cz←(2 2⍴2)⌺∆z
   (,[2+⍳3]cz)+.×⍉⍪w}
-  ∆C1←{w←⍺⊃W ⋄ x←⍺⊃Z ⋄ ∆z←⍵ ⋄ _←⍺ ∆(⍉,[⍳2]x)+.×,[⍳2]∆z ⋄ ∆z+.×⍉w}
-  ∆LA←{⍺≥≢Z:⍵
-   down←(⍺+6)∇(⍺+3)∆CV(⍺+4)∆CV(⍺+5)∆UP ⍵↑[2]⍨-2÷⍨⊃⌽⍴⍵
-   (⍺+0)∆CV(⍺+1)∆CV(⍵ ∆CC⍨⍺+2)+(⍺+2)∆MX down}
+ ∆C1←{w←⍺⊃W ⋄ x←⍺⊃Z ⋄ ∆z←⍵
+  _←⍺ ∆(⍉,[⍳2]x)+.×,[⍳2]∆z
+  ∆z+.×⍉w}
+ ∆LA←{⍺≥≢Z:⍵ ⋄ in←⍵↑[2]⍨-2÷⍨⊃⌽⍴⍵
+  d←(⍺+6)∇(⍺+3)∆CV(⍺+4)∆CV(⍺+5)∆UP in
+  (⍺+0)∆CV(⍺+1)∆CV(⍵ ∆CC⍨⍺+2)+(⍺+2)∆MX d
+ }
  3 ∆LA 0 ∆CV 1 ∆CV 2 ∆C1 Y∆-(~Y),[1.5]Y}
@@
 E←{-+⌿,⍟(⍺×⍵[;;1])+(~⍺)×⍵[;;0]}
-
- RUN←{Y Y∆(Y E Y∆)⊣(Y←⌊0.5+nm↑⍵↓⍨2÷⍨(⍴⍵)-nm←2↑⍴Y∆)BCK⊢Y∆←FWD ⍺}
-
-:EndNamespace
-\end{verbatim}
-\caption{Complete APL U-Net Implementation}
-
-\end{figure*}
-
-\begin{figure*}
+RUN←{Y←⌊0.5+nm↑⍵↓⍨2÷⍨(⍴⍵)-nm←2↑⍴Y∆←FWD ⍺
+ Y Y∆(Y E Y∆)⊣Y BCK Y∆}
+\end{verbatim}
+
+\section{PyTorch Reference Implementation}
 \begin{verbatim}
@@
 
-   def __init__(self, in_channels, out_channels):
+ def __init__(
+   self, in_channels, out_channels):
   super().__init__()
@@
     nn.ReLU(inplace=True),
-            nn.Conv2d(out_channels, out_channels, 
+    nn.Conv2d(
+     out_channels, out_channels, 
      kernel_size=(3, 3), bias=False),
@@
   return self.path(x)
-\end{verbatim}
-\caption{PyTorch Reference Implementation, Preamble and Helpers}
-
-\end{figure*}
-\begin{figure*}
-\begin{verbatim}
+
 class Down(nn.Module):
@@
   self.path = nn.Sequential(
-            nn.MaxPool2d(kernel_size=(2, 2), stride=2),
-            TwoConv(in_channels, 2 * in_channels),
+   nn.MaxPool2d(
+    kernel_size=(2, 2), stride=2),
+    TwoConv(
+     in_channels, 2 * in_channels
+    ),
    )
@@
   self.convolutions = 
-            TwoConv(in_channels, in_channels // 2)
+   TwoConv(
+    in_channels, in_channels // 2
+   )
 
@@
   upped = self.upsampling(x_in)
-        cropped = torchvision.transforms.functional.center_crop(
+  cropped = 
+   torchvision.transforms.
+    functional.center_crop(
      x_to_crop, upped.shape[-2:]
@@
   return self.convolutions(x)
-\end{verbatim}
-\caption{PyTorch Reference Implementation, Down and Up passes}
-\end{figure*}
-
-\begin{figure*}
-\begin{verbatim}
+
 class USegment(nn.Module):
 
-    def __init__(self, in_channels, bottom_u=None):
+ def __init__(
+   self, in_channels, bottom_u=None
+  ):
   super().__init__()
 
-        # Default value for the bottom U.
   if bottom_u is None:
@@
  def forward(self, x):
-        return self.up(x, self.bottom_u(self.down(x)))
+  return self.up(
+   x, self.bottom_u(self.down(x))
+  )
 
@@
    self.u,
-            nn.Conv2d(64, 2, kernel_size=1, bias=False),
+   nn.Conv2d(
+    64, 2, kernel_size=1, bias=False
+   ),
   )
@@
 \end{verbatim}
-\caption{PyTorch Reference Implementation, Wiring}
-
-\end{figure*}
-
-\begin{table*}
-\begin{centering}
-\begin{tabular}{rr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}lr@{\extracolsep{0pt}.}l}
- &
-\multicolumn{2}{c}{0} &
-\multicolumn{2}{c}{1} &
-\multicolumn{2}{c}{2} &
-\multicolumn{2}{c}{3} &
-\multicolumn{2}{c}{4} &
-\multicolumn{2}{c}{5} &
-\multicolumn{2}{c}{6} &
-\multicolumn{2}{c}{7} &
-\multicolumn{2}{c}{8} &
-\multicolumn{2}{c}{9} &
-\multicolumn{2}{c}{} &
-\multicolumn{2}{c}{\emph{Avg}}\tabularnewline
-\cmidrule{2-25} \cmidrule{4-25} \cmidrule{6-25} \cmidrule{8-25} \cmidrule{10-25} \cmidrule{12-25} \cmidrule{14-25} \cmidrule{16-25} \cmidrule{18-25} \cmidrule{20-25} \cmidrule{22-25} \cmidrule{24-25} 
-\noun{\footnotesize{}Dyalog} &
-{\footnotesize{}80}&{\footnotesize{}90} &
-{\footnotesize{}77}&{\footnotesize{}56} &
-{\footnotesize{}78}&{\footnotesize{}32} &
-{\footnotesize{}81}&{\footnotesize{}33} &
-{\footnotesize{}77}&{\footnotesize{}95} &
-{\footnotesize{}80}&{\footnotesize{}51} &
-{\footnotesize{}81}&{\footnotesize{}08} &
-{\footnotesize{}81}&{\footnotesize{}87} &
-{\footnotesize{}80}&{\footnotesize{}23} &
-{\footnotesize{}79}&{\footnotesize{}42} &
-\multicolumn{2}{c}{} &
-{\footnotesize{}79}&{\footnotesize{}92}\tabularnewline
-\noun{\footnotesize{}Co-dfns (CPU)} &
-{\footnotesize{}34}&{\footnotesize{}11} &
-{\footnotesize{}33}&{\footnotesize{}66} &
-{\footnotesize{}35}&{\footnotesize{}98} &
-{\footnotesize{}37}&{\footnotesize{}72} &
-{\footnotesize{}38}&{\footnotesize{}20} &
-{\footnotesize{}38}&{\footnotesize{}64} &
-{\footnotesize{}38}&{\footnotesize{}11} &
-{\footnotesize{}38}&{\footnotesize{}46} &
-{\footnotesize{}38}&{\footnotesize{}57} &
-{\footnotesize{}38}&{\footnotesize{}76} &
-\multicolumn{2}{c}{} &
-{\footnotesize{}37}&{\footnotesize{}22}\tabularnewline
-\noun{\footnotesize{}Co-dfns (GPU)} &
-{\footnotesize{}8}&{\footnotesize{}39} &
-{\footnotesize{}8}&{\footnotesize{}20} &
-{\footnotesize{}8}&{\footnotesize{}62} &
-{\footnotesize{}8}&{\footnotesize{}63} &
-{\footnotesize{}8}&{\footnotesize{}49} &
-{\footnotesize{}8}&{\footnotesize{}39} &
-{\footnotesize{}8}&{\footnotesize{}19} &
-{\footnotesize{}8}&{\footnotesize{}63} &
-{\footnotesize{}8}&{\footnotesize{}51} &
-{\footnotesize{}8}&{\footnotesize{}59} &
-\multicolumn{2}{c}{} &
-{\footnotesize{}8}&{\footnotesize{}46}\tabularnewline
-\noun{\footnotesize{}PyTorch (CPU)} &
-{\footnotesize{}17}&{\footnotesize{}27} &
-{\footnotesize{}17}&{\footnotesize{}34} &
-{\footnotesize{}17}&{\footnotesize{}21} &
-{\footnotesize{}16}&{\footnotesize{}60} &
-{\footnotesize{}17}&{\footnotesize{}46} &
-{\footnotesize{}17}&{\footnotesize{}01} &
-{\footnotesize{}17}&{\footnotesize{}33} &
-{\footnotesize{}17}&{\footnotesize{}04} &
-{\footnotesize{}17}&{\footnotesize{}36} &
-{\footnotesize{}17}&{\footnotesize{}33} &
-\multicolumn{2}{c}{} &
-{\footnotesize{}17}&{\footnotesize{}22}\tabularnewline
-\noun{\footnotesize{}PyTorch (SMP)} &
-{\footnotesize{}6}&{\footnotesize{}31} &
-{\footnotesize{}6}&{\footnotesize{}42} &
-{\footnotesize{}6}&{\footnotesize{}45} &
-{\footnotesize{}6}&{\footnotesize{}19} &
-{\footnotesize{}6}&{\footnotesize{}67} &
-{\footnotesize{}6}&{\footnotesize{}49} &
-{\footnotesize{}6}&{\footnotesize{}43} &
-{\footnotesize{}6}&{\footnotesize{}38} &
-{\footnotesize{}6}&{\footnotesize{}36} &
-{\footnotesize{}6}&{\footnotesize{}55} &
-\multicolumn{2}{c}{} &
-{\footnotesize{}6}&{\footnotesize{}42}\tabularnewline
-\noun{\footnotesize{}PyTorch (GPU)} &
-{\footnotesize{}3}&{\footnotesize{}50} &
-{\footnotesize{}3}&{\footnotesize{}50} &
-{\footnotesize{}3}&{\footnotesize{}47} &
-{\footnotesize{}3}&{\footnotesize{}44} &
-{\footnotesize{}3}&{\footnotesize{}44} &
-{\footnotesize{}3}&{\footnotesize{}44} &
-{\footnotesize{}3}&{\footnotesize{}44} &
-{\footnotesize{}3}&{\footnotesize{}44} &
-{\footnotesize{}3}&{\footnotesize{}46} &
-{\footnotesize{}3}&{\footnotesize{}46} &
-\multicolumn{2}{c}{} &
-{\footnotesize{}3}&{\footnotesize{}46}\tabularnewline
-\cmidrule{2-25} \cmidrule{4-25} \cmidrule{6-25} \cmidrule{8-25} \cmidrule{10-25} \cmidrule{12-25} \cmidrule{14-25} \cmidrule{16-25} \cmidrule{18-25} \cmidrule{20-25} \cmidrule{22-25} \cmidrule{24-25} 
-\end{tabular}
-\par\end{centering}
-\caption{Raw Performance Timings (Secs)}
-
-\end{table*}
+\balance
 
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{2C__Users_arcfi_OneDrive_Documents_Co-dfns_Papers_2023-array-unet_Paper_bibliography}
+\bibliography{bibliography}
 
